\section{Review Exercises}

\subsection{Elementary computational exercises}

\begin{enumerate}
\item Solve by elimination:
\[
\begin{cases}
x+y=7,\\
2x-y=2.
\end{cases}
\]
Write the augmented matrix and verify the solution in both equations.

\item Solve by elimination:
\[
\begin{cases}
3x+2y=4,\\
5x-2y=12.
\end{cases}
\]
Choose the elimination step so that no fractions appear before the final line.

\item Classify and solve when possible:
\[
\begin{cases}
x+2y=3,\\
2x+4y=6.
\end{cases}
\]
Write the solution set using one parameter.

\item Classify:
\[
\begin{cases}
x-3y=2,\\
2x-6y=7.
\end{cases}
\]
Show the contradictory row produced by elimination.

\item Solve
\[
\begin{cases}
2x+y=5,\\
x-y=1
\end{cases}
\]
with Cramer's rule. Compute $\det A$, $\det A_x$, and $\det A_y$ separately.

\item Solve
\[
\begin{cases}
2x+y=5,\\
x+y=3
\end{cases}
\]
using the inverse-matrix method. Compute $A^{-1}$ first and verify the solution by substitution.

\item For the parameter $a$, classify
\[
\begin{cases}
x+y=3,\\
2x+ay=6.
\end{cases}
\]
Find all exceptional values and determine the solution type in every case.

\item For the parameter $b$, classify
\[
\begin{cases}
x-2y=1,\\
3x-6y=b.
\end{cases}
\]
State when the system has no solution and when it has infinitely many solutions.

\item Find the intersection point of the lines
\[
y=2x-3,
\qquad
y=-x+6
\]
by solving the corresponding system. Plot both lines and mark the solution.

\item Reduce
\[
\left[\begin{array}{cc|c}
1&2&5\\
3&6&15
\end{array}\right]
\]
to echelon form, identify the pivot and free variable, and write the full solution set.
\end{enumerate}

\subsection{Conceptual and reasoning exercises}

\begin{enumerate}
\item Explain why a solution of a system must satisfy every equation simultaneously. How is this represented geometrically for two equations in two unknowns?

\item Describe the geometric meaning of exactly one solution, no solution, and infinitely many solutions for two lines in the plane.

\item Explain why elementary row operations preserve the solution set. Relate each operation to an equivalent transformation of the equations.

\item A row
\[
[\,0\quad0\mid5\,]
\]
appears during elimination. Explain why it proves inconsistency. What different information is carried by the row $[\,0\quad0\mid0\,]$?

\item Explain why a missing pivot produces a free variable and infinitely many solutions when no contradiction is present.

\item Compare Gaussian elimination, Cramer's rule, and the inverse-matrix method. Which method is the main classification tool, and under what condition do the other two apply?

\item Explain why $\det A\neq0$ guarantees one solution for every right-hand side, while $\det A=0$ does not by itself distinguish no solution from infinitely many solutions.

\item In a parameter-dependent system, why must values that make a proposed divisor zero be separated before division? Give a simple example.

\item Two equations have proportional left-hand sides. Explain how the right-hand sides determine whether the corresponding lines coincide or are distinct and parallel.

\item Give a complete strategy for a $2\times2$ system: translate it into matrix form, choose a method, classify the result, and interpret the solution set geometrically.
\end{enumerate}
