\subsection{Elementary Graphs and Their Intersections}

Once coordinates are available, an equation can be read as a condition selecting points of the plane. At this stage we do not yet develop a general theory of lines or parabolas; we use simple examples to practise the Cartesian language.

\begin{examplebox}[A line from a value table]
For
\[
y=x+2,
\]
choosing several values of $x$ gives
\[
\begin{array}{c|ccccc}
x&-2&-1&0&1&2\\ \hline
y&0&1&2&3&4
\end{array}
\]
Plotting these points reveals a straight line. The equation is a rule for generating all points of the graph.
\end{examplebox}

\begin{examplebox}[A parabola and a common point problem]
For
\[
y=x^2
\qquad\text{and}\qquad
y=x+2,
\]
a common point must satisfy both conditions. Hence
\[
x^2=x+2,
\]
so
\[
(x-2)(x+1)=0.
\]
The graphs meet at
\[
(-1,1)
\qquad\text{and}\qquad
(2,4).
\]
\end{examplebox}

\begin{center}
\begin{tikzpicture}[x=0.95cm,y=0.8cm,>=Latex]
  \draw[step=1cm,gray!20,very thin] (-3.2,-0.8) grid (3.4,5.3);
  \draw[->] (-3.3,0) -- (3.5,0) node[right] {$x$};
  \draw[->] (0,-0.9) -- (0,5.4) node[above] {$y$};
  \draw[blue!75!black,thick,domain=-2.3:2.3,samples=100] plot (\x,{\x*\x});
  \draw[orange!85!black,thick,domain=-3:3,samples=2] plot (\x,{\x+2});
  \fill (-1,1) circle (2.3pt) node[above left] {$(-1,1)$};
  \fill (2,4) circle (2.3pt) node[above right] {$(2,4)$};
  \node[blue!75!black] at (2.55,5.0) {$y=x^2$};
  \node[orange!85!black] at (-2.2,0.45) {$y=x+2$};
\end{tikzpicture}
\end{center}

\begin{interpretationbox}[Intersection means simultaneous satisfaction]
A plotted point belongs to a graph when its coordinates satisfy the graph's equation. An intersection point belongs to both graphs, so its coordinates must satisfy both equations simultaneously. This principle will later organise line intersections, systems of equations, and intersections of more complicated curves.
\end{interpretationbox}
