\section{Constructing Cartesian Coordinates from Euclidean Geometry}

The preceding section established the Euclidean objects and constructions
available before coordinates. We now make the additional choices that turn that
plane into a Cartesian coordinate plane: axes, an origin, directions, and a
common numerical scale.

\begin{corebox}[The central invention]
A Cartesian coordinate system converts geometric position into arithmetic data.
After choosing two perpendicular axes, an origin, orientations, and a common
unit, every point can be described by an ordered pair of real numbers. Geometry
can then be studied with algebra without ceasing to be geometry.
\end{corebox}

\input{chapters/chapter_01/cartesian_coordinates/plane_before_coordinates}
\input{chapters/chapter_01/cartesian_coordinates/first_straight_line}
\input{chapters/chapter_01/cartesian_coordinates/origin_on_the_line}
\input{chapters/chapter_01/cartesian_coordinates/perpendicular_axis}
\input{chapters/chapter_01/cartesian_coordinates/positive_and_negative_directions}
\input{chapters/chapter_01/cartesian_coordinates/common_unit_length}
\input{chapters/chapter_01/cartesian_coordinates/refining_the_scale}
\input{chapters/chapter_01/cartesian_coordinates/orthogonal_projection}
\input{chapters/chapter_01/cartesian_coordinates/euclidean_distance_formula}

\begin{summarybox}[Cartesian space]
The Cartesian plane is produced by a sequence of explicit choices: two
perpendicular axes, their intersection as origin, positive orientations, and a
common unit. Orthogonal projection then assigns an ordered pair $(x,y)$ to every
point. The distance formula is the algebraic expression of the earlier
Euclidean notion of length.
\end{summarybox}
