\section{Euler--Lagrange Field Equations}
\label{sec:euler_lagrange_field_equations}

We now derive the differential equation satisfied by a stationary field configuration. The calculation is the spacetime extension of the mechanical derivation, but it contains several derivative directions and a boundary surrounding an entire spacetime region.

Every essential step will be displayed: the varied field, the variation of its derivatives, spacetime integration by parts, the boundary term, and the use of arbitrary interior variations.

\subsection{Statement of the variational problem}

Consider one real field \(\phi(x)\) with action

\[
S[\phi]
=
\int_{\Omega}
\mathcal{L}(\phi,\partial_{\mu}\phi,x)
\,\mathrm{d}^{4}x.
\]

We introduce the family

\[
\phi_{\varepsilon}(x)
=
\phi(x)+\varepsilon\eta(x),
\]

where \(\eta\) is smooth and satisfies the chosen boundary condition. For the standard fixed-boundary problem,

\[
\eta=0
\qquad
\text{on }\partial\Omega.
\]

The stationary-action condition is

\[
\left.
\frac{\mathrm{d}}{\mathrm{d}\varepsilon}
S[\phi_{\varepsilon}]
\right|_{\varepsilon=0}
=0
\]

for every admissible \(\eta(x)\).

\subsection{The first variation}

From the chain rule,

\[
\delta S
=
\int_{\Omega}
\left[
\frac{\partial\mathcal{L}}{\partial\phi}
\eta
+
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
\partial_{\mu}\eta
\right]
\,\mathrm{d}^{4}x.
\]

Define

\[
P^{\mu}
:=
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}.
\]

This notation is temporary and simplifies the integration-by-parts step. The first variation becomes

\[
\delta S
=
\int_{\Omega}
\left[
\frac{\partial\mathcal{L}}{\partial\phi}
\eta
+
P^{\mu}\partial_{\mu}\eta
\right]
\,\mathrm{d}^{4}x.
\]

\subsection{Spacetime integration by parts}

The product rule gives

\[
\partial_{\mu}(P^{\mu}\eta)
=
(\partial_{\mu}P^{\mu})\eta
+
P^{\mu}\partial_{\mu}\eta.
\]

Therefore

\[
P^{\mu}\partial_{\mu}\eta
=
\partial_{\mu}(P^{\mu}\eta)
-
(\partial_{\mu}P^{\mu})\eta.
\]

Substitution into the first variation gives

\[
\begin{aligned}
\delta S
&=
\int_{\Omega}
\partial_{\mu}(P^{\mu}\eta)
\,\mathrm{d}^{4}x\\
&\quad+
\int_{\Omega}
\left[
\frac{\partial\mathcal{L}}{\partial\phi}
-
\partial_{\mu}P^{\mu}
\right]
\eta
\,\mathrm{d}^{4}x.
\end{aligned}
\]

The first integral is a spacetime divergence. By the divergence theorem,

\[
\int_{\Omega}
\partial_{\mu}(P^{\mu}\eta)
\,\mathrm{d}^{4}x
=
\int_{\partial\Omega}
P^{\mu}n_{\mu}\eta
\,\mathrm{d}\Sigma,
\]

where \(n_{\mu}\) is normal to the boundary and \(\mathrm{d}\Sigma\) is the induced boundary measure.

Thus

\[
\boxed{
\begin{aligned}
\delta S
&=
\int_{\partial\Omega}
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
n_{\mu}\eta
\,\mathrm{d}\Sigma\\
&\quad+
\int_{\Omega}
\left[
\frac{\partial\mathcal{L}}{\partial\phi}
-
\partial_{\mu}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
\right)
\right]
\eta
\,\mathrm{d}^{4}x.
\end{aligned}
}.
\]

The first line is the boundary contribution. The second is the interior contribution.

\subsection{Using the boundary condition}

For fixed field values on \(\partial\Omega\),

\[
\eta|_{\partial\Omega}=0.
\]

Therefore

\[
\int_{\partial\Omega}
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
n_{\mu}\eta
\,\mathrm{d}\Sigma
=0.
\]

The variation reduces to

\[
\delta S
=
\int_{\Omega}
\left[
\frac{\partial\mathcal{L}}{\partial\phi}
-
\partial_{\mu}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
\right)
\right]
\eta
\,\mathrm{d}^{4}x.
\]

The boundary term has not been ignored. It vanishes because the admissible variations were required to vanish on the boundary.

\subsection{The fundamental lemma in spacetime}

Suppose a continuous function \(F(x)\) satisfies

\[
\int_{\Omega}
F(x)\eta(x)
\,\mathrm{d}^{4}x
=0
\]

for every smooth compactly supported function \(\eta\). Then

\[
F(x)=0
\]

at every interior point.

The reason is local. If \(F\) were positive or negative in a small neighbourhood, one could choose \(\eta\) supported in that neighbourhood with the same sign, producing a nonzero integral.

Applying this result to \(\delta S=0\) gives

\[
\frac{\partial\mathcal{L}}{\partial\phi}
-
\partial_{\mu}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
\right)
=0.
\]

Equivalently,

\[
\boxed{
\partial_{\mu}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
\right)
-
\frac{\partial\mathcal{L}}{\partial\phi}
=0
}.
\]

This is the Euler--Lagrange field equation for one real field.

\subsection{Comparison with the mechanical equation}

The mechanical Euler--Lagrange equation is

\[
\frac{\mathrm{d}}{\mathrm{d}t}
\left(
\frac{\partial L}{\partial\dot q}
\right)
-
\frac{\partial L}{\partial q}
=0.
\]

The field equation is

\[
\partial_{\mu}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
\right)
-
\frac{\partial\mathcal{L}}{\partial\phi}
=0.
\]

The correspondence is direct:

\[
q(t)
\longrightarrow
\phi(x),
\]

\[
\dot q
\longrightarrow
\partial_{\mu}\phi,
\]

and

\[
\frac{\mathrm{d}}{\mathrm{d}t}
\longrightarrow
\partial_{\mu}
\quad\text{with summation over }\mu.
\]

The single time derivative in mechanics becomes a spacetime divergence.

\subsection{Time-and-space form}

For a density written as

\[
\mathcal{L}
=
\mathcal{L}(
\phi,
\partial_{t}\phi,
\partial_{1}\phi,
\partial_{2}\phi,
\partial_{3}\phi,
x),
\]

the field equation expands to

\[
\boxed{
\frac{\partial}{\partial t}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{t}\phi)}
\right)
+
\sum_{i=1}^{3}
\frac{\partial}{\partial x^{i}}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{i}\phi)}
\right)
-
\frac{\partial\mathcal{L}}{\partial\phi}
=0
}.
\]

No metric convention is needed for this formula. The signs are determined by the signs already present in \(\mathcal{L}\).

\subsection{Example: kinetic, gradient, potential, and source terms}

Consider

\[
\mathcal{L}
=
\frac{\rho}{2}(\partial_{t}\phi)^{2}
-
\frac{\kappa}{2}|\nabla\phi|^{2}
-
V(\phi)
+
J\phi.
\]

First,

\[
\frac{\partial\mathcal{L}}
{\partial(\partial_{t}\phi)}
=
\rho\partial_{t}\phi,
\]

so

\[
\frac{\partial}{\partial t}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{t}\phi)}
\right)
=
\rho\partial_{t}^{2}\phi.
\]

For each spatial derivative,

\[
\frac{\partial\mathcal{L}}
{\partial(\partial_{i}\phi)}
=
-\kappa\partial_{i}\phi.
\]

Therefore

\[
\sum_{i=1}^{3}
\partial_{i}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{i}\phi)}
\right)
=
-\kappa\nabla^{2}\phi.
\]

Finally,

\[
\frac{\partial\mathcal{L}}{\partial\phi}
=
-V'(\phi)+J.
\]

The Euler--Lagrange equation is

\[
\rho\partial_{t}^{2}\phi
-
\kappa\nabla^{2}\phi
+
V'(\phi)
-
J
=0.
\]

Thus

\[
\boxed{
\rho\phi_{tt}
-
\kappa\nabla^{2}\phi
+
V'(\phi)
=
J
}.
\]

This single calculation displays the role of every standard term:

\begin{itemize}
    \item the time-derivative term produces local acceleration;
    \item the gradient term produces the Laplacian;
    \item the potential produces the local force \(V'(\phi)\);
    \item the source appears on the right-hand side.
\end{itemize}

\subsection{Functional derivative notation}

The coefficient of \(\delta\phi\) after integration by parts is called the functional derivative of the action:

\[
\boxed{
\frac{\delta S}{\delta\phi(x)}
=
\frac{\partial\mathcal{L}}{\partial\phi}
-
\partial_{\mu}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
\right)
}.
\]

The first variation may then be written, after the boundary term has been controlled, as

\[
\delta S
=
\int_{\Omega}
\frac{\delta S}{\delta\phi(x)}
\delta\phi(x)
\,\mathrm{d}^{4}x.
\]

The field equation is

\[
\frac{\delta S}{\delta\phi(x)}=0.
\]

This notation is compact, but it should not hide the integration-by-parts calculation from which it arises.

\subsection{Structural checks}

A derived field equation should be checked systematically.

\begin{enumerate}
    \item Every term must have the same physical dimensions.
    \item The equation should have the same number and type of free indices on both sides.
    \item The highest derivative order should agree with the derivative order of the Lagrangian density.
    \item Removing a source or potential should give the expected simpler equation.
    \item Constant fields should behave consistently when all derivative terms vanish.
\end{enumerate}

For the example above, setting \(V=0\) and \(J=0\) gives

\[
\rho\phi_{tt}-\kappa\nabla^{2}\phi=0,
\]

a wave equation. Setting the time derivatives to zero gives a static elliptic equation.

\subsection{Common mistakes}

The main errors in the derivation are:

\begin{enumerate}
    \item omitting the variation of \(\partial_{\mu}\phi\);
    \item failing to sum over the derivative index \(\mu\);
    \item treating \(\delta\phi\) and \(\partial_{\mu}\delta\phi\) as independent;
    \item losing the minus sign introduced by integration by parts;
    \item discarding the boundary term without a stated boundary condition;
    \item applying a metric sign convention before one has been defined;
    \item confusing the partial derivative of \(\mathcal{L}\) with the functional derivative of \(S\).
\end{enumerate}

\subsection{What has been established}

For a local action

\[
S[\phi]
=
\int_{\Omega}
\mathcal{L}(\phi,\partial_{\mu}\phi,x)
\,\mathrm{d}^{4}x,
\]

stationarity under arbitrary admissible variations gives

\[
\partial_{\mu}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
\right)
-
\frac{\partial\mathcal{L}}{\partial\phi}
=0.
\]

The equation is local because the variation can be localized around any interior point. The next section generalizes the derivation to several interacting fields and coupled equations.