\section{Boundary Terms in Spacetime}
\label{sec:boundary_terms_in_spacetime}

The derivation of a field equation always separates the variation into an interior term and a boundary term. The interior term produces the differential equation. The boundary term determines which boundary data are compatible with the variational principle.

A boundary term is not automatically zero. It vanishes only after the admissible variations, fall-off conditions, or natural boundary conditions have been specified.

\subsection{The general boundary contribution}

For one field with action

\[
S[\phi]
=
\int_{\Omega}
\mathcal{L}(\phi,\partial_{\mu}\phi,x)
\,\mathrm{d}^{4}x,
\]

the first variation is

\[
\begin{aligned}
\delta S
&=
\int_{\Omega}
\left[
\frac{\partial\mathcal{L}}{\partial\phi}
-
\partial_{\mu}
\left(
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
\right)
\right]
\delta\phi
\,\mathrm{d}^{4}x\\
&\quad+
\int_{\partial\Omega}
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
n_{\mu}\delta\phi
\,\mathrm{d}\Sigma.
\end{aligned}
\]

The second integral is the boundary contribution. It pairs the boundary variation \(\delta\phi\) with the normal component of

\[
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}.
\]

The variational problem is well posed only when this boundary contribution is controlled.

\subsection{The geometry of the boundary}

For a region

\[
\Omega=[t_{1},t_{2}]\times\Sigma,
\]

the boundary generally contains three parts:

\begin{enumerate}
    \item the initial spatial surface at \(t=t_{1}\);
    \item the final spatial surface at \(t=t_{2}\);
    \item the timelike side boundary generated by \(\partial\Sigma\) during the time interval.
\end{enumerate}

Thus boundary conditions may be required both in time and in space.

Fixing the initial and final field configurations means

\[
\delta\phi(\mathbf{x},t_{1})=0,
\qquad
\delta\phi(\mathbf{x},t_{2})=0.
\]

Fixing the field on the spatial boundary means

\[
\delta\phi=0
\qquad
\text{on }\partial\Sigma
\]

throughout the time interval.

These conditions play different physical roles and should not be combined without explanation.

\subsection{Dirichlet boundary conditions}

A Dirichlet condition prescribes the field value on the boundary:

\[
\phi|_{\partial\Omega}
=
f(x),
\]

where \(f\) is given.

Every varied field must have the same boundary value, so

\[
\boxed{
\delta\phi|_{\partial\Omega}=0
}.
\]

The boundary integral then vanishes directly.

This is the field-theoretic analogue of fixing the endpoints of a particle trajectory.

\subsection{Natural boundary conditions}

Suppose the field value is not fixed on part of the boundary. Then \(\delta\phi\) is arbitrary there. Stationarity requires the coefficient of that arbitrary boundary variation to vanish:

\[
\boxed{
 n_{\mu}
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi)}
=0
\qquad
\text{on the free part of }\partial\Omega
}.
\]

This is a natural boundary condition. It is derived from the action rather than imposed before the variation.

For a standard gradient energy, the natural condition is often a Neumann condition involving the normal derivative of the field.

\subsection{Neumann and mixed conditions}

A Neumann condition prescribes a normal derivative:

\[
\frac{\partial\phi}{\partial n}
=
g(x)
\qquad
\text{on }\partial\Omega.
\]

Here

\[
\frac{\partial\phi}{\partial n}
:=
n^{\mu}\partial_{\mu}\phi.
\]

A mixed or Robin condition combines the field and its normal derivative:

\[
a(x)\phi
+
b(x)\frac{\partial\phi}{\partial n}
=
c(x).
\]

Such conditions may arise after adding an explicit boundary term to the action. The complete variational principle then includes both bulk and boundary contributions.

\subsection{Explicit example: the vibrating string}

Consider

\[
S[u]
=
\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
\left[
\frac{\mu}{2}u_{t}^{2}
-
\frac{T}{2}u_{x}^{2}
\right]
\,\mathrm{d}x\,\mathrm{d}t.
\]

The first variation is

\[
\delta S
=
\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
\left[
\mu u_{t}\partial_{t}(\delta u)
-
T u_{x}\partial_{x}(\delta u)
\right]
\,\mathrm{d}x\,\mathrm{d}t.
\]

Integrating the time term by parts gives

\[
\begin{aligned}
&\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
\mu u_{t}\partial_{t}(\delta u)
\,\mathrm{d}x\,\mathrm{d}t\\
&\quad=
\left[
\int_{0}^{L}
\mu u_{t}\delta u
\,\mathrm{d}x
\right]_{t_{1}}^{t_{2}}
-
\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
\mu u_{tt}\delta u
\,\mathrm{d}x\,\mathrm{d}t.
\end{aligned}
\]

Integrating the spatial term by parts gives

\[
\begin{aligned}
&-
\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
T u_{x}\partial_{x}(\delta u)
\,\mathrm{d}x\,\mathrm{d}t\\
&\quad=
-
\int_{t_{1}}^{t_{2}}
\left[
T u_{x}\delta u
\right]_{0}^{L}
\,\mathrm{d}t
+
\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
T u_{xx}\delta u
\,\mathrm{d}x\,\mathrm{d}t.
\end{aligned}
\]

Therefore

\[
\begin{aligned}
\delta S
&=
\left[
\int_{0}^{L}
\mu u_{t}\delta u
\,\mathrm{d}x
\right]_{t_{1}}^{t_{2}}\\
&\quad-
\int_{t_{1}}^{t_{2}}
\left[
T u_{x}\delta u
\right]_{0}^{L}
\,\mathrm{d}t\\
&\quad+
\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
(-\mu u_{tt}+T u_{xx})
\delta u
\,\mathrm{d}x\,\mathrm{d}t.
\end{aligned}
\]

This formula displays the time boundary, the spatial boundary, and the interior equation separately.

\subsection{Fixed initial and final configurations}

If the initial and final string shapes are fixed, then

\[
\delta u(x,t_{1})=0,
\qquad
\delta u(x,t_{2})=0.
\]

The time-boundary contribution vanishes:

\[
\left[
\int_{0}^{L}
\mu u_{t}\delta u
\,\mathrm{d}x
\right]_{t_{1}}^{t_{2}}
=0.
\]

This condition does not require \(u_{t}\) to vanish at the initial or final time. It requires only the variation of the field value to vanish there.

\subsection{Fixed string endpoints}

For a string attached at both ends, the endpoint displacement is prescribed:

\[
u(0,t)=0,
\qquad
u(L,t)=0.
\]

Therefore

\[
\delta u(0,t)=0,
\qquad
\delta u(L,t)=0.
\]

The spatial boundary term vanishes independently of the value of \(u_{x}\).

\subsection{Free string endpoints}

If an endpoint is free to move transversely, \(\delta u\) is arbitrary there. Stationarity then requires

\[
T u_{x}=0
\]

at that endpoint. For nonzero tension,

\[
\boxed{
u_{x}=0
\qquad
\text{at a free endpoint}
}.
\]

This condition has a direct physical interpretation. The transverse component of the tension at the endpoint must vanish.

The variational principle therefore produces both the interior wave equation and the physically appropriate natural boundary condition.

\subsection{Boundary forces}

An external force applied at the endpoint can be included through a boundary action. For example, a time-dependent endpoint force \(F(t)\) at \(x=L\) may contribute

\[
S_{\mathrm{boundary}}
=
\int_{t_{1}}^{t_{2}}
F(t)u(L,t)
\,\mathrm{d}t.
\]

Its variation is

\[
\delta S_{\mathrm{boundary}}
=
\int_{t_{1}}^{t_{2}}
F(t)\delta u(L,t)
\,\mathrm{d}t.
\]

Combining this with the bulk boundary term gives a force-balance condition at \(x=L\). The precise sign depends on the orientation convention for the endpoint normal.

This example shows that boundary terms can encode actual interactions rather than merely mathematical corrections.

\subsection{Fields on unbounded regions}

If space is unbounded, the boundary is formally at spatial infinity. One must then impose fall-off conditions strong enough to make the surface integral vanish.

For example, a scalar field may be required to satisfy

\[
\phi(\mathbf{x},t)\longrightarrow 0
\]

and

\[
|\nabla\phi|
\longrightarrow 0
\]

sufficiently rapidly as \(|\mathbf{x}|\to\infty\).

The required rate depends on the spatial dimension and on the surface area growth. Saying only that the field becomes small is not always sufficient; the complete surface integral must tend to zero.

\subsection{Adding a total divergence}

If

\[
\mathcal{L}'
=
\mathcal{L}
+
\partial_{\mu}K^{\mu},
\]

then

\[
S'-S
=
\int_{\partial\Omega}
K^{\mu}n_{\mu}
\,\mathrm{d}\Sigma.
\]

The interior Euler--Lagrange equations are unchanged when the variation of this boundary contribution vanishes. However, the canonical boundary quantities and the natural boundary conditions may change.

Thus equivalent bulk equations do not imply that all boundary formulations are identical.

\subsection{Higher-derivative actions}

If the density depends on second derivatives,

\[
\mathcal{L}
=
\mathcal{L}(
\phi,
\partial_{\mu}\phi,
\partial_{\mu}\partial_{\nu}\phi),
\]

then two integrations by parts are needed. The boundary contribution generally contains both \(\delta\phi\) and derivatives of \(\delta\phi\).

Additional boundary data are therefore required. The present chapter focuses on first-derivative densities, for which fixing \(\phi\) on the boundary is sufficient for the standard derivation.

\subsection{Diagnostic questions}

Before removing a boundary term, one should ask:

\begin{enumerate}
    \item What is the complete boundary of the integration region?
    \item Are the initial and final field configurations fixed?
    \item Is the field fixed on the spatial boundary?
    \item If the field is free, what natural condition follows?
    \item Are there external boundary interactions?
    \item On an unbounded region, do the fall-off conditions make the surface integral vanish?
    \item Has a total divergence changed the boundary formulation?
\end{enumerate}

\subsection{What has been established}

The first variation of a field action contains a bulk term and a surface term. Dirichlet conditions eliminate the surface term by setting \(\delta\phi=0\) on the boundary. Free boundary values instead produce natural conditions involving the normal field momentum.

For the vibrating string, the same variation produces the interior wave equation, fixed-end conditions, or the free-end condition \(u_{x}=0\), depending on the admissible boundary variations.

The next section carries out the wave-equation derivation as a complete worked field-theory example.