\section{The Wave Equation from an Action}
\label{sec:wave_equation_from_action}

The vibrating string provides the simplest complete example of a dynamical field equation derived from an action. Its Lagrangian density contains a time-derivative term describing inertia and a spatial-gradient term describing elastic restoring forces.

We will derive the equation, identify the boundary assumptions, verify a plane-wave solution, and interpret the coefficients.

\subsection{The field and the model}

Let

\[
u=u(x,t)
\]

be the small transverse displacement of a string whose equilibrium position is the interval

\[
0\leq x\leq L.
\]

We assume:

\begin{enumerate}
    \item the displacement is small;
    \item the string has constant mass density per unit length \(\mu\);
    \item the tension \(T\) is constant;
    \item the motion is transverse;
    \item damping and external forces are absent.
\end{enumerate}

The field history is defined on

\[
\Omega=[t_{1},t_{2}]\times[0,L].
\]

\subsection{The Lagrangian density}

The kinetic-energy density is

\[
\mathcal{T}
=
\frac{\mu}{2}u_{t}^{2},
\]

where

\[
u_{t}=\frac{\partial u}{\partial t}.
\]

For small slopes, the elastic-energy density is

\[
\mathcal{V}
=
\frac{T}{2}u_{x}^{2},
\]

where

\[
u_{x}=\frac{\partial u}{\partial x}.
\]

The Lagrangian density is kinetic minus potential energy density:

\[
\boxed{
\mathcal{L}
=
\frac{\mu}{2}u_{t}^{2}
-
\frac{T}{2}u_{x}^{2}
}.
\]

The action is

\[
\boxed{
S[u]
=
\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
\left[
\frac{\mu}{2}u_{t}^{2}
-
\frac{T}{2}u_{x}^{2}
\right]
\,\mathrm{d}x\,\mathrm{d}t
}.
\]

\subsection{Derivation using the Euler--Lagrange field equation}

The density does not depend explicitly on \(u\), so

\[
\frac{\partial\mathcal{L}}{\partial u}=0.
\]

For the time derivative,

\[
\frac{\partial\mathcal{L}}{\partial u_{t}}
=
\mu u_{t}.
\]

Therefore

\[
\frac{\partial}{\partial t}
\left(
\frac{\partial\mathcal{L}}{\partial u_{t}}
\right)
=
\mu u_{tt}.
\]

For the spatial derivative,

\[
\frac{\partial\mathcal{L}}{\partial u_{x}}
=
-Tu_{x}.
\]

Therefore

\[
\frac{\partial}{\partial x}
\left(
\frac{\partial\mathcal{L}}{\partial u_{x}}
\right)
=
-Tu_{xx}.
\]

The field Euler--Lagrange equation is

\[
\frac{\partial}{\partial t}
\left(
\frac{\partial\mathcal{L}}{\partial u_{t}}
\right)
+
\frac{\partial}{\partial x}
\left(
\frac{\partial\mathcal{L}}{\partial u_{x}}
\right)
-
\frac{\partial\mathcal{L}}{\partial u}
=0.
\]

Substitution gives

\[
\mu u_{tt}-Tu_{xx}=0.
\]

Thus

\[
\boxed{
\mu
\frac{\partial^{2}u}{\partial t^{2}}
=
T
\frac{\partial^{2}u}{\partial x^{2}}
}.
\]

Defining

\[
c^{2}=\frac{T}{\mu},
\]

we obtain

\[
\boxed{
\frac{\partial^{2}u}{\partial t^{2}}
=
c^{2}
\frac{\partial^{2}u}{\partial x^{2}}
}.
\]

This is the one-dimensional wave equation.

\subsection{Direct variation as a check}

Let

\[
u_{\varepsilon}=u+\varepsilon\eta.
\]

Then

\[
\delta u=\eta,
\qquad
\delta u_{t}=\eta_{t},
\qquad
\delta u_{x}=\eta_{x}.
\]

The first variation is

\[
\delta S
=
\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
\left(
\mu u_{t}\eta_{t}
-
T u_{x}\eta_{x}
\right)
\,\mathrm{d}x\,\mathrm{d}t.
\]

After integration by parts in time and space,

\[
\begin{aligned}
\delta S
&=
\left[
\int_{0}^{L}
\mu u_{t}\eta
\,\mathrm{d}x
\right]_{t_{1}}^{t_{2}}\\
&\quad-
\int_{t_{1}}^{t_{2}}
\left[
T u_{x}\eta
\right]_{0}^{L}
\,\mathrm{d}t\\
&\quad+
\int_{t_{1}}^{t_{2}}
\int_{0}^{L}
(-\mu u_{tt}+Tu_{xx})
\eta
\,\mathrm{d}x\,\mathrm{d}t.
\end{aligned}
\]

For fixed initial and final configurations and fixed string endpoints, all boundary variations vanish. Stationarity for arbitrary interior \(\eta\) gives

\[
-\mu u_{tt}+Tu_{xx}=0,
\]

which is equivalent to the equation obtained from the compact formula.

\subsection{Physical interpretation}

The term

\[
\mu u_{tt}
\]

is mass per unit length multiplied by transverse acceleration. It represents local inertia.

The term

\[
T u_{xx}
\]

is proportional to local curvature. A straight segment has \(u_{xx}=0\) and no net transverse restoring force. A curved segment is accelerated toward a straighter configuration.

The equation therefore expresses local force balance:

\[
\text{inertial response}
=
\text{elastic restoring force}.
\]

\subsection{Dimensional check}

The dimensions of tension are

\[
[T]=\mathrm{N}=\mathrm{kg\,m\,s^{-2}},
\]

and the dimensions of linear mass density are

\[
[\mu]=\mathrm{kg\,m^{-1}}.
\]

Therefore

\[
\left[\frac{T}{\mu}\right]
=
\mathrm{m^{2}\,s^{-2}}.
\]

Hence

\[
[c]
=
\mathrm{m\,s^{-1}},
\]

so \(c\) has the dimensions of speed.

The wave speed is

\[
\boxed{
c=\sqrt{\frac{T}{\mu}}
}.
\]

Increasing the tension increases the propagation speed, while increasing the mass density decreases it.

\subsection{Plane-wave verification}

Consider the complex trial solution

\[
u(x,t)
=
A e^{i(kx-\omega t)}.
\]

Its second derivatives are

\[
u_{tt}
=
-\omega^{2}u,
\]

and

\[
u_{xx}
=
-k^{2}u.
\]

Substitution into

\[
u_{tt}=c^{2}u_{xx}
\]

gives

\[
-\omega^{2}u
=
-c^{2}k^{2}u.
\]

For a nonzero wave amplitude,

\[
\boxed{
\omega^{2}=c^{2}k^{2}
}.
\]

Thus

\[
\omega=c|k|
\]

for positive frequency. The phase speed is

\[
\frac{\omega}{|k|}=c.
\]

The real and imaginary parts of the complex exponential are real solutions because the wave equation is linear with real coefficients.

\subsection{General travelling waves}

The wave equation admits solutions of the form

\[
u(x,t)=f(x-ct),
\]

where \(f\) is twice differentiable. Indeed,

\[
u_{t}=-cf'(x-ct),
\qquad
u_{tt}=c^{2}f''(x-ct),
\]

while

\[
u_{x}=f'(x-ct),
\qquad
u_{xx}=f''(x-ct).
\]

Therefore

\[
u_{tt}=c^{2}u_{xx}.
\]

Similarly,

\[
u(x,t)=g(x+ct)
\]

is a left-moving solution. The general solution on an unbounded line can be written as

\[
\boxed{
u(x,t)=f(x-ct)+g(x+ct)
}.
\]

Boundary conditions restrict which combinations are allowed on a finite interval.

\subsection{Boundary conditions}

For fixed endpoints,

\[
u(0,t)=0,
\qquad
u(L,t)=0.
\]

For free endpoints, stationarity gives

\[
u_{x}(0,t)=0,
\qquad
u_{x}(L,t)=0.
\]

Periodic boundary conditions take the form

\[
u(0,t)=u(L,t),
\]

and

\[
u_{x}(0,t)=u_{x}(L,t).
\]

The differential equation alone does not determine a unique physical motion. Initial data and boundary data are part of the model.

\subsection{Energy density}

The energy density associated with the string is

\[
\boxed{
\mathcal{E}
=
\frac{\mu}{2}u_{t}^{2}
+
\frac{T}{2}u_{x}^{2}
}.
\]

The signs differ from the Lagrangian density because energy is kinetic plus potential, whereas the Lagrangian is kinetic minus potential.

The total energy is

\[
E(t)
=
\int_{0}^{L}
\mathcal{E}(x,t)
\,\mathrm{d}x.
\]

Differentiating,

\[
\frac{\mathrm{d}E}{\mathrm{d}t}
=
\int_{0}^{L}
\left(
\mu u_{t}u_{tt}
+
T u_{x}u_{xt}
\right)
\,\mathrm{d}x.
\]

Using the field equation and integrating the second term by parts gives

\[
\boxed{
\frac{\mathrm{d}E}{\mathrm{d}t}
=
\left[T u_{x}u_{t}\right]_{0}^{L}
}.
\]

The change of energy inside the string is determined by energy flow through its endpoints. For fixed endpoints, \(u_{t}=0\) there, and the total energy is conserved.

\subsection{Adding an external source}

A distributed external force density \(J(x,t)\) can be included through

\[
\mathcal{L}
=
\frac{\mu}{2}u_{t}^{2}
-
\frac{T}{2}u_{x}^{2}
+
J u.
\]

The source is prescribed, so \(\delta J=0\). The field equation becomes

\[
\boxed{
\mu u_{tt}-Tu_{xx}=J
}.
\]

The homogeneous wave equation describes free propagation. The sourced equation describes waves generated or driven by an external influence.

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item using energy density instead of Lagrangian density in the action;
    \item losing the minus sign in the gradient term;
    \item forgetting to vary both \(u_{t}\) and \(u_{x}\);
    \item discarding time and spatial boundary terms without separate justifications;
    \item assuming a plane wave is allowed without checking the boundary conditions;
    \item writing \(c=T/\mu\) instead of \(c^{2}=T/\mu\);
    \item confusing the source-free and driven equations.
\end{enumerate}

\subsection{What has been established}

The action

\[
S[u]
=
\int\!\int
\left[
\frac{\mu}{2}u_{t}^{2}
-
\frac{T}{2}u_{x}^{2}
\right]
\,\mathrm{d}x\,\mathrm{d}t
\]

produces the wave equation

\[
u_{tt}=c^{2}u_{xx},
\qquad
c^{2}=\frac{T}{\mu}.
\]

The derivation identifies the interior dynamics and the boundary conditions within one variational calculation. Plane waves satisfy the dispersion relation \(\omega^{2}=c^{2}k^{2}\), and the associated energy changes only through boundary flux.

The next section turns to static variational problems and derives the Laplace and Poisson equations from energy functionals.