\section{Laplace and Poisson Equations}
\label{sec:laplace_and_poisson_equations}

Not every variational problem describes time evolution. Static field configurations may be selected by making an energy functional stationary. The resulting Euler--Lagrange equation is then an equation in space rather than spacetime.

The Laplace and Poisson equations provide the basic examples. They appear in electrostatics, gravitation, steady heat flow, incompressible potential flow, and equilibrium elasticity.

\subsection{A static energy functional}

Let \(\phi(\mathbf{x})\) be a real scalar field on a spatial region \(\Sigma\). Consider

\[
\boxed{
E[\phi]
=
\int_{\Sigma}
\left[
\frac{\kappa}{2}|\nabla\phi|^{2}
-
\rho(\mathbf{x})\phi
\right]
\,\mathrm{d}^{3}x
}.
\]

Here

\begin{itemize}
    \item \(\kappa>0\) is a constant;
    \item \(\rho(\mathbf{x})\) is a prescribed source;
    \item the gradient term penalizes spatial variation;
    \item the source term favours field values aligned with \(\rho\).
\end{itemize}

The field \(\phi\) is varied, while \(\rho\) is held fixed.

\subsection{Variation of the gradient term}

Introduce

\[
\phi_{\varepsilon}
=
\phi+\varepsilon\eta.
\]

Then

\[
\delta\phi=\eta,
\qquad
\delta(\nabla\phi)=\nabla\eta.
\]

The variation of the gradient contribution is

\[
\begin{aligned}
\delta
\left(
\frac{\kappa}{2}|\nabla\phi|^{2}
\right)
&=
\kappa\nabla\phi\cdot\nabla\eta.
\end{aligned}
\]

The source contribution varies as

\[
\delta(-\rho\phi)
=
-\rho\eta.
\]

Therefore

\[
\delta E
=
\int_{\Sigma}
\left[
\kappa\nabla\phi\cdot\nabla\eta
-
\rho\eta
\right]
\,\mathrm{d}^{3}x.
\]

\subsection{Spatial integration by parts}

The identity

\[
\nabla\cdot(\eta\nabla\phi)
=
\nabla\eta\cdot\nabla\phi
+
\eta\nabla^{2}\phi
\]

implies

\[
\nabla\phi\cdot\nabla\eta
=
\nabla\cdot(\eta\nabla\phi)
-
\eta\nabla^{2}\phi.
\]

Thus

\[
\begin{aligned}
\delta E
&=
\kappa
\int_{\partial\Sigma}
\eta
\frac{\partial\phi}{\partial n}
\,\mathrm{d}A\\
&\quad-
\int_{\Sigma}
\left(
\kappa\nabla^{2}\phi+ho
\right)
\eta
\,\mathrm{d}^{3}x.
\end{aligned}
\]

Here

\[
\frac{\partial\phi}{\partial n}
=
\mathbf{n}\cdot\nabla\phi
\]

is the outward normal derivative.

For fixed boundary values,

\[
\eta=0
\qquad
\text{on }\partial\Sigma,
\]

so the surface term vanishes. Stationarity for arbitrary interior \(\eta\) gives

\[
\kappa\nabla^{2}\phi+ho=0.
\]

Therefore

\[
\boxed{
\nabla^{2}\phi
=
-\frac{\rho}{\kappa}
}.
\]

This is the Poisson equation for the chosen sign convention.

\subsection{The Laplace equation}

In a source-free region,

\[
\rho=0.
\]

The field equation becomes

\[
\boxed{
\nabla^{2}\phi=0
}.
\]

This is the Laplace equation. A function satisfying it is called harmonic.

The distinction between the two equations is therefore simple:

\[
\text{Poisson equation: source present},
\]

\[
\text{Laplace equation: source absent}.
\]

A field may satisfy the Poisson equation in a region containing sources and the Laplace equation in neighbouring source-free regions.

\subsection{Electrostatic interpretation}

For an electrostatic potential \(\Phi\), choose

\[
\kappa=\varepsilon_{0}
\]

and let \(\rho\) denote the electric charge density. The functional is

\[
E[\Phi]
=
\int_{\Sigma}
\left[
\frac{\varepsilon_{0}}{2}|\nabla\Phi|^{2}
-
\rho\Phi
\right]
\,\mathrm{d}^{3}x.
\]

Stationarity gives

\[
\boxed{
\nabla^{2}\Phi
=
-\frac{\rho}{\varepsilon_{0}}
}.
\]

Using

\[
\mathbf{E}=-\nabla\Phi,
\]

we find

\[
\nabla\cdot\mathbf{E}
=
-\nabla^{2}\Phi
=
\frac{\rho}{\varepsilon_{0}},
\]

which is Gauss's law in differential form.

This provides an early example of a familiar field equation emerging from a variational functional.

\subsection{Natural boundary condition}

If the boundary value of \(\phi\) is free, then \(\eta\) is arbitrary on \(\partial\Sigma\). Stationarity requires

\[
\boxed{
\kappa
\frac{\partial\phi}{\partial n}
=0
\qquad
\text{on the free boundary}
}.
\]

For nonzero \(\kappa\),

\[
\frac{\partial\phi}{\partial n}=0.
\]

This is the homogeneous Neumann condition. It says that the field has no normal gradient through the boundary.

A nonzero prescribed normal derivative can be generated by adding an appropriate boundary source term to the functional.

\subsection{Dirichlet and Neumann data}

Two common boundary-value problems are:

\begin{enumerate}
    \item Dirichlet problem: specify \(\phi\) on \(\partial\Sigma\);
    \item Neumann problem: specify \(\partial\phi/\partial n\) on \(\partial\Sigma\).
\end{enumerate}

For the Dirichlet problem, admissible variations vanish on the boundary. For the Neumann problem, the field value may vary, but the normal derivative is constrained.

A pure Neumann problem determines \(\phi\) only up to an additive constant, because

\[
\nabla^{2}(\phi+C)
=
\nabla^{2}\phi.
\]

Additional information, such as the average value of \(\phi\), is needed to fix that constant.

\subsection{Compatibility condition for a Neumann problem}

Integrate the Poisson equation over \(\Sigma\):

\[
\int_{\Sigma}
\nabla^{2}\phi
\,\mathrm{d}^{3}x
=
-\frac{1}{\kappa}
\int_{\Sigma}
\rho
\,\mathrm{d}^{3}x.
\]

By the divergence theorem,

\[
\int_{\partial\Sigma}
\frac{\partial\phi}{\partial n}
\,\mathrm{d}A
=
-\frac{1}{\kappa}
\int_{\Sigma}
\rho
\,\mathrm{d}^{3}x.
\]

Thus prescribed Neumann data cannot be arbitrary. Their total flux must be compatible with the total source inside the region.

For homogeneous Neumann data,

\[
\frac{\partial\phi}{\partial n}=0,
\]

the compatibility condition is

\[
\int_{\Sigma}
\rho
\,\mathrm{d}^{3}x=0.
\]

\subsection{One-dimensional worked example}

Consider the interval

\[
0\leq x\leq L
\]

with constant source \(\rho_{0}\). The Poisson equation is

\[
\frac{\mathrm{d}^{2}\phi}{\mathrm{d}x^{2}}
=
-\frac{\rho_{0}}{\kappa}.
\]

Integrating once,

\[
\frac{\mathrm{d}\phi}{\mathrm{d}x}
=
-\frac{\rho_{0}}{\kappa}x+C_{1}.
\]

Integrating again,

\[
\phi(x)
=
-\frac{\rho_{0}}{2\kappa}x^{2}
+C_{1}x+C_{2}.
\]

Impose Dirichlet conditions

\[
\phi(0)=0,
\qquad
\phi(L)=0.
\]

The first condition gives

\[
C_{2}=0.
\]

The second gives

\[
0
=
-\frac{\rho_{0}}{2\kappa}L^{2}
+C_{1}L,
\]

so

\[
C_{1}
=
\frac{\rho_{0}L}{2\kappa}.
\]

Therefore

\[
\boxed{
\phi(x)
=
\frac{\rho_{0}}{2\kappa}
x(L-x)
}.
\]

The solution is zero at both endpoints and reaches an extremum in the middle of the interval.

\subsection{Verification of the worked solution}

Differentiate twice:

\[
\phi'(x)
=
\frac{\rho_{0}}{2\kappa}(L-2x),
\]

and

\[
\phi''(x)
=
-\frac{\rho_{0}}{\kappa}.
\]

Thus the Poisson equation is satisfied. Also,

\[
\phi(0)=0,
\qquad
\phi(L)=0.
\]

Both the differential equation and the boundary conditions have been checked.

\subsection{Uniqueness for the Dirichlet problem}

Suppose \(\phi_{1}\) and \(\phi_{2}\) satisfy the same Poisson equation and the same Dirichlet boundary data. Define

\[
w=\phi_{1}-\phi_{2}.
\]

Then

\[
\nabla^{2}w=0
\]

and

\[
w=0
\qquad
\text{on }\partial\Sigma.
\]

Using integration by parts,

\[
\int_{\Sigma}
|\nabla w|^{2}
\,\mathrm{d}^{3}x
=
\int_{\partial\Sigma}
w\frac{\partial w}{\partial n}
\,\mathrm{d}A
-
\int_{\Sigma}
w\nabla^{2}w
\,\mathrm{d}^{3}x.
\]

Both terms on the right vanish, so

\[
\int_{\Sigma}|\nabla w|^{2}\,\mathrm{d}^{3}x=0.
\]

Therefore

\[
\nabla w=0,
\]

so \(w\) is constant. Since \(w=0\) on the boundary,

\[
w=0
\]

throughout the region. Hence

\[
\boxed{
\phi_{1}=\phi_{2}
}.
\]

The Dirichlet solution is unique under the stated regularity assumptions.

\subsection{Minimum-energy interpretation}

Let \(\phi_{*}\) satisfy the Poisson equation with fixed boundary values. Consider

\[
\phi=\phi_{*}+\eta,
\qquad
\eta|_{\partial\Sigma}=0.
\]

Expanding the functional gives

\[
E[\phi]
=
E[\phi_{*}]
+
\delta E[\phi_{*};\eta]
+
\frac{\kappa}{2}
\int_{\Sigma}
|\nabla\eta|^{2}
\,\mathrm{d}^{3}x.
\]

The first variation vanishes because \(\phi_{*}\) is stationary. Since \(\kappa>0\),

\[
E[\phi]-E[\phi_{*}]
=
\frac{\kappa}{2}
\int_{\Sigma}
|\nabla\eta|^{2}
\,\mathrm{d}^{3}x
\geq 0.
\]

Thus the stationary solution is a minimum of the functional for fixed Dirichlet data.

This is stronger than stationarity and follows from the positive quadratic gradient term.

\subsection{Relation to time-dependent field theory}

The static functional may be obtained from a dynamical density by restricting attention to time-independent fields. For

\[
\mathcal{L}
=
\frac{\rho_{m}}{2}\phi_{t}^{2}
-
\frac{\kappa}{2}|\nabla\phi|^{2}
+
\rho\phi,
\]

the field equation is

\[
\rho_{m}\phi_{tt}
-
\kappa\nabla^{2}\phi
=
\rho.
\]

For a static configuration,

\[
\phi_{tt}=0,
\]

so

\[
\nabla^{2}\phi
=
-\frac{\rho}{\kappa}.
\]

The Poisson equation is therefore the equilibrium limit of a broader dynamical theory.

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item confusing the energy functional with a time-dependent action;
    \item losing the minus sign produced by spatial integration by parts;
    \item varying the prescribed source \(\rho\);
    \item solving the differential equation without imposing boundary conditions;
    \item assuming a Neumann solution is unique without fixing its additive constant;
    \item ignoring the compatibility condition for prescribed flux;
    \item failing to verify both the equation and the boundary data.
\end{enumerate}

\subsection{What has been established}

The static functional

\[
E[\phi]
=
\int_{\Sigma}
\left[
\frac{\kappa}{2}|\nabla\phi|^{2}
-
\rho\phi
\right]
\,\mathrm{d}^{3}x
\]

produces the Poisson equation

\[
\nabla^{2}\phi=-\frac{\rho}{\kappa}.
\]

In source-free regions this reduces to the Laplace equation. Boundary variations distinguish Dirichlet, Neumann, and natural boundary conditions. For fixed Dirichlet data and \(\kappa>0\), the stationary solution is unique and minimizes the energy functional.

The final section adds local potentials and interactions, derives nonlinear field equations, and introduces canonical field momentum and the Hamiltonian density.