\section{Review Exercises}

\subsection{Elementary computational and graphical exercises}

\begin{enumerate}

\item Write a parametric equation of the line through $P=(2,-1)$ with direction
$\mathbf v=(3,4)$. Find the points corresponding to $t=-1,0,2$ and plot them.

\item Find a normal equation and an implicit equation of the line through
$P=(1,3)$ with normal vector $\mathbf n=(2,-1)$. Check that $P$ satisfies the
resulting equation.

\item Convert
\[
2x-3y=6
\]
to slope-intercept form. Find both axis intercepts and give one direction
vector and one normal vector.

\item Find the line through
\[
A=(-1,2),\qquad B=(3,-4)
\]
in parametric, implicit, and slope-intercept form. Verify that both points
satisfy every description.

\item Find and plot the intersection of
\[
L_1:y=2x-1,
\qquad
L_2:x+y=5.
\]
Verify the point in both equations.

\item Determine whether each pair of lines is parallel, perpendicular,
identical, or neither:
\begin{enumerate}
\item $2x-y=1$ and $4x-2y=7$;
\item $x+2y=3$ and $2x-y=4$;
\item $3x+y=5$ and $6x+2y=10$.
\end{enumerate}

\item Compute the distance from $P=(3,1)$ to the line
\[
2x-y-4=0.
\]
Find the foot of the perpendicular and verify the distance directly from the
coordinates of the two points.

\item Write a parametric equation of the plane through
\[
P=(1,2,0)
\]
with direction vectors
\[
\mathbf u=(1,0,1),\qquad \mathbf v=(0,2,-1).
\]
Produce four points of the plane.

\item For the plane from the previous exercise, compute
\[
\mathbf n=\mathbf u\times\mathbf v
\]
and derive an implicit equation. Verify that the four previously found points
satisfy it.

\item Find the plane through
\[
P=(1,0,0),\qquad Q=(0,1,0),\qquad R=(0,0,1).
\]
Give both parametric and normal forms, and compute the distance from the origin
to this plane.

\end{enumerate}

\subsection{Development and reasoning exercises}

\begin{enumerate}

\item A line is written as
\[
L=\{P_0+t\mathbf v:t\in\mathbb R\}.
\]
Explain why replacing $\mathbf v$ by $2\mathbf v$ or by $-\mathbf v$ produces
the same geometric line, although the parameter traverses it differently.

\item Explain why slope-intercept form cannot represent a vertical line without
a special case, while parametric and normal forms handle it naturally. Give one
vertical line in all applicable forms.

\item Starting with a direction vector
\[
\mathbf v=(a,b),
\]
construct a normal vector and derive an implicit equation of the line through a
point $P_0=(x_0,y_0)$. Explain why the direction and normal descriptions encode
the same line.

\item For
\[
L=\{P_0+t\mathbf v:t\in\mathbb R\}
\]
and a point $P$, derive the formula for the nearest point
\[
Q=P_0+\operatorname{proj}_{\mathbf v}(P-P_0).
\]
Explain why every other point of the line is farther from $P$.

\item A student says that two lines are perpendicular whenever their direction
vectors have zero dot product. Explain why this is sufficient in the plane but
would require an additional intersection condition for geometric lines in
space. Use a drawing or concrete example to clarify the distinction.

\item Compare
\[
P=P_0+t\mathbf v
\]
with
\[
P=P_0+s\mathbf u+t\mathbf v.
\]
Use one concrete line and one concrete plane to explain why one independent
direction produces a one-dimensional set and two nonparallel directions
produce a two-dimensional set.

\item The plane generated by $\mathbf u$ and $\mathbf v$ uses
$\mathbf u\times\mathbf v$ as a normal. Verify the perpendicularity by dot
products in one numerical example and explain why reversing the order changes
the normal vector but not the plane.

\item Three points determine a plane only when they are not collinear. Give one
set of three collinear points and one set of three noncollinear points. Explain
what happens to
\[
(Q-P)\times(R-P)
\]
in each case.

\item Determine the relative position of
\[
\Pi_1:x+y+z=1,
\qquad
\Pi_2:x-y+z=3.
\]
Find a direction vector of their intersection line and one point on it. Explain
why the cross product of the normals gives the correct direction.

\item Give a strategy for choosing between parametric and normal descriptions
when solving each of the following tasks: constructing an object from points or
directions, testing membership, finding an intersection, checking
parallelism or perpendicularity, and computing distance. Support the strategy
with one line example and one plane example.

\end{enumerate}
