\section{Why this chapter is necessary}

This chapter collects the minimal mathematical language that will be used throughout these notes. It is not a replacement for a mathematics course. The reader is expected to have already met elementary algebra, functions, analytic geometry, vectors, matrices, derivatives, and integrals. However, experience shows that having passed a mathematics course does not always mean being able to use mathematics fluently in a new context.

The purpose of this chapter is therefore practical. We recall the mathematical objects and notations that will appear later, before they are used in physical models. At this stage we do not develop the physical applications in detail. Those applications will appear in later chapters, especially in kinematics and dynamics. Here the goal is to make the mathematical language readable.

Mathematics in these notes should not be treated as a collection of symbolic tricks. A formula is not only something into which numbers are substituted. A formula expresses a relation. A function describes dependence. A derivative describes local change. An integral describes accumulation. A vector describes magnitude and direction. A matrix can describe a transformation or organize several quantities at once.

The reader should use this chapter as a short toolbox. The tools introduced here will return later with physical meaning.

\section{Scientific notation}

Physics and science in general often deal with very large and very small numbers. Writing such numbers in ordinary decimal notation is inconvenient and often makes the scale of the quantity difficult to see. For this reason we use scientific notation.

A number written in scientific notation has the form
\[
a \times 10^n,
\]
where \(n\) is an integer and usually
\[
1 \leq |a| < 10.
\]
The number \(a\) tells us the significant digits, while \(10^n\) tells us the scale.

For example,
\[
2.7 \times 10^{17}
\]
means
\[
270000000000000000.
\]
The exponent \(17\) tells us that the decimal point has been moved seventeen places to the right.

Similarly,
\[
1.602 \times 10^{-19}
\]
means
\[
0.0000000000000000001602.
\]
The exponent \(-19\) tells us that the decimal point has been moved nineteen places to the left.

This notation is not optional. It is a basic part of scientific language. Without it, many numbers used in science would be almost unreadable.

For example, the elementary charge is approximately
\[
e = 1.602 \times 10^{-19}\,\mathrm{C}.
\]
Written without scientific notation, this is
\[
e = 0.0000000000000000001602\,\mathrm{C}.
\]
The scientific form is much easier to read and compare.

The Avogadro constant is approximately
\[
N_A = 6.022 \times 10^{23}\,\mathrm{mol}^{-1}.
\]
Written in full decimal form, this is approximately
\[
N_A = 602200000000000000000000\,\mathrm{mol}^{-1}.
\]
The scientific notation immediately shows that this is a number of order \(10^{23}\).

The speed of light in vacuum is approximately
\[
c = 3.00 \times 10^8\,\mathrm{m/s}.
\]
In ordinary decimal notation, this is
\[
c = 300000000\,\mathrm{m/s}.
\]

Scientific notation also makes multiplication and division easier. We use the rules
\[
10^a \cdot 10^b = 10^{a+b}
\]
and
\[
\frac{10^a}{10^b} = 10^{a-b}.
\]
For example,
\[
(2 \times 10^3)(3 \times 10^4)
= 6 \times 10^{3+4}
= 6 \times 10^7.
\]
Similarly,
\[
\frac{8 \times 10^6}{2 \times 10^2}
= 4 \times 10^{6-2}
= 4 \times 10^4.
\]

The exponent of ten gives the order of magnitude. In many scientific situations, understanding the order of magnitude is more important than knowing many decimal digits. A result of order \(10^{-19}\) and a result of order \(10^{19}\) describe completely different scales.

The reader is expected to understand notation such as
\[
10^{-6}, \qquad 10^3, \qquad 2.7 \times 10^{17}, \qquad 6.022 \times 10^{23}.
\]
Without this basic skill, it is not possible to read many scientific formulas or numerical examples.

\section{Numbers, quantities, and units}

A number by itself is often not enough. In science, we usually work with quantities. A quantity has a numerical value, a unit, and a meaning.

For example, the expression
\[
5
\]
is only a number. It becomes a physical quantity only when we know what it represents:
\[
5\,\mathrm{m}, \qquad 5\,\mathrm{s}, \qquad 5\,\mathrm{kg}, \qquad 5\,\mathrm{C}.
\]
These expressions do not mean the same thing. The unit is part of the information.

Units must be treated consistently. We can add quantities with the same unit:
\[
2\,\mathrm{m} + 3\,\mathrm{m} = 5\,\mathrm{m}.
\]
But the expression
\[
2\,\mathrm{m} + 3\,\mathrm{s}
\]
is not a meaningful sum, because metres and seconds describe different kinds of quantities.

Units also help us check formulas. If the left-hand side of an equation has a certain unit, then the right-hand side must have the same unit. This simple observation is one of the most useful ways to detect errors.

For example, if a formula gives a quantity measured in metres, every term that is added to produce that quantity must also have the unit of metres. If a calculation produces seconds where metres are expected, something is wrong.

This type of check does not solve the problem for us, but it protects us from many meaningless results.

\section{Variables, constants, and parameters}

Mathematical notation uses symbols. These symbols may play different roles. Some symbols represent variables, some represent constants, and some represent parameters.

A variable is a quantity that is allowed to change. For example, in the expression
\[
f(x) = x^2 + 1,
\]
the symbol \(x\) is the variable. Different values of \(x\) give different values of \(f(x)\).

A constant has a fixed value. For example, in
\[
f(x) = 3x + 2,
\]
the numbers \(3\) and \(2\) are constants.

A parameter is a symbol that is treated as fixed in one situation, but may have different values in different situations. For example,
\[
f(x) = ax + b
\]
describes a whole family of linear functions. For a particular function, \(a\) and \(b\) are fixed. But if we choose different values of \(a\) and \(b\), we get a different function.

For example, if
\[
a = 2, \qquad b = 5,
\]
then
\[
f(x) = 2x + 5.
\]
If
\[
a = -1, \qquad b = 0,
\]
then
\[
f(x) = -x.
\]

When reading a formula, we should ask:
\begin{itemize}
    \item Which symbols are variables?
    \item Which symbols are constants?
    \item Which symbols are parameters?
    \item What values are allowed?
    \item What changes when one symbol changes?
    \item What remains fixed?
\end{itemize}

This distinction is essential for understanding mathematical models.

\section{Functions}

A function assigns an output to an input. If \(f\) is a function and \(x\) is an input, then
\[
f(x)
\]
denotes the output corresponding to \(x\).

For example, let
\[
f(x) = x^2 + 1.
\]
Then
\[
f(0) = 0^2 + 1 = 1,
\]
\[
f(1) = 1^2 + 1 = 2,
\]
and
\[
f(3) = 3^2 + 1 = 10.
\]

The notation \(f(x)\) should not be read as multiplication of \(f\) and \(x\). It means: the value of the function \(f\) at the argument \(x\).

A function may also depend on more than one variable. For example,
\[
g(x,y) = x^2 + y^2
\]
has two inputs. If
\[
x = 3, \qquad y = 4,
\]
then
\[
g(3,4) = 3^2 + 4^2 = 25.
\]

Functions are important because they describe dependence. When we write
\[
y = f(x),
\]
we say that \(y\) depends on \(x\). The question is not only how to compute values, but also how to understand the dependence.

For example, if
\[
f(x) = 2x + 1,
\]
then increasing \(x\) by \(1\) increases \(f(x)\) by \(2\). If
\[
f(x) = x^2,
\]
then the change in \(f(x)\) depends on the current value of \(x\). The function is no longer linear.

\section{Graphs}

A graph is a visual representation of a relation. For a function
\[
y = f(x),
\]
we usually place \(x\) on the horizontal axis and \(y\) on the vertical axis. Each point on the graph has coordinates
\[
(x,f(x)).
\]

A graph is not only a picture. It contains information. From a graph we can often see:
\begin{itemize}
    \item where the function is increasing,
    \item where it is decreasing,
    \item where it is constant,
    \item where it changes rapidly,
    \item where it changes slowly,
    \item whether it is linear or nonlinear,
    \item whether it has zeros,
    \item whether it has maximum or minimum values.
\end{itemize}

For example, the graph of
\[
f(x) = 2x + 1
\]
is a straight line. The graph of
\[
f(x) = x^2
\]
is a parabola.

The reader should become comfortable moving between formulas and graphs. A formula gives a symbolic description. A graph gives a visual description. They are two ways of representing the same relation.

\section{Cartesian coordinates}

To describe points in a line, plane, or space, we use coordinates. The most common coordinate system is the Cartesian coordinate system.

On a line, a point is described by one coordinate:
\[
x.
\]
In a plane, a point is described by two coordinates:
\[
(x,y).
\]
In space, a point is described by three coordinates:
\[
(x,y,z).
\]

For example, the point
\[
(2,3)
\]
in the plane has first coordinate \(2\) and second coordinate \(3\). The point
\[
(1,-2,5)
\]
in space has coordinates \(1\), \(-2\), and \(5\).

This is naturally connected with the way we describe the world around us. We live in space, and many situations require us to say where something is. A coordinate system gives a mathematical way of doing this.

However, coordinates are not the point itself. Coordinates depend on a choice of origin, axes, orientation, and units. The same point can have different coordinates in different coordinate systems.

This distinction is important. Geometry concerns objects such as points, lines, planes, and curves. Coordinates are a method of describing these objects using numbers.

\section{Curves and parametric descriptions}

A curve can be described in different ways. One way is to give an equation such as
\[
y = f(x).
\]
For example,
\[
y = x^2
\]
describes a parabola in the plane.

Another way is to use a parameter. A parameter is an additional variable that describes how the coordinates are generated. For example, we can write
\[
x = t, \qquad y = t^2.
\]
As \(t\) changes, the point
\[
(x,y) = (t,t^2)
\]
moves along the parabola.

A parametric curve in the plane can be written as
\[
\mathbf{r}(t) = (x(t),y(t)).
\]
For example,
\[
\mathbf{r}(t) = (\cos t,\sin t)
\]
describes the unit circle, because
\[
\cos^2 t + \sin^2 t = 1.
\]

At this stage, the parameter \(t\) is only a mathematical parameter. In later chapters it may be interpreted in a physical way, but here we only need the mathematical idea: a curve can be described by giving its coordinates as functions of a parameter.

\section{Vectors}

A vector is a mathematical object with magnitude and direction. In coordinates, a vector in the plane may be written as
\[
\mathbf{v} = (v_x,v_y),
\]
and a vector in space as
\[
\mathbf{v} = (v_x,v_y,v_z).
\]

The numbers \(v_x\), \(v_y\), and \(v_z\) are called components. They depend on the coordinate system.

For example,
\[
\mathbf{v} = (3,4)
\]
has magnitude
\[
|\mathbf{v}| = \sqrt{3^2 + 4^2} = 5.
\]

Vectors can be added component by component:
\[
(1,2) + (3,4) = (4,6).
\]
They can be multiplied by a scalar:
\[
2(1,3) = (2,6).
\]

The zero vector is
\[
\mathbf{0} = (0,0)
\]
in the plane, or
\[
\mathbf{0} = (0,0,0)
\]
in space.

The dot product of two vectors is defined by
\[
\mathbf{a}\cdot\mathbf{b}
= a_xb_x + a_yb_y
\]
in two dimensions, and by
\[
\mathbf{a}\cdot\mathbf{b}
= a_xb_x + a_yb_y + a_zb_z
\]
in three dimensions.

The dot product is also related to the angle between vectors:
\[
\mathbf{a}\cdot\mathbf{b}
= |\mathbf{a}|\,|\mathbf{b}|\cos\theta.
\]
This formula shows that the dot product contains information about both lengths and direction.

Vectors will be used later whenever direction matters. Here the important point is only the mathematical structure: components, magnitude, addition, scalar multiplication, and dot product.

\section{Matrices}

A matrix is a rectangular array of numbers. For example,
\[
A =
\begin{pmatrix}
1 & 2 \\
3 & 4
\end{pmatrix}.
\]

Matrices can be added when they have the same size:
\[
\begin{pmatrix}
1 & 2 \\
3 & 4
\end{pmatrix}
+
\begin{pmatrix}
5 & 6 \\
7 & 8
\end{pmatrix}
=
\begin{pmatrix}
6 & 8 \\
10 & 12
\end{pmatrix}.
\]

A matrix can be multiplied by a scalar:
\[
2
\begin{pmatrix}
1 & 2 \\
3 & 4
\end{pmatrix}
=
\begin{pmatrix}
2 & 4 \\
6 & 8
\end{pmatrix}.
\]

A matrix can also multiply a vector. For example,
\[
\begin{pmatrix}
1 & 2 \\
3 & 4
\end{pmatrix}
\begin{pmatrix}
x \\
y
\end{pmatrix}
=
\begin{pmatrix}
x + 2y \\
3x + 4y
\end{pmatrix}.
\]

Matrix multiplication is not the same as ordinary multiplication of numbers. In general,
\[
AB \neq BA.
\]
This means that the order of multiplication matters.

Matrices are useful because they can represent linear transformations. For example, the matrix
\[
R(\theta) =
\begin{pmatrix}
\cos\theta & -\sin\theta \\
\sin\theta & \cos\theta
\end{pmatrix}
\]
represents a rotation in the plane by an angle \(\theta\). If
\[
\mathbf{x} =
\begin{pmatrix}
x \\
y
\end{pmatrix},
\]
then
\[
R(\theta)\mathbf{x}
\]
is the rotated vector.

In these notes, matrices will not be developed in detail. The reader should remember only the basic idea: matrices organize several numbers and can describe transformations acting on vectors or systems of quantities.

\section{Derivatives}

The derivative describes how a function changes locally. If
\[
y = f(x),
\]
then the derivative of \(f\) with respect to \(x\) is written as
\[
f'(x)
\]
or
\[
\frac{df}{dx}.
\]

Intuitively, the derivative tells us how much the value of the function changes when the input changes by a very small amount. It is the local rate of change of the function.

For example, if
\[
f(x) = x^2,
\]
then
\[
f'(x) = 2x.
\]
At \(x = 1\),
\[
f'(1) = 2.
\]
At \(x = 3\),
\[
f'(3) = 6.
\]
This means that the function \(x^2\) changes more rapidly near \(x=3\) than near \(x=1\).

If
\[
f(x) = 3x + 2,
\]
then
\[
f'(x) = 3.
\]
The derivative is constant because the function is linear and has the same slope everywhere.

If
\[
f(x) = C,
\]
where \(C\) is a constant, then
\[
f'(x) = 0.
\]
A constant function does not change.

The derivative also has a geometric interpretation. It is the slope of the tangent line to the graph of the function. If the derivative is positive, the function is increasing. If the derivative is negative, the function is decreasing. If the derivative is zero, the function is locally flat.

Some basic differentiation rules are:
\[
\frac{d}{dx}x^n = nx^{n-1},
\]
\[
\frac{d}{dx}(f(x)+g(x)) = f'(x)+g'(x),
\]
\[
\frac{d}{dx}(cf(x)) = c f'(x),
\]
where \(c\) is a constant.

For example,
\[
\frac{d}{dx}(3x^2 + 5x - 7)
= 6x + 5.
\]

In later chapters, derivatives will receive physical interpretations. In this chapter, the important point is purely mathematical: a derivative measures local change.

\section{Integrals}

An integral describes accumulation. It can be understood as a limiting sum or, geometrically, as the signed area under a graph.

The definite integral of \(f(x)\) from \(a\) to \(b\) is written as
\[
\int_a^b f(x)\,dx.
\]

For example,
\[
\int_0^3 2x\,dx
=
\left[x^2\right]_0^3
=
9 - 0
=
9.
\]

The symbol \(dx\) indicates the variable of integration. In the expression
\[
\int_a^b f(x)\,dx,
\]
we integrate with respect to \(x\).

The indefinite integral is written as
\[
\int f(x)\,dx.
\]
It gives a family of functions whose derivative is \(f(x)\). For example,
\[
\int 2x\,dx = x^2 + C,
\]
because
\[
\frac{d}{dx}(x^2 + C) = 2x.
\]

The constant \(C\) appears because the derivative of a constant is zero. Therefore many functions can have the same derivative.

Some basic integration rules are:
\[
\int x^n\,dx = \frac{x^{n+1}}{n+1} + C
\]
for \(n \neq -1\),
\[
\int (f(x)+g(x))\,dx
=
\int f(x)\,dx + \int g(x)\,dx,
\]
and
\[
\int c f(x)\,dx
=
c\int f(x)\,dx,
\]
where \(c\) is a constant.

For example,
\[
\int (3x^2 + 5x - 7)\,dx
=
x^3 + \frac{5}{2}x^2 - 7x + C.
\]

In later chapters, integrals will receive physical interpretations. Here the important mathematical idea is that integration describes accumulation and is closely connected with differentiation.

\section{Differentiation and integration as inverse operations}

Differentiation and integration are closely related. In a basic sense, they are inverse operations.

If
\[
F'(x) = f(x),
\]
then
\[
\int f(x)\,dx = F(x) + C.
\]

For example,
\[
\frac{d}{dx}x^3 = 3x^2.
\]
Therefore,
\[
\int 3x^2\,dx = x^3 + C.
\]

For definite integrals, the connection is expressed by the fundamental theorem of calculus:
\[
\int_a^b f(x)\,dx = F(b)-F(a),
\]
where
\[
F'(x) = f(x).
\]

For example, if
\[
f(x) = 2x,
\]
then we can choose
\[
F(x) = x^2.
\]
Thus
\[
\int_1^4 2x\,dx
=
F(4)-F(1)
=
4^2 - 1^2
=
15.
\]

This relation between local change and accumulation is one of the central ideas of calculus.

\section{Simple differential equations}

A differential equation is an equation involving an unknown function and one or more of its derivatives. It describes a relation between a function and its change.

For example,
\[
\frac{dy}{dx} = 3
\]
asks for a function whose derivative is \(3\). The general solution is
\[
y(x) = 3x + C.
\]

Another example is
\[
\frac{dy}{dx} = 2x.
\]
A solution is
\[
y(x) = x^2 + C,
\]
because
\[
\frac{d}{dx}(x^2 + C) = 2x.
\]

A slightly different example is
\[
\frac{dy}{dx} = ky,
\]
where \(k\) is a constant. This equation says that the derivative of the function is proportional to the function itself. One family of solutions is
\[
y(x) = Ce^{kx}.
\]
Indeed,
\[
\frac{d}{dx}Ce^{kx} = kCe^{kx} = ky(x).
\]

The constant \(C\) is determined by an additional condition. For example, if
\[
y(0)=2,
\]
then
\[
2 = Ce^{k\cdot 0} = C,
\]
so
\[
y(x)=2e^{kx}.
\]

In these notes, differential equations will appear later as mathematical descriptions of change. Here the important point is only the basic structure: a differential equation contains derivatives and asks for an unknown function.

\section{Approximation and limiting cases}

Mathematics often uses exact expressions, but in applications we frequently use approximations. An approximation replaces a complicated expression with a simpler one under specified conditions.

For example, for small \(x\) measured in radians,
\[
\sin x \approx x.
\]
Also, for small \(x\),
\[
(1+x)^n \approx 1 + nx.
\]

Approximations are useful only when we know when they are valid. The expression
\[
\sin x \approx x
\]
is useful for small \(x\), but not for large \(x\).

A limiting case is a way of checking an expression by asking what happens when a variable becomes very small, very large, or takes a special value.

For example, consider
\[
f(x) = \frac{x}{1+x}.
\]
If \(x\) is very small, then
\[
f(x) \approx x.
\]
If \(x\) is very large, then
\[
f(x) \approx 1.
\]
These limiting behaviours help us understand the structure of the function.

In later chapters, approximations and limiting cases will help us check whether mathematical models make sense.

\section{What the reader should remember}

This chapter is only a mathematical preparation. It does not contain the physical theory itself. Its purpose is to make the later chapters readable.

The reader should remember the following points:
\begin{itemize}
    \item Scientific notation is necessary for reading very large and very small numbers.
    \item A quantity has a numerical value, a unit, and a meaning.
    \item Variables, constants, and parameters play different roles in formulas.
    \item A function describes dependence.
    \item A graph represents a relation visually.
    \item Cartesian coordinates describe points after a coordinate system has been chosen.
    \item Parametric descriptions allow us to represent curves using an additional variable.
    \item Vectors have components, magnitude, and direction.
    \item Matrices organize numbers and can represent transformations.
    \item A derivative describes local change.
    \item An integral describes accumulation.
    \item Differentiation and integration are inverse operations in the sense of calculus.
    \item A differential equation is an equation for an unknown function involving derivatives.
    \item Approximations and limiting cases help us understand expressions and check results.
\end{itemize}

The physical meaning of these tools will be developed later. In particular, kinematics and dynamics will show how this mathematical language becomes a way of describing motion, interaction, and change.
