\section{Review Exercises}

\subsection{Elementary computational exercises}

\begin{enumerate}
\item For
\[
A=\begin{pmatrix}2&-1&4\\0&3&5\end{pmatrix},
\]
state its size and identify $a_{12}$, $a_{23}$, the second row, and the third column.

\item Compute $A+B$, $A-B$, and $3A$ for
\[
A=\begin{pmatrix}1&2\\-1&3\end{pmatrix},
\qquad
B=\begin{pmatrix}4&-2\\5&0\end{pmatrix}.
\]

\item Decide which sums are defined and compute those that are:
\[
\begin{pmatrix}1&2\\3&4\end{pmatrix}+
\begin{pmatrix}0&1\\-1&2\end{pmatrix},
\qquad
\begin{pmatrix}1&2&3\end{pmatrix}+
\begin{pmatrix}1\\2\\3\end{pmatrix}.
\]

\item Compute both $AB$ and $BA$ for
\[
A=\begin{pmatrix}1&2\\0&-1\end{pmatrix},
\qquad
B=\begin{pmatrix}3&0\\4&2\end{pmatrix}.
\]

\item For
\[
A=\begin{pmatrix}1&2&-1\\0&3&4\end{pmatrix},
\qquad
x=\begin{pmatrix}2\\-1\\3\end{pmatrix},
\]
compute $Ax$ and display each row--column product.

\item Compute $A^{\mathsf T}$ and $(A^{\mathsf T})^{\mathsf T}$ for
\[
A=\begin{pmatrix}1&-2&0\\4&3&5\end{pmatrix}.
\]

\item For
\[
D=\begin{pmatrix}2&0&0\\0&-1&0\\0&0&4\end{pmatrix},
\qquad
x=\begin{pmatrix}3\\2\\-1\end{pmatrix},
\]
compute $Dx$ and describe what happens to each coordinate.

\item Starting from
\[
A=\begin{pmatrix}1&2&-1\\2&4&3\\0&1&5\end{pmatrix},
\]
perform $R_2\leftarrow R_2-2R_1$ and then undo the operation.

\item Apply in order to
\[
A=\begin{pmatrix}1&2\\3&7\end{pmatrix}
\]
the operations $R_2\leftarrow R_2-3R_1$ and $R_1\leftarrow R_1-2R_2$. Record every intermediate matrix.

\item Compute $Ax$ for $x=(4,-1)^{\mathsf T}$ when
\[
A=I,
\qquad
A=\begin{pmatrix}0&1\\1&0\end{pmatrix},
\qquad
A=\begin{pmatrix}2&0\\0&3\end{pmatrix}.
\]
Identify the action visible in each result.
\end{enumerate}

\subsection{Development and reasoning exercises}

\begin{enumerate}
\item Let
\[
A=\begin{pmatrix}1&2\\0&1\end{pmatrix},
\quad
B=\begin{pmatrix}2&0\\1&1\end{pmatrix},
\quad
x=\begin{pmatrix}1\\-1\end{pmatrix}.
\]
Compute $Bx$, then $A(Bx)$, then $AB$ and $(AB)x$. Explain how the calculation demonstrates composition and associativity.

\item A linear action sends
\[
e_1\mapsto\begin{pmatrix}2\\1\end{pmatrix},
\qquad
e_2\mapsto\begin{pmatrix}-1\\3\end{pmatrix}.
\]
Construct its matrix and use it to find the image of $(4,-2)^{\mathsf T}$. Explain why the columns contain the images of the basis vectors.

\item The triangle has vertices $(0,0)$, $(1,0)$, and $(0,1)$. Apply
\[
A=\begin{pmatrix}2&1\\0&1\end{pmatrix}
\]
to all three vertices, plot the original and transformed triangles, and describe the visible geometric effect.

\item Compute $AB$ and $BA$ for
\[
A=\begin{pmatrix}1&1\\0&1\end{pmatrix},
\qquad
B=\begin{pmatrix}0&-1\\1&0\end{pmatrix}.
\]
Apply both products to $e_1$ and explain geometrically why changing the order changes the result.

\item Reduce
\[
M=\begin{pmatrix}1&2&1\\2&4&3\\3&6&4\end{pmatrix}
\]
with elementary row operations until the dependence between the rows is visible. State the relation you discover and explain what the arithmetic has exposed.

\item Write the elementary matrix that performs $R_2\leftarrow R_2-3R_1$ on a $2\times2$ matrix. Find its inverse elementary matrix and verify that their product is the identity.

\item For matrices of sizes $2\times3$ and $3\times4$, write a symbolic row--column entry of the product and use it to justify why the result has size $2\times4$.

\item Explain why matrix addition is positionwise while multiplication combines rows and columns. Include a numerical example showing that entrywise multiplication would not represent composition.

\item Explain the three meanings of a row operation in this course: a calculation on entries, a reversible change of representation, and a tool that reveals a relation. Illustrate all three on one matrix.

\item Given a matrix expression, formulate a decision procedure: inspect dimensions, determine which operations are defined, preserve multiplication order, exploit visible structure, and verify the result on a test vector.
\end{enumerate}
