\chapter{Special Relativity: Mathematical Structure and Physical Consequences}

\section{Aim and Scope}

This chapter is a self-contained lecture note on special relativity.  It is
written without graphical TikZ input files.  Places where a spacetime diagram
would be pedagogically useful are marked by explicit placeholders.  The purpose
is to establish a reliable mathematical and conceptual text first; graphical
material can later be added one diagram at a time after separate compilation
checks.

Special relativity is a theory of spacetime structure.  Its most important
message is not only that moving clocks and rods behave in unfamiliar ways.  The
deeper statement is that space and time coordinates are frame-dependent, while a
certain quadratic spacetime interval is invariant.  Once this invariant
structure is accepted, the familiar results -- relativity of simultaneity, time
dilation, length contraction, velocity addition, and relativistic energy and
momentum -- become consequences of one coherent geometry.

Throughout the chapter we use the following notation.  An inertial frame is
denoted by \(S\), another frame moving with constant velocity \(v\) along the
positive \(x\)-axis is denoted by \(S'\), and
\[
\beta=\frac{v}{c},
\qquad
\gamma=\frac{1}{\sqrt{1-\beta^2}}.
\]
Here \(c\) is the speed of light in vacuum.  We use coordinates
\[
(ct,x,y,z),
\]
so that the time coordinate has the same physical dimension as the spatial
coordinates.  Unless explicitly stated otherwise, the metric convention is
\[
\Delta s^2=c^2\Delta t^2-\Delta x^2-\Delta y^2-\Delta z^2.
\]

\begin{remarkbox}
This chapter deliberately contains no active spacetime diagrams.  Figure slots
are written as ordinary boxed text rather than TikZ code.  This keeps the build
stable while the mathematical content is developed.
\end{remarkbox}

\clearpage

\section{Events, Coordinates, and Inertial Frames}

An \emph{event} is something localized in space and time.  Examples include a
flash of light, the emission of a particle, the arrival of a signal, or one tick
of a clock.  In an inertial frame \(S\), an event is assigned four coordinates
\[
( c\Delta t,\Delta x,\Delta y,\Delta z),
\]
but a single event itself is normally written as
\[
(ct,x,y,z).
\]
The same physical event may have different coordinates in a different inertial
frame \(S'\).  The transformation between inertial coordinates is the main
mathematical object of special relativity.

An inertial frame is a frame in which a free particle moves with constant
velocity.  Equivalently, it is a frame without acceleration or rotation.  The
principle of relativity states that no inertial frame is physically preferred.
The laws of physics must have the same form in all inertial frames.

For most derivations it is enough to use the standard configuration:
\begin{itemize}
    \item the axes of \(S\) and \(S'\) are parallel,
    \item \(S'\) moves with velocity \(v\) along the \(x\)-axis of \(S\),
    \item the origins coincide at \(t=t'=0\),
    \item transverse coordinates are unchanged: \(y'=y\), \(z'=z\).
\end{itemize}
Then the nontrivial part of the theory is the relation between \((ct,x)\) and
\((ct',x')\).

\begin{definitionbox}
A \emph{worldline} is the set of events occupied by an object as time evolves.
In a spacetime diagram with vertical coordinate \(ct\) and horizontal coordinate
\(x\), a particle at rest in frame \(S\) has a vertical worldline.  A particle
moving with constant velocity has a straight tilted worldline.
\end{definitionbox}

\begin{center}
\fbox{\parbox{0.84\linewidth}{\centering
Placeholder for a later spacetime diagram: events, worldlines, and inertial axes.}}
\end{center}

\clearpage

\section{Postulates of Special Relativity}

Special relativity is based on two physical postulates.

\begin{enumerate}
    \item \textbf{Principle of relativity.}  The laws of physics have the same
    form in all inertial frames.
    \item \textbf{Invariance of the speed of light.}  Light in vacuum propagates
    with the same speed \(c\) in every inertial frame, independently of the
    motion of the source or the observer.
\end{enumerate}

The second postulate is incompatible with Galilean transformations.  In
Newtonian kinematics one writes
\[
x'=x-vt,
\qquad
 t'=t.
\]
If a light ray satisfies \(x=ct\) in frame \(S\), then the Galilean formulas give
\[
x'=x-vt=(c-v)t,
\qquad
 t'=t,
\]
and therefore
\[
\frac{x'}{t'}=c-v.
\]
The light speed would depend on the observer's motion.  This contradicts the
second postulate.  The transformation law between inertial frames must therefore
be different.

A crucial assumption, usually left implicit, is that transformations between
inertial frames are linear.  Linearity follows from homogeneity of space and
time: the same experiment performed at another place or another time should not
change the transformation rule.  It also ensures that uniform motion is mapped
to uniform motion.

\begin{remarkbox}
The postulates do not say that all observers see the same coordinates.  They say
that the laws of physics have the same form and that the speed \(c\) is
invariant.  Coordinate time and coordinate distance become frame-dependent
quantities.
\end{remarkbox}

\clearpage

\section{The Spacetime Interval}

The central invariant of special relativity is the spacetime interval.  For two
events with coordinate differences
\[
( c\Delta t,\Delta x,\Delta y,\Delta z),
\]
define
\[
\Delta s^2=c^2\Delta t^2-\Delta x^2-\Delta y^2-\Delta z^2.
\]
This quantity is invariant under Lorentz transformations.  In other words, if
the same two events are described in another inertial frame \(S'\), then
\[
\Delta s'^2=\Delta s^2.
\]
The interval classifies separations between events.

\begin{center}
\begin{tabular}{lll}
\toprule
Type & Condition & Physical meaning \\
\midrule
Timelike & \(\Delta s^2>0\) & A massive particle can connect the events. \\
Lightlike & \(\Delta s^2=0\) & A light signal can connect the events. \\
Spacelike & \(\Delta s^2<0\) & No causal signal with speed \(\leq c\) can connect them. \\
\bottomrule
\end{tabular}
\end{center}

For timelike separation, the \emph{proper time} \(\Delta\tau\) is defined by
\[
\Delta s^2=c^2\Delta\tau^2.
\]
Hence
\[
\Delta\tau
=\sqrt{\Delta t^2-\frac{\Delta x^2+\Delta y^2+\Delta z^2}{c^2}}.
\]
This is the time measured by an ideal clock that travels from the first event to
the second event along the corresponding worldline.

For spacelike separation, one may define a proper distance \(\Delta\ell\) by
\[
\Delta\ell^2=-\Delta s^2.
\]
This is the spatial distance between the events in a frame where the two events
are simultaneous.

\begin{center}
\fbox{\parbox{0.84\linewidth}{\centering
Placeholder for a later light-cone diagram: timelike, lightlike, and spacelike regions.}}
\end{center}

\clearpage

\section{Derivation of the Lorentz Transformation}

We now derive the Lorentz transformation in one spatial dimension.  Because the
origin of \(S'\) moves according to \(x=vt\) in frame \(S\), the coordinate
\(x'\) must vanish when \(x=vt\).  Therefore linearity suggests
\[
x'=A(x-vt),
\]
where \(A\) may depend on \(v\), but not on \(x\) or \(t\).

Similarly, by symmetry, the inverse transformation must have the same form with
\(v\) replaced by \(-v\):
\[
x=A(x'+vt').
\]
The transformation must preserve the light rays \(x=ct\) and \(x=-ct\).  The
standard Lorentz transformation satisfying these requirements is
\[
x'=\gamma(x-vt),
\]
\[
t'=\gamma\left(t-\frac{vx}{c^2}\right).
\]
In terms of \(ct\), it becomes
\[
x'=\gamma(x-\beta ct),
\qquad
ct'=\gamma(ct-\beta x).
\]
The inverse transformation is
\[
x=\gamma(x'+vt'),
\]
\[
t=\gamma\left(t'+\frac{vx'}{c^2}\right),
\]
or equivalently
\[
x=\gamma(x'+\beta ct'),
\qquad
ct=\gamma(ct'+\beta x').
\]

Let us verify interval invariance in one spatial dimension.  We compute
\[
(c t')^2-(x')^2
=\gamma^2\left[(ct-\beta x)^2-(x-\beta ct)^2\right].
\]
Expanding the bracket gives
\[
(ct-\beta x)^2-(x-\beta ct)^2
=c^2t^2-2\beta ctx+\beta^2x^2-x^2+2\beta ctx-\beta^2c^2t^2.
\]
The mixed terms cancel, so
\[
(ct-\beta x)^2-(x-\beta ct)^2
=(1-\beta^2)c^2t^2-(1-\beta^2)x^2.
\]
Thus
\[
(c t')^2-(x')^2
=\gamma^2(1-\beta^2)(c^2t^2-x^2).
\]
Since
\[
\gamma^2(1-\beta^2)=1,
\]
we obtain
\[
(c t')^2-(x')^2=c^2t^2-x^2.
\]
This proves the invariance of the interval in the \((ct,x)\)-plane.

\clearpage

\section{Matrix Form and the Metric}

Introduce the two-component spacetime column vector
\[
X=\begin{pmatrix} ct \\ x \end{pmatrix}.
\]
A Lorentz boost along the \(x\)-axis can be written as
\[
X'=\Lambda X,
\]
where
\[
\Lambda=
\begin{pmatrix}
\gamma & -\gamma\beta \\
-\gamma\beta & \gamma
\end{pmatrix}.
\]
The interval in one spatial dimension is
\[
\Delta s^2=X^{T}\eta X,
\qquad
\eta=\begin{pmatrix}1&0\\0&-1\end{pmatrix}.
\]
The invariance condition is
\[
\Lambda^{T}\eta\Lambda=\eta.
\]
This equation is the analogue of the Euclidean rotation condition
\[
R^{T}IR=I,
\]
but with the Minkowski metric \(\eta\) instead of the Euclidean identity matrix.

In four-dimensional notation one writes
\[
x^\mu=(ct,x,y,z),
\]
with Greek indices \(\mu,\nu=0,1,2,3\).  With the chosen sign convention,
\[
\eta_{\mu\nu}=\mathrm{diag}(1,-1,-1,-1).
\]
The interval is
\[
\Delta s^2=\eta_{\mu\nu}\Delta x^\mu\Delta x^\nu.
\]
The Lorentz transformation satisfies
\[
\eta_{\rho\sigma}\Lambda^\rho{}_{\mu}\Lambda^\sigma{}_{\nu}=\eta_{\mu\nu}.
\]
This compact expression is the algebraic form of spacetime geometry.

\clearpage

\section{Rapidity and Hyperbolic Rotations}

The Lorentz transformation resembles a rotation, but it preserves
\((ct)^2-x^2\) instead of \(x^2+y^2\).  This is why hyperbolic functions appear.
Define the rapidity \(\varphi\) by
\[
\tanh\varphi=\beta=\frac{v}{c}.
\]
Then
\[
\cosh\varphi=\gamma,
\qquad
\sinh\varphi=\gamma\beta.
\]
The Lorentz transformation becomes
\[
ct'=ct\cosh\varphi-x\sinh\varphi,
\]
\[
x'=x\cosh\varphi-ct\sinh\varphi.
\]
This is a hyperbolic rotation in the \((ct,x)\)-plane.  It preserves
\[
(ct')^2-(x')^2=(ct)^2-x^2.
\]

The advantage of rapidity is that rapidities add.  If one boost has rapidity
\(\varphi_1\), and a second boost in the same direction has rapidity
\(\varphi_2\), then the combined boost has rapidity
\[
\varphi=\varphi_1+\varphi_2.
\]
Velocity itself does not add linearly; rapidity does.

\begin{center}
\fbox{\parbox{0.84\linewidth}{\centering
Placeholder for a later diagram: Lorentz axes as hyperbolic rotations toward the light cone.}}
\end{center}

\clearpage

\section{Relativity of Simultaneity}

The time transformation is
\[
t'=\gamma\left(t-\frac{vx}{c^2}\right).
\]
For two events, subtract the equations:
\[
\Delta t'=\gamma\left(\Delta t-\frac{v\Delta x}{c^2}\right).
\]
Suppose that the two events are simultaneous in frame \(S\).  Then
\(\Delta t=0\), and therefore
\[
\Delta t'=-\gamma\frac{v\Delta x}{c^2}.
\]
If \(v\neq0\) and \(\Delta x\neq0\), then \(\Delta t'\neq0\).  Thus two
spatially separated events that are simultaneous in one inertial frame are not
generally simultaneous in another inertial frame.

This is not an optical delay effect.  It remains after one corrects for signal
travel time.  It is a statement about the coordinate assignment used by inertial
frames.

\begin{examplebox}
Consider two lightning strikes hitting the front and rear of a moving train.  A
platform observer may judge the strikes simultaneous.  The train observer uses a
different simultaneity slice and assigns different times to the same two events.
Both descriptions are consistent because simultaneity at a distance is not
absolute.
\end{examplebox}

\begin{center}
\fbox{\parbox{0.84\linewidth}{\centering
Placeholder for a later diagram: two different simultaneity slices through the same spacetime.}}
\end{center}

\clearpage

\section{Time Dilation}

Time dilation concerns two events occurring at the same place in the rest frame
of a clock.  Suppose the clock is at rest in \(S'\).  Then two ticks of the
clock satisfy
\[
\Delta x'=0.
\]
Use the inverse Lorentz transformation for time:
\[
\Delta t=\gamma\left(\Delta t'+\frac{v\Delta x'}{c^2}\right).
\]
Since \(\Delta x'=0\), this reduces to
\[
\Delta t=\gamma\Delta t'.
\]
The time measured by the clock in its own rest frame is the proper time
\(\Delta\tau\), so
\[
\Delta t'=\Delta\tau.
\]
Therefore
\[
\boxed{\Delta t=\gamma\Delta\tau.}
\]
Because \(\gamma\geq1\), the coordinate time measured in the frame where the
clock is moving is larger than the proper time measured by the clock itself.
This is the statement that a moving clock is measured to run slow.

The same result follows from the interval.  For a clock moving with speed \(u\)
in frame \(S\), we have
\[
\Delta x=u\Delta t.
\]
Then
\[
c^2\Delta\tau^2=c^2\Delta t^2-\Delta x^2
=c^2\Delta t^2-u^2\Delta t^2
=c^2\Delta t^2\left(1-\frac{u^2}{c^2}\right).
\]
Thus
\[
\Delta\tau=\Delta t\sqrt{1-\frac{u^2}{c^2}}
=\frac{\Delta t}{\gamma_u}.
\]
Equivalently,
\[
\Delta t=\gamma_u\Delta\tau.
\]

\begin{center}
\fbox{\parbox{0.84\linewidth}{\centering
Placeholder for a later diagram: two ticks on one moving clock worldline.}}
\end{center}

\clearpage

\section{Length Contraction}

Length contraction concerns a rod at rest in frame \(S'\).  Let its endpoints
have fixed coordinates \(x'_1\) and \(x'_2\), so its proper length is
\[
L_0=x'_2-x'_1.
\]
To measure the rod's length in frame \(S\), one must record the positions of the
two endpoints at the same time in \(S\).  Hence the two measurement events
satisfy
\[
\Delta t=0.
\]
The spatial Lorentz transformation gives
\[
\Delta x'=\gamma(\Delta x-v\Delta t).
\]
With \(\Delta t=0\), this becomes
\[
\Delta x'=\gamma\Delta x.
\]
Since \(\Delta x'=L_0\) and \(\Delta x=L\), we get
\[
L_0=\gamma L.
\]
Therefore
\[
\boxed{L=\frac{L_0}{\gamma}.}
\]
A moving rod is shorter along the direction of motion than its proper length.
Lengths perpendicular to the direction of motion are unchanged by a boost in the
\(x\)-direction.

The simultaneity condition is essential.  If the two endpoints are not measured
at the same time in the measuring frame, the result is not a length measurement
in that frame.

\begin{remarkbox}
Length contraction is not mechanical compression.  The rod is not squeezed in
its own rest frame.  It has length \(L_0\) there.  A different frame measures a
shorter length because it uses a different simultaneity condition for comparing
endpoint events.
\end{remarkbox}

\begin{center}
\fbox{\parbox{0.84\linewidth}{\centering
Placeholder for a later diagram: endpoint worldlines and simultaneous measurement slices.}}
\end{center}

\clearpage

\section{Velocity Addition}

Let an object move with velocity \(u'=dx'/dt'\) in frame \(S'\).  Frame \(S'\)
moves with velocity \(v\) relative to frame \(S\).  From the inverse Lorentz
transformation,
\[
dx=\gamma(dx'+vdt'),
\]
\[
dt=\gamma\left(dt'+\frac{v}{c^2}dx'\right).
\]
Therefore
\[
u=\frac{dx}{dt}
=\frac{dx'+vdt'}{dt'+(v/c^2)dx'}.
\]
Dividing numerator and denominator by \(dt'\) gives
\[
\boxed{u=\frac{u'+v}{1+u'v/c^2}.}
\]
This formula replaces the Galilean rule \(u=u'+v\).

If \(u'=c\), then
\[
u=\frac{c+v}{1+v/c}=c.
\]
Thus light has speed \(c\) in both frames.  If \(|u'|<c\) and \(|v|<c\), then
\(|u|<c\).  Relativistic velocity addition never pushes a subluminal speed past
\(c\).

For three-dimensional motion with a boost along \(x\), the components transform
as
\[
u_x=\frac{u'_x+v}{1+u'_xv/c^2},
\]
\[
u_y=\frac{u'_y}{\gamma(1+u'_xv/c^2)},
\qquad
u_z=\frac{u'_z}{\gamma(1+u'_xv/c^2)}.
\]
The transverse components are affected because time transforms.

\clearpage

\section{Proper Time and Worldline Length}

For a general worldline \(\mathbf{x}(t)\), define the ordinary speed
\[
u(t)=\left|\frac{d\mathbf{x}}{dt}\right|.
\]
The proper time increment is obtained from the interval:
\[
c^2d\tau^2=c^2dt^2-d\mathbf{x}^2.
\]
Since
\[
d\mathbf{x}^2=u(t)^2dt^2,
\]
we find
\[
d\tau=dt\sqrt{1-\frac{u(t)^2}{c^2}}.
\]
The total proper time accumulated along the worldline is therefore
\[
\tau=\int d\tau
=\int_{t_1}^{t_2}\sqrt{1-\frac{u(t)^2}{c^2}}\,dt.
\]
This formula gives the elapsed time on an ideal clock carried along the
worldline.

For constant speed \(u\), it reduces to
\[
\Delta\tau=\Delta t\sqrt{1-\frac{u^2}{c^2}}
=\frac{\Delta t}{\gamma_u}.
\]
The twin paradox is resolved by this formula: different worldlines connecting
the same two events can have different accumulated proper times.  The traveling
twin follows a non-inertial path if departure and reunion occur at the same
place, and the proper time integral differs from that of the stay-at-home twin.

\clearpage

\section{Four-Velocity and Four-Momentum}

The position four-vector is
\[
x^\mu=(ct,x,y,z).
\]
The proper time allows us to define the four-velocity
\[
U^\mu=\frac{dx^\mu}{d\tau}.
\]
Because
\[
\frac{dt}{d\tau}=\gamma_u,
\]
we obtain
\[
U^\mu=\gamma_u(c,u_x,u_y,u_z).
\]
Its invariant norm is
\[
U^\mu U_\mu=c^2.
\]
This is the relativistic replacement for the Newtonian velocity vector as the
geometrically natural velocity object.

The four-momentum of a particle with rest mass \(m\) is
\[
p^\mu=mU^\mu.
\]
Therefore
\[
p^\mu=(\gamma mc,\gamma m\mathbf{u}).
\]
Writing
\[
p^\mu=\left(\frac{E}{c},\mathbf{p}\right),
\]
we identify
\[
E=\gamma mc^2,
\qquad
\mathbf{p}=\gamma m\mathbf{u}.
\]
The invariant norm of four-momentum is
\[
p^\mu p_\mu=m^2c^2.
\]
In terms of energy and momentum,
\[
\frac{E^2}{c^2}-p^2=m^2c^2.
\]
Multiplying by \(c^2\) gives the energy-momentum relation
\[
\boxed{E^2=p^2c^2+m^2c^4.}
\]
For a particle at rest, \(p=0\), hence
\[
\boxed{E_0=mc^2.}
\]
For a massless particle, \(m=0\), hence
\[
E=pc.
\]

\clearpage

\section{Relativistic Energy at Low Velocity}

The relativistic energy is
\[
E=\gamma mc^2.
\]
For \(u\ll c\), write
\[
\gamma=\frac{1}{\sqrt{1-u^2/c^2}}.
\]
Using the binomial approximation
\[
(1-\epsilon)^{-1/2}\approx 1+\frac{1}{2}\epsilon
\qquad (\epsilon\ll1),
\]
we obtain
\[
\gamma\approx 1+\frac{1}{2}\frac{u^2}{c^2}.
\]
Therefore
\[
E=\gamma mc^2
\approx mc^2+\frac{1}{2}mu^2.
\]
The first term is the rest energy, and the second term is the Newtonian kinetic
energy.  Thus special relativity contains Newtonian mechanics as a low-speed
approximation.

The relativistic kinetic energy is defined as
\[
K=E-mc^2=(\gamma-1)mc^2.
\]
At low speeds,
\[
K\approx\frac{1}{2}mu^2.
\]
At speeds close to \(c\), \(\gamma\) grows without bound, so the kinetic energy
needed to accelerate a massive particle toward \(c\) grows without bound.
This is why massive particles cannot be accelerated to the speed of light.

\clearpage

\section{Muon Lifetime Example}

Cosmic rays striking the upper atmosphere produce muons.  A muon at rest has a
mean lifetime approximately
\[
\tau_0\approx 2.2\,\mu\mathrm{s}.
\]
Classically, even if a muon moved almost at the speed of light, it would travel
roughly
\[
c\tau_0\approx (3.0\times10^8\,\mathrm{m/s})(2.2\times10^{-6}\,\mathrm{s})
\approx 660\,\mathrm{m}.
\]
Many muons are produced several kilometres above Earth's surface, yet many are
detected near the ground.  Special relativity explains this.

In the Earth frame, the muon moves rapidly, so its lifetime is time-dilated:
\[
\Delta t=\gamma\tau_0.
\]
If \(\gamma=10\), then
\[
\Delta t\approx 22\,\mu\mathrm{s}.
\]
The distance traveled can then be of order
\[
L\approx c\Delta t\approx 6.6\,\mathrm{km}.
\]
This is compatible with atmospheric distances.

In the muon rest frame, the muon lifetime is simply \(\tau_0\).  Instead, the
atmosphere moves toward the muon and is length-contracted:
\[
L=\frac{L_0}{\gamma}.
\]
The two descriptions use different coordinates, but they agree on the physical
fact: the muon can reach the detector.

\begin{center}
\fbox{\parbox{0.84\linewidth}{\centering
Placeholder for a later diagram: muon worldline and atmospheric distance in two frames.}}
\end{center}

\clearpage

\section{Causal Structure}

The sign of the interval determines causal relationships.  If two events are
timelike separated, then there exists an inertial frame in which they occur at
the same spatial point.  Their time order is the same in every inertial frame.
A massive particle can travel from one event to the other if the second lies in
the future light cone of the first.

If two events are lightlike separated, only a signal moving at speed \(c\) can
connect them.  Their interval is zero:
\[
\Delta s^2=0.
\]
If two events are spacelike separated, no signal moving at or below \(c\) can
connect them.  Different inertial frames may disagree about their temporal
order, but this does not violate causality because neither event can influence
the other.

This is the relativistic replacement for Newtonian absolute time.  Absolute
simultaneity disappears, but causal order for physically connectable events is
preserved.

\clearpage

\section{Common Misinterpretations}

\subsection{Moving clocks do not malfunction}

Time dilation is not a statement about defective clocks.  It applies to all
ideal clocks: atomic clocks, particle lifetimes, mechanical clocks, and
biological aging.  The reason is geometric: clocks measure proper time along
their worldlines.

\subsection{Length contraction is not compression}

A moving rod is not physically squeezed in its own rest frame.  It has its
proper length \(L_0\) there.  A different frame measures a shorter length because
it uses a different simultaneity condition for comparing endpoint events.

\subsection{Relativity of simultaneity is not an optical illusion}

Observers may visually see events after light travel delays.  Relativity of
simultaneity is a deeper statement: after correcting for signal travel time,
different inertial frames still assign different time coordinates to separated
events.

\subsection{Nothing massive reaches the speed of light}

The speed \(c\) is not merely the speed of light.  It is the invariant speed of
spacetime.  Massive particles have timelike worldlines and speeds below \(c\).
Massless particles in vacuum move at \(c\).

\clearpage

\section{Formula Summary}

The Lorentz factor is
\[
\gamma=\frac{1}{\sqrt{1-v^2/c^2}}.
\]
The Lorentz transformation is
\[
x'=\gamma(x-vt),
\qquad
 t'=\gamma\left(t-\frac{vx}{c^2}\right).
\]
The inverse transformation is
\[
x=\gamma(x'+vt'),
\qquad
 t=\gamma\left(t'+\frac{vx'}{c^2}\right).
\]
The invariant interval is
\[
\Delta s^2=c^2\Delta t^2-\Delta x^2-\Delta y^2-\Delta z^2.
\]
The proper time for timelike separation is
\[
\Delta\tau=\sqrt{\Delta t^2-\frac{\Delta x^2+\Delta y^2+\Delta z^2}{c^2}}.
\]
Time dilation is
\[
\Delta t=\gamma\Delta\tau.
\]
Length contraction is
\[
L=\frac{L_0}{\gamma}.
\]
Relativity of simultaneity is expressed by
\[
\Delta t'=\gamma\left(\Delta t-\frac{v\Delta x}{c^2}\right).
\]
Velocity addition in one dimension is
\[
u=\frac{u'+v}{1+u'v/c^2}.
\]
Relativistic momentum and energy are
\[
\mathbf{p}=\gamma m\mathbf{u},
\qquad
E=\gamma mc^2.
\]
The invariant energy-momentum relation is
\[
E^2=p^2c^2+m^2c^4.
\]

\begin{summarybox}
Special relativity is the geometry of Minkowski spacetime.  Lorentz
transformations preserve the spacetime interval.  Time dilation, length
contraction, relativity of simultaneity, velocity addition, and relativistic
energy-momentum are not separate tricks; they are consequences of the same
invariant spacetime structure.
\end{summarybox}
