\section{Confidence intervals from sampling behaviour}

Suppose a statistic \(T\) estimates an unknown parameter \(\theta\). If the
sampling distribution of \(T\) is approximately normal, then
\[
\frac{T-\theta}{\operatorname{SE}(T)}
\approx N(0,1).
\]
For a standard normal random variable, approximately \(95\%\) of the probability
lies between \(-1.96\) and \(1.96\). Therefore
\[
P\left(
-1.96
\leq
\frac{T-\theta}{\operatorname{SE}(T)}
\leq
1.96
\right)
\approx 0.95.
\]

This is a probability statement about the random statistic \(T\), made before the
sample is observed. It says that repeated samples usually produce statistics
within about two standard errors of the parameter.

The same inequalities can be rearranged:
\[
P\left(
T-1.96\operatorname{SE}(T)
\leq
\theta
\leq
T+1.96\operatorname{SE}(T)
\right)
\approx 0.95.
\]
This rearrangement is the source of a confidence interval.

\subsection*{From a random interval to an observed interval}

Before the sample is collected, both endpoints
\[
T-1.96\operatorname{SE}(T)
\qquad\text{and}\qquad
T+1.96\operatorname{SE}(T)
\]
are random. Different samples produce different intervals.

After the data are observed, the statistic becomes \(t_{\mathrm{obs}}\), and the
random interval becomes one numerical interval:
\[
\left[
 t_{\mathrm{obs}}-1.96\widehat{\operatorname{SE}}(T),
 t_{\mathrm{obs}}+1.96\widehat{\operatorname{SE}}(T)
\right].
\]

\begin{definitionbox}
A \textbf{confidence interval} is an interval computed by a procedure designed to
capture the unknown parameter with a stated long-run frequency under repeated
sampling.
\end{definitionbox}

A \(95\%\) confidence procedure is intended to produce intervals containing the
true parameter in approximately \(95\%\) of repeated samples, provided the model
and approximation are appropriate.

\subsection*{What the confidence level means}

The confidence level describes the procedure, not a probability distribution for
the fixed parameter.

Imagine repeatedly performing the following steps:

\begin{enumerate}
    \item draw a new sample of size \(n\);
    \item compute the statistic and its estimated standard error;
    \item construct the confidence interval;
    \item record whether the interval contains the true parameter.
\end{enumerate}

A well-calibrated \(95\%\) procedure should succeed in approximately \(95\%\) of
these repetitions.

Once one particular sample has been observed, the resulting interval either
contains the parameter or it does not. In the classical interpretation, the
parameter is fixed and the interval-producing procedure is random.

\begin{remarkbox}
It is common to say informally that we are ``\(95\%\) confident'' that the
parameter lies in the reported interval. The precise repeated-sampling meaning is
that the procedure used to construct the interval has approximately \(95\%\)
coverage under the assumed model.
\end{remarkbox}

\subsection*{A confidence interval for a population mean}

Suppose
\[
X_1,\ldots,X_n
\]
are independent observations with population mean \(\mu\), and suppose the sample
is large enough for a normal approximation to the sample mean.

Since
\[
\overline X
\approx
N\left(\mu,\frac{\sigma^2}{n}\right),
\]
a \(95\%\) confidence interval has the general form
\[
\overline X
\pm
1.96\frac{\sigma}{\sqrt n}.
\]

Usually \(\sigma\) is unknown, so it is replaced by the sample standard deviation
\(S\):
\[
\overline X
\pm
1.96\frac{S}{\sqrt n}.
\]
For large samples this gives a useful approximation. For small samples from a
normal population, a more accurate procedure uses the Student \(t\) distribution,
but that refinement is not developed here.

\begin{examplebox}
Suppose a sample of size \(100\) gives
\[
\bar{x}=52.4
\qquad\text{and}\qquad
s=10.
\]
The estimated standard error is
\[
\frac{s}{\sqrt n}
=
\frac{10}{10}
=1.
\]
An approximate \(95\%\) confidence interval for \(\mu\) is
\[
52.4\pm1.96,
\]
so
\[
[50.44,54.36].
\]
The interval reports both an estimate and the sampling uncertainty associated
with that estimate.
\end{examplebox}

\subsection*{A confidence interval for a proportion}

For independent Bernoulli observations, the sample proportion satisfies
\[
\widehat p
\approx
N\left(p,\frac{p(1-p)}{n}\right)
\]
when the sample is sufficiently large and the expected numbers of successes and
failures are not too small.

Because \(p\) is unknown, a simple approximate standard error replaces it by
\(\widehat p\):
\[
\widehat{\operatorname{SE}}(\widehat p)
=
\sqrt{\frac{\widehat p(1-\widehat p)}{n}}.
\]
This gives the familiar approximate interval
\[
\widehat p
\pm
1.96
\sqrt{\frac{\widehat p(1-\widehat p)}{n}}.
\]

\begin{examplebox}
Suppose \(240\) successes are observed in \(400\) independent trials. Then
\[
\widehat p=\frac{240}{400}=0.6.
\]
The estimated standard error is
\[
\sqrt{\frac{0.6(0.4)}{400}}
\approx0.0245.
\]
An approximate \(95\%\) confidence interval is
\[
0.6\pm1.96(0.0245),
\]
which is approximately
\[
[0.552,0.648].
\]
The observed proportion is \(0.6\), but the interval expresses that other nearby
values of the underlying probability are also compatible with the sampling
variation.
\end{examplebox}

The simple interval above is useful for explaining the logic, although more
accurate intervals may be preferred when the sample is small or the observed
proportion is close to \(0\) or \(1\).

\subsection*{Precision and sample size}

A confidence interval becomes narrower when the standard error becomes smaller.
For means and proportions, the standard error decreases at the rate
\[
\frac1{\sqrt n}.
\]
Thus larger samples usually give more precise intervals.

But precision should not be confused with correctness. A very narrow confidence
interval may still be misleading if:

\begin{itemize}
    \item the observations are not representative of the intended population;
    \item the measurements are systematically biased;
    \item the independence assumption is inappropriate;
    \item the probability model does not describe the data-generating process;
    \item the normal approximation is poor.
\end{itemize}

Increasing \(n\) reduces random sampling error. It does not automatically remove
systematic error or repair a defective model.

\subsection*{Confidence intervals as model compatibility}

A confidence interval can be read as a set of parameter values that are reasonably
compatible with the observed statistic under the assumed sampling model.

Values near the centre of the interval make the observed statistic unsurprising.
Values far outside the interval would place the observed statistic several
standard errors away from what those values predict.

This interpretation connects confidence intervals with hypothesis tests. A
hypothesis test begins with one proposed parameter value and asks whether the
observed statistic would be unusually far from the value predicted by that
hypothesis.

\begin{summarybox}
A confidence interval is obtained by turning a probability statement about a
random statistic into an interval-valued procedure for an unknown parameter. Its
validity comes from the sampling distribution and its long-run coverage, not from
the interval formula alone.
\end{summarybox}
