\section{Standard errors and normal approximations}

A statistic varies from sample to sample. The standard deviation of its sampling
distribution measures the typical size of this variation.

\begin{definitionbox}
The \textbf{standard error} of a statistic \(T\) is the standard deviation of its
sampling distribution:
\[
\operatorname{SE}(T)=\sqrt{\Var(T)}.
\]
\end{definitionbox}

The term does not describe an error caused by incorrect computation. It describes
the random variation produced by the sampling process.

\subsection*{Standard deviation and standard error}

The standard deviation of the population describes variation among individual
observations. The standard error describes variation among values of a statistic
computed from repeated samples.

For example, if individual observations have standard deviation \(\sigma\), then
\[
\operatorname{SD}(X_i)=\sigma.
\]
For the sample mean,
\[
\operatorname{SE}(\overline X_n)
=
\operatorname{SD}(\overline X_n)
=
\frac{\sigma}{\sqrt n}.
\]

The two quantities answer different questions:

\begin{center}
\begin{tabular}{lll}
\toprule
Quantity & What varies? & What it measures \\
\midrule
\(\sigma\) & individual observations & spread of the population \\
\(\sigma/\sqrt n\) & sample means & sampling variability of the mean \\
\bottomrule
\end{tabular}
\end{center}

A population may remain highly variable while the sample mean is estimated with
considerably greater stability.

\subsection*{Standard errors of the main statistics}

For independent observations with common variance \(\sigma^2\),
\[
\operatorname{SE}(\overline X_n)=\frac{\sigma}{\sqrt n}.
\]

For a sample proportion based on independent Bernoulli observations,
\[
\operatorname{SE}(\widehat p_n)
=
\sqrt{\frac{p(1-p)}{n}}.
\]

Both formulas contain \(1/\sqrt n\). This is the characteristic scale of
large-sample fluctuations for averages.

If the sample size is multiplied by four, the standard error is divided by two.
If the sample size is multiplied by one hundred, the standard error is divided by
ten.

\begin{remarkbox}
Larger samples reduce random sampling variability, but with diminishing returns.
Substantially greater precision may require a much larger sample.
\end{remarkbox}

\subsection*{Normal approximation in standard-error units}

The central limit theorem suggests that, for large samples,
\[
\frac{\overline X_n-\mu}
{\operatorname{SE}(\overline X_n)}
\approx N(0,1).
\]
This can be written informally as
\[
\overline X_n
\approx
\mu
+
\operatorname{SE}(\overline X_n)Z,
\qquad Z\sim N(0,1).
\]

The formula separates three components:

\begin{itemize}
    \item \(\mu\), the centre of the sampling distribution;
    \item \(\operatorname{SE}(\overline X_n)\), the scale of the random
    fluctuation;
    \item \(Z\), the approximately standard normal shape of that fluctuation.
\end{itemize}

The same idea applies to a sample proportion:
\[
\frac{\widehat p_n-p}
{\operatorname{SE}(\widehat p_n)}
\approx N(0,1).
\]

A difference should therefore be interpreted relative to its standard error. A
numerical difference of \(0.02\) may be negligible in one problem and highly
unusual in another, depending on the sampling variability.

\subsection*{Typical normal ranges}

For a standard normal random variable \(Z\), approximately
\[
P(-1.96\leq Z\leq 1.96)=0.95.
\]
Consequently, when a normal approximation is appropriate,
\[
P\left(
-1.96
\leq
\frac{T-\theta}{\operatorname{SE}(T)}
\leq
1.96
\right)
\approx 0.95.
\]

This is equivalent to
\[
P\left(
\theta-1.96\operatorname{SE}(T)
\leq T\leq
\theta+1.96\operatorname{SE}(T)
\right)
\approx 0.95.
\]

Thus the statistic usually lies within about two standard errors of its centre.
This is a property of the sampling procedure before the data are observed.

\begin{examplebox}
Suppose
\[
\mu=100,
\qquad
\sigma=15,
\qquad
n=100.
\]
Then
\[
\operatorname{SE}(\overline X)=\frac{15}{10}=1.5.
\]
The central limit theorem suggests that approximately \(95\%\) of sample means
lie in
\[
100\pm 1.96(1.5),
\]
which is approximately
\[
[97.06,102.94].
\]
This interval describes the long-run behaviour of the random sample mean when
the population mean is \(100\).
\end{examplebox}

\subsection*{Known and estimated standard errors}

The formulas above may contain unknown parameters. For the mean,
\[
\operatorname{SE}(\overline X)=\frac{\sigma}{\sqrt n}
\]
depends on the population standard deviation \(\sigma\). For a proportion,
\[
\operatorname{SE}(\widehat p)=
\sqrt{\frac{p(1-p)}{n}}
\]
depends on the unknown probability \(p\).

In practice, unknown quantities are often replaced by estimates computed from
the sample. For example,
\[
\widehat{\operatorname{SE}}(\overline X)
=
\frac{S}{\sqrt n},
\]
where \(S\) is the sample standard deviation, and
\[
\widehat{\operatorname{SE}}(\widehat p)
=
\sqrt{\frac{\widehat p(1-\widehat p)}{n}}.
\]

The estimated standard error is itself computed from random data, so it also
varies from sample to sample. For sufficiently large samples, it often provides a
good approximation to the true standard error.

This replacement should not be hidden. The theoretical standard error belongs to
the model, while the estimated standard error is obtained from the observed
sample.

\subsection*{What a normal approximation permits}

Once a statistic is approximately normal after standardization, probabilities
can be interpreted through distances measured in standard errors.

For a proposed parameter value \(\theta_0\), the standardized difference
\[
Z_{\mathrm{obs}}
=
\frac{T_{\mathrm{obs}}-\theta_0}
{\operatorname{SE}_{\theta_0}(T)}
\]
measures how far the observed statistic lies from the value predicted by the
model.

A value near zero is typical. A value far into a tail of the standard normal
distribution is unusual.

The same approximation can also be rearranged to produce a range of parameter
values compatible with the observed statistic. This leads to confidence
intervals.

\begin{summarybox}
A standard error is the standard deviation of a sampling distribution. It gives
the natural unit for measuring the difference between an observed statistic and
a parameter value. Together with the central limit theorem, it turns random
sampling variation into a scale that can be used for confidence intervals and
hypothesis tests.
\end{summarybox}
