\section{Statistical hypotheses and model compatibility}

A confidence interval produces a range of parameter values compatible with an
observed statistic. A hypothesis test begins with one proposed parameter value
and asks whether the observed statistic is compatible with it.

Suppose a model contains a parameter \(\theta\). A null hypothesis has the form
\[
H_0:\theta=\theta_0.
\]
The hypothesis specifies a model under which the sampling distribution of a
statistic can be studied.

The central question is:

\begin{center}
If \(H_0\) were true, would the observed statistic be ordinary or unusually
extreme?
\end{center}

\subsection*{The null distribution}

Let \(T\) be a statistic whose distribution depends on \(\theta\). Under
\[
H_0:\theta=\theta_0,
\]
the model determines a particular sampling distribution for \(T\). After
observing the sample, we compute
\[
t_{\mathrm{obs}}.
\]
We then locate this value within the distribution predicted by \(H_0\).

The reasoning is
\[
H_0
\longrightarrow
\text{sampling distribution under }H_0
\longrightarrow
\text{observed statistic}
\longrightarrow
\text{assessment of compatibility}.
\]

A value near the centre of the null distribution is not surprising. A value far
into a tail is difficult to explain as ordinary sampling variation under the
proposed model.

\begin{remarkbox}
A probability model may produce rare outcomes. An unusual observation is evidence
against a model, not a logical contradiction.
\end{remarkbox}

\subsection*{A standardized test statistic}

When a normal approximation is appropriate, the distance between the observed
statistic and the null value is measured in standard-error units:
\[
Z
=
\frac{T-\theta_0}{\operatorname{SE}_{H_0}(T)}.
\]
After observation,
\[
z_{\mathrm{obs}}
=
\frac{t_{\mathrm{obs}}-\theta_0}
{\operatorname{SE}_{H_0}(T)}.
\]

Under the null hypothesis, this standardized statistic is approximately
\[
N(0,1).
\]
Values near zero are typical. Large positive or negative values are unusual.
The sign gives the direction of the difference, while the magnitude gives the
distance from the null model in standard-error units.

The raw difference
\[
t_{\mathrm{obs}}-\theta_0
\]
is not sufficient by itself. The same numerical difference may be ordinary when
the standard error is large and highly unusual when it is small.

\subsection*{The p-value}

\begin{definitionbox}
The \textbf{p-value} is the probability, calculated under the null hypothesis, of
obtaining a test statistic at least as extreme as the observed one.
\end{definitionbox}

For a two-sided normal test,
\[
p\text{-value}
=
P\left(|Z|\geq |z_{\mathrm{obs}}|\right),
\qquad Z\sim N(0,1).
\]

A small \(p\)-value means that the observed result, or something still more
extreme, would be rare under the null model. It therefore provides evidence
against \(H_0\).

A large \(p\)-value means that the observed result is not unusual under \(H_0\).
It does not prove that the null hypothesis is true.

In particular,
\[
p\text{-value}
\neq
P(H_0\text{ is true}\mid\text{data}).
\]
The p-value is a probability about hypothetical data under the null model, not a
probability assigned to the hypothesis itself.

\subsection*{A test for a population proportion}

Suppose independent Bernoulli observations have unknown success probability
\(p\). Consider
\[
H_0:p=p_0.
\]
The sample proportion is
\[
\widehat p=\frac1n\sum_{i=1}^nX_i.
\]
Under the null hypothesis,
\[
\E_{H_0}[\widehat p]=p_0
\]
and
\[
\operatorname{SE}_{H_0}(\widehat p)
=
\sqrt{\frac{p_0(1-p_0)}{n}}.
\]
Therefore the standardized test statistic is
\[
Z
=
\frac{\widehat p-p_0}
{\sqrt{p_0(1-p_0)/n}}.
\]
For a sufficiently large sample, its distribution under \(H_0\) is approximately
standard normal.

\begin{examplebox}
Suppose \(n=400\) independent trials produce \(240\) successes. Then
\[
\widehat p=0.6.
\]
We examine
\[
H_0:p=0.5.
\]
Under this hypothesis,
\[
\operatorname{SE}_{H_0}(\widehat p)
=
\sqrt{\frac{0.5(0.5)}{400}}
=0.025.
\]
Thus
\[
z_{\mathrm{obs}}
=
\frac{0.6-0.5}{0.025}
=4.
\]
A value four standard errors from the null centre is very unusual under a standard
normal approximation. The two-sided p-value is very small, so the data provide
strong evidence against \(H_0:p=0.5\).
\end{examplebox}

The conclusion is not a proof with absolute certainty. It says that the observed
proportion is difficult to explain as ordinary sampling variation under the
proposed value \(p=0.5\).

\subsection*{Significance levels and decisions}

A testing procedure may specify a significance level \(\alpha\), often
\[
\alpha=0.05.
\]
A conventional decision rule is
\[
p\text{-value}\leq\alpha
\quad\Longrightarrow\quad
\text{reject }H_0.
\]
If the p-value is larger than \(\alpha\), we say that the data do not provide
sufficient evidence to reject \(H_0\).

This is not the same as accepting or proving the null hypothesis. A test may fail
to detect a real difference when the sample is too small or the statistic is too
variable.

Because the decision is based on a random sample, errors are possible. A type I
error occurs when a true null hypothesis is rejected. A type II error occurs when
a false null hypothesis is not rejected.

The significance level controls the probability of a type I error under the null
model. Larger samples generally make it easier to detect systematic departures
from the null value.

\subsection*{Statistical and practical importance}

A very large sample may make a small difference statistically detectable. This
does not imply that the difference is important in practice.

Statistical significance asks whether the observed difference is large relative
to sampling variation. Practical importance asks whether the size and
consequences of the difference matter in the context of the problem.

\begin{remarkbox}
A complete analysis should report the estimated effect and its uncertainty, not
only whether a p-value crossed a conventional threshold.
\end{remarkbox}

\subsection*{Confidence intervals and tests}

For a two-sided large-sample procedure, a parameter value \(\theta_0\) is usually
rejected at significance level \(0.05\) precisely when it lies outside the
corresponding approximate \(95\%\) confidence interval.

The confidence interval asks
\[
\text{Which parameter values are compatible with the observed statistic?}
\]
The hypothesis test asks
\[
\text{Is this particular parameter value compatible with the observed statistic?}
\]
Both use the same sampling distribution and the same standardized distance.

\subsection*{Every conclusion depends on the model}

A small p-value may indicate that the parameter value in \(H_0\) is wrong. It may
also reveal a failure of another assumption used to construct the null
distribution. Dependence, biased sampling, measurement error, or an inappropriate
probability law can also make a statistic look unusual.

A statistical conclusion should therefore be read conditionally:

\begin{center}
Under the stated sampling model and assumptions, the observed data are more or
less compatible with the proposed hypothesis.
\end{center}

\begin{summarybox}
A hypothesis test compares an observed statistic with the sampling distribution
predicted by a proposed model. A small p-value means that the observed result
would be unusual under that model. Testing does not remove uncertainty; it uses
probability to measure how strongly the data challenge a specified hypothesis.
\end{summarybox}
