\section{Cumulative distribution functions}

A probability mass function records probability at individual support values. A
cumulative distribution function stores the same law by accumulating probability
up to a threshold.

\begin{definitionbox}
For a real-valued random variable \(X\), the \textbf{cumulative distribution
function}, or \textbf{CDF}, is
\[
F_X(t)=P(X\leq t),
\qquad t\in\mathbb R.
\]
\end{definitionbox}

If \(X\) is discrete with probability mass function \(p_X\), then
\[
F_X(t)=\sum_{x\leq t}p_X(x).
\]
Thus the CDF is the accumulated PMF.

\subsection*{Example: an indicator variable}

Let \(I_A\) be the indicator of an event \(A\), and write \(p=P(A)\). Then
\[
F_{I_A}(t)=
\begin{cases}
0, & t<0,\\
1-p, & 0\leq t<1,\\
1, & t\geq1.
\end{cases}
\]
The jump at \(0\) has size \(1-p\), and the jump at \(1\) has size \(p\).

\subsection*{Jumps and probability masses}

For every real number \(x\),
\[
P(X=x)=F_X(x)-F_X(x^-),
\]
where
\[
F_X(x^-)=\lim_{t\uparrow x}F_X(t)
\]
is the left-hand limit. The probability mass at a point is therefore the size of
the jump of the CDF at that point.

\subsection*{Basic properties of every CDF}

Every cumulative distribution function satisfies:
\begin{itemize}
    \item \(0\leq F_X(t)\leq1\) for every \(t\in\mathbb R\);
    \item \(F_X\) is non-decreasing;
    \item \(F_X\) is right-continuous;
    \item \(\lim_{t\to-\infty}F_X(t)=0\);
    \item \(\lim_{t\to\infty}F_X(t)=1\).
\end{itemize}

Right-continuity means
\[
\lim_{s\downarrow t}F_X(s)=F_X(t).
\]
For a discrete random variable, the CDF is constant between support values and
jumps upward at values with positive probability. The right-continuous convention
includes the mass at the point where the jump occurs.

\subsection*{Computing probabilities from a CDF}

Once \(F_X\) is known,
\[
P(X>t)=1-F_X(t).
\]
For \(a<b\),
\[
P(a<X\leq b)=F_X(b)-F_X(a).
\]
Endpoint conventions matter in the discrete case because an endpoint may carry
positive probability.

\begin{theorembox}
For a discrete random variable, the PMF and CDF describe the same probability
law. The CDF accumulates the masses, and the masses are recovered from its jumps.
\end{theorembox}
