\section{The Euclidean Geometry Available Before Coordinates}

Before introducing coordinates, we recall the geometric world that is already
available. At this stage there are no ordered pairs, no axes, and no vectors.
There are points, straight lines, segments, circles, and angles, together with
relations such as incidence, betweenness, congruence, perpendicularity, and
parallelism.

\begin{corebox}[What must be retained]
Geometry does not begin with coordinates. Points, lines, circles, right angles,
parallelism, congruence, and straightedge-and-compass constructions are already
meaningful before any point is assigned a pair of numbers. Coordinates will add
an algebraic language to this existing geometry.
\end{corebox}

\begin{remarkbox}
Coordinates will later translate this geometry into arithmetic. They do not
create the geometry. We can already construct and compare geometric objects
before assigning numbers to all points and angles.
\end{remarkbox}

\input{chapters/chapter_01/euclidean_prerequisites/levels_of_geometric_statements}
\input{chapters/chapter_01/euclidean_prerequisites/euclids_five_postulates}
\input{chapters/chapter_01/euclidean_prerequisites/straightedge_and_compass_rules}
\input{chapters/chapter_01/euclidean_prerequisites/derived_constructions}
\input{chapters/chapter_01/euclidean_prerequisites/parallel_lines_and_proportions}
\input{chapters/chapter_01/euclidean_prerequisites/trigonometry_before_coordinates}
\input{chapters/chapter_01/euclidean_prerequisites/numerical_angle_measure}
