\subsection{Sine and Cosine Before Coordinates}

Take a right triangle containing an acute angle $\alpha$. Let $h$ denote the
hypotenuse, let $a$ be the leg adjacent to $\alpha$, and let $b$ be the leg
opposite $\alpha$.

\begin{center}
\begin{tikzpicture}[scale=0.95]
    \coordinate (A) at (0,0);
    \coordinate (B) at (4.8,0);
    \coordinate (C) at (4.8,2.9);
    \coordinate (D) at (3.0,0);
    \coordinate (E) at (3.0,1.8125);

    \draw[subset] (A) -- (B) -- (C) -- cycle;
    \draw (B) ++(-0.3,0) -- ++(0,0.3) -- ++(0.3,0);

    \draw[orange!85!black,line width=1pt] (D) -- (E);
    \draw (D) ++(-0.22,0) -- ++(0,0.22) -- ++(0.22,0);
    \draw[orange!85!black,->] (3.13,0.63) -- (3.13,1.03);
    \draw[orange!85!black,->] (4.93,1.05) -- (4.93,1.45);

    \draw (0.85,0) arc[start angle=0,end angle=31.1,radius=0.85];
    \node at (1.08,0.28) {$\alpha$};

    \node[below] at (3.9,0) {$a$};
    \node[right] at (4.8,2.05) {$b$};
    \node[above left] at (3.9,2.35) {$h$};

    \node[below] at (1.5,0) {$a'$};
    \node[left,orange!85!black] at (3.0,0.92) {$b'$};
    \node[above left] at (1.55,0.98) {$h'$};

    \node[orange!85!black,align=left,anchor=west] at (5.35,1.45)
        {$DE\parallel BC$\\[-1pt]\small Thales' theorem};
\end{tikzpicture}

\[
\frac{a'}{a}=\frac{b'}{b}=\frac{h'}{h},
\qquad
\frac{a'}{h'}=\frac{a}{h},
\qquad
\frac{b'}{h'}=\frac{b}{h}.
\]
\end{center}

\begin{proofbox}[Why the ratios depend only on the angle]
All right triangles containing the same acute angle are similar. Corresponding
side lengths therefore differ by one common scale factor, which cancels in the
ratios $a/h$ and $b/h$. These ratios depend on the angle, not on the size of the
triangle chosen to represent it.
\end{proofbox}

\begin{definitionbox}
For an acute geometric angle $\alpha$,
\[
\boxed{
\cos\alpha=\frac{a}{h},
\qquad
\sin\alpha=\frac{b}{h}
}.
\]
\end{definitionbox}

\begin{definitionbox}
For a right angle,
\[
\boxed{\cos\alpha=0}.
\]
If $\alpha$ is obtuse and $\beta$ is its acute supplement, define
\[
\boxed{\cos\alpha=-\cos\beta}.
\]
Thus cosine is positive for acute angles, zero for a right angle, and negative
for obtuse angles.
\end{definitionbox}

\begin{interpretationbox}[Trigonometry is geometric before it is coordinate-based]
Sine and cosine are ratios determined by similarity. Their definitions do not
require Cartesian axes or ordered pairs. Coordinates will later provide a new
way to calculate and use these same geometric quantities.
\end{interpretationbox}

\begin{derivationbox}[The fundamental identity]
For an acute angle, the Pythagorean theorem gives
\[
a^2+b^2=h^2.
\]
Dividing by $h^2$ yields
\[
\left(\frac{a}{h}\right)^2+
\left(\frac{b}{h}\right)^2=1.
\]
Substituting the definitions of cosine and sine gives the identity below.
\end{derivationbox}

\begin{theorembox}
For every angle in the geometric range considered here,
\[
\boxed{\cos^2\alpha+\sin^2\alpha=1}.
\]
For acute angles this is a direct reformulation of the Pythagorean theorem.
\end{theorembox}
