\section{Beyond a Change of Basis: Polar Coordinates}

The preceding section changed the basis of the plane. In every such change we
kept two fixed, nonparallel vector directions and described a vector by the
signed amounts of those two directions. The coordinate grid could become
slanted, but it still consisted of two families of parallel straight lines.

There is a more radical possibility. We may keep the same Euclidean plane and
describe a point by quantities that are not coefficients of two fixed basis
vectors. Polar coordinates provide the first important example.

\begin{remarkbox}
A change from one vector basis to another is a \emph{linear} change of
coordinates. Polar coordinates are not obtained from a new vector basis. Their
coordinate curves are rays and circles, and the directions used to read the
coordinates vary from point to point.
\end{remarkbox}

\subsection*{Two different questions about the same point}

The Cartesian description of a point uses two signed coordinate values obtained
from projections parallel to the coordinate directions:
\[
P=(x,y).
\]
This is natural when two fixed straight directions are important. We may instead
ask:
\begin{enumerate}
    \item How far is the point from the origin?
    \item In which direction do we move from the origin to reach it?
\end{enumerate}

These questions lead to the polar pair
\[
(r,\theta),
\]
where, for a point different from the origin,
\[
r>0,
\qquad
0\leq\theta<2\pi.
\]
The number $r$ is the Euclidean distance from the origin, and $\theta$ is the
angle measured counterclockwise from the positive $x$-axis.

The origin is handled separately. In the usual non-unique polar notation every
pair $(0,\theta)$ represents the origin. In this course we choose the single
representative
\[
O=(0,0)
\]
so that every point has exactly one allowed polar pair. Thus
\[
\mathcal P
=
\{(0,0)\}
\cup
\bigl((0,\infty)\times[0,2\pi)\bigr).
\]

\subsection*{Rectangular and polar grids}

The two descriptions below refer to the same Euclidean plane. What changes is
the family of curves used to locate a point.

\begin{center}
\begin{tikzpicture}[scale=0.78,>=stealth]
    % Cartesian grid
    \begin{scope}[xshift=0cm]
        \draw[gray!35,step=1] (-3.2,-3.2) grid (3.2,3.2);
        \draw[axis,->] (-3.4,0) -- (3.5,0) node[right] {$x$};
        \draw[axis,->] (0,-3.4) -- (0,3.5) node[above] {$y$};
        \coordinate (P1) at (2.4,1.8);
        \draw[helper,dashed] (P1) -- (2.4,0);
        \draw[helper,dashed] (P1) -- (0,1.8);
        \fill[geometricSubsetColor] (P1) circle (2.3pt) node[above right] {$P$};
        \node[align=center] at (0,-4.15)
            {Cartesian grid\\two families of parallel lines};
    \end{scope}

    % Polar grid
    \begin{scope}[xshift=9.2cm]
        \draw[axis,->] (-3.4,0) -- (3.5,0) node[right] {$x$};
        \draw[axis,->] (0,-3.4) -- (0,3.5) node[above] {$y$};
        \foreach \r in {1,2,3} {
            \draw[gray!45] (0,0) circle (\r);
        }
        \foreach \a in {0,30,...,330} {
            \draw[gray!45] (0,0) -- ({3.2*cos(\a)},{3.2*sin(\a)});
        }
        \coordinate (P2) at (2.4,1.8);
        \draw[very thick,->] (0,0) -- (P2) node[midway,above left] {$r$};
        \draw (0.8,0) arc[start angle=0,end angle=36.87,radius=0.8];
        \node at (1.05,0.34) {$\theta$};
        \fill[geometricSubsetColor] (P2) circle (2.3pt) node[above right] {$P$};
        \node[align=center] at (0,-4.15)
            {polar grid\\rays and concentric circles};
    \end{scope}
\end{tikzpicture}
\end{center}

In the Cartesian grid, keeping $x$ fixed gives a vertical straight line and
keeping $y$ fixed gives a horizontal straight line. In the polar grid, keeping
$\theta$ fixed gives a ray from the origin, while keeping $r$ fixed gives a
circle. This is the visible reason polar coordinates are not merely a slanted
basis.

\subsection*{Converting between the descriptions}

Let $P$ have polar coordinates $(r,\theta)$. The right triangle determined by
$P$ gives
\[
\boxed{
x=r\cos\theta,
\qquad
y=r\sin\theta
}.
\]
Thus the position vector of $P$ may be written as
\[
\overrightarrow{OP}
=
r(\cos\theta,\sin\theta).
\]
This resembles a scalar multiple of a unit vector, but the direction
$(\cos\theta,\sin\theta)$ depends on the coordinate $\theta$. There are not two
fixed basis vectors whose constant coefficients are $r$ and $\theta$.

Conversely, if $(x,y)\neq(0,0)$, then
\[
\boxed{r=\sqrt{x^2+y^2}}.
\]
The polar angle is the unique number $\theta\in[0,2\pi)$ satisfying
\[
\boxed{
\cos\theta=\frac{x}{r},
\qquad
\sin\theta=\frac{y}{r}
}.
\]
These two equations determine both the quadrant and the exact direction. The
single expression $\arctan(y/x)$ is insufficient because it loses quadrant
information and is undefined when $x=0$. In computation the correct angle is
usually obtained with the two-argument function $\operatorname{atan2}(y,x)$,
with a negative output shifted by $2\pi$.

\begin{examplebox}
For
\[
(r,\theta)=\left(2,\frac{\pi}{3}\right),
\]
we obtain
\[
x=2\cos\frac{\pi}{3}=1,
\qquad
y=2\sin\frac{\pi}{3}=\sqrt3.
\]
Thus the same point is written as
\[
P=(1,\sqrt3)
\]
in Cartesian coordinates and as
\[
P=\left(2,\frac{\pi}{3}\right)
\]
in polar coordinates.
\end{examplebox}

\subsection*{Why the transformation is nonlinear}

For a change of vector basis, multiplying a vector by $\lambda$ multiplies each
of its coordinates by $\lambda$, and the coordinates of a sum are obtained by
adding coordinate pairs. Polar coordinates do not behave this way.

For example, the points with polar pairs
\[
(1,0)
\qquad\text{and}\qquad
\left(1,\frac{\pi}{2}\right)
\]
have Cartesian position vectors $(1,0)$ and $(0,1)$. Their vector sum is
$(1,1)$, whose polar coordinates are
\[
\left(\sqrt2,\frac{\pi}{4}\right),
\]
not
\[
\left(2,\frac{\pi}{2}\right).
\]
Therefore polar coordinate pairs cannot be added componentwise to add vectors.
The formulas
\[
x=r\cos\theta,
\qquad
y=r\sin\theta
\]
contain trigonometric functions and describe a genuinely nonlinear change of
coordinates.

\subsection*{A circle becomes simple}

Polar coordinates are useful when distance from the origin is the main
geometric feature. Let
\[
C_R
=
\{(x,y)\in\mathbb R^2:x^2+y^2=R^2\},
\qquad R>0.
\]
Substituting
\[
x=r\cos\theta,
\qquad
y=r\sin\theta
\]
gives
\[
r^2\cos^2\theta+r^2\sin^2\theta=R^2.
\]
Since $\cos^2\theta+\sin^2\theta=1$, this reduces to
\[
\boxed{r=R}.
\]
Thus
\[
x^2+y^2=R^2
\qquad\text{and}\qquad
r=R
\]
describe the same geometric circle. The set has not changed; only the coordinate
language has changed.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[->,thick] (-3.2,0) -- (3.4,0) node[right] {$x$};
    \draw[->,thick] (0,-3.2) -- (0,3.4) node[above] {$y$};
    \draw[very thick] (0,0) circle (2);
    \draw[thick] (0,0) -- ({2*cos(35)},{2*sin(35)})
        node[midway,above left] {$R$};
    \fill (0,0) circle (1.7pt) node[below left] {$O$};
    \node[above right] at (1.4,1.4) {$C_R$};
\end{tikzpicture}
\end{center}

\begin{summarybox}
A vector basis uses fixed straight directions and leads to linear coordinate
changes. Polar coordinates instead use a distance and an angle:
\[
(r,\theta)\longleftrightarrow(r\cos\theta,r\sin\theta).
\]
Their coordinate curves are circles and rays, their directions vary across the
plane, and vector addition is not performed by adding polar coordinate pairs.
This is our first example of a nonlinear coordinate system.
\end{summarybox}
