\section{Review Exercises}

\subsection{Elementary computational exercises}

\begin{enumerate}
\item Determine whether
\[
A=\begin{pmatrix}2&1\\3&2\end{pmatrix}
\]
is invertible. If so, compute $A^{-1}$ with the $2\times2$ formula and verify both products.

\item Compute the inverse of
\[
A=\begin{pmatrix}1&3\\2&5\end{pmatrix}
\]
by Gauss--Jordan elimination, recording every row operation.

\item Compute the inverse of
\[
D=\begin{pmatrix}2&0\\0&-4\end{pmatrix}
\]
and explain the visible coordinate action that is being reversed.

\item Decide which matrices are invertible and compute the inverse when it exists:
\[
A=\begin{pmatrix}1&2\\2&4\end{pmatrix},
\quad
B=\begin{pmatrix}0&1\\1&0\end{pmatrix},
\quad
C=\begin{pmatrix}3&0\\0&1\end{pmatrix}.
\]

\item For
\[
A=\begin{pmatrix}2&-1\\1&1\end{pmatrix},
\qquad
b=\begin{pmatrix}5\\1\end{pmatrix},
\]
compute $A^{-1}$ and solve $Ax=b$.

\item Solve $AX=B$ for
\[
A=\begin{pmatrix}1&1\\0&2\end{pmatrix},
\qquad
B=\begin{pmatrix}3&1\\4&2\end{pmatrix}.
\]

\item Solve $XA=B$ for the same matrices and compare the multiplication order with the previous exercise.

\item For
\[
A=\begin{pmatrix}a&b\\c&d\end{pmatrix},
\]
write the cofactor matrix, the adjugate, and the inverse formula. Specialise to $A=\begin{pmatrix}2&3\\1&4\end{pmatrix}$.

\item Let
\[
A=\begin{pmatrix}1&2\\0&1\end{pmatrix},
\qquad
B=\begin{pmatrix}2&0\\1&1\end{pmatrix}.
\]
Compute $A^{-1}$, $B^{-1}$, $(AB)^{-1}$, and $B^{-1}A^{-1}$.

\item During Gauss--Jordan elimination the left block reduces to
\[
\begin{pmatrix}1&0\\0&0\end{pmatrix}.
\]
Explain the computational conclusion and confirm it for $A=\begin{pmatrix}1&2\\2&4\end{pmatrix}$ using the determinant.
\end{enumerate}

\subsection{Development and reasoning exercises}

\begin{enumerate}
\item Invert
\[
A=\begin{pmatrix}
1&1&0\\
0&1&1\\
1&0&1
\end{pmatrix}
\]
by Gauss--Jordan elimination. Before every major step, state the pivot goal; after the calculation, verify one product $AA^{-1}$.

\item Invert the triangular matrix
\[
T=\begin{pmatrix}
1&2&0\\
0&1&3\\
0&0&2
\end{pmatrix}
\]
by Gauss--Jordan elimination. Explain how the zero pattern reduces the work without changing the method.

\item Apply Gauss--Jordan elimination to
\[
S=\begin{pmatrix}
1&2&3\\
2&4&6\\
0&1&1
\end{pmatrix}.
\]
Show where a required pivot fails, identify the row relation, and connect the failure with $\det S=0$.

\item For
\[
A=\begin{pmatrix}
2&0&0\\
0&-1&0\\
0&0&4
\end{pmatrix},
\]
find the inverse directly from structure and again by Gauss--Jordan. Compare the two approaches and explain why structural inspection should come first.

\item Compute the inverse of
\[
A=\begin{pmatrix}2&1\\1&1\end{pmatrix}
\]
with both the formula and Gauss--Jordan. Match the corresponding steps and explain what each method makes most visible.

\item Prove that the Gauss--Jordan construction
\[
[A\mid I]\longrightarrow[I\mid A^{-1}]
\]
works by interpreting row operations as multiplication by elementary matrices. Use a concrete $2\times2$ example inside the argument.

\item Explain structurally why
\[
A\operatorname{adj}(A)=(\det A)I.
\]
Verify the identity for one $2\times2$ matrix and state why the adjugate method is conceptually useful even when it is not efficient.

\item Prove
\[
(AB)^{-1}=B^{-1}A^{-1}
\]
and illustrate the reversed order using two simple geometric actions, such as a scaling and a coordinate swap.

\item A student writes $AX=B\Rightarrow X=BA^{-1}$. Diagnose the error, solve one numerical equation $AX=B$, and explain why the correct side of multiplication follows from the order of actions.

\item Give a complete inversion strategy: inspect structure, test existence, choose between a direct formula and Gauss--Jordan, perform and record reversible steps, verify the answer, and interpret a failure. Apply the strategy briefly to one matrix of your choice.
\end{enumerate}
