\section{Review Exercises}

\subsection{Elementary computational exercises}

\begin{enumerate}
\item For
\[
f(x)=x^2-1,
\]
compute the average rate of change on $[1,3]$ and on $[2,2.5]$. Interpret each result as a secant slope.

\item Use the difference quotient to compute the derivative of
\[
f(x)=x^2
\]
at a general point $a$. Then evaluate $f'(3)$ and write the tangent-line equation at $(3,9)$.

\item Differentiate:
\[
f(x)=4x^3-3x^2+2x-7,
\qquad
g(x)=3e^x-2\sin x+\cos x.
\]

\item Differentiate using the product, quotient, or chain rule as appropriate:
\[
f(x)=x^2e^x,
\qquad
g(x)=\frac{x+1}{x-2},
\qquad
h(x)=(x^2+1)^4.
\]

\item Find the tangent and normal lines to
\[
f(x)=x^3-x
\]
at $x=1$. Verify that their slopes have product $-1$.

\item For
\[
f(x)=x^3-3x,
\]
find all critical points, determine the intervals on which $f$ is increasing or decreasing, and classify the local extrema.

\item Find the absolute maximum and minimum of
\[
f(x)=x^2-4x+1
\]
on $[0,5]$. Check the critical points and both endpoints.

\item For
\[
f(x)=x^4-4x^2,
\]
compute $f'$ and $f''$. Determine the intervals of increase and decrease, the local extrema, the intervals of concavity, and the inflection points.

\item A particle has position
\[
s(t)=t^3-6t^2+9t,
\qquad 0\le t\le4.
\]
Find velocity and acceleration, determine when the particle is at rest, and identify the intervals of positive and negative velocity.

\item A rectangle has perimeter $20$. Let one side have length $x$.
\begin{enumerate}
\item write the area as a function of $x$ and determine its physical domain;
\item find the critical point;
\item compare it with the endpoint behaviour and determine the maximum-area rectangle.
\end{enumerate}
\end{enumerate}

\subsection{Conceptual and reasoning exercises}

\begin{enumerate}
\item Explain how the derivative connects an average rate of change, a limiting secant slope, a tangent slope, and an instantaneous rate.

\item Why does differentiability at a point imply continuity there? Give an example showing that continuity does not imply differentiability.

\item Explain why the units of a derivative are ``output units per input unit''. Give one geometric and one physical example.

\item Interpret the product rule and chain rule structurally. Why does the chain rule multiply by the rate of change of the inner function?

\item Explain how the mean value theorem connects local derivative information with finite changes over an interval.

\item Why does $f'(x)>0$ throughout an interval imply that $f$ is increasing there? State clearly where the mean value theorem enters the argument.

\item A critical point need not be a local extremum. Give an example and explain why solving $f'(x)=0$ is only the beginning of the analysis.

\item Explain what new information is supplied by the second derivative. Distinguish concavity, inflection points, and local extrema.

\item In an optimisation problem, why must the model, domain, critical points, and endpoints all be checked? Explain how ignoring the domain can produce a mathematically correct but physically meaningless answer.

\item Describe a complete qualitative analysis of a differentiable function: domain, zeros, first derivative, critical points, monotonicity, extrema, second derivative, concavity, inflection points, and long-term behaviour. Explain how these pieces form one coherent picture rather than a checklist of unrelated calculations.
\end{enumerate}
