\section{Scalar and Vector Potentials}
\label{sec:scalar_vector_potentials}

The electric and magnetic fields are the directly measurable quantities of classical electromagnetism, but the action principle is most naturally formulated in terms of potentials. The scalar potential and vector potential also encode two of Maxwell's equations automatically and make the gauge structure of the theory visible.

\subsection{Why introduce potentials?}

The electromagnetic fields satisfy the homogeneous relations

\[
\nabla\cdot\mathbf{B}=0
\]

and

\[
\nabla\times\mathbf{E}
+
\frac{\partial\mathbf{B}}{\partial t}
=0.
\]

These equations constrain the possible electric and magnetic fields. Potentials provide a systematic way to solve the constraints before the remaining Maxwell equations are imposed.

The basic definitions are

\[
\boxed{
\mathbf{B}
=
\nabla\times\mathbf{A}
}
\]

and

\[
\boxed{
\mathbf{E}
=
-\nabla\Phi
-
\frac{\partial\mathbf{A}}{\partial t}
}.
\]

Here

\begin{itemize}
    \item \(\Phi(\mathbf{x},t)\) is the scalar potential;
    \item \(\mathbf{A}(\mathbf{x},t)\) is the vector potential.
\end{itemize}

The scalar potential contributes the irrotational part of the electric field. The time derivative of the vector potential supplies the electric field produced by time-dependent magnetic configurations.

\subsection{Electrostatic scalar potential}

In electrostatics,

\[
\frac{\partial\mathbf{B}}{\partial t}=0
\]

and Faraday's law becomes

\[
\nabla\times\mathbf{E}=0.
\]

On a simply connected region, every curl-free vector field can be written locally as the gradient of a scalar function. Therefore

\[
\boxed{
\mathbf{E}
=
-\nabla\Phi
}.
\]

The minus sign is conventional and physically useful. A positive test charge accelerates toward lower electric potential energy.

For a charge \(q\), the potential energy is

\[
U=q\Phi,
\]

and the force is

\[
\mathbf{F}
=
-\nabla U
=
-q\nabla\Phi
=
q\mathbf{E}.
\]

\subsection{Potential difference}

For two points \(P_1\) and \(P_2\) in an electrostatic field,

\[
\Phi(P_2)-\Phi(P_1)
=
-\int_{P_1}^{P_2}
\mathbf{E}\cdot\mathrm{d}\boldsymbol{\ell}.
\]

Because

\[
\nabla\times\mathbf{E}=0,
\]

the integral is path independent in a simply connected region. Only potential differences are measurable. Adding a constant,

\[
\Phi
\longrightarrow
\Phi+C,
\]

does not change the electric field.

\subsection{Example: uniform electric field}

Let

\[
\mathbf{E}
=
E_0\mathbf{e}_x.
\]

A corresponding scalar potential is

\[
\boxed{
\Phi(x)
=
-E_0x
}.
\]

Indeed,

\[
-\nabla\Phi
=
-\frac{\partial\Phi}{\partial x}
\mathbf{e}_x
=
E_0\mathbf{e}_x.
\]

The more general potential

\[
\Phi(x)
=
-E_0x+C
\]

produces the same field.

\subsection{Example: point-charge potential}

For a point charge \(q\) at the origin, the electrostatic potential in SI units is

\[
\boxed{
\Phi(r)
=
\frac{1}{4\pi\varepsilon_0}
\frac{q}{r}
}.
\]

Taking the gradient gives

\[
\mathbf{E}
=
-\nabla\Phi
=
\frac{1}{4\pi\varepsilon_0}
\frac{q}{r^2}
\mathbf{e}_r.
\]

The inverse-distance potential therefore produces the inverse-square electric field.

\subsection{Magnetic vector potential}

The magnetic Gauss law is

\[
\nabla\cdot\mathbf{B}=0.
\]

On a sufficiently regular simply connected region, a divergence-free field can be written as a curl:

\[
\boxed{
\mathbf{B}
=
\nabla\times\mathbf{A}
}.
\]

This representation is natural because the divergence of every curl vanishes identically:

\[
\nabla\cdot
\left(
\nabla\times\mathbf{A}
\right)
=0.
\]

The vector potential is not unique. If

\[
\mathbf{A}'
=
\mathbf{A}
+
\nabla\chi,
\]

then

\[
\nabla\times\mathbf{A}'
=
\nabla\times\mathbf{A}
+
\nabla\times\nabla\chi
=
\mathbf{B}.
\]

This nonuniqueness is the first appearance of gauge freedom.

\subsection{Example: uniform magnetic field}

Let

\[
\mathbf{B}
=
B_0\mathbf{e}_z.
\]

One useful vector potential is the symmetric gauge

\[
\boxed{
\mathbf{A}
=
\frac{1}{2}
\mathbf{B}\times\mathbf{r}
}.
\]

Explicitly,

\[
\mathbf{A}
=
\frac{B_0}{2}
\left(
-y\mathbf{e}_x
+x\mathbf{e}_y
\right).
\]

Its curl is

\[
\nabla\times\mathbf{A}
=
B_0\mathbf{e}_z.
\]

Another valid choice is

\[
\mathbf{A}'
=
B_0x\mathbf{e}_y.
\]

The difference is a gradient:

\[
\mathbf{A}'-\mathbf{A}
=
\nabla
\left(
\frac{1}{2}B_0xy
\right).
\]

The two vector potentials describe the same magnetic field.

\subsection{Time-dependent fields}

When the magnetic field changes with time, the electric field is generally not curl free. A scalar potential alone is insufficient. The combination

\[
\mathbf{E}
=
-\nabla\Phi
-
\frac{\partial\mathbf{A}}{\partial t}
\]

contains both the electrostatic and induction contributions.

The vector potential is therefore not merely a convenient representation of \(\mathbf{B}\). Its time dependence contributes directly to the physical electric field.

\subsection{Units}

In SI units,

\[
[\Phi]
=
\mathrm{V},
\]

and

\[
[\mathbf{A}]
=
\mathrm{V\,s\,m^{-1}}
=
\mathrm{T\,m}.
\]

The terms in the electric-field formula have the same units:

\[
[\nabla\Phi]
=
\mathrm{V\,m^{-1}},
\]

and

\[
\left[
\frac{\partial\mathbf{A}}{\partial t}
\right]
=
\mathrm{V\,m^{-1}}.
\]

Likewise,

\[
[\nabla\times\mathbf{A}]
=
\mathrm{T}.
\]

\subsection{Local and global existence}

The statements

\[
\nabla\times\mathbf{E}=0
\quad\Longrightarrow\quad
\mathbf{E}=-\nabla\Phi
\]

and

\[
\nabla\cdot\mathbf{B}=0
\quad\Longrightarrow\quad
\mathbf{B}=\nabla\times\mathbf{A}
\]

are always valid locally under standard regularity assumptions. Their global validity can depend on the topology of the region.

For example, a region containing a removed line or excluded flux tube may admit fields whose local potentials cannot be combined into one globally single-valued potential. These topological subtleties do not alter the local differential identities, but they matter in global gauge theory.

\subsection{Potentials and physical observables}

Classically, forces on a particle depend on

\[
\mathbf{E}
\quad\text{and}\quad
\mathbf{B},
\]

not on a particular choice of \(\Phi\) and \(\mathbf{A}\). Different potentials can represent the same electromagnetic field.

This does not make the potentials mathematically dispensable. They are the natural variables for

\begin{itemize}
    \item the electromagnetic action;
    \item Lorentz-covariant notation;
    \item the construction of the field tensor;
    \item gauge transformations;
    \item coupling to charged matter.
\end{itemize}

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item using \(\mathbf{E}=-\nabla\Phi\) for a general time-dependent field;
    \item assuming that the scalar or vector potential is unique;
    \item forgetting the minus sign in the electrostatic relation;
    \item confusing a constant shift of \(\Phi\) with a change of the electric field;
    \item claiming global existence of a potential without checking the topology of the domain;
    \item treating different gauge choices as different physical electromagnetic fields.
\end{enumerate}

\subsection{What has been established}

The electromagnetic fields can be represented by a scalar potential and a vector potential:

\[
\mathbf{E}
=
-\nabla\Phi
-
\frac{\partial\mathbf{A}}{\partial t},
\qquad
\mathbf{B}
=
\nabla\times\mathbf{A}.
\]

The potentials are not unique, but their derivatives determine the measurable fields. The next section verifies directly that these definitions reproduce the homogeneous Maxwell equations.
