\section{Electric and Magnetic Components of the Field Tensor}
\label{sec:electric_magnetic_tensor_components}

The six independent components of the antisymmetric tensor \(F_{\mu\nu}\) are the three electric-field components and the three magnetic-field components. Because signs depend on index position and metric convention, the identification must be derived rather than memorized in isolation.

\subsection{Conventions}

We use

\[
A^{\mu}
=
\left(
\frac{\Phi}{c},
A_x,
A_y,
A_z
\right),
\]

\[
A_{\mu}
=
\left(
\frac{\Phi}{c},
-A_x,
-A_y,
-A_z
\right),
\]

and

\[
\partial_{\mu}
=
\left(
\frac{1}{c}\partial_t,
\partial_x,
\partial_y,
\partial_z
\right).
\]

The field tensor is

\[
F_{\mu\nu}
=
\partial_{\mu}A_{\nu}
-
\partial_{\nu}A_{\mu}.
\]

\subsection{Time-space components}

For a spatial index \(i\),

\[
\begin{aligned}
F_{0i}
&=
\partial_0A_i
-
\partial_iA_0\\
&=
\frac{1}{c}
\frac{\partial}{\partial t}
(-A_i)
-
\partial_i
\left(
\frac{\Phi}{c}
\right)\\
&=
-
\frac{1}{c}
\left(
\frac{\partial A_i}{\partial t}
+
\partial_i\Phi
\right).
\end{aligned}
\]

Since

\[
E_i
=
-
\partial_i\Phi
-
\frac{\partial A_i}{\partial t},
\]

we obtain

\[
\boxed{
F_{0i}
=
\frac{E_i}{c}
}.
\]

By antisymmetry,

\[
\boxed{
F_{i0}
=
-
\frac{E_i}{c}
}.
\]

Raising both indices changes the sign of a time-space component:

\[
\boxed{
F^{0i}
=
-
\frac{E_i}{c}
},
\qquad
\boxed{
F^{i0}
=
\frac{E_i}{c}
}.
\]

\subsection{Purely spatial components}

For spatial indices \(i,j\),

\[
\begin{aligned}
F_{ij}
&=
\partial_iA_j
-
\partial_jA_i\\
&=
-
\partial_iA_j^{\mathrm{(3D)}}
+
\partial_jA_i^{\mathrm{(3D)}}.
\end{aligned}
\]

Using

\[
B_k
=
\epsilon_{kij}
\partial_iA_j^{\mathrm{(3D)}},
\]

we find

\[
\boxed{
F_{ij}
=
-\epsilon_{ijk}B_k
}.
\]

Raising two spatial indices produces two minus signs, so

\[
\boxed{
F^{ij}=F_{ij}
=
-\epsilon_{ijk}B_k
}.
\]

Explicitly,

\[
F_{12}=-B_z,
\qquad
F_{23}=-B_x,
\qquad
F_{31}=-B_y.
\]

\subsection{Matrix with lowered indices}

The complete covariant matrix is

\[
\boxed{
F_{\mu\nu}
=
\begin{pmatrix}
0 & E_x/c & E_y/c & E_z/c\\
-E_x/c & 0 & -B_z & B_y\\
-E_y/c & B_z & 0 & -B_x\\
-E_z/c & -B_y & B_x & 0
\end{pmatrix}
}.
\]

The matrix is manifestly antisymmetric.

\subsection{Matrix with raised indices}

Raising both indices gives

\[
\boxed{
F^{\mu\nu}
=
\begin{pmatrix}
0 & -E_x/c & -E_y/c & -E_z/c\\
E_x/c & 0 & -B_z & B_y\\
E_y/c & B_z & 0 & -B_x\\
E_z/c & -B_y & B_x & 0
\end{pmatrix}
}.
\]

Only the time-space entries change sign when both indices are raised.

\subsection{Recovering the fields from the tensor}

The electric field is extracted using

\[
\boxed{
E_i
=
cF_{0i}
=
-cF^{0i}
}.
\]

The magnetic field is extracted using

\[
\boxed{
B_i
=
-
\frac{1}{2}
\epsilon_{ijk}F_{jk}
}.
\]

To verify the second formula, substitute

\[
F_{jk}
=
-\epsilon_{jkl}B_l.
\]

Then

\[
-
\frac{1}{2}
\epsilon_{ijk}F_{jk}
=
\frac{1}{2}
\epsilon_{ijk}
\epsilon_{jkl}B_l
=
B_i.
\]

\subsection{The dual tensor in components}

With

\[
\widetilde{F}^{\mu\nu}
=
\frac{1}{2}
\epsilon^{\mu\nu\rho\sigma}
F_{\rho\sigma}
\]

and

\[
\epsilon^{0123}=+1,
\]

the contravariant dual tensor is

\[
\boxed{
\widetilde{F}^{\mu\nu}
=
\begin{pmatrix}
0 & -B_x & -B_y & -B_z\\
B_x & 0 & E_z/c & -E_y/c\\
B_y & -E_z/c & 0 & E_x/c\\
B_z & E_y/c & -E_x/c & 0
\end{pmatrix}
}.
\]

The dual operation exchanges the electric and magnetic sectors, together with the signs and factors of \(c\) required by the chosen units and orientation.

\subsection{Example: purely electric field}

Let

\[
\mathbf{E}
=
E_0\mathbf{e}_x,
\qquad
\mathbf{B}=0.
\]

Then

\[
F^{\mu\nu}
=
\begin{pmatrix}
0 & -E_0/c & 0 & 0\\
E_0/c & 0 & 0 & 0\\
0 & 0 & 0 & 0\\
0 & 0 & 0 & 0
\end{pmatrix}.
\]

Only the \(0\)-\(1\) spacetime plane is involved.

\subsection{Example: purely magnetic field}

Let

\[
\mathbf{B}
=
B_0\mathbf{e}_z,
\qquad
\mathbf{E}=0.
\]

Then

\[
F^{\mu\nu}
=
\begin{pmatrix}
0 & 0 & 0 & 0\\
0 & 0 & -B_0 & 0\\
0 & B_0 & 0 & 0\\
0 & 0 & 0 & 0
\end{pmatrix}.
\]

The magnetic field occupies the spatial \(x\)-\(y\) plane perpendicular to \(\mathbf{B}\).

\subsection{Lorentz mixing of electric and magnetic fields}

For a boost with velocity

\[
\mathbf{v}
=
v\mathbf{e}_x,
\]

the tensor transformation law gives

\[
\boxed{
E'_x=E_x
},
\]

\[
\boxed{
E'_y
=
\gamma
\left(
E_y-vB_z
\right)
},
\]

\[
\boxed{
E'_z
=
\gamma
\left(
E_z+vB_y
\right)
}.
\]

The magnetic components transform as

\[
\boxed{
B'_x=B_x
},
\]

\[
\boxed{
B'_y
=
\gamma
\left(
B_y+
\frac{v}{c^2}E_z
\right)
},
\]

\[
\boxed{
B'_z
=
\gamma
\left(
B_z-
\frac{v}{c^2}E_y
\right)
}.
\]

The components parallel to the boost remain unchanged, while transverse electric and magnetic components mix.

\subsection{Vector form of the boost transformation}

Separate each field into components parallel and perpendicular to \(\mathbf{v}\). Then

\[
\boxed{
\mathbf{E}'_{\parallel}
=
\mathbf{E}_{\parallel}
},
\]

\[
\boxed{
\mathbf{B}'_{\parallel}
=
\mathbf{B}_{\parallel}
},
\]

\[
\boxed{
\mathbf{E}'_{\perp}
=
\gamma
\left(
\mathbf{E}_{\perp}
+
\mathbf{v}\times\mathbf{B}
\right)
},
\]

and

\[
\boxed{
\mathbf{B}'_{\perp}
=
\gamma
\left(
\mathbf{B}_{\perp}
-
\frac{1}{c^2}
\mathbf{v}\times\mathbf{E}
\right)
}.
\]

These relations show directly that the division into electric and magnetic fields depends on the observer, while the tensor \(F_{\mu\nu}\) is the frame-independent geometric object.

\subsection{Nonrelativistic limits}

When

\[
v\ll c,
\]

we have

\[
\gamma\approx1.
\]

To first order in \(v/c\),

\[
\mathbf{E}'
\approx
\mathbf{E}
+
\mathbf{v}\times\mathbf{B},
\]

and

\[
\mathbf{B}'
\approx
\mathbf{B}
-
\frac{1}{c^2}
\mathbf{v}\times\mathbf{E}.
\]

The electric correction \(\mathbf{v}\times\mathbf{B}\) is familiar from the magnetic part of the Lorentz force under a change of frame.

\subsection{Index-position diagnostic}

With the conventions of this chapter,

\[
F_{0i}
=
\frac{E_i}{c},
\]

but

\[
F^{0i}
=
-
\frac{E_i}{c}.
\]

By contrast,

\[
F_{ij}=F^{ij}.
\]

Whenever a sign appears inconsistent, the first diagnostic step is to check whether the indices are raised or lowered and which metric signature is being used.

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item using the same matrix for \(F_{\mu\nu}\) and \(F^{\mu\nu}\);
    \item forgetting the factor \(1/c\) in the electric components;
    \item assigning \(F_{12}=+B_z\) with the conventions used here;
    \item treating \(\mathbf{E}\) and \(\mathbf{B}\) as independently transforming three-vectors under Lorentz boosts;
    \item losing the antisymmetry when writing the matrix;
    \item applying formulas from a different metric convention without translating the signs.
\end{enumerate}

\subsection{What has been established}

The electromagnetic tensor contains the fields according to

\[
F_{0i}
=
\frac{E_i}{c},
\qquad
F_{ij}
=
-\epsilon_{ijk}B_k.
\]

The explicit matrices display all signs and factors of \(c\). Lorentz boosts mix transverse electric and magnetic components, confirming that \(F_{\mu\nu}\), rather than \(\mathbf{E}\) or \(\mathbf{B}\) separately, is the fundamental relativistic field. The next section constructs scalar invariants from this tensor.
