\section{Review Exercises}

\subsection{Elementary computational and graphical exercises}

\begin{enumerate}

\item Plot the points
\[
A=(3,2),\qquad B=(-2,4),\qquad C=(-3,-1),\qquad D=(2,-3).
\]
For each point, state its quadrant and draw its orthogonal projections onto both
axes.

\item For $P=(-4,3)$, write the two projection points on the coordinate axes and
compute the distances from $P$ to the $x$-axis, the $y$-axis, and the origin.

\item A point has horizontal coordinate $-5$ and vertical coordinate $2$. Construct
it from the axes by orthogonal projection. Repeat for a point with coordinates
$(0,-4)$ and explain what is special about its position.

\item Compute and illustrate with right triangles the distances
\[
d\bigl((1,2),(4,6)\bigr),
\qquad
d\bigl((-2,1),(4,1)\bigr),
\qquad
d\bigl((-3,-2),(1,1)\bigr).
\]

\item Find the midpoint of the segment joining $A=(-2,1)$ and $B=(4,5)$. Verify
that the midpoint is equally distant from both endpoints.

\item A point has coordinates $P=(6,-4)$ when one unit on each axis has its
original length. Find its coordinates when one new unit is equal to two old
units and the origin and axis directions remain unchanged.

\item A point has coordinates $P=(5,3)$. Move the origin to the old point $(2,1)$
without changing the units or axis directions. Compute the coordinates of $P$ in
the new system.

\item A point has coordinates $P=(4,-2)$. Reverse only the positive direction of
the $x$-axis, leaving the origin, unit, and $y$-axis unchanged. Find the new
coordinates and draw both descriptions of the same point.

\item Prepare a value table for
\[
y=-2x+4
\]
at five convenient integer values of $x$. Plot the points, draw the line, and
find its intersections with both coordinate axes.

\item Plot on one coordinate system
\[
y=x+2,
\qquad
y=x^2.
\]
Find their common points algebraically and verify each point in both equations.

\end{enumerate}

\subsection{Development and reasoning exercises}

\begin{enumerate}

\item Describe how to construct a Cartesian coordinate system starting only from a
plane, one chosen point, one chosen line through that point, and the ability to
construct perpendicular lines. Identify every additional choice that must be
made before points can receive numerical coordinates.

\item Explain why a geometric point is not the same object as an ordered pair.
Choose one fixed point and describe it in two coordinate systems with different
origins. State precisely what changes and what remains unchanged.

\item Derive
\[
d(P,Q)=\sqrt{(x_2-x_1)^2+(y_2-y_1)^2}
\]
from the horizontal and vertical displacements between $P=(x_1,y_1)$ and
$Q=(x_2,y_2)$. Include a diagram and explain where perpendicularity is used.

\item A drawing uses twice the scale horizontally as vertically. Explain why the
picture of
\[
x^2+y^2=4
\]
may look like an ellipse even though the equation describes a circle. Explain
what information belongs to the geometry and what information belongs only to
the drawing convention.

\item Find all points on the $x$-axis whose distance from $P=(2,3)$ is $5$.
Before calculating, predict geometrically how many solutions can occur. Then
solve and explain why the algebra agrees with the picture.

\item A point is described by $(6,2)$. A new coordinate system is obtained by
moving the origin to $(2,-1)$, reversing the $x$-axis direction, and replacing
the old unit by a unit twice as long. Compute the new coordinates by treating
these three changes one at a time, and explain why the order of the reasoning
matters.

\item Three vertices of an axis-aligned rectangle are
\[
A=(-2,-1),\qquad B=(3,-1),\qquad C=(3,2).
\]
Find the fourth vertex and compute the side and diagonal lengths. Explain how
the coordinate description makes the rectangle structure visible without
measuring angles numerically.

\item The line $y=2x-3$ and the parabola $y=x^2-2$ are drawn in the same coordinate
system. Find their common points and explain why solving
\[
2x-3=x^2-2
\]
is exactly the algebraic version of asking for points that belong to both
curves.

\item Orthogonal projection assigns to every point of the plane one number on each
axis. Explain why two different points cannot receive the same ordered pair.
Then explain why every ordered pair of real numbers determines exactly one point
of the plane.

\item The construction of a scale begins with integer marks and repeated
bisection. Explain what points can be labelled after finitely many bisections and
why an additional continuity assumption is needed to regard every point of the
axis as having a real coordinate. Keep the argument geometric rather than
set-theoretic.

\end{enumerate}
