\section{Another Way to Describe a Point: Polar Coordinates}

The Cartesian description of a point uses two signed distances:

\[
P=(x,y).
\]

This is natural when horizontal and vertical directions are important. It is not,
however, the only possible way to describe position in the plane. We may instead
ask two different questions:

\begin{enumerate}
    \item How far is the point from the origin?
    \item In which direction do we move from the origin to reach it?
\end{enumerate}

These questions lead to \emph{polar coordinates}. For a point different from the
origin, we use a pair

\[
(r,\theta),
\]

where

\[
r>0,
\qquad
0\leq \theta<2\pi.
\]

The number $r$ is the distance from the origin, and $\theta$ is the angle measured
counterclockwise from the positive $x$-axis. The interval for the angle is
half-open: it contains $0$, but it does not contain $2\pi$. This choice prevents
the same direction from being written twice.

The origin is handled separately. We assign to it the single polar coordinate
pair

\[
O=(0,0).
\]

Thus the set of coordinate pairs used in this course is

\[
\mathcal{P}
=
\{(0,0)\}
\cup
\bigl((0,\infty)\times[0,2\pi)\bigr).
\]

This should be read as follows: either we have the distinguished pair $(0,0)$,
which represents the origin, or we have a positive radius together with an angle
from $0$ up to, but not including, $2\pi$.

\begin{center}
\begin{tikzpicture}[scale=1.05]
    \draw[->] (-0.5,0) -- (4.8,0) node[right] {$x$};
    \draw[->] (0,-0.5) -- (0,3.8) node[above] {$y$};
    \coordinate (O) at (0,0);
    \coordinate (P) at (3.6,2.4);
    \draw[thick] (O) -- (P) node[midway,above left] {$r$};
    \draw[dashed] (P) -- (3.6,0);
    \draw[dashed] (P) -- (0,2.4);
    \draw (0.9,0) arc[start angle=0,end angle=33.69,radius=0.9];
    \node at (1.15,0.35) {$\theta$};
    \fill (P) circle (2pt) node[above right] {$P$};
    \fill (O) circle (1.5pt) node[below left] {$O$};
\end{tikzpicture}
\end{center}

The Cartesian and polar descriptions refer to the same point. The right triangle
in the figure gives

\[
x=r\cos\theta,
\qquad
y=r\sin\theta.
\]

Conversely, if the Cartesian coordinates are known, then

\[
r=\sqrt{x^2+y^2}.
\]

For a point different from the origin, the angle is chosen uniquely from the
interval $[0,2\pi)$ so that its sine and cosine have the correct signs. For this
reason, the formula $\theta=\arctan(y/x)$ must be used with care: by itself it
does not distinguish all four quadrants.

\begin{examplebox}
Consider the point whose polar coordinates are

\[
(r,\theta)=\left(2,\frac{\pi}{3}\right).
\]

Its Cartesian coordinates are

\[
x=2\cos\frac{\pi}{3}=1,
\qquad
y=2\sin\frac{\pi}{3}=\sqrt{3}.
\]

Thus the same point may be written as

\[
P=\left(1,\sqrt{3}\right)
\]

in Cartesian coordinates and as

\[
P=\left(2,\frac{\pi}{3}\right)
\]

in polar coordinates.
\end{examplebox}

Polar coordinates are especially convenient when distance from the origin is the
main geometric feature. Let

\[
C_R
=
\{(x,y)\in\mathbb{R}^2:x^2+y^2=R^2\},
\qquad R>0.
\]

This is the set of all points in the Cartesian plane whose coordinates satisfy
the condition

\[
x^2+y^2=R^2.
\]

Geometrically, $C_R$ is the circle of radius $R$ centred at the origin. After
substituting

\[
x=r\cos\theta,
\qquad
y=r\sin\theta,
\]

the Cartesian condition becomes

\[
r^2\cos^2\theta+r^2\sin^2\theta=R^2.
\]

Since $\cos^2\theta+\sin^2\theta=1$, this reduces to

\[
\boxed{r=R}.
\]

In polar coordinates, the same circle is therefore represented by the set

\[
\{(R,\theta):0\leq\theta<2\pi\}.
\]

The set $C_R$ has not changed. The conditions

\[
x^2+y^2=R^2
\qquad\text{and}\qquad
r=R,
\quad 0\leq\theta<2\pi,
\]

describe the same geometric subset of the plane, but they use different
coordinates. In Cartesian coordinates we express a relation between horizontal
and vertical position. In polar coordinates we state directly that every point
of the set has the same distance from the origin.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[->,thick] (-3.2,0) -- (3.4,0) node[right] {$x$};
    \draw[->,thick] (0,-3.2) -- (0,3.4) node[above] {$y$};
    \draw[very thick] (0,0) circle (2);
    \draw[thick] (0,0) -- ({2*cos(35)},{2*sin(35)})
        node[midway,above left] {$R$};
    \fill (0,0) circle (1.7pt) node[below left] {$O$};
    \node[above right] at (1.4,1.4) {$C_R$};
\end{tikzpicture}
\end{center}

\begin{remarkbox}
The restriction $0\leq\theta<2\pi$ is part of the definition used here. The
angle $2\pi$ is not included, because it points in the same direction as $0$.
We do not identify or glue the endpoints of the interval. We simply choose one
representative angle for each direction.
\end{remarkbox}
