\section{Constructing Cartesian Coordinates from Euclidean Geometry}

We do not begin with ordered pairs of numbers.  We begin with a Euclidean
plane: a geometric plane in which points, straight lines, perpendicular lines,
congruent segments, angle bisection, and the Pythagorean theorem are already
available.  Our aim is to construct a numerical description of this geometry.

\begin{remarkbox}
The Cartesian coordinate system does not create the Euclidean plane.  It is a
method for encoding an already existing Euclidean geometry by numbers.
\end{remarkbox}

\subsection{The Euclidean Plane Before Coordinates}

At the beginning there are points and geometric relations, but no distinguished
origin, no coordinate axes, and no numerical labels attached to points.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \fill[gray!8] (-4,-2.2) rectangle (4,2.2);
    \foreach \x/\y in {-3.2/1.2,-2.1/-0.8,-0.6/1.5,0.8/-1.1,2.1/0.9,3.1/-0.4} {
        \fill (\x,\y) circle (1.7pt);
    }
    \node at (0,-2.55) {a Euclidean plane with no chosen coordinates};
\end{tikzpicture}
\end{center}

\begin{definitionbox}
For this construction we assume the elementary operations of Euclidean
geometry:
\begin{itemize}
    \item through two distinct points one may draw a unique straight line;
    \item through a point one may construct a line perpendicular to a given line;
    \item a segment length may be transferred with a compass;
    \item a segment may be bisected;
    \item lengths in a right triangle satisfy the Pythagorean theorem.
\end{itemize}
\end{definitionbox}

These assumptions supply the geometry from which coordinates will be built.

\subsection{Choosing the First Axis and the Origin}

Choose an arbitrary straight line and call it the future $x$-axis.  Then choose
an arbitrary point $O$ on this line.  This point will become the origin.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[subset] (-4.2,0) -- (4.2,0);
    \fill (0,0) circle (2pt);
    \node[below] at (0,0) {$O$};
    \node[subset label,right] at (4.2,0) {first axis};
\end{tikzpicture}
\end{center}

Nothing forces this particular choice.  A different line or a different point
would produce a different coordinate system on the same plane.

\subsection{Constructing the Perpendicular Axis}

Through $O$, construct the unique line perpendicular to the first axis.  This
line becomes the $y$-axis.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-4.2,0) -- (4.4,0) node[right] {$x$};
    \draw[axis] (0,-3.2) -- (0,3.4) node[above] {$y$};
    \fill (0,0) circle (2pt);
    \node[below left] at (0,0) {$O$};
    \draw (0.25,0) -- (0.25,0.25) -- (0,0.25);
\end{tikzpicture}
\end{center}

The two axes divide the plane into four right angles.  At this stage the axes
have directions, but they still have no numerical scale.

\begin{definitionbox}
A rectangular Cartesian frame consists of
\begin{itemize}
    \item an origin $O$;
    \item two perpendicular axes through $O$;
    \item a chosen positive direction on each axis;
    \item a common unit length.
\end{itemize}
\end{definitionbox}

\subsection{Choosing and Transferring a Unit Length}

Choose a segment of positive length and declare it to have length $1$.  Using a
compass, transfer this same length from $O$ in both directions along both axes.
This produces the marks $1$, $-1$ on the $x$-axis and $1$, $-1$ on the
$y$-axis.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-4.2,0) -- (4.4,0) node[right] {$x$};
    \draw[axis] (0,-3.2) -- (0,3.4) node[above] {$y$};
    \foreach \x in {-3,-2,-1,1,2,3} {
        \draw (\x,0.09) -- (\x,-0.09) node[below=3pt] {$\x$};
    }
    \foreach \y in {-2,-1,1,2} {
        \draw (0.09,\y) -- (-0.09,\y) node[left=3pt] {$\y$};
    }
    \node[below left] at (0,0) {$0$};
    \draw[<->,gray] (0,2.65) -- (1,2.65) node[midway,above] {unit length};
\end{tikzpicture}
\end{center}

Repeated transfer of the unit segment constructs all integer marks.  The sign
records on which side of the origin a point lies.

\begin{remarkbox}
Using the same compass opening on both axes is essential.  It makes one unit on
the $x$-axis congruent to one unit on the $y$-axis.  Without this agreement, the
coordinate formula for Euclidean distance would contain different scale
factors in the two directions.
\end{remarkbox}

\subsection{Bisecting the Unit and Approximating Positions}

A unit segment can be bisected geometrically.  Repeating the construction gives
marks at
\[
\frac12,\quad \frac14,\quad \frac34,\quad \frac18,\quad \frac38,
\quad \ldots
\]
and similarly on the negative side.

\begin{center}
\begin{tikzpicture}[scale=1.1]
    \draw[axis] (-0.4,0) -- (4.4,0);
    \foreach \x/\lab in {0/0,1/{\frac14},2/{\frac12},3/{\frac34},4/1} {
        \draw (\x,0.1) -- (\x,-0.1) node[below=4pt] {$\lab$};
    }
    \draw[gray] (0,0.45) -- (4,0.45);
    \draw[gray] (0,0.38) -- (0,0.52);
    \draw[gray] (2,0.38) -- (2,0.52);
    \draw[gray] (4,0.38) -- (4,0.52);
    \node[above] at (2,0.52) {successive bisections};
\end{tikzpicture}
\end{center}

After $n$ bisections, one obtains all dyadic positions
\[
\frac{k}{2^n},
\qquad k\in\mathbb Z.
\]
These marks can approximate any position on the axis with arbitrary precision.

\begin{remarkbox}
Repeated bisection does not construct every real number exactly.  It constructs
the dyadic rational numbers.  The claim needed here is approximation: for every
point on the axis and every desired precision, sufficiently fine bisection
places dyadic marks on both sides of that point as closely as required.
Identifying all axis positions with real numbers additionally uses the usual
continuity and completeness assumptions of Euclidean measurement.
\end{remarkbox}

Thus each axis becomes a geometric copy of the real number line once an origin,
a direction, and a unit have been fixed.

\subsection{Projecting a Point onto the Axes}

Let $P$ be any point of the plane.  Through $P$, construct a line perpendicular
to the $x$-axis.  Let its intersection with the $x$-axis be $P_x$.  Through
$P$, construct a line perpendicular to the $y$-axis, and call the intersection
with the $y$-axis $P_y$.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-1.2,0) -- (5.0,0) node[right] {$x$};
    \draw[axis] (0,-1.2) -- (0,4.0) node[above] {$y$};
    \coordinate (P) at (3.2,2.4);
    \coordinate (Px) at (3.2,0);
    \coordinate (Py) at (0,2.4);
    \draw[helper] (P) -- (Px);
    \draw[helper] (P) -- (Py);
    \fill (P) circle (2.2pt) node[above right] {$P$};
    \fill (Px) circle (1.8pt) node[below] {$P_x$};
    \fill (Py) circle (1.8pt) node[left] {$P_y$};
    \draw (2.95,0) -- (2.95,0.25) -- (3.2,0.25);
    \draw (0,2.15) -- (0.25,2.15) -- (0.25,2.4);
\end{tikzpicture}
\end{center}

The signed number assigned to $P_x$ is called the $x$-coordinate of $P$.  The
signed number assigned to $P_y$ is called the $y$-coordinate of $P$.

\begin{definitionbox}
If the orthogonal projections of $P$ onto the two scaled axes have numerical
labels $x$ and $y$, then the Cartesian coordinates of $P$ are the ordered pair
\[
P\longleftrightarrow (x,y).
\]
\end{definitionbox}

The order matters: the first number comes from projection onto the $x$-axis,
and the second from projection onto the $y$-axis.

\subsection{Why Every Point Has One Coordinate Pair}

The Euclidean parallel and perpendicular constructions guarantee that each
point has exactly one perpendicular projection onto each axis.  Therefore each
point produces one ordered pair $(x,y)$.

Conversely, given two axis positions labelled $x$ and $y$, draw through the
$x$-position a line perpendicular to the $x$-axis and through the $y$-position
a line perpendicular to the $y$-axis.  These two lines meet at exactly one
point.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-1.0,0) -- (4.8,0) node[right] {$x$};
    \draw[axis] (0,-1.0) -- (0,3.8) node[above] {$y$};
    \draw[helper] (3,0) -- (3,2.5);
    \draw[helper] (0,2.5) -- (3,2.5);
    \fill (3,0) circle (1.7pt) node[below] {$x$};
    \fill (0,2.5) circle (1.7pt) node[left] {$y$};
    \fill (3,2.5) circle (2.2pt) node[above right] {$(x,y)$};
\end{tikzpicture}
\end{center}

\begin{theorembox}
Once the rectangular frame and its common unit are fixed, there is a one-to-one
correspondence
\[
\boxed{
\text{points of the Euclidean plane}
\quad\longleftrightarrow\quad
\text{ordered pairs }(x,y)\in\mathbb R^2
}.
\]
The ordered pairs are numerical labels for geometric points; they are not the
starting point of the construction.
\end{theorembox}

\subsection{The Distance Formula Comes from Pythagoras}

Let
\[
P=(x_1,y_1),
\qquad
Q=(x_2,y_2).
\]
The horizontal separation of the points has length
\[
|x_2-x_1|,
\]
and the vertical separation has length
\[
|y_2-y_1|.
\]
These lengths form the legs of a right triangle whose hypotenuse is the segment
$PQ$.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-0.8,0) -- (5.2,0) node[right] {$x$};
    \draw[axis] (0,-0.8) -- (0,4.5) node[above] {$y$};
    \coordinate (P) at (1,1);
    \coordinate (Q) at (4,3.5);
    \coordinate (R) at (4,1);
    \draw[subset] (P) -- (Q);
    \draw[helper] (P) -- (R) node[midway,below] {$|x_2-x_1|$};
    \draw[helper] (R) -- (Q) node[midway,right] {$|y_2-y_1|$};
    \draw (3.72,1) -- (3.72,1.28) -- (4,1.28);
    \fill (P) circle (2pt) node[below left] {$P$};
    \fill (Q) circle (2pt) node[above right] {$Q$};
\end{tikzpicture}
\end{center}

By the Pythagorean theorem,
\[
|PQ|^2
=
(x_2-x_1)^2+(y_2-y_1)^2.
\]
Therefore
\begin{theorembox}
The Euclidean distance expressed in Cartesian coordinates is
\[
\boxed{
d_E(P,Q)
=
\sqrt{(x_2-x_1)^2+(y_2-y_1)^2}
}.
\]
This formula does not define a new geometry independently of the original
plane.  It is the numerical expression of the Euclidean length already present
in that plane.
\end{theorembox}

\subsection{Angles Are Inherited, Not Created by Coordinates}

Angles and perpendicularity belong to the Euclidean geometry with which we
started.  The axes were chosen to be perpendicular before any coordinates were
assigned.  Coordinates allow us to calculate angle relations, but they do not
create the notion of angle.

For example, two perpendicular lines through a point divide a full turn into
four equal right angles.  Cartesian coordinates encode this fact through the
perpendicular axes and, later, through algebraic criteria for slopes.

\begin{remarkbox}
The logical direction in this construction is
\[
\text{Euclidean geometry}
\longrightarrow
\text{perpendicular axes and a unit}
\longrightarrow
\text{coordinates}
\longrightarrow
\text{algebraic formulas for Euclidean geometry}.
\]
It is not
\[
\mathbb R^2
\longrightarrow
\text{geometry without further assumptions}.
\]
\end{remarkbox}

\begin{summarybox}
A Cartesian coordinate system is constructed on a Euclidean plane by:
\begin{enumerate}
    \item choosing a line and an origin;
    \item constructing the perpendicular axis through the origin;
    \item choosing positive directions and one common unit length;
    \item transferring and bisecting that unit to scale the axes;
    \item assigning to every point its two orthogonal projections;
    \item reading the two signed axis positions as $(x,y)$;
    \item translating Euclidean theorems, such as Pythagoras, into coordinate
          formulas.
\end{enumerate}
The resulting identification with $\mathbb R^2$ is a coordinate representation
of the original Euclidean plane.
\end{summarybox}
