\section{Constructing Cartesian Coordinates from Euclidean Geometry}

We begin with a Euclidean plane, not with ordered pairs and not with a coordinate
system that has already been drawn. Points, straight lines, right angles,
perpendicularity, congruent segments, and the Pythagorean theorem belong to the
geometry available before coordinates are introduced. Our task is to construct,
step by step, the additional choices that turn this plane into a Cartesian
coordinate plane.

\begin{remarkbox}
Coordinates do not create Euclidean geometry. They arise after we choose and
construct a particular system for assigning numerical labels to points of an
already existing Euclidean plane.
\end{remarkbox}

\subsection{The Plane Before Coordinates}

At first there is only a Euclidean plane. There is no distinguished straight
line, no origin, no horizontal or vertical direction, no positive direction,
and no numerical scale.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \fill[gray!8] (-4,-2.1) rectangle (4,2.1);
    \foreach \x/\y in {-3.1/1.1,-2.0/-0.7,-0.5/1.4,0.9/-1.0,2.0/0.8,3.0/-0.3} {
        \fill (\x,\y) circle (1.7pt);
    }
    \node at (0,-2.45) {a Euclidean plane with no chosen coordinates};
\end{tikzpicture}
\end{center}

\begin{definitionbox}
For the construction below we use standard Euclidean operations and theorems:
\begin{itemize}
    \item constructing the straight line through two distinct points;
    \item choosing a point on a constructed line;
    \item constructing a right angle, equivalently constructing the unique
          perpendicular to a line through a chosen point;
    \item transferring a segment length with a compass;
    \item bisecting a segment;
    \item using the continuity of the Euclidean line when passing from a dense
          family of constructed marks to all real positions;
    \item applying the Pythagorean theorem to a right triangle;
    \item later, constructing and comparing circles and arcs and applying the
          law of cosines in a Euclidean triangle.
\end{itemize}
\end{definitionbox}

A right angle is available from the beginning as a Euclidean construction. What
is postponed is its numerical measurement in radians or degrees.

\subsection{First Construction: A Straight Line}

Choose two distinct points of the plane and construct the unique straight line
through them. At this stage it is only a chosen Euclidean line. It is not yet an
axis, because there is not yet a second line, an origin, a direction, or a
scale.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \fill[gray!8] (-4,-2.1) rectangle (4,2.1);
    \coordinate (A) at (-2.8,-0.7);
    \coordinate (B) at (2.7,0.8);
    \draw[subset] (-3.8,-0.97) -- (3.8,1.10);
    \fill (A) circle (2pt) node[below left] {$A$};
    \fill (B) circle (2pt) node[above right] {$B$};
    \node at (0,-2.45) {the first constructed line};
\end{tikzpicture}
\end{center}

\subsection{Second Construction: A Point on the Line}

Choose an arbitrary point $O$ on the constructed line. This point will later
become the common zero of both coordinate axes, but it becomes an origin only
after the rest of the coordinate structure has been constructed.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \fill[gray!8] (-4,-2.1) rectangle (4,2.1);
    \draw[subset] (-3.8,-0.95) -- (3.8,0.95);
    \fill (0,0) circle (2.2pt);
    \node[below right] at (0,0) {$O$};
    \node at (0,-2.45) {a point chosen on the first line};
\end{tikzpicture}
\end{center}

\subsection{Third Construction: The Perpendicular Line}

Through $O$, construct the unique line perpendicular to the first line. The two
constructed lines now intersect at a right angle. Only now do we have the
geometric framework from which two coordinate axes can be made.

We call the first line the $x$-axis, the perpendicular line the $y$-axis, and
their intersection $O$ the origin.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-4.2,0) -- (4.4,0) node[right] {$x$};
    \draw[axis] (0,-3.1) -- (0,3.3) node[above] {$y$};
    \fill (0,0) circle (2pt);
    \node[below left] at (0,0) {$O$};
    \draw (0.26,0) -- (0.26,0.26) -- (0,0.26);
\end{tikzpicture}
\end{center}

The axes are perpendicular because this was required in their geometric
construction. No coordinate formula and no numerical angle measure has been
used.

\subsection{Choosing Positive and Negative Directions}

Each axis is divided by $O$ into two opposite rays. Choose one ray on the
$x$-axis as positive and one ray on the $y$-axis as positive. The opposite rays
are then called negative.

These choices are not forced by the Euclidean plane. Reversing one of them would
produce a different coordinate system on the same geometric plane.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis,->] (-4.2,0) -- (4.4,0) node[right] {positive $x$};
    \draw[axis,->] (0,-3.1) -- (0,3.3) node[above] {positive $y$};
    \fill (0,0) circle (2pt);
    \node[below left] at (0,0) {$O$};
    \node[below] at (-2.7,0) {negative $x$};
    \node[left] at (0,-2.2) {negative $y$};
\end{tikzpicture}
\end{center}

\subsection{Constructing a Common Unit Length}

Choose a positive segment length and call it one unit. Using a compass, transfer
this same length from $O$ repeatedly along both axes and in both directions.
The first marks are labelled $1$ and $-1$; repetition constructs the integer
marks
\[
\ldots,-3,-2,-1,0,1,2,3,\ldots
\]
on each axis.

\begin{center}
\begin{tikzpicture}[scale=0.82]
    \draw[axis] (-4.2,0) -- (4.4,0) node[right] {$x$};
    \draw[axis] (0,-4.1) -- (0,4.3) node[above] {$y$};

    % Four successive compass placements.  The same colour denotes the same
    % iteration on the two axes.
    \draw[blue!70,dashed,line width=0.8pt] (0,0) circle (1);
    \draw[orange!85!black,dashed,line width=0.8pt] (1,0) circle (1);
    \draw[orange!85!black,dashed,line width=0.8pt] (0,1) circle (1);
    \draw[green!55!black,dashed,line width=0.8pt] (2,0) circle (1);
    \draw[green!55!black,dashed,line width=0.8pt] (0,2) circle (1);
    \draw[violet!75,dashed,line width=0.8pt] (3,0) circle (1);
    \draw[violet!75,dashed,line width=0.8pt] (0,3) circle (1);

    \foreach \x in {-3,-2,-1,1,2,3} {
        \draw (\x,0.09) -- (\x,-0.09) node[below=3pt] {$\x$};
    }
    \foreach \y in {-3,-2,-1,1,2,3} {
        \draw (0.09,\y) -- (-0.09,\y) node[left=3pt] {$\y$};
    }
    \node[below left] at (0,0) {$0$};

    \fill[blue!70] (0,0) circle (1.8pt);
    \fill[orange!85!black] (1,0) circle (1.8pt);
    \fill[orange!85!black] (0,1) circle (1.8pt);
    \fill[green!55!black] (2,0) circle (1.8pt);
    \fill[green!55!black] (0,2) circle (1.8pt);
    \fill[violet!75] (3,0) circle (1.8pt);
    \fill[violet!75] (0,3) circle (1.8pt);

    \draw[<->,gray] (0,-3.65) -- (1,-3.65);
    \node[gray,below] at (0.5,-3.65) {one common unit};
\end{tikzpicture}
\end{center}

\begin{remarkbox}
The same unit must be used on both axes. A segment labelled as one unit on the
$x$-axis must be congruent to a segment labelled as one unit on the $y$-axis.
This common scale is what later allows the Pythagorean theorem to produce the
standard coordinate formula for Euclidean distance.
\end{remarkbox}

\subsection{Refining the Scale by Bisection}

The compass and straightedge construction does not stop at integer marks.
Bisecting each unit segment constructs halves. Bisecting again constructs
quarters, then eighths, and so on. Repeating this process on every integer
interval and on both sides of the origin gives all dyadic marks
\[
\frac{k}{2^n},
\qquad k\in\mathbb Z,
\quad n\in\mathbb N.
\]

\begin{center}
\begin{tikzpicture}[scale=1.1]
    \draw[axis] (-0.4,0) -- (4.4,0);
    \foreach \x/\lab in {0/0,1/{\frac14},2/{\frac12},3/{\frac34},4/1} {
        \draw (\x,0.1) -- (\x,-0.1) node[below=4pt] {$\lab$};
    }
    \node[above] at (2,0.35) {successive bisections of one unit};
\end{tikzpicture}
\end{center}

The dyadic marks are dense on the axis: every interval, however small, contains
such marks. They therefore provide arbitrarily fine numerical approximations to
positions on the line.

To pass from this dense constructed scale to a coordinate for every point of an
axis, we use the usual Euclidean continuity assumption: the line has no gaps,
and its positions can be put into one-to-one correspondence with the real
numbers. After this step, every point of each axis has a unique signed real
label, and every real number labels exactly one point of that axis.

\subsection{Assigning Coordinates by Orthogonal Projection}

Only after the axes, directions, common unit, integer marks, refined scale, and
real labels have been constructed can an arbitrary point of the plane be given
coordinates.

Let $P$ be any point of the plane. Through $P$, construct the line perpendicular
to the $x$-axis. It meets the $x$-axis at a unique point $P_x$. Through $P$,
construct the line perpendicular to the $y$-axis. It meets the $y$-axis at a
unique point $P_y$.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-1.1,0) -- (5.0,0) node[right] {$x$};
    \draw[axis] (0,-1.1) -- (0,4.0) node[above] {$y$};
    \coordinate (P) at (3.2,2.4);
    \coordinate (Px) at (3.2,0);
    \coordinate (Py) at (0,2.4);
    \draw[helper] (P) -- (Px);
    \draw[helper] (P) -- (Py);
    \fill (P) circle (2.2pt) node[above right] {$P$};
    \fill (Px) circle (1.8pt) node[below] {$P_x$};
    \fill (Py) circle (1.8pt) node[left] {$P_y$};
\end{tikzpicture}
\end{center}

If the signed real labels of $P_x$ and $P_y$ are $x$ and $y$, respectively, we
assign to $P$ the ordered pair
\[
P\longleftrightarrow(x,y).
\]

The construction is unambiguous because the perpendicular through a given point
to a given line is unique. Conversely, given real numbers $x$ and $y$, locate
the corresponding points on the two axes and construct through them lines
perpendicular to their respective axes. Those two lines meet in exactly one
point of the plane.

\begin{theorembox}
Once the two perpendicular axes, their positive directions, and a common unit
are fixed, orthogonal projection gives a one-to-one correspondence
\[
\boxed{
\text{points of the Euclidean plane}
\longleftrightarrow
\text{pairs }(x,y)\in\mathbb R^2
}.
\]
The ordered pair is the numerical address of a geometric point in the chosen
coordinate system. It is not the point itself and it is not the source of the
geometry.
\end{theorembox}

\subsection{The Coordinate Formula for Euclidean Distance}

The Euclidean distance between two points is the length of the straight segment
joining them. Coordinates now allow this geometric length to be calculated.

For points
\[
P=(x_1,y_1),
\qquad
Q=(x_2,y_2),
\]
their projections show that the horizontal and vertical separations have
lengths
\[
|x_2-x_1|
\qquad\text{and}\qquad
|y_2-y_1|.
\]
Together with the segment $PQ$, these two perpendicular segments form a right
triangle. Here ``right'' refers to the Euclidean right angle already used in the
construction; no numerical measure of that angle is required.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-0.8,0) -- (5.2,0) node[right] {$x$};
    \draw[axis] (0,-0.8) -- (0,4.5) node[above] {$y$};
    \coordinate (P) at (1,1);
    \coordinate (Q) at (4,3.5);
    \coordinate (R) at (4,1);
    \draw[subset] (P) -- (Q);
    \draw[helper] (P) -- (R) node[midway,below] {$|x_2-x_1|$};
    \draw[helper] (R) -- (Q) node[midway,right] {$|y_2-y_1|$};
    \draw (3.72,1) -- (3.72,1.28) -- (4,1.28);
    \fill (P) circle (2pt) node[below left] {$P$};
    \fill (Q) circle (2pt) node[above right] {$Q$};
\end{tikzpicture}
\end{center}

By the Pythagorean theorem,
\[
d_E(P,Q)^2
=
(x_2-x_1)^2+(y_2-y_1)^2,
\]
and therefore
\[
\boxed{
d_E(P,Q)
=
\sqrt{(x_2-x_1)^2+(y_2-y_1)^2}
}.
\]

The formula does not create a new notion of distance. It is the numerical
coordinate expression of the Euclidean segment length already present in the
plane.

\subsection{Angles and the Calculation of Their Measure}

Right angles have already been used as Euclidean constructions: they are what
make the axes perpendicular and what make the triangle in the distance formula a
right triangle. We now make the general language of angles explicit and define
how to assign numerical measures to angles.

\begin{definitionbox}
Let $r_1$ and $r_2$ be two rays with the same initial point $O$. The rays,
together with a choice of one of the two regions between them, determine an
angle $\alpha$ with vertex $O$.
\end{definitionbox}

The angle is the chosen geometric region. It is neither the small arc drawn
near the vertex nor a number. The arc is only a graphical mark indicating
which region has been selected.

\begin{center}
\begin{tikzpicture}[scale=1.0]
    \coordinate (O) at (0,0);
    \coordinate (A) at (3.4,0);
    \coordinate (B) at (1.7,2.8);
    \fill[blue!8] (O) -- (A) arc[start angle=0,end angle=58.7,radius=3.4] -- cycle;
    \draw[axis] (O) -- (A) node[right] {$r_1$};
    \draw[axis] (O) -- (B) node[above] {$r_2$};
    \draw[subset] (0.9,0) arc[start angle=0,end angle=58.7,radius=0.9];
    \node[subset label] at (0.95,0.48) {$\alpha$};
    \fill (O) circle (2pt) node[below left] {$O$};
\end{tikzpicture}
\end{center}

\subsubsection*{Why circles give a consistent measure}

Draw a circle centred at $O$. The rays cut from the circle an arc corresponding
to the chosen angle.

\begin{theorembox}
For every Euclidean circle, the ratio of circumference $C$ to diameter $D$ is
the same constant:
\[
\frac{C}{D}=\pi.
\]
This constant is called $\pi$. Since $D=2r$, every circle of radius $r$ has
circumference
\[
\boxed{C=2\pi r}.
\]
\end{theorembox}

The constancy of $C/D$ follows from similarity: enlarging a circle by a factor
$\lambda$ multiplies every length, including its diameter, circumference, and
corresponding arc lengths, by $\lambda$.

If the same two rays meet two circles of radii $r_1$ and $r_2$ and cut arc
lengths $s_1$ and $s_2$, then
\[
\frac{s_1}{s_2}=\frac{r_1}{r_2},
\]
and therefore
\[
\frac{s_1}{r_1}=\frac{s_2}{r_2}.
\]
Thus the quotient $s/r$ depends only on the angle, not on the chosen circle.

\begin{definitionbox}
Let $s$ be the length of the selected arc on a circle of radius $r$ centred at
the vertex. The radian measure of $\alpha$ is
\[
\boxed{m_{\mathrm{rad}}(\alpha)=\frac{s}{r}}.
\]
\end{definitionbox}

For a full turn, $s=2\pi r$, so
\[
m_{\mathrm{rad}}(\text{full turn})=2\pi.
\]
A half-turn has radian measure $\pi$, and a quarter-turn has radian measure
$\pi/2$.

Degrees are a second numerical scale. A full turn is assigned $360^\circ$, so
\[
2\pi\text{ radians}=360^\circ
\]
and
\[
\boxed{
m_{\deg}(\alpha)
=
\frac{180^\circ}{\pi}\,m_{\mathrm{rad}}(\alpha)
}.
\]

\subsubsection*{The law of cosines is inherited from Euclidean geometry}

Directly measuring an arc is often difficult. We therefore use a theorem that
belongs to the Euclidean geometry with which we started.

Choose points
\[
A\in r_1\setminus\{O\},
\qquad
B\in r_2\setminus\{O\}.
\]
The points form the Euclidean triangle $OAB$. Before writing the formula, we
identify the angle and the three side lengths occurring in it.

\begin{center}
\begin{tikzpicture}[scale=0.92]
    \coordinate (O) at (0,0);
    \coordinate (A) at (4.4,0);
    \coordinate (B) at (1.55,3.0);

    \fill[blue!6] (O) -- (1.0,0)
        arc[start angle=0,end angle=62.7,radius=1.0] -- cycle;

    \draw[subset] (O) -- (A)
        node[midway,below=5pt] {$d_E(O,A)$};
    \draw[subset] (O) -- (B)
        node[midway,left=7pt] {$d_E(O,B)$};
    \draw[subset] (A) -- (B)
        node[midway,above right=3pt] {$d_E(A,B)$};

    \draw[subset] (0.78,0)
        arc[start angle=0,end angle=62.7,radius=0.78];
    \node[subset label] at (0.88,0.42) {$\alpha$};

    \fill (O) circle (2.1pt) node[below left] {$O$};
    \fill (A) circle (2.1pt) node[below right] {$A$};
    \fill (B) circle (2.1pt) node[above] {$B$};

    \node[align=left,anchor=west] at (5.0,2.35) {
        adjacent to $\alpha$:\\
        $d_E(O,A)$ and $d_E(O,B)$
    };
    \node[align=left,anchor=west] at (5.0,0.65) {
        opposite $\alpha$:\\
        $d_E(A,B)$
    };
\end{tikzpicture}
\end{center}

The side $AB$ is opposite the angle $\alpha$, while $OA$ and $OB$ are adjacent
to it. With these symbols fixed by the diagram, the classical law of cosines,
proved geometrically without Cartesian coordinates, states that
\[
d_E(A,B)^2
=
d_E(O,A)^2+d_E(O,B)^2
-2d_E(O,A)d_E(O,B)
\cos\bigl(m_{\mathrm{rad}}(\alpha)\bigr).
\]

\begin{remarkbox}
The law of cosines is not created or justified by the coordinate system in this
lecture. It is an earlier theorem of Euclidean geometry. Coordinates enter only
when we calculate the three side lengths
\[
d_E(O,A),\qquad d_E(O,B),\qquad d_E(A,B)
\]
from the coordinates of $O$, $A$, and $B$.
\end{remarkbox}

Solving the Euclidean law of cosines for the angle measure gives
\begin{theorembox}
\[
\boxed{
m_{\mathrm{rad}}(\alpha)
=
\arccos\left(
\frac{
d_E(O,A)^2+d_E(O,B)^2-d_E(A,B)^2
}{
2d_E(O,A)d_E(O,B)
}
\right)
}.
\]
The distance formula supplies the three numerical lengths needed in this
expression.
\end{theorembox}

The logical direction is therefore
\[
\begin{aligned}
&\text{Euclidean geometry and the law of cosines}\\
&\qquad\longrightarrow\text{Cartesian coordinates}\\
&\qquad\longrightarrow\text{numerical side lengths}\\
&\qquad\longrightarrow\text{calculation of the angle measure}.
\end{aligned}
\]

\begin{examplebox}
Let
\[
O=(0,0),\qquad A=(1,0),\qquad B=(1,1).
\]
The coordinate distance formula gives
\[
d_E(O,A)=1,
\qquad
d_E(O,B)=\sqrt2,
\qquad
d_E(A,B)=1.
\]
The Euclidean law of cosines then gives
\[
\cos\bigl(m_{\mathrm{rad}}(\alpha)\bigr)
=
\frac{1^2+(\sqrt2)^2-1^2}{2\cdot1\cdot\sqrt2}
=
\frac{1}{\sqrt2}.
\]
Hence
\[
m_{\mathrm{rad}}(\alpha)=\frac{\pi}{4},
\qquad
m_{\deg}(\alpha)=45^\circ.
\]
The angle is the geometric region determined by the rays $OA$ and $OB$; the
numbers $\pi/4$ and $45^\circ$ are two measures of that angle.
\end{examplebox}

\begin{definitionbox}
A \emph{right angle} is the constructible angle formed by two perpendicular
rays. Equivalently, two perpendicular lines meet so that the four angles around
their intersection are congruent; each of these angles is called a right angle.
\end{definitionbox}

This name was already used earlier in the construction of the axes and in the
Pythagorean triangle for the distance formula. The new information obtained now
is numerical: if $\alpha$ is a right angle, then
\[
m_{\mathrm{rad}}(\alpha)=\frac{\pi}{2},
\qquad
m_{\deg}(\alpha)=90^\circ.
\]
A right angle is not either of these numbers; it is a geometric angle having
these measures in the two measurement systems.

\begin{summarybox}
The construction proceeds in the following order:
\begin{enumerate}
    \item begin with a Euclidean plane containing no chosen coordinates;
    \item construct one straight line;
    \item choose a point $O$ on that line;
    \item construct through $O$ the perpendicular line, and only then name the
          two lines as coordinate axes and $O$ as their origin;
    \item choose positive directions on both axes;
    \item choose one common unit length and transfer it with a compass in both
          positive and negative directions to construct the integer marks;
    \item repeatedly bisect intervals to construct the dense family of dyadic
          marks;
    \item use Euclidean continuity to identify every axis position with a real
          number;
    \item assign to every point its unique pair of orthogonal projections;
    \item derive the coordinate distance formula from the resulting right
          triangle and the Pythagorean theorem;
    \item only after this construction introduce the numerical measurement of
          general angles.
\end{enumerate}
\end{summarybox}