\subsection{From construction to recognition}

The geometric definitions above answer the question \emph{what the curves are}.
A different problem begins when an equation is given first:

\begin{center}
\emph{Which geometric curve, if any, is hidden in this algebraic condition?}
\end{center}

When no mixed term is present, an equation of the form
\[
Ax^2+Cy^2+Dx+Ey+F=0
\]
can be analysed by completing squares. The signs of the squared terms then have a
direct geometric meaning:
\[
\begin{array}{c|c}
\text{same nonzero sign}&\text{ellipse type or a degenerate case},\\
\text{opposite signs}&\text{hyperbola type or intersecting lines},\\
\text{one quadratic direction absent}&\text{parabola type or a degenerate case}.
\end{array}
\]
This reasoning is already available with ordinary algebra.

The mixed term $Bxy$ indicates that the chosen coordinate axes may not follow the
natural symmetry directions of the curve. Rotating the axes can remove that term.
The quantity
\[
B^2-4AC
\]
packages the sign information in a way that survives such rotations. We will use
its sign as a recognition criterion, but its full structural explanation belongs
to the later study of matrices and quadratic forms.

\begin{remarkbox}
There is no contradiction in using a theorem before giving its deepest possible
explanation, provided its status is stated honestly. Here the geometric curves
have already been constructed; the discriminant is introduced as an efficient
recognition tool whose conceptual compression will be revisited later.
\end{remarkbox}
