\section{Reading Shape, Motion, and Orbit Geometry}

The same conic can now be described in three complementary ways:

\[
\begin{array}{c|c}
\text{Cartesian equation} & \text{tests points and exposes axes},\\
\text{parameterisation} & \text{moves a point along the curve},\\
\text{polar equation from a focus} & \text{exposes focal distance and eccentricity}.
\end{array}
\]

The purpose of this section is not to classify every possible quadratic equation.
It is to learn how to move between these descriptions and how to read geometry
from each one.

\subsection{A visual dictionary of the standard forms}

\begin{center}
\begin{tikzpicture}[scale=0.63,>=stealth]
    \begin{scope}[xshift=-6.6cm,yshift=3.2cm]
        \draw[axis,->] (-2.4,0)--(2.6,0);
        \draw[axis,->] (0,-2.2)--(0,2.4);
        \draw[subset] (0,0) circle (1.55);
        \node at (0,-2.8) {$x^2+y^2=R^2$};
    \end{scope}
    \begin{scope}[xshift=0cm,yshift=3.2cm]
        \draw[axis,->] (-2.8,0)--(3.0,0);
        \draw[axis,->] (0,-2.2)--(0,2.4);
        \draw[subset] (0,0) ellipse (2.15 and 1.25);
        \node at (0,-2.8) {$\frac{x^2}{a^2}+\frac{y^2}{b^2}=1$};
    \end{scope}
    \begin{scope}[xshift=6.6cm,yshift=3.2cm]
        \draw[axis,->] (-1.4,0)--(3.6,0);
        \draw[axis,->] (0,-2.2)--(0,2.4);
        \draw[subset,domain=-2:2,samples=100] plot ({0.55*\x*\x},{\x});
        \node at (0.9,-2.8) {$y^2=4px$};
    \end{scope}
    \begin{scope}[yshift=-3.2cm]
        \draw[axis,->] (-4.1,0)--(4.3,0);
        \draw[axis,->] (0,-2.7)--(0,2.9);
        \draw[helper,dashed] (-3.8,-2.2)--(3.8,2.2);
        \draw[helper,dashed] (-3.8,2.2)--(3.8,-2.2);
        \draw[subset,domain=1.25:3.7,samples=100]
            plot (\x,{0.85*sqrt(\x*\x-1)});
        \draw[subset,domain=1.25:3.7,samples=100]
            plot (\x,{-0.85*sqrt(\x*\x-1)});
        \draw[subset,domain=-3.7:-1.25,samples=100]
            plot (\x,{0.85*sqrt(\x*\x-1)});
        \draw[subset,domain=-3.7:-1.25,samples=100]
            plot (\x,{-0.85*sqrt(\x*\x-1)});
        \node at (0,-3.35) {$\frac{x^2}{a^2}-\frac{y^2}{b^2}=1$};
    \end{scope}
\end{tikzpicture}
\end{center}

\begin{corebox}[Fast recognition without losing geometry]
After completing squares and moving constants, equal signs on two squared
coordinates indicate ellipse-type geometry; opposite signs indicate a
hyperbola; one squared coordinate and one linear coordinate indicate a
parabola. The equation should then be rewritten in a standard form from which
centre, vertex, axes, or asymptotes can be read.
\end{corebox}

\subsection{Completing squares as geometric reconstruction}

Consider
\[
x^2+4y^2-6x+8y-3=0.
\]
Group the coordinate terms:
\[
(x^2-6x)+4(y^2+2y)=3.
\]
Complete squares:
\[
(x-3)^2-9+4\bigl((y+1)^2-1\bigr)=3.
\]
Therefore
\[
(x-3)^2+4(y+1)^2=16,
\]
and hence
\[
\boxed{\frac{(x-3)^2}{16}+\frac{(y+1)^2}{4}=1}.
\]

\begin{center}
\begin{tikzpicture}[scale=0.78,>=stealth]
    \draw[axis,->] (-2.0,0)--(8.5,0) node[right] {$x$};
    \draw[axis,->] (0,-4.2)--(0,3.2) node[above] {$y$};
    \draw[gray!35,step=1] (-1.5,-3.8) grid (8.0,2.8);
    \coordinate (C) at (3,-1);
    \draw[subset] (C) ellipse (4 and 2);
    \draw[helper] (C)--++(4,0) node[midway,below] {$a=4$};
    \draw[helper] (C)--++(0,2) node[midway,left] {$b=2$};
    \fill[geometricSubsetColor] (C) circle (2.2pt)
        node[below left] {$(3,-1)$};
\end{tikzpicture}
\end{center}

\begin{interpretationbox}[What the algebra did]
Completing squares did not create a new curve. It translated the coordinate
language until the centre and semi-axes became visible.
\end{interpretationbox}

\subsection{From Cartesian ellipse to eccentricity}

For
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}=1,
\qquad a\geq b,
\]
define
\[
c=\sqrt{a^2-b^2}.
\]
The eccentricity is
\[
\boxed{e=\frac ca=\sqrt{1-\frac{b^2}{a^2}}}.
\]
Therefore
\[
\boxed{b^2=a^2(1-e^2)}.
\]
The focal parameter in the polar equation is
\[
\boxed{p=a(1-e^2)=\frac{b^2}{a}}.
\]

Substituting into the polar equation gives
\[
\boxed{r(\theta)=\frac{a(1-e^2)}{1+e\cos\theta}}.
\]
This is the same ellipse, now viewed from one focus instead of from its centre.

\begin{center}
\begin{tikzpicture}[scale=0.84,>=stealth]
    \draw[axis,->] (-5.0,0)--(5.4,0) node[right] {major axis};
    \draw[axis,->] (0,-3.0)--(0,3.2);
    \draw[subset] (0,0) ellipse (4.0 and 2.25);
    \coordinate (F) at (-3.31,0);
    \coordinate (C) at (0,0);
    \coordinate (P) at (2.0,1.95);
    \draw[blue!70!black,line width=1.1pt] (F)--(P)
        node[midway,above left] {$r$};
    \draw[orange!85!black,line width=1pt] (C)--(P)
        node[midway,below right] {centre-based view};
    \fill[geometricSubsetColor] (F) circle (2.2pt) node[below] {focus};
    \fill[geometricSubsetColor] (C) circle (2.2pt) node[below right] {centre};
    \fill[geometricSubsetColor] (P) circle (2.2pt) node[above right] {$P$};
\end{tikzpicture}
\end{center}

\begin{interpretationbox}[Two natural viewpoints]
The centre-based equation is symmetric and efficient for drawing the ellipse.
The focus-based polar equation is efficient for dynamics, because a force
source located at the focus acts along the radial direction.
\end{interpretationbox}

\subsection{Nearest and farthest distance on an ellipse}

For
\[
r(\theta)=\frac{p}{1+e\cos\theta},
\qquad 0<e<1,
\]
the nearest point occurs at $\theta=0$:
\[
\boxed{r_{\min}=\frac{p}{1+e}=a(1-e)}.
\]
The farthest point occurs at $\theta=\pi$:
\[
\boxed{r_{\max}=\frac{p}{1-e}=a(1+e)}.
\]

Thus
\[
\boxed{a=\frac{r_{\min}+r_{\max}}2},
\qquad
e=\frac{r_{\max}-r_{\min}}{r_{\max}+r_{\min}}.
\]
These formulas reconstruct the size and eccentricity of an ellipse directly
from its extreme focal distances.

\begin{center}
\begin{tikzpicture}[scale=0.84,>=stealth]
    \draw[axis,->] (-4.8,0)--(5.1,0);
    \draw[axis,->] (0,-2.8)--(0,3.0);
    \draw[subset] (0.6,0) ellipse (3.7 and 2.0);
    \coordinate (F) at (-2.45,0);
    \coordinate (Near) at (-3.1,0);
    \coordinate (Far) at (4.3,0);
    \draw[blue!70!black,line width=1.1pt,<->] (F)--(Near)
        node[midway,below] {$a(1-e)$};
    \draw[orange!85!black,line width=1.1pt,<->] (F)--(Far)
        node[midway,above] {$a(1+e)$};
    \fill[geometricSubsetColor] (F) circle (2.2pt) node[below] {$F$};
\end{tikzpicture}
\end{center}

\subsection{How eccentricity changes the orbit}

Fix the focal parameter $p$ and vary $e$ in
\[
r=\frac{p}{1+e\cos\theta}.
\]

\begin{center}
\begin{tikzpicture}[scale=0.78,>=stealth]
    \draw[axis,->] (-5.0,0)--(5.4,0) node[right] {polar axis};
    \draw[axis,->] (0,-3.8)--(0,4.0);
    \draw[blue!70!black,line width=1.1pt]
        plot[domain=0:360,samples=180]
        ({2.2*cos(\x)/(1+0.25*cos(\x))},{2.2*sin(\x)/(1+0.25*cos(\x))});
    \draw[orange!85!black,line width=1.1pt]
        plot[domain=0:360,samples=180]
        ({2.2*cos(\x)/(1+0.65*cos(\x))},{2.2*sin(\x)/(1+0.65*cos(\x))});
    \fill[geometricSubsetColor] (0,0) circle (2.2pt) node[below left] {focus};
    \node[blue!70!black] at (-3.3,2.7) {$e=0.25$};
    \node[orange!85!black] at (3.5,3.1) {$e=0.65$};
\end{tikzpicture}
\end{center}

For $e$ close to $0$, the ellipse is nearly circular. Increasing $e$ makes the
nearest and farthest distances more unequal. At $e=1$, the farthest distance
ceases to be finite and the curve becomes a parabola. For $e>1$, only a range of
angles gives positive $r$, producing a hyperbolic branch.

\subsection{The angle shift and orientation of the orbit}

The polar equation
\[
r=\frac{p}{1+e\cos\theta}
\]
places the nearest point on the positive polar axis. A rotated conic is described
by
\[
\boxed{r=\frac{p}{1+e\cos(\theta-\theta_0)}}.
\]
The parameter $\theta_0$ rotates the entire curve without changing its shape.

\begin{center}
\begin{tikzpicture}[scale=0.78,>=stealth]
    \draw[axis,->] (-4.6,0)--(4.9,0) node[right] {$x$};
    \draw[axis,->] (0,-3.7)--(0,3.9) node[above] {$y$};
    \draw[subset,rotate=28] (0.6,0) ellipse (3.4 and 1.8);
    \draw[blue!70!black,line width=1.1pt,->] (0,0)--(3.2,1.7)
        node[midway,above] {$\theta_0$ direction};
    \fill[geometricSubsetColor] (0,0) circle (2.2pt) node[below left] {focus};
\end{tikzpicture}
\end{center}

\begin{interpretationbox}[Shape versus orientation]
The parameters $p$ and $e$ determine size and shape. The angle $\theta_0$
determines orientation. This separation is exactly what is useful when an orbit
can point in any direction in the plane.
\end{interpretationbox}

\subsection{Why this prepares mechanics}

In a central-force problem, the force always points toward or away from one
fixed point. Placing the polar origin at that point makes the radial distance
$r$ and polar angle $\theta$ the natural variables.

The equation
\[
\boxed{r(\theta)=\frac{p}{1+e\cos(\theta-\theta_0)}}
\]
then describes the possible geometric shapes of trajectories:
\[
\begin{array}{c|c|c}
e & \text{curve} & \text{qualitative motion}\\
\hline
0\leq e<1 & \text{circle or ellipse} & \text{bounded trajectory},\\
e=1 & \text{parabola} & \text{threshold escape},\\
e>1 & \text{hyperbola} & \text{unbounded passage}.
\end{array}
\]

\begin{corebox}[The mathematical preparation]
This chapter does not derive the orbit equation from Newton's laws. It builds
the geometric language needed for that derivation: polar coordinates, radial
distance, eccentricity, orientation, and the common conic equation.
\end{corebox}

\subsection{What remains of general quadratic classification}

A general second-degree equation
\[
Ax^2+Bxy+Cy^2+Dx+Ey+F=0
\]
may describe a conic, a pair of lines, a point, or no real points. For the
present course, the essential practical procedure is:

\begin{enumerate}
    \item collect the $x$ and $y$ terms;
    \item complete squares when there is no mixed term;
    \item move constants and normalise;
    \item read the geometric data from the resulting standard form;
    \item check separately whether the equation is degenerate.
\end{enumerate}

A mixed term $Bxy$ signals that the natural axes may be rotated relative to the
chosen Cartesian axes. A full systematic treatment belongs with quadratic
forms and diagonalisation; it is not needed for the focal and polar geometry
developed here.

\begin{summarybox}[Three descriptions, one curve]
Cartesian equations reveal centres, vertices, axes, and asymptotes.
Parameterisations describe traversal. Polar equations centred at a focus reveal
radial distance, eccentricity, and orientation. Analytic geometry is the art of
choosing the description that makes the geometric question simplest.
\end{summarybox}