\subsection{Linear Combinations: What Addition and Scaling Can Build}

Addition and scalar multiplication were introduced as two separate operations,
but they are most useful when they are combined.

\begin{definitionbox}
Given vectors $\mathbf u$ and $\mathbf v$ and real numbers $a$ and $b$, the
vector
\[
\boxed{a\mathbf u+b\mathbf v}
\]
is called a \emph{linear combination} of $\mathbf u$ and $\mathbf v$.
\end{definitionbox}

\begin{interpretationbox}[How a linear combination is constructed]
First scale $\mathbf u$ by $a$ and $\mathbf v$ by $b$; then add the resulting
displacements. A linear combination is therefore not a new operation. It is a
controlled use of addition and scalar multiplication.
\end{interpretationbox}

\begin{center}
\begin{tikzpicture}[scale=0.9,>=stealth]
    \coordinate (O) at (0,0);
    \coordinate (AU) at (3.4,0.9);
    \coordinate (BV) at (1.0,2.7);
    \coordinate (S) at (4.4,3.6);

    \draw[blue!70!black,line width=1.1pt,->] (O) -- (AU)
        node[midway,below] {$a\mathbf u$};
    \draw[orange!85!black,line width=1.1pt,->] (AU) -- (S)
        node[midway,right] {$b\mathbf v$};
    \draw[very thick,->] (O) -- (S)
        node[midway,above left] {$a\mathbf u+b\mathbf v$};
    \draw[helper,dashed] (O) -- (BV);
    \draw[helper,dashed] (BV) -- (S);
\end{tikzpicture}
\end{center}

\begin{corebox}[One direction versus two directions]
A single nonzero vector generates only the line
\[
\{\lambda\mathbf v:\lambda\in\mathbb R\}.
\]
Two parallel vectors still generate only one line. Two nonparallel vectors
provide two independent directions and can generate the entire plane.
\end{corebox}

\begin{proofbox}[Why nonparallel directions reach the plane]
If $\mathbf u$ and $\mathbf v$ are parallel, every scalar multiple of either
vector lies on the same line, so every sum $a\mathbf u+b\mathbf v$ remains on
that line. If they are not parallel, the parallelogram construction allows a
target displacement to be decomposed into one component along each direction.
The two coefficients record those two signed motions.
\end{proofbox}

\begin{theorembox}
Two nonparallel vectors $\mathbf u$ and $\mathbf v$ determine every vector in
the plane uniquely in the form
\[
\boxed{\mathbf w=a\mathbf u+b\mathbf v}.
\]
The numbers $a$ and $b$ record how much motion occurs in each chosen direction.
\end{theorembox}

\begin{interpretationbox}[Bridge to a basis]
A basis is not an unrelated new object. It is a pair of independent directions
whose linear combinations describe every vector uniquely. Basis coordinates are
precisely the coefficients in that reconstruction.
\end{interpretationbox}
