\subsection{Stretching, Shrinking, and Reversing a Displacement}

Addition combines different displacements. A second operation appears when the
same displacement is repeated.

\begin{corebox}[The geometric idea]
Scalar multiplication changes how much of one displacement is taken. It may
stretch, shrink, erase, or reverse a vector, but it does not introduce a new
line of direction.
\end{corebox}

\begin{derivationbox}[From repetition to continuous scaling]
If
\[
\mathbf v=(v_1,v_2),
\]
then performing it twice gives
\[
\mathbf v+\mathbf v=(2v_1,2v_2),
\]
and performing it three times gives
\[
3\mathbf v=(3v_1,3v_2).
\]
Both components must be multiplied by the same factor if the direction is to
remain unchanged. Extending repetition from integers to arbitrary real factors
leads to the general operation.
\end{derivationbox}

\begin{definitionbox}
For a real number $\lambda$ and a vector $\mathbf v=(v_1,v_2)$,
\[
\boxed{
\lambda\mathbf v=(\lambda v_1,\lambda v_2)
}.
\]
The number $\lambda$ is called a \emph{scalar}.
\end{definitionbox}

\begin{center}
\begin{tikzpicture}[scale=0.9,>=stealth]
    \coordinate (O) at (0,0);
    \coordinate (Vhalf) at (1.2,0.7);
    \coordinate (V) at (2.4,1.4);
    \coordinate (TwoV) at (4.8,2.8);
    \coordinate (MinusV) at (-2.4,-1.4);

    \draw[green!55!black,line width=1pt,->] (O) -- (Vhalf)
        node[midway,above left] {$\tfrac12\mathbf v$};
    \draw[blue!70!black,line width=1.1pt,->] (O) -- (V)
        node[midway,below right] {$\mathbf v$};
    \draw[orange!85!black,line width=1.1pt,->] (O) -- (TwoV)
        node[midway,above left] {$2\mathbf v$};
    \draw[violet!75,line width=1.1pt,->] (O) -- (MinusV)
        node[midway,above left] {$-\mathbf v$};

    \draw[gray,dashed] (-3.0,-1.75) -- (5.4,3.15);
    \fill (O) circle (2pt) node[below right] {$O$};
\end{tikzpicture}
\end{center}

\begin{interpretationbox}[How the scalar changes the vector]
\begin{itemize}
    \item if $\lambda>1$, the vector is stretched;
    \item if $0<\lambda<1$, it is shortened;
    \item if $\lambda=0$, it collapses to the zero vector;
    \item if $\lambda<0$, it is also reversed.
\end{itemize}
The scalar changes the amount and possibly the orientation of the displacement,
not the underlying line on which it lies.
\end{interpretationbox}

\begin{theorembox}
Two nonzero vectors $\mathbf u$ and $\mathbf v$ are parallel if and only if one
is a scalar multiple of the other:
\[
\boxed{
\mathbf u\parallel\mathbf v
\iff
\mathbf u=\lambda\mathbf v
\text{ for some }\lambda\in\mathbb R\setminus\{0\}
}.
\]
\end{theorembox}

\begin{remarkbox}
Fractions and irrational factors are natural extensions of repetition. The
vector $\tfrac12\mathbf v$ must be repeated twice to produce $\mathbf v$, while
$\sqrt2\,\mathbf v$ has the same direction and $\sqrt2$ times the length.
Scalar multiplication turns discrete repetition into continuous scaling.
\end{remarkbox}
