\section{Vectors from Coordinate Displacements}

A coordinate pair tells us where a point is. It does not yet tell us how one
point is related to another. To describe motion from one point to another we
need a new object: not another position, but a record of the change of position.

Choose
\[
A=(1,2),
\qquad
B=(4,6).
\]
To travel from $A$ to $B$, the horizontal coordinate must increase by $3$ and
the vertical coordinate by $4$. The pair
\[
(3,4)
\]
therefore records an instruction:

\begin{center}
move three units in the positive $x$ direction and four units in the positive
$y$ direction.
\end{center}

This instruction can be carried out from $A$, but it can also be carried out
from any other point. The starting point may change while the displacement
remains the same. That distinction is the idea from which vectors arise.

\noindent\textbf{What a vector is.} A point records a position. A vector records a displacement: a change of
position that can be translated to a different starting point without changing
the displacement itself. Vector operations describe how such displacements are
combined, reversed, and rescaled.

The development below follows a fixed question-driven order. We first decide
when two arrows represent the same displacement. We then ask how successive
displacements combine, what it means to repeat or reverse one, what addition and
scaling can build, how to measure length and direction, and how two
displacements can be compared geometrically. The coordinate formulas will be
consequences of those questions, not isolated rules introduced for memorisation.

\input{chapters/chapter_02/vectors_from_displacements/vector_as_coordinate_change}
\input{chapters/chapter_02/vectors_from_displacements/vector_addition}
\input{chapters/chapter_02/vectors_from_displacements/scalar_multiplication}
\input{chapters/chapter_02/vectors_from_displacements/linear_combinations}
\input{chapters/chapter_02/vectors_from_displacements/length_and_inclination}
\input{chapters/chapter_02/vectors_from_displacements/parallelism_and_perpendicularity}
\input{chapters/chapter_02/vectors_from_displacements/dot_product_properties}
\input{chapters/chapter_02/vectors_from_displacements/angle_between_vectors}
\input{chapters/chapter_02/vectors_from_displacements/cosine_theorem}
\input{chapters/chapter_02/vectors_from_displacements/projection}
\input{chapters/chapter_02/vectors_from_displacements/orthogonal_decomposition}
\input{chapters/chapter_02/vectors_from_displacements/three_dimensions}

\noindent\textbf{Vector toolkit.} A vector is a displacement, not a point. The essential operations are addition
and scalar multiplication; together they form linear combinations. Length and
the dot product encode metric information, while parallelism, perpendicularity,
angles, projections, and orthogonal decompositions translate geometric
questions into coordinate calculations. The same ideas extend from the plane to
three-dimensional space.
