\subsection{From Cartesian Points to Free Vectors}

A point of the Cartesian plane is an ordered pair of real numbers. Let
\[
P=(p_1,p_2),
\qquad
Q=(q_1,q_2).
\]
Before vectors are introduced, we first define several arithmetic operations on
such points. These definitions are possible because a point of the Cartesian
plane is represented by an ordered pair of numbers.

The point
\[
O=(0,0)
\]
plays the role of zero. Addition of points is defined coordinatewise:
\[
\boxed{
P+Q=(p_1+q_1,\,p_2+q_2)
}.
\]
The point opposite to $P$ is
\[
\boxed{
-P=(-p_1,-p_2)
},
\]
so that
\[
P+(-P)=O.
\]
Subtraction of points is defined by adding the opposite point:
\[
\boxed{
Q-P=Q+(-P)=(q_1-p_1,\,q_2-p_2)
}.
\]
For a real number $\lambda$, multiplication of a point by a number is also
defined coordinatewise:
\[
\boxed{
\lambda P=(\lambda p_1,\,\lambda p_2)
}.
\]
At this stage these are arithmetic operations on ordered pairs. They are useful
computational rules, but their geometric meaning has not yet been established.
The first expression to acquire such a meaning will be the difference $Q-P$.
It will describe the directed change from the point $P$ to the point $Q$.

\subsubsection*{An ordered pair of points determines an anchored vector}

The ordered pair $(P,Q)$ determines an arrow beginning at $P$ and ending at
$Q$. We denote it by
\[
\overrightarrow{PQ}
\]
and call it an \emph{anchored vector}. The order matters: $P$ is its initial
point and $Q$ is its endpoint.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-0.5,0) -- (7.0,0) node[right] {$x$};
    \draw[axis] (0,-0.5) -- (0,5.2) node[above] {$y$};

    \coordinate (P) at (1.2,1.1);
    \coordinate (Q) at (5.2,4.0);
    \coordinate (R) at (5.2,1.1);

    \draw[helper,dashed] (P) -- (R);
    \draw[helper,dashed] (R) -- (Q);
    \draw[subset,->] (P) -- (Q)
        node[pos=0.46,above left=3pt] {$\overrightarrow{PQ}$};

    \fill[geometricSubsetColor] (P) circle (2pt)
        node[below left=4pt] {$P=(p_1,p_2)$};
    \fill[geometricSubsetColor] (Q) circle (2pt)
        node[above right=7pt] {$Q=(q_1,q_2)$};

    \node[below=5pt] at (3.2,1.1) {$q_1-p_1$};
    \node[left=5pt,rotate=90] at (5.2,2.55) {$q_2-p_2$};
\end{tikzpicture}
\end{center}

The coordinates of the anchored vector are the coordinate differences between
its endpoint and its initial point:
\[
\boxed{
[\overrightarrow{PQ}]
=
Q-P
=
(q_1-p_1,\,q_2-p_2)
}.
\]
Its length is the distance between the two points:
\[
\boxed{
|\overrightarrow{PQ}|=\varrho(P,Q)
}.
\]
Reversing the order reverses the arrow and changes the signs of both
coordinates:
\[
[\overrightarrow{QP}]
=P-Q
=-(Q-P).
\]

\subsubsection*{When do arrows drawn in different places represent the same vector?}

We first define equality geometrically. Consider two anchored vectors
\[
\overrightarrow{PQ}
\qquad\text{and}\qquad
\overrightarrow{P'Q'}.
\]
They are called equal when the midpoint of the pair $P,Q'$ is also the midpoint
of the pair $Q,P'$. Equivalently, the four points form a parallelogram in which
the two arrows are opposite directed sides.

\begin{center}
\begin{tikzpicture}[scale=0.95]
    \coordinate (P) at (0.4,0.5);
    \coordinate (Q) at (3.5,1.7);
    \coordinate (Pp) at (1.7,3.0);
    \coordinate (Qp) at (4.8,4.2);
    \coordinate (M) at (2.6,2.35);

    \draw[helper] (P) -- (Pp) -- (Qp) -- (Q) -- cycle;
    \draw[helper,dashed] (P) -- (Qp);
    \draw[helper,dashed] (Q) -- (Pp);

    \draw[subset,->] (P) -- (Q)
        node[midway,below] {$\overrightarrow{PQ}$};
    \draw[subset,->] (Pp) -- (Qp)
        node[midway,above] {$\overrightarrow{P'Q'}$};

    \fill (P) circle (2pt) node[below left] {$P$};
    \fill (Q) circle (2pt) node[below right] {$Q$};
    \fill (Pp) circle (2pt) node[above left] {$P'$};
    \fill (Qp) circle (2pt) node[above right] {$Q'$};
    \fill[geometricSubsetColor] (M) circle (2.2pt)
        node[right] {common midpoint};
\end{tikzpicture}
\end{center}

The midpoint condition is
\[
\frac12(P+Q')=\frac12(Q+P').
\]
Rearranging this equality gives
\[
Q-P=Q'-P'.
\]
We have therefore obtained a coordinate criterion for the geometric relation:
\[
\boxed{
\overrightarrow{PQ}=\overrightarrow{P'Q'}
\quad\Longleftrightarrow\quad
Q-P=Q'-P'
}.
\]
Thus two anchored vectors are equal precisely when they produce the same change
in every coordinate. Their initial points may be different.

\begin{center}
\begin{tikzpicture}[scale=0.85]
    \draw[axis] (-0.5,0) -- (8.0,0) node[right] {$x$};
    \draw[axis] (0,-0.5) -- (0,5.5) node[above] {$y$};

    \coordinate (A) at (0.8,0.8);
    \coordinate (B) at (3.4,2.3);
    \coordinate (C) at (2.6,0.6);
    \coordinate (D) at (5.2,2.1);
    \coordinate (E) at (4.5,2.8);
    \coordinate (F) at (7.1,4.3);

    \draw[subset,->] (A) -- (B) node[midway,above left] {$(2.6,1.5)$};
    \draw[subset,->] (C) -- (D) node[midway,below right] {$(2.6,1.5)$};
    \draw[subset,->] (E) -- (F) node[midway,above left] {$(2.6,1.5)$};

    \fill (A) circle (1.8pt);
    \fill (B) circle (1.8pt);
    \fill (C) circle (1.8pt);
    \fill (D) circle (1.8pt);
    \fill (E) circle (1.8pt);
    \fill (F) circle (1.8pt);
\end{tikzpicture}
\end{center}

\subsubsection*{Free vectors}

Equality divides all anchored vectors into disjoint classes. Each class contains
all arrows representing one and the same directed change. Such a class is called
a \emph{free vector}.

If $\overrightarrow{PQ}$ is one representative of a free vector $\mathbf v$, we
may write
\[
\mathbf v=[\overrightarrow{PQ}].
\]
All representatives have the same coordinate differences. Consequently a free
vector is completely described by one coordinate pair:
\[
\boxed{
\mathbf v=[v_1,v_2]
}.
\]
Every anchored vector satisfying
\[
Q-P=(v_1,v_2)
\]
is a representative of this same free vector.

A particularly convenient representative begins at the origin. If
\[
O=(0,0),
\qquad
V=(v_1,v_2),
\]
then
\[
[\overrightarrow{OV}]=V-O=(v_1,v_2).
\]
Thus the point $V$ and the free vector $\mathbf v$ use the same numbers, but they
are not the same type of object: $V$ is a point, whereas $\mathbf v$ is a class
of equal anchored arrows.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \draw[axis] (-0.5,0) -- (7.0,0) node[right] {$x$};
    \draw[axis] (0,-0.5) -- (0,5.2) node[above] {$y$};

    \coordinate (O) at (0,0);
    \coordinate (V) at (3.0,2.0);
    \coordinate (P) at (2.3,1.0);
    \coordinate (Q) at (5.3,3.0);

    \draw[subset,->] (O) -- (V)
        node[midway,above left] {$\overrightarrow{OV}$};
    \draw[subset,->] (P) -- (Q)
        node[midway,below right] {$\overrightarrow{PQ}$};

    \fill (O) circle (2pt) node[below left] {$O$};
    \fill[geometricSubsetColor] (V) circle (2pt)
        node[above left] {$V=(v_1,v_2)$};
    \fill (P) circle (1.8pt) node[below left] {$P$};
    \fill (Q) circle (1.8pt) node[above right] {$Q$};

    \node[align=left,anchor=west] at (4.0,0.55)
        {$V-O=Q-P=(v_1,v_2)$};
\end{tikzpicture}
\end{center}

Only now, after free vectors have been constructed, do we define operations on
them. Each operation will first be described using representatives and then
written in coordinates.
