\section{Real functions and basic examples}

In this course we will often study functions whose inputs and outputs are real numbers. Such functions are called real functions of a real variable. They are central in calculus because they allow us to describe quantities that depend on one real parameter.

A typical notation is
\[
f:D\to\mathbb{R},
\qquad
D\subseteq\mathbb{R}.
\]
This means that the inputs are taken from a set \(D\) of real numbers, and the outputs are real numbers.

\subsection*{Natural domains}

Sometimes the domain is given explicitly. Sometimes it is understood from the formula. When a formula is given and no domain is stated, we often look for the largest set of real inputs for which the formula makes sense. This is called the natural domain.

\begin{examplebox}
For
\[
f(x)=2x+1,
\]
the natural domain is \(\mathbb{R}\). For
\[
g(x)=\frac1x,
\]
the natural domain is \(\mathbb{R}\setminus\{0\}\). For
\[
h(x)=\sqrt{x},
\]
the natural domain over the real numbers is \([0,\infty)\).
\end{examplebox}

These examples show that the formula is not the whole story. We must also ask where the formula is meaningful.

\subsection*{Affine and linear functions}

A function of the form
\[
f(x)=mx+b
\]
is called an affine function. Its graph is a line. The number \(m\) is the slope, and \(b=f(0)\) is the vertical intercept.

In elementary school terminology, functions of the form \(mx+b\) are often called linear. In the strict language of linear algebra, however, a real function of one variable is linear only when
\[
f(x)=mx,
\]
so that \(f(0)=0\).

\begin{remarkbox}
This distinction matters because this book also studies matrices and linear algebra. We use \emph{affine} for \(f(x)=mx+b\), and reserve \emph{linear} in the strict sense for \(f(x)=mx\).
\end{remarkbox}

\begin{examplebox}
The function
\[
f(x)=3x-2
\]
is affine but not linear in the strict sense, because
\[
f(0)=-2\ne0.
\]
Its slope is \(3\), so the output increases by \(3\) whenever the input increases by \(1\).
\end{examplebox}

\subsection*{Quadratic, reciprocal, and square root functions}

A quadratic function has the form
\[
f(x)=ax^2+bx+c,
\qquad a\ne0.
\]
Its graph is a parabola. The reciprocal function \(1/x\) illustrates a domain exclusion and a vertical asymptote, while \(\sqrt{x}\) illustrates a function whose real domain begins at an endpoint.

\begin{examplebox}
For \(f(x)=x^2\),
\[
f(-2)=f(2)=4.
\]
For \(g(x)=1/x\), the value \(g(0)\) is not defined. For \(h(x)=\sqrt{x}\), negative real inputs are excluded.
\end{examplebox}

\subsection*{Angles and trigonometric functions}

Calculus formulas use angles measured in radians. For a circle of radius \(r\), an angle \(\theta\) measured in radians satisfies
\[
s=r\theta,
\]
where \(s\) is the corresponding arc length. In particular,
\[
2\pi\text{ radians}=360^\circ,
\qquad
\pi\text{ radians}=180^\circ.
\]

On the unit circle, the point determined by an angle \(x\) has coordinates
\[
(\cos x,\sin x).
\]
Both sine and cosine are defined on \(\mathbb{R}\), take values in \([-1,1]\), and satisfy
\[
\sin(x+2\pi)=\sin x,
\qquad
\cos(x+2\pi)=\cos x,
\]
and
\[
\sin^2x+\cos^2x=1.
\]

The tangent function is
\[
\tan x=\frac{\sin x}{\cos x}.
\]
It is defined only when \(\cos x\ne0\), so its natural domain is
\[
\mathbb{R}\setminus
\left\{
\frac{\pi}{2}+k\pi:k\in\mathbb{Z}
\right\}.
\]

\begin{remarkbox}
The standard limit
\[
\lim_{x\to0}\frac{\sin x}{x}=1
\]
and the usual derivative formulas for trigonometric functions have their standard form only when angles are measured in radians.
\end{remarkbox}

\subsection*{Exponential functions}

For \(a>0\), \(a\ne1\), the exponential function with base \(a\) is
\[
f(x)=a^x.
\]
Its domain is \(\mathbb{R}\), and its values are positive. The exponent laws include
\[
a^{x+y}=a^xa^y,
\qquad
a^{x-y}=\frac{a^x}{a^y},
\qquad
(a^x)^y=a^{xy}.
\]
If \(a>1\), the function is increasing. If \(0<a<1\), it is decreasing.

A distinguished base is
\[
e\approx2.71828.
\]
The natural exponential function \(e^x\) has a central role in calculus because its derivative is itself. It also appears in the standard limit
\[
\lim_{x\to0}(1+x)^{1/x}=e.
\]

\begin{examplebox}
The model
\[
N(t)=N_0e^{kt}
\]
describes exponential growth when \(k>0\) and exponential decay when \(k<0\). Its initial value is \(N(0)=N_0\).
\end{examplebox}

\subsection*{Logarithmic functions}

The logarithm with base \(a>0\), \(a\ne1\), is the inverse of the exponential function \(a^x\). The statement
\[
y=\log_a x
\]
means
\[
a^y=x.
\]
Because exponential values are positive, logarithms are defined only for \(x>0\).

The logarithm with base \(e\) is the natural logarithm \(\ln x\). The inverse relationships are
\[
\ln(e^x)=x
\]
for every real \(x\), and
\[
e^{\ln x}=x
\]
for \(x>0\). The basic laws include
\[
\ln(xy)=\ln x+\ln y,
\qquad
\ln\frac{x}{y}=\ln x-\ln y,
\]
and
\[
\ln(x^p)=p\ln x,
\]
whenever all expressions are defined.

\begin{remarkbox}
The condition \(x>0\) is essential for \(\ln x\). In antiderivatives, \(\ln|x|\) appears because its derivative is \(1/x\) on both intervals \((-\infty,0)\) and \((0,\infty)\).
\end{remarkbox}

\subsection*{Inverse trigonometric functions}

Trigonometric functions are periodic, so restricted domains are needed before inverse functions can be defined. The standard choices are
\[
\arcsin:[-1,1]\to
\left[-\frac{\pi}{2},\frac{\pi}{2}\right],
\]
\[
\arccos:[-1,1]\to[0,\pi],
\]
and
\[
\arctan:\mathbb{R}\to
\left(-\frac{\pi}{2},\frac{\pi}{2}\right).
\]
For example,
\[
\arcsin\frac12=\frac{\pi}{6},
\qquad
\arctan1=\frac{\pi}{4}.
\]

\subsection*{Piecewise functions}

A function may use different formulas on different parts of its domain. Such functions are called piecewise-defined functions.

\begin{examplebox}
Define
\[
f(x)=
\begin{cases}
x+1, & x<0,\\
x^2, & x\ge0.
\end{cases}
\]
Then \(f(-2)=-1\), while \(f(3)=9\). The input determines which rule applies.
\end{examplebox}

\begin{summarybox}
When studying a real function, ask:
\begin{itemize}
\item What is the domain?
\item Is the function algebraic, trigonometric, exponential, logarithmic, or piecewise?
\item If its graph is a line, is the function affine or linear in the strict sense?
\item Are angles measured in radians?
\item Are logarithmic arguments positive?
\item What restricted range defines an inverse trigonometric function?
\item Which rule applies to a piecewise-defined input?
\end{itemize}
\end{summarybox}

Real functions are the main objects of calculus. Their study begins with reading domains, formulas, inverse relationships, and graph structure carefully.
