\section{Review Exercises}

\subsection{Elementary computational exercises}

\begin{enumerate}
\item Compute
\[
\det\begin{pmatrix}3&-1\\2&4\end{pmatrix},
\qquad
\det\begin{pmatrix}5&10\\1&2\end{pmatrix}.
\]
Identify the visible row relation in the second matrix.

\item Determine all values of $t$ for which
\[
\det\begin{pmatrix}t&2\\3&t\end{pmatrix}=0.
\]

\item For
\[
A=\begin{pmatrix}1&2\\3&4\end{pmatrix},
\]
compute the determinant after swapping the rows, multiplying the first row by $5$, and replacing $R_2$ by $R_2-3R_1$.

\item Compute the determinants of the triangular matrices
\[
\begin{pmatrix}2&1&4\\0&-3&5\\0&0&6\end{pmatrix},
\qquad
\begin{pmatrix}1&0&0\\2&4&0\\-1&3&5\end{pmatrix}.
\]

\item Compute
\[
\det\begin{pmatrix}1&2&0\\-1&3&4\\2&0&5\end{pmatrix}
\]
by expanding along the first row.

\item Compute
\[
\det\begin{pmatrix}2&0&1\\0&3&0\\4&0&-1\end{pmatrix}
\]
by choosing the row or column with the most zeros.

\item Use one row operation to simplify and compute
\[
\det\begin{pmatrix}1&2&1\\2&5&3\\0&1&2\end{pmatrix}.
\]
Track the determinant correctly.

\item Decide without full expansion which determinants are zero:
\[
\det\begin{pmatrix}1&2\\2&4\end{pmatrix},
\quad
\det\begin{pmatrix}1&2&3\\2&4&6\\0&1&1\end{pmatrix},
\quad
\det\begin{pmatrix}1&0&2\\0&1&3\\0&0&4\end{pmatrix}.
\]

\item For
\[
A=\begin{pmatrix}1&2\\0&3\end{pmatrix},
\qquad
B=\begin{pmatrix}2&-1\\4&1\end{pmatrix},
\]
compute $\det A$, $\det B$, $AB$, and $\det(AB)$.

\item Let
\[
A=\begin{pmatrix}1&2\\2&4\end{pmatrix}.
\]
Compute $Ae_1$ and $Ae_2$, show that the outputs are parallel, and connect this observation with $\det A=0$.
\end{enumerate}

\subsection{Development and reasoning exercises}

\begin{enumerate}
\item Compute
\[
\det\begin{pmatrix}1&2&0\\-1&3&4\\2&0&5\end{pmatrix}
\]
first by the direct $3\times3$ formula and then by Laplace expansion. Compare the work and explain which structure should guide the choice.

\item Use row operations to compute
\[
\det\begin{pmatrix}1&2&1\\2&5&3\\3&7&5\end{pmatrix}.
\]
Record every operation and explain its effect on the determinant.

\item Compute
\[
\det\begin{pmatrix}
2&0&1&0\\
0&3&0&0\\
4&0&-1&2\\
0&0&0&5
\end{pmatrix}
\]
by exploiting zeros. Explain why a full expansion would obscure the structure.

\item Determine all values of $t$ for which
\[
\det\begin{pmatrix}
1&t&0\\
2&2&1\\
0&1&t
\end{pmatrix}=0.
\]
For every exceptional value, identify a relation among rows, columns, or outputs when possible.

\item Let
\[
A=\begin{pmatrix}2&1\\1&1\end{pmatrix}.
\]
Map the unit square by $A$, compute the area of the resulting parallelogram, and compare it with $|\det A|$. Repeat with a singular matrix and describe the collapse.

\item Prove directly from the $2\times2$ formula that swapping two rows changes the sign of the determinant and that adding a multiple of one row to the other leaves it unchanged.

\item A matrix has two proportional rows. Give a proof of $\det A=0$ using row operations and a second proof using the interpretation of restricted outputs.

\item Explain the identity
\[
\det(AB)=\det A\det B
\]
as successive area or volume scaling. Verify the interpretation with one explicit pair of $2\times2$ matrices.

\item Compare four strategies for a determinant: direct small formula, Laplace expansion, triangular structure, and row reduction. Give one example where each is the natural choice.

\item Design a complete pre-calculation inspection routine for determinants. It must check size, zeros, proportional rows or columns, triangular structure, useful row operations, and the possibility that a parameter creates exceptional cases.
\end{enumerate}
