\section{Acceleration}

\subsection{Why velocity is not the end of the description}

Velocity describes how position changes. But velocity itself may also change.

A car may move faster and faster. A body may slow down. A point moving along a circle may keep the same speed but continuously change direction. In each of these cases, velocity changes.

Acceleration describes the change of velocity with respect to time.

\subsection{Average acceleration}

Suppose the velocity at time \(t_1\) is \(\mathbf{v}(t_1)\), and the velocity at time \(t_2\) is \(\mathbf{v}(t_2)\). The change in velocity is

\[
\Delta \mathbf{v}
=
\mathbf{v}(t_2)-\mathbf{v}(t_1).
\]

The average acceleration over the interval is

\[
\mathbf{a}_{\mathrm{avg}}
=
\frac{\Delta \mathbf{v}}{\Delta t}
=
\frac{\mathbf{v}(t_2)-\mathbf{v}(t_1)}{t_2-t_1}.
\]

Average acceleration describes how velocity changes over a time interval.

\subsection{Instantaneous acceleration}

Instantaneous acceleration is the derivative of velocity with respect to time:

\[
\mathbf{a}(t)
=
\frac{d\mathbf{v}}{dt}.
\]

Since velocity is the derivative of position,

\[
\mathbf{v}(t)=\frac{d\mathbf{r}}{dt},
\]

we also have

\[
\mathbf{a}(t)
=
\frac{d^2\mathbf{r}}{dt^2}.
\]

Thus acceleration is the second derivative of position with respect to time.

In coordinates,

\[
\mathbf{a}(t)
=
\left(
\frac{d^2x}{dt^2},
\frac{d^2y}{dt^2},
\frac{d^2z}{dt^2}
\right).
\]

In two dimensions,

\[
\mathbf{a}(t)
=
\left(
\frac{d^2x}{dt^2},
\frac{d^2y}{dt^2}
\right).
\]

In one dimension,

\[
a(t)=\frac{d^2x}{dt^2}.
\]

\subsection{Acceleration is also a vector}

Acceleration is a vector. It has magnitude and direction. It describes how the velocity vector changes.

This point is important. Acceleration does not only mean that speed increases. Acceleration means that velocity changes. Since velocity is a vector, it can change in two ways:

\begin{itemize}
    \item its magnitude can change,
    \item its direction can change.
\end{itemize}

Therefore a body may be accelerating even if its speed is constant, provided that the direction of velocity changes.

This is one of the most common conceptual difficulties in kinematics. Constant speed does not necessarily mean zero acceleration.

\subsection{Example: motion along a line}

Consider

\[
x(t)=x_0+at.
\]

We already found that

\[
v(t)=a.
\]

The acceleration is

\[
a_{\mathrm{kin}}(t)=\frac{dv}{dt}.
\]

Since \(v(t)\) is constant,

\[
a_{\mathrm{kin}}(t)=0.
\]

Here we use the notation \(a_{\mathrm{kin}}\) for acceleration in this example to avoid confusion with the parameter \(a\) in the formula for \(x(t)\). This is a useful warning: the same letter can be used for different things in different contexts, but in one calculation we must keep the meaning clear.

\subsection{Example: parabolic position}

Consider again

\[
\mathbf{r}(t)=(t,t^2).
\]

We found

\[
\mathbf{v}(t)=(1,2t).
\]

Therefore

\[
\mathbf{a}(t)
=
\frac{d\mathbf{v}}{dt}
=
(0,2).
\]

The acceleration is constant. It points in the positive \(y\)-direction.

This does not come from looking only at the geometric curve. It comes from how the point moves along the curve as time changes.

\subsection{Example: circular motion}

Consider again

\[
\mathbf{r}(t)=(R\cos t,R\sin t).
\]

We found

\[
\mathbf{v}(t)=(-R\sin t,R\cos t).
\]

Now we differentiate again:

\[
\mathbf{a}(t)
=
(-R\cos t,-R\sin t).
\]

But

\[
\mathbf{r}(t)=(R\cos t,R\sin t).
\]

Therefore

\[
\mathbf{a}(t)=-\mathbf{r}(t).
\]

This means that the acceleration points toward the center of the circle. The velocity is tangent to the circle, but the acceleration points inward.

This is a major conceptual point. In uniform circular motion, speed may be constant, but acceleration is not zero, because the direction of velocity is changing.

\subsection{Units of acceleration}

Acceleration has units of velocity divided by time. If velocity is measured in metres per second and time in seconds, acceleration is measured in

\[
\mathrm{m/s^2}.
\]

This also follows from

\[
\mathbf{a}(t)=\frac{d^2\mathbf{r}}{dt^2}.
\]

Position contributes metres, and two differentiations with respect to time contribute two factors of seconds in the denominator.

The unit \(\mathrm{m/s^2}\) means metres per second per second. It describes how velocity changes each second.
