\section{What kinematics studies}

Kinematics is the part of mechanics that describes motion. At this stage we do not ask why the motion happens. We do not introduce forces, interactions, or causes of motion. Those questions belong to dynamics. In kinematics we ask a more basic question:

\[
\text{How can motion be described mathematically?}
\]

To answer this question, we need the mathematical language recalled in the previous chapter: coordinates, functions, parametric curves, vectors, derivatives, and integrals. The new idea is not a new mathematical object. The new idea is interpretation. We now use mathematical objects to describe the motion of a body in space and time.

The central object of kinematics is the position of a moving point as a function of time. Once position is known, we can ask how it changes. This leads to velocity. Once velocity is known, we can ask how velocity changes. This leads to acceleration. The structure is therefore

\[
\text{position} \quad \longrightarrow \quad \text{velocity} \quad \longrightarrow \quad \text{acceleration}.
\]

Mathematically, this structure is built using derivatives. Later we will also move in the opposite direction:

\[
\text{acceleration} \quad \longrightarrow \quad \text{velocity} \quad \longrightarrow \quad \text{position}.
\]

Mathematically, this opposite direction is built using integrals and initial conditions.

This chapter should be read slowly. The goal is not only to remember definitions. The goal is to understand how a mathematical description of motion is constructed.

\section{Position as a function of time}

\subsection{Position in a coordinate system}

To describe motion, we must first describe position. A position is described only after we choose a coordinate system. In one dimension, the position of a point can be represented by one coordinate:

\[
x.
\]

In two dimensions, the position is represented by two coordinates:

\[
(x,y).
\]

In three dimensions, the position is represented by three coordinates:

\[
(x,y,z).
\]

It is often convenient to write position as a vector. In three dimensions we write

\[
\mathbf{r} = (x,y,z).
\]

In two dimensions we write

\[
\mathbf{r} = (x,y).
\]

The symbol \(\mathbf{r}\) represents the position vector. It points from the origin of the chosen coordinate system to the point whose position we want to describe.

The position vector depends on the coordinate system. If we choose a different origin or different axes, the coordinates of the same physical point may change. This is not a contradiction. Coordinates are not the point itself. Coordinates are a numerical description of the point after a coordinate system has been chosen.

\subsection{Motion as changing position}

A body is in motion when its position changes in time. This means that the coordinates of the point are functions of time.

In one dimension we write

\[
x = x(t).
\]

In two dimensions we write

\[
x = x(t), \qquad y = y(t),
\]
or, in vector form,

\[
\mathbf{r}(t) = (x(t),y(t)).
\]

In three dimensions we write

\[
\mathbf{r}(t) = (x(t),y(t),z(t)).
\]

This notation is important. The symbol \(t\) represents time. The expression \(\mathbf{r}(t)\) means that position depends on time. If we choose a particular value of time, we obtain the position of the body at that time.

For example, if

\[
\mathbf{r}(t) = (t,t^2),
\]

then at \(t=0\),

\[
\mathbf{r}(0) = (0,0).
\]

At \(t=1\),

\[
\mathbf{r}(1) = (1,1).
\]

At \(t=2\),

\[
\mathbf{r}(2) = (2,4).
\]

Thus one formula describes many positions. It tells us where the point is at each time.

\subsection{Trajectory}

The set of all positions occupied by a moving point is called its trajectory. The trajectory is the curve traced by the moving point in space.

It is important to distinguish between the motion and the trajectory. The trajectory is the geometric curve. The motion contains more information, because it tells us where the point is at each moment of time.

Two different motions may have the same trajectory. For example,

\[
\mathbf{r}_1(t) = (\cos t,\sin t)
\]
and
\[
\mathbf{r}_2(t) = (\cos 2t,\sin 2t)
\]

both describe points moving on the unit circle, because in both cases

\[
x^2+y^2 = 1.
\]

However, these two motions are not identical. The second parametrization goes around the same circle differently as time changes. The path is the same, but the time description is different.

This is one reason why parametric descriptions are so important in physics. Geometry alone tells us the shape of the path. Kinematics needs more: it needs the position as a function of time.

\subsection{Motion along a straight line}

The simplest example is motion along a straight line. In one dimension we may write

\[
x(t) = x_0 + at,
\]

where \(x_0\) and \(a\) are constants. At this stage \(a\) is only a parameter in the formula. We will interpret such parameters more deeply after introducing velocity.

If \(x_0=1\) and \(a=2\), then

\[
x(t) = 1+2t.
\]

At \(t=0\),

\[
x(0)=1.
\]

At \(t=1\),

\[
x(1)=3.
\]

At \(t=2\),

\[
x(2)=5.
\]

The position changes along one axis. The trajectory is a line. The formula tells us not only the line on which the point lies, but also the position of the point at every time.

In two dimensions, a straight-line motion can be written as

\[
\mathbf{r}(t) = \mathbf{r}_0 + t\mathbf{a},
\]
where

\[
\mathbf{r}_0 = (x_0,y_0)
\]
and

\[
\mathbf{a} = (a_x,a_y).
\]

In coordinates this means

\[
x(t)=x_0+a_xt,
\]
\[
y(t)=y_0+a_yt.
\]

As \(t\) changes, the point moves along a straight line in the plane.

\subsection{Motion along a parabola}

Consider the position

\[
\mathbf{r}(t) = (t,t^2).
\]

This means

\[
x(t)=t,
\]
and

\[
y(t)=t^2.
\]

If we eliminate the parameter \(t\), we get

\[
t=x,
\]
and therefore

\[
y=x^2.
\]

So the trajectory is a parabola.

The parametric description

\[
\mathbf{r}(t)=(t,t^2)
\]

contains time information. The equation

\[
y=x^2
\]

contains only the shape of the trajectory. It tells us which curve is traced, but not how the point moves along the curve as time passes.

Another parametrization may give the same parabola with a different time description. For example,

\[
\mathbf{r}(t) = (2t,4t^2)
\]

also satisfies

\[
y=x^2,
\]

because if \(x=2t\), then \(t=x/2\), and

\[
y=4t^2=4\left(\frac{x}{2}\right)^2=x^2.
\]

The same geometric curve can therefore be produced by different time parametrizations.

\subsection{Motion along a circle}

Now consider

\[
\mathbf{r}(t)=(R\cos t,R\sin t),
\]

where \(R>0\) is a constant.

Then

\[
x(t)=R\cos t,
\]
and

\[
y(t)=R\sin t.
\]

We can check what geometric curve this describes. We compute

\[
x^2(t)+y^2(t)
=
R^2\cos^2 t+R^2\sin^2 t.
\]

Using

\[
\cos^2 t+\sin^2 t=1,
\]

we obtain

\[
x^2(t)+y^2(t)=R^2.
\]

Thus the point moves along a circle of radius \(R\) centered at the origin.

Again, the equation

\[
x^2+y^2=R^2
\]

describes the circle as a geometric object. The parametric formula

\[
\mathbf{r}(t)=(R\cos t,R\sin t)
\]

describes the position on the circle at each time.

We may also write

\[
\mathbf{r}(t)=(R\cos(\omega t),R\sin(\omega t)),
\]

where \(\omega\) is a constant. This describes the same circle, but the time dependence is different. The parameter \(\omega\) controls how rapidly the argument of the trigonometric functions changes. We will return to the meaning of this when we discuss velocity and circular motion.

\subsection{What position tells us and what it does not tell us}

A function \(\mathbf{r}(t)\) is the complete kinematic description of position. It tells us where the point is at every time. From it, we can find the trajectory by looking at the set of all points \(\mathbf{r}(t)\).

However, position alone is not the end of the description. Once we know how position depends on time, we naturally ask how fast the position changes and in what direction it changes. This question leads to velocity.

Before moving on, the reader should remember the following points:

\begin{itemize}
    \item Motion is described by position as a function of time.
    \item In one dimension, position is written as \(x(t)\).
    \item In two or three dimensions, position is written as a vector \(\mathbf{r}(t)\).
    \item The trajectory is the curve traced by the moving point.
    \item The trajectory is not the same as the full motion.
    \item Different time parametrizations may describe the same geometric curve.
    \item In physics, the time dependence is essential.
\end{itemize}

\section{Velocity}

\subsection{Why position is not enough}

Knowing the position of a moving point at every time gives a complete description of where the point is. But we often want to describe motion more locally. We want to know how the position is changing at a given moment.

If the position changes a lot during a short time interval, we say that the motion is fast. If the position changes only a little during the same time interval, we say that the motion is slow. If the position changes in a different direction, the motion has a different direction.

Velocity is the mathematical object that describes this local change of position.

\subsection{Average velocity}

Let the position of a moving point be \(\mathbf{r}(t)\). Suppose we compare two moments of time, \(t_1\) and \(t_2\). The change in position is

\[
\Delta \mathbf{r} = \mathbf{r}(t_2)-\mathbf{r}(t_1).
\]

The change in time is

\[
\Delta t = t_2-t_1.
\]

The average velocity over this time interval is defined as

\[
\mathbf{v}_{\mathrm{avg}}
=
\frac{\Delta \mathbf{r}}{\Delta t}
=
\frac{\mathbf{r}(t_2)-\mathbf{r}(t_1)}{t_2-t_1}.
\]

This is a vector. Its direction is the direction of the displacement \(\Delta \mathbf{r}\), and its magnitude tells us how much displacement occurs per unit time during the interval.

In one dimension, this becomes

\[
v_{\mathrm{avg}}
=
\frac{x(t_2)-x(t_1)}{t_2-t_1}.
\]

Average velocity depends on the interval. If we choose a large interval, we describe the overall change during that interval. If we choose a smaller interval, we get a more local description.

\subsection{Instantaneous velocity}

To describe the motion at one moment, we take the limit of the average velocity as the time interval becomes very small. This gives instantaneous velocity:

\[
\mathbf{v}(t)
=
\frac{d\mathbf{r}}{dt}.
\]

If

\[
\mathbf{r}(t)=(x(t),y(t),z(t)),
\]

then

\[
\mathbf{v}(t)
=
\left(
\frac{dx}{dt},
\frac{dy}{dt},
\frac{dz}{dt}
\right).
\]

In two dimensions,

\[
\mathbf{v}(t)
=
\left(
\frac{dx}{dt},
\frac{dy}{dt}
\right).
\]

In one dimension,

\[
v(t)=\frac{dx}{dt}.
\]

Velocity is therefore the derivative of position with respect to time. This is the first major connection between kinematics and calculus.

\subsection{Velocity as a vector attached to the moving point}

Velocity is a vector quantity. This means that it has magnitude and direction.

At each time \(t\), the point has a position \(\mathbf{r}(t)\) and a velocity \(\mathbf{v}(t)\). It is useful to imagine the velocity vector as attached to the point at its current position.

The velocity vector is tangent to the trajectory. This statement is important. The trajectory tells us the path. The velocity tells us the local direction in which the point is moving along that path.

For example, if a point moves along a circle, the velocity is not directed toward the center of the circle. The velocity is tangent to the circle. The direction of velocity changes from point to point, even if the length of the velocity vector remains constant.

\subsection{Speed}

The magnitude of the velocity vector is called speed:

\[
v(t)=|\mathbf{v}(t)|.
\]

This notation can be slightly confusing because the same letter \(v\) is often used for both velocity and speed. In these notes, bold symbols such as \(\mathbf{v}\) denote vectors, while non-bold symbols such as \(v\) may denote magnitudes.

If

\[
\mathbf{v}(t)=(v_x(t),v_y(t),v_z(t)),
\]

then

\[
|\mathbf{v}(t)|
=
\sqrt{v_x^2(t)+v_y^2(t)+v_z^2(t)}.
\]

Speed is not a vector. It is a non-negative number. It tells us how fast the point is moving, but not in which direction.

Velocity tells us both: how fast and in which direction.

\subsection{Example: motion along a line}

Consider the one-dimensional position

\[
x(t)=x_0+at,
\]

where \(x_0\) and \(a\) are constants.

The velocity is

\[
v(t)=\frac{dx}{dt}.
\]

Therefore

\[
v(t)=a.
\]

The velocity is constant. This means that the position changes at the same rate at every time.

If

\[
x(t)=1+2t,
\]

then

\[
v(t)=2.
\]

The number \(2\) is not only the coefficient in the formula. It is the rate at which position changes with respect to time.

\subsection{Example: motion along a parabola}

Consider

\[
\mathbf{r}(t)=(t,t^2).
\]

Then

\[
x(t)=t,
\]
and

\[
y(t)=t^2.
\]

The velocity is

\[
\mathbf{v}(t)
=
\left(
\frac{dx}{dt},
\frac{dy}{dt}
\right)
=
(1,2t).
\]

At \(t=0\),

\[
\mathbf{v}(0)=(1,0).
\]

At \(t=1\),

\[
\mathbf{v}(1)=(1,2).
\]

At \(t=2\),

\[
\mathbf{v}(2)=(1,4).
\]

The trajectory is a parabola, but the velocity tells us how the point moves along the parabola. The horizontal component is constant, while the vertical component changes.

This example also shows that a point on a trajectory and a velocity vector are different objects. The point tells us where the body is. The velocity vector tells us how that position is changing at that moment.

\subsection{Example: motion along a circle}

Consider

\[
\mathbf{r}(t)=(R\cos t,R\sin t).
\]

The velocity is

\[
\mathbf{v}(t)
=
\frac{d\mathbf{r}}{dt}
=
(-R\sin t,R\cos t).
\]

The speed is

\[
|\mathbf{v}(t)|
=
\sqrt{(-R\sin t)^2+(R\cos t)^2}.
\]

Thus

\[
|\mathbf{v}(t)|
=
\sqrt{R^2\sin^2 t+R^2\cos^2 t}
=
R.
\]

The speed is constant. However, the velocity vector is not constant, because its direction changes with time.

This is an important point. Constant speed does not mean constant velocity. Velocity is a vector. If the direction changes, the velocity changes.

We can also check that the velocity is tangent to the circle. The position vector is

\[
\mathbf{r}(t)=(R\cos t,R\sin t),
\]

and the velocity vector is

\[
\mathbf{v}(t)=(-R\sin t,R\cos t).
\]

Their dot product is

\[
\mathbf{r}(t)\cdot\mathbf{v}(t)
=
R\cos t(-R\sin t)+R\sin t(R\cos t).
\]

Therefore

\[
\mathbf{r}(t)\cdot\mathbf{v}(t)=0.
\]

The position vector and velocity vector are perpendicular. Since the position vector points radially outward from the center of the circle, the velocity vector is tangent to the circle.

\subsection{Example: recognizing a special point on a trajectory}

Consider a motion in the plane described by

\[
\mathbf{r}(t)=(t, -t^2+4t).
\]

The trajectory is a parabola because

\[
x=t,
\]

so

\[
y=-x^2+4x.
\]

The velocity is

\[
\mathbf{v}(t)=
\left(
\frac{dx}{dt},
\frac{dy}{dt}
\right)
=
(1,-2t+4).
\]

The vertical component of velocity is

\[
v_y(t)=-2t+4.
\]

At the highest point of this parabola, the vertical component of velocity should be zero. We solve

\[
-2t+4=0.
\]

Thus

\[
t=2.
\]

The position at this time is

\[
\mathbf{r}(2)=(2,-4+8)=(2,4).
\]

So the highest point is \((2,4)\).

This example shows that velocity is not introduced only to compute another formula. Velocity helps us understand the structure of motion. It can identify special points on a trajectory, such as points where the motion changes from upward to downward.

\subsection{Units of velocity}

Velocity has units of position divided by time. If position is measured in metres and time in seconds, velocity is measured in

\[
\mathrm{m/s}.
\]

This follows directly from the definition

\[
\mathbf{v}(t)=\frac{d\mathbf{r}}{dt}.
\]

The numerator has the unit of length, and the denominator has the unit of time.

Whenever a velocity is computed, its unit should be checked. If a calculation gives a velocity in metres or seconds instead of metres per second, something has gone wrong.

\section{Acceleration}

\subsection{Why velocity is not the end of the description}

Velocity describes how position changes. But velocity itself may also change.

A car may move faster and faster. A body may slow down. A point moving along a circle may keep the same speed but continuously change direction. In each of these cases, velocity changes.

Acceleration describes the change of velocity with respect to time.

\subsection{Average acceleration}

Suppose the velocity at time \(t_1\) is \(\mathbf{v}(t_1)\), and the velocity at time \(t_2\) is \(\mathbf{v}(t_2)\). The change in velocity is

\[
\Delta \mathbf{v}
=
\mathbf{v}(t_2)-\mathbf{v}(t_1).
\]

The average acceleration over the interval is

\[
\mathbf{a}_{\mathrm{avg}}
=
\frac{\Delta \mathbf{v}}{\Delta t}
=
\frac{\mathbf{v}(t_2)-\mathbf{v}(t_1)}{t_2-t_1}.
\]

Average acceleration describes how velocity changes over a time interval.

\subsection{Instantaneous acceleration}

Instantaneous acceleration is the derivative of velocity with respect to time:

\[
\mathbf{a}(t)
=
\frac{d\mathbf{v}}{dt}.
\]

Since velocity is the derivative of position,

\[
\mathbf{v}(t)=\frac{d\mathbf{r}}{dt},
\]

we also have

\[
\mathbf{a}(t)
=
\frac{d^2\mathbf{r}}{dt^2}.
\]

Thus acceleration is the second derivative of position with respect to time.

In coordinates,

\[
\mathbf{a}(t)
=
\left(
\frac{d^2x}{dt^2},
\frac{d^2y}{dt^2},
\frac{d^2z}{dt^2}
\right).
\]

In two dimensions,

\[
\mathbf{a}(t)
=
\left(
\frac{d^2x}{dt^2},
\frac{d^2y}{dt^2}
\right).
\]

In one dimension,

\[
a(t)=\frac{d^2x}{dt^2}.
\]

\subsection{Acceleration is also a vector}

Acceleration is a vector. It has magnitude and direction. It describes how the velocity vector changes.

This point is important. Acceleration does not only mean that speed increases. Acceleration means that velocity changes. Since velocity is a vector, it can change in two ways:

\begin{itemize}
    \item its magnitude can change,
    \item its direction can change.
\end{itemize}

Therefore a body may be accelerating even if its speed is constant, provided that the direction of velocity changes.

This is one of the most common conceptual difficulties in kinematics. Constant speed does not necessarily mean zero acceleration.

\subsection{Example: motion along a line}

Consider

\[
x(t)=x_0+at.
\]

We already found that

\[
v(t)=a.
\]

The acceleration is

\[
a_{\mathrm{kin}}(t)=\frac{dv}{dt}.
\]

Since \(v(t)\) is constant,

\[
a_{\mathrm{kin}}(t)=0.
\]

Here we use the notation \(a_{\mathrm{kin}}\) for acceleration in this example to avoid confusion with the parameter \(a\) in the formula for \(x(t)\). This is a useful warning: the same letter can be used for different things in different contexts, but in one calculation we must keep the meaning clear.

\subsection{Example: parabolic position}

Consider again

\[
\mathbf{r}(t)=(t,t^2).
\]

We found

\[
\mathbf{v}(t)=(1,2t).
\]

Therefore

\[
\mathbf{a}(t)
=
\frac{d\mathbf{v}}{dt}
=
(0,2).
\]

The acceleration is constant. It points in the positive \(y\)-direction.

This does not come from looking only at the geometric curve. It comes from how the point moves along the curve as time changes.

\subsection{Example: circular motion}

Consider again

\[
\mathbf{r}(t)=(R\cos t,R\sin t).
\]

We found

\[
\mathbf{v}(t)=(-R\sin t,R\cos t).
\]

Now we differentiate again:

\[
\mathbf{a}(t)
=
(-R\cos t,-R\sin t).
\]

But

\[
\mathbf{r}(t)=(R\cos t,R\sin t).
\]

Therefore

\[
\mathbf{a}(t)=-\mathbf{r}(t).
\]

This means that the acceleration points toward the center of the circle. The velocity is tangent to the circle, but the acceleration points inward.

This is a major conceptual point. In uniform circular motion, speed may be constant, but acceleration is not zero, because the direction of velocity is changing.

\subsection{Units of acceleration}

Acceleration has units of velocity divided by time. If velocity is measured in metres per second and time in seconds, acceleration is measured in

\[
\mathrm{m/s^2}.
\]

This also follows from

\[
\mathbf{a}(t)=\frac{d^2\mathbf{r}}{dt^2}.
\]

Position contributes metres, and two differentiations with respect to time contribute two factors of seconds in the denominator.

The unit \(\mathrm{m/s^2}\) means metres per second per second. It describes how velocity changes each second.

\section{Moving between position, velocity, and acceleration}

\subsection{The derivative direction}

So far we have moved in one direction:

\[
\mathbf{r}(t)
\quad \longrightarrow \quad
\mathbf{v}(t)
\quad \longrightarrow \quad
\mathbf{a}(t).
\]

This direction uses derivatives:

\[
\mathbf{v}(t)=\frac{d\mathbf{r}}{dt},
\]
and

\[
\mathbf{a}(t)=\frac{d\mathbf{v}}{dt}
=
\frac{d^2\mathbf{r}}{dt^2}.
\]

If we know position as a function of time, we can compute velocity and acceleration by differentiating. No additional information is needed for this direction. The function \(\mathbf{r}(t)\) already contains all information about the motion.

For example, if

\[
\mathbf{r}(t)=(t,t^2),
\]

then

\[
\mathbf{v}(t)=(1,2t),
\]

and

\[
\mathbf{a}(t)=(0,2).
\]

Everything follows by differentiation.

\subsection{The integral direction}

We can also move in the opposite direction. If we know acceleration, we may want to recover velocity. If we know velocity, we may want to recover position. This direction uses integration:

\[
\mathbf{a}(t)
\quad \longrightarrow \quad
\mathbf{v}(t)
\quad \longrightarrow \quad
\mathbf{r}(t).
\]

But integration introduces constants. These constants are not mathematical noise. They represent missing information about the motion.

If

\[
\mathbf{v}(t)=\frac{d\mathbf{r}}{dt},
\]

then

\[
\mathbf{r}(t)
=
\int \mathbf{v}(t)\,dt.
\]

More precisely,

\[
\mathbf{r}(t)
=
\mathbf{r}(t_0)+\int_{t_0}^{t}\mathbf{v}(\tau)\,d\tau.
\]

The symbol \(\tau\) is used as a dummy variable of integration. The upper limit \(t\) is the time at which we want to find the position. The lower limit \(t_0\) is the initial time.

This formula says that the position at time \(t\) equals the initial position plus the accumulated effect of velocity from \(t_0\) to \(t\).

Similarly, if

\[
\mathbf{a}(t)=\frac{d\mathbf{v}}{dt},
\]

then

\[
\mathbf{v}(t)
=
\mathbf{v}(t_0)+\int_{t_0}^{t}\mathbf{a}(\tau)\,d\tau.
\]

This formula says that the velocity at time \(t\) equals the initial velocity plus the accumulated effect of acceleration from \(t_0\) to \(t\).

\subsection{Why initial conditions are necessary}

Suppose we know only that

\[
v(t)=3.
\]

Then

\[
x(t)=\int 3\,dt=3t+C.
\]

The constant \(C\) is not determined by the velocity. It tells us the initial position.

If we know that

\[
x(0)=2,
\]

then

\[
2=3\cdot 0+C,
\]

so

\[
C=2.
\]

Therefore

\[
x(t)=3t+2.
\]

Without the initial condition, we would know the way position changes, but not the actual starting position.

Now suppose we know only that acceleration is constant:

\[
a(t)=A.
\]

Then velocity is

\[
v(t)=At+C_1.
\]

The constant \(C_1\) is determined by the initial velocity. If

\[
v(0)=v_0,
\]

then

\[
C_1=v_0,
\]

so

\[
v(t)=v_0+At.
\]

To recover position, we integrate again:

\[
x(t)=\int (v_0+At)\,dt
=
v_0t+\frac{1}{2}At^2+C_2.
\]

The constant \(C_2\) is determined by the initial position. If

\[
x(0)=x_0,
\]

then

\[
C_2=x_0.
\]

Therefore

\[
x(t)=x_0+v_0t+\frac{1}{2}At^2.
\]

This formula is often presented as a standard kinematic formula. Here we see where it comes from: it is the result of integrating constant acceleration twice and using initial conditions.

\subsection{Vector form}

The same idea works in vector form.

If acceleration is known, then

\[
\mathbf{v}(t)
=
\mathbf{v}(t_0)+\int_{t_0}^{t}\mathbf{a}(\tau)\,d\tau.
\]

If velocity is known, then

\[
\mathbf{r}(t)
=
\mathbf{r}(t_0)+\int_{t_0}^{t}\mathbf{v}(\tau)\,d\tau.
\]

If acceleration is constant and equal to \(\mathbf{A}\), then

\[
\mathbf{v}(t)=\mathbf{v}_0+\mathbf{A}t,
\]

assuming \(t_0=0\). Integrating again gives

\[
\mathbf{r}(t)=\mathbf{r}_0+\mathbf{v}_0t+\frac{1}{2}\mathbf{A}t^2.
\]

This is the vector version of the same relation.

\subsection{What is known and what is missing}

It is useful to summarize the logic.

If we know position \(\mathbf{r}(t)\), then we can find velocity and acceleration by differentiation:

\[
\mathbf{r}(t)
\longrightarrow
\mathbf{v}(t)
\longrightarrow
\mathbf{a}(t).
\]

No initial conditions are needed because the position function already contains the full motion.

If we know velocity \(\mathbf{v}(t)\), then we can find acceleration by differentiation. But to find position, we must integrate and we need an initial position:

\[
\mathbf{v}(t)
\longrightarrow
\mathbf{a}(t),
\]

but

\[
\mathbf{v}(t)
\longrightarrow
\mathbf{r}(t)
\]

requires \(\mathbf{r}(t_0)\).

If we know acceleration \(\mathbf{a}(t)\), then we can recover velocity and position only by integration. We need initial velocity and initial position:

\[
\mathbf{a}(t)
\longrightarrow
\mathbf{v}(t)
\longrightarrow
\mathbf{r}(t),
\]

but this requires

\[
\mathbf{v}(t_0)
\]
and
\[
\mathbf{r}(t_0).
\]

This is not a technical detail. It expresses an important fact about motion: knowing how something changes is not the same as knowing where it is or what its current value is. To reconstruct the motion, we need both a law of change and initial information.

\subsection{The central structure of kinematics}

The central structure of kinematics is therefore:

\[
\boxed{
\mathbf{v}(t)=\frac{d\mathbf{r}}{dt}
}
\]

and

\[
\boxed{
\mathbf{a}(t)=\frac{d\mathbf{v}}{dt}
=
\frac{d^2\mathbf{r}}{dt^2}
}
\]

together with the inverse relations

\[
\boxed{
\mathbf{v}(t)
=
\mathbf{v}(t_0)+\int_{t_0}^{t}\mathbf{a}(\tau)\,d\tau
}
\]

and

\[
\boxed{
\mathbf{r}(t)
=
\mathbf{r}(t_0)+\int_{t_0}^{t}\mathbf{v}(\tau)\,d\tau
}.
\]

These formulas should not be treated as isolated recipes. They express the whole logic of kinematics:

\begin{itemize}
    \item position describes where the body is,
    \item velocity describes how position changes,
    \item acceleration describes how velocity changes,
    \item derivatives move from position to velocity and acceleration,
    \item integrals move back from acceleration or velocity to velocity or position,
    \item initial conditions supply the information lost when differentiating.
\end{itemize}

This is the mathematical language that will be used again in later chapters. Dynamics will ask why acceleration appears. But before we ask why motion changes, we must first know how to describe motion and its changes precisely.
