\section{Minkowski Spacetime and Events}
\label{sec:minkowski_spacetime_and_events}

Classical field theory describes quantities defined throughout space and time. In a relativistic theory, however, space and time cannot be treated as two unrelated backgrounds. Observers in relative motion disagree about spatial distances and time intervals, yet they agree on the speed of light and on the causal order of events that can influence one another.

The appropriate arena is Minkowski spacetime. Its basic elements are events, inertial coordinate systems, and worldlines. The metric structure that measures spacetime intervals will be introduced in the next section; here we first establish the geometric and physical objects that the metric will act upon.

\subsection{Events as points in spacetime}

An event is something that occurs at one place and one time. Examples include

\begin{itemize}
    \item the emission of a light pulse;
    \item a particle passing a detector;
    \item two particles colliding;
    \item a field reaching a specified value at a specified location.
\end{itemize}

An observer using an inertial coordinate system assigns four coordinates to an event:

\[
(t,x,y,z).
\]

It is convenient to combine them into the spacetime coordinate array

\[
\boxed{
x^{\mu}
=
(x^{0},x^{1},x^{2},x^{3})
=
(ct,x,y,z)
}.
\]

The index

\[
\mu=0,1,2,3
\]

labels the four spacetime components. The zeroth coordinate is

\[
x^{0}=ct,
\]

so all four components have the dimensions of length.

At this stage, the upper index on \(x^{\mu}\) is part of the notation for a contravariant four-vector. The distinction between upper and lower indices will be developed systematically later in the chapter.

\subsection{Why use \(ct\) rather than \(t\)?}

Time and space have different physical units. Multiplying time by the speed of light gives

\[
[ct]=\mathrm{m},
\]

so the coordinates

\[
ct,
\qquad
x,
\qquad
y,
\qquad
z
\]

can be compared within one geometric expression.

This convention does not claim that time is identical to space. The distinction is encoded by the spacetime metric, which assigns a different sign to the temporal component than to the spatial components.

If natural units are later adopted, one sets

\[
c=1,
\]

and then writes

\[
x^{\mu}=(t,\mathbf{x}).
\]

Until the final section of this chapter, the factor \(c\) will usually remain explicit.

\subsection{Inertial reference frames}

An inertial reference frame is a coordinate system in which a free particle moves with constant velocity. If no force acts, its spatial trajectory has the form

\[
\mathbf{x}(t)
=
\mathbf{x}_{0}
+
\mathbf{v}t.
\]

Two inertial frames may move uniformly relative to one another. Their coordinates are related by a Lorentz transformation rather than by the Galilean transformation of nonrelativistic mechanics.

The two postulates underlying special relativity are:

\begin{enumerate}
    \item the laws of physics have the same form in every inertial frame;
    \item the speed of light in vacuum has the same value \(c\) in every inertial frame.
\end{enumerate}

The second statement forces a modification of the classical relation between space and time. If a light pulse satisfies

\[
|\Delta\mathbf{x}|=c\Delta t
\]

in one inertial frame, it must satisfy the corresponding relation in every inertial frame.

\subsection{Coordinate descriptions and physical events}

The event itself is independent of coordinates. Different observers assign different coordinate quadruples to the same event.

Suppose observer \(S\) assigns

\[
x^{\mu}=(ct,x,y,z)
\]

to an event, while observer \(S'\) assigns

\[
x'^{\mu}=(ct',x',y',z').
\]

The coordinates differ, but the event is the same point of spacetime. A Lorentz transformation relates the two descriptions:

\[
x'^{\mu}
=
\Lambda^{\mu}{}_{\nu}x^{\nu}.
\]

The matrix \(\Lambda^{\mu}{}_{\nu}\) will be constructed explicitly in the section on Lorentz transformations.

This distinction is fundamental:

\begin{quote}
Coordinates describe an event; they are not the event itself.
\end{quote}

The same principle applies to fields. A scalar field assigns one physical value to an event, while its coordinate formula changes when the coordinate system changes.

\subsection{Worldlines}

The history of a point particle is represented by a curve in spacetime called its worldline. In a chosen inertial frame, the worldline may be written as

\[
x^{\mu}(t)
=
\bigl(ct,\mathbf{x}(t)\bigr).
\]

A particle at rest at spatial position \(\mathbf{x}_{0}\) has

\[
\mathbf{x}(t)=\mathbf{x}_{0},
\]

so its worldline is parallel to the time axis.

A particle moving uniformly with velocity \(\mathbf{v}\) has

\[
\mathbf{x}(t)
=
\mathbf{x}_{0}+\mathbf{v}t.
\]

In a one-dimensional spacetime diagram with horizontal axis \(x\) and vertical axis \(ct\), the worldline is a straight line. Its slope satisfies

\[
\frac{\Delta(ct)}{\Delta x}
=
\frac{c}{v}.
\]

For a material particle,

\[
|v|<c,
\]

so its worldline is steeper than a light ray in such a diagram.

\subsection{Light rays and the light cone}

A light signal moving in one spatial dimension obeys

\[
\Delta x=\pm c\Delta t.
\]

Since

\[
\Delta x^{0}=c\Delta t,
\]

this becomes

\[
\Delta x=\pm\Delta x^{0}.
\]

In a spacetime diagram, light rays form lines at \(45^{\circ}\) when the axes \(ct\) and \(x\) are drawn with the same scale.

In three spatial dimensions, the set of all light signals emitted from one event forms a light cone. The cone divides spacetime into four regions:

\begin{itemize}
    \item the future light cone;
    \item the past light cone;
    \item the region outside the cone in spatial directions;
    \item the cone itself.
\end{itemize}

An event inside the future light cone can be influenced by a signal emitted at the original event without exceeding the speed of light. An event inside the past light cone can influence the original event. Events outside the cone cannot be connected to the original event by a material particle or signal moving at or below \(c\).

\subsection{Causal classification without the metric notation}

Let two events have coordinate differences

\[
\Delta t=t_{2}-t_{1},
\qquad
\Delta\mathbf{x}=\mathbf{x}_{2}-\mathbf{x}_{1}.
\]

Compare

\[
c^{2}(\Delta t)^{2}
\]

with

\[
|\Delta\mathbf{x}|^{2}.
\]

Three cases occur.

\begin{enumerate}
    \item If
    \[
    c^{2}(\Delta t)^{2}
    >
    |\Delta\mathbf{x}|^{2},
    \]
    the separation is timelike. A massive particle moving slower than light can travel from one event to the other.

    \item If
    \[
    c^{2}(\Delta t)^{2}
    =
    |\Delta\mathbf{x}|^{2},
    \]
    the separation is null or lightlike. Only a signal moving at speed \(c\) can connect the events.

    \item If
    \[
    c^{2}(\Delta t)^{2}
    <
    |\Delta\mathbf{x}|^{2},
    \]
    the separation is spacelike. Connecting the events would require speed greater than \(c\).
\end{enumerate}

The next section will combine these three comparisons into one invariant quantity called the spacetime interval.

\subsection{Worked example: can one event influence another?}

Event \(A\) occurs at

\[
t_{A}=0,
\qquad
x_{A}=0.
\]

Event \(B\) occurs at

\[
t_{B}=2.0\,\mu\mathrm{s},
\qquad
x_{B}=400\,\mathrm{m}.
\]

The coordinate differences are

\[
\Delta t=2.0\times10^{-6}\,\mathrm{s},
\]

and

\[
\Delta x=400\,\mathrm{m}.
\]

A light signal travels during this time the distance

\[
c\Delta t
=
(3.00\times10^{8}\,\mathrm{m\,s^{-1}})
(2.0\times10^{-6}\,\mathrm{s})
=
600\,\mathrm{m}.
\]

Since

\[
|\Delta x|=400\,\mathrm{m}<600\,\mathrm{m}=c\Delta t,
\]

the separation is timelike. A signal moving at speed

\[
v
=
\frac{\Delta x}{\Delta t}
=
2.0\times10^{8}\,\mathrm{m\,s^{-1}}
\]

can connect the events, and this speed satisfies

\[
v<c.
\]

Therefore event \(A\) can causally influence event \(B\).

\subsection{Worked example: spacelike separation}

Suppose two events satisfy

\[
\Delta t=1.0\,\mu\mathrm{s},
\qquad
|\Delta x|=500\,\mathrm{m}.
\]

Light travels only

\[
c\Delta t
=
300\,\mathrm{m}
\]

during this time. The speed required to connect the events would be

\[
v
=
\frac{500\,\mathrm{m}}{1.0\times10^{-6}\,\mathrm{s}}
=
5.0\times10^{8}\,\mathrm{m\,s^{-1}},
\]

which is greater than \(c\). The separation is spacelike, so no causal influence can travel from one event to the other within special relativity.

\subsection{Simultaneity is frame dependent}

Two events are simultaneous in a frame \(S\) if

\[
\Delta t=0.
\]

For events separated in space, a Lorentz transformation generally gives

\[
\Delta t'\neq0.
\]

Thus distant simultaneity is not an observer-independent relation. This does not create a causal contradiction because spacelike-separated events cannot influence one another.

Events occurring at the same spacetime point are different. If

\[
\Delta\mathbf{x}=0
\]

and

\[
\Delta t=0,
\]

then the two descriptions refer to one event, and every inertial observer agrees that the events coincide.

\subsection{Fields on Minkowski spacetime}

A scalar field is a function

\[
\phi:\mathbb{M}^{1,3}\longrightarrow\mathbb{R},
\]

where \(\mathbb{M}^{1,3}\) denotes four-dimensional Minkowski spacetime.

In coordinates, one writes

\[
\phi=\phi(x^{\mu})
=
\phi(ct,x,y,z).
\]

A vector field assigns a four-vector to every event:

\[
A^{\mu}=A^{\mu}(x).
\]

The central requirement of relativistic field theory is that physical laws should be expressible in a form that is valid in every inertial frame. Covariant notation is the language used to make this requirement visible.

\subsection{Structural checks}

A spacetime description should pass several immediate checks.

\begin{enumerate}
    \item Every event must be assigned four coordinates in a chosen frame.
    \item The zeroth coordinate \(x^{0}=ct\) must have the same dimensions as the spatial coordinates.
    \item A material worldline must remain inside the light cone.
    \item A light ray must lie on the light cone.
    \item A causal conclusion must not depend on one particular inertial coordinate system.
\end{enumerate}

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item identifying an event with its coordinates rather than treating coordinates as one description of the event;
    \item using \(t\) as the zeroth component while treating all components as having units of length;
    \item drawing a light ray at \(45^{\circ}\) when the time and space axes use different scales;
    \item assuming that all observers agree on distant simultaneity;
    \item confusing a worldline with the spatial path seen at one time;
    \item treating spacelike separation as merely a large distance rather than a comparison between distance and elapsed time.
\end{enumerate}

\subsection{What has been established}

Minkowski spacetime is the arena of special-relativistic field theory. Its points are events, and an inertial observer assigns coordinates

\[
x^{\mu}=(ct,\mathbf{x}).
\]

A particle traces a worldline, light signals generate the light cone, and pairs of events are classified as timelike, null, or spacelike according to whether their spatial separation is smaller than, equal to, or greater than the distance light can travel during the elapsed time.

The next section introduces the Minkowski metric, combines this causal classification into the invariant spacetime interval, and defines proper time.