\section{Four-Derivatives and the d'Alembert Operator}
\label{sec:four_derivatives_and_dalembert_operator}

Relativistic field equations contain derivatives with respect to both time and space. To write them covariantly, these derivatives must be assembled into an object with a definite Lorentz-transformation rule.

The resulting object is the four-gradient. Contracting the four-gradient with itself produces the d'Alembert operator, the relativistic analogue of the Laplacian.

\subsection{Coordinates and partial derivatives}

The spacetime coordinates are

\[
x^{\mu}=(ct,x,y,z).
\]

The derivative with respect to the zeroth coordinate is

\[
\frac{\partial}{\partial x^{0}}
=
\frac{\partial}{\partial(ct)}
=
\frac{1}{c}
\frac{\partial}{\partial t}.
\]

The spatial derivatives are

\[
\frac{\partial}{\partial x^{1}}
=
\frac{\partial}{\partial x},
\qquad
\frac{\partial}{\partial x^{2}}
=
\frac{\partial}{\partial y},
\qquad
\frac{\partial}{\partial x^{3}}
=
\frac{\partial}{\partial z}.
\]

\subsection{The covariant four-gradient}

Define

\[
\boxed{
\partial_{\mu}
:=
\frac{\partial}{\partial x^{\mu}}
}.
\]

Its components are

\[
\boxed{
\partial_{\mu}
=
\left(
\frac{1}{c}\frac{\partial}{\partial t},
\nabla
\right)
}.
\]

Here

\[
\nabla
=
\left(
\frac{\partial}{\partial x},
\frac{\partial}{\partial y},
\frac{\partial}{\partial z}
\right).
\]

The lower index is essential. By the chain rule, derivatives with respect to contravariant coordinates transform as covectors.

\subsection{Transformation law from the chain rule}

Let

\[
x'^{\mu}
=
\Lambda^{\mu}{}_{\nu}x^{\nu}.
\]

The inverse relation is

\[
x^{\nu}
=
(\Lambda^{-1})^{\nu}{}_{\mu}x'^{\mu}.
\]

Using the chain rule,

\[
\begin{aligned}
\partial'_{\mu}
&=
\frac{\partial}{\partial x'^{\mu}}\\
&=
\frac{\partial x^{\nu}}
{\partial x'^{\mu}}
\frac{\partial}{\partial x^{\nu}}\\
&=
(\Lambda^{-1})^{\nu}{}_{\mu}
\partial_{\nu}.
\end{aligned}
\]

Thus

\[
\boxed{
\partial'_{\mu}
=
(\Lambda^{-1})^{\nu}{}_{\mu}
\partial_{\nu}
}.
\]

The four-gradient is a covector operator.

\subsection{The contravariant four-gradient}

Raise the index with the inverse metric:

\[
\boxed{
\partial^{\mu}
=
\eta^{\mu\nu}\partial_{\nu}
}.
\]

With

\[
\eta^{\mu\nu}
=
\operatorname{diag}(1,-1,-1,-1),
\]

we obtain

\[
\boxed{
\partial^{\mu}
=
\left(
\frac{1}{c}\frac{\partial}{\partial t},
-\nabla
\right)
}.
\]

Thus

\[
\partial^{0}=\partial_{0},
\qquad
\partial^{i}=-\partial_{i}.
\]

The sign difference between \(\partial_{\mu}\) and \(\partial^{\mu}\) is a direct consequence of the metric convention.

\subsection{Derivative of a scalar field}

Let \(\phi(x)\) be a Lorentz scalar:

\[
\phi'(x')=\phi(x).
\]

Its gradient

\[
\partial_{\mu}\phi
\]

is a covector. The raised gradient

\[
\partial^{\mu}\phi
\]

is a contravariant four-vector.

In components,

\[
\boxed{
\partial_{\mu}\phi
=
\left(
\frac{1}{c}\phi_{t},
\nabla\phi
\right)
}
\]

and

\[
\boxed{
\partial^{\mu}\phi
=
\left(
\frac{1}{c}\phi_{t},
-\nabla\phi
\right)
}.
\]

The quantity

\[
\partial_{\mu}\phi\,
\partial^{\mu}\phi
\]

is a Lorentz scalar.

Expanding,

\[
\boxed{
\partial_{\mu}\phi\,
\partial^{\mu}\phi
=
\frac{1}{c^{2}}
\left(
\frac{\partial\phi}{\partial t}
\right)^{2}
-
|\nabla\phi|^{2}
}.
\]

This combination will become the kinetic term of the relativistic scalar-field Lagrangian.

\subsection{Four-divergence}

For a contravariant four-vector field \(A^{\mu}(x)\), its divergence is

\[
\boxed{
\partial_{\mu}A^{\mu}
}.
\]

The derivative index is contracted with the vector index, so the result is a Lorentz scalar.

If

\[
A^{\mu}=(A^{0},\mathbf{A}),
\]

then

\[
\boxed{
\partial_{\mu}A^{\mu}
=
\frac{1}{c}
\frac{\partial A^{0}}{\partial t}
+
\nabla\cdot\mathbf{A}
}.
\]

For the four-current

\[
J^{\mu}=(c\rho,\mathbf{J}),
\]

we find

\[
\partial_{\mu}J^{\mu}
=
\frac{1}{c}
\frac{\partial(c\rho)}{\partial t}
+
\nabla\cdot\mathbf{J}.
\]

Therefore

\[
\boxed{
\partial_{\mu}J^{\mu}=0
}
\]

is equivalent to the continuity equation

\[
\boxed{
\frac{\partial\rho}{\partial t}
+
\nabla\cdot\mathbf{J}
=0
}.
\]

This is an example of a familiar three-dimensional conservation law written as one covariant scalar equation.

\subsection{The d'Alembert operator}

Contract the two four-derivatives:

\[
\boxed{
\Box
:=
\partial_{\mu}\partial^{\mu}
=
\eta^{\mu\nu}
\partial_{\mu}\partial_{\nu}
}.
\]

The symbol \(\Box\) is read as d'Alembertian or wave operator.

Expanding the components,

\[
\begin{aligned}
\Box
&=
\partial_{0}\partial^{0}
+
\partial_{1}\partial^{1}
+
\partial_{2}\partial^{2}
+
\partial_{3}\partial^{3}\\
&=
\frac{1}{c^{2}}
\frac{\partial^{2}}{\partial t^{2}}
-
\frac{\partial^{2}}{\partial x^{2}}
-
\frac{\partial^{2}}{\partial y^{2}}
-
\frac{\partial^{2}}{\partial z^{2}}.
\end{aligned}
\]

Thus

\[
\boxed{
\Box
=
\frac{1}{c^{2}}
\frac{\partial^{2}}{\partial t^{2}}
-
\nabla^{2}
}.
\]

With the opposite metric signature, the overall sign of \(\Box\) is often defined oppositely. The convention must always be checked.

\subsection{Covariant wave equation}

The ordinary wave equation is

\[
\frac{\partial^{2}\phi}{\partial t^{2}}
=
c^{2}\nabla^{2}\phi.
\]

Move all terms to one side and divide by \(c^{2}\):

\[
\frac{1}{c^{2}}
\frac{\partial^{2}\phi}{\partial t^{2}}
-
\nabla^{2}\phi
=0.
\]

Therefore

\[
\boxed{
\Box\phi=0
}.
\]

The compact equation is Lorentz covariant because \(\Box\phi\) is a scalar when \(\phi\) is a scalar field.

\subsection{Why \(\Box\) is Lorentz invariant}

The operator is a contraction:

\[
\Box
=
\partial_{\mu}\partial^{\mu}.
\]

The first derivative transforms covariantly and the second contravariantly. Their contracted index disappears, leaving a scalar operator.

More explicitly,

\[
\partial'_{\mu}\partial'^{\mu}
=
\partial_{\mu}\partial^{\mu}.
\]

Therefore if

\[
\Box\phi=0
\]

holds in one inertial frame, the same covariant equation holds in every inertial frame.

\subsection{Worked example: applying \(\Box\)}

Let

\[
\phi(x,t)
=
A\cos(kx-\omega t)
\]

in one spatial dimension. Then

\[
\frac{\partial^{2}\phi}{\partial t^{2}}
=
-\omega^{2}\phi,
\]

and

\[
\frac{\partial^{2}\phi}{\partial x^{2}}
=
-k^{2}\phi.
\]

Therefore

\[
\begin{aligned}
\Box\phi
&=
\frac{1}{c^{2}}
(-\omega^{2}\phi)
-
(-k^{2}\phi)\\
&=
\left(
-\frac{\omega^{2}}{c^{2}}
+k^{2}
\right)
\phi.
\end{aligned}
\]

The equation

\[
\Box\phi=0
\]

requires

\[
\boxed{
\omega^{2}=c^{2}k^{2}
}.
\]

The covariant equation therefore contains the familiar dispersion relation of a massless wave.

\subsection{The wave four-vector}

A plane-wave phase may be written as

\[
\mathbf{k}\cdot\mathbf{x}-\omega t.
\]

Define the contravariant wave four-vector

\[
\boxed{
k^{\mu}
=
\left(
\frac{\omega}{c},
\mathbf{k}
\right)
}.
\]

Its covariant components are

\[
\boxed{
k_{\mu}
=
\left(
\frac{\omega}{c},
-\mathbf{k}
\right)
}.
\]

Then

\[
k_{\mu}x^{\mu}
=
\omega t-
\mathbf{k}\cdot\mathbf{x}.
\]

Therefore

\[
\mathbf{k}\cdot\mathbf{x}-\omega t
=
-k_{\mu}x^{\mu}.
\]

A plane wave may be written as

\[
\phi(x)
=
A e^{-ik_{\mu}x^{\mu}}.
\]

The phase is a Lorentz scalar.

\subsection{Derivative of a plane wave}

For

\[
\phi(x)
=
A e^{-ik_{\nu}x^{\nu}},
\]

we have

\[
\partial_{\mu}\phi
=
-ik_{\mu}\phi.
\]

Applying another derivative,

\[
\begin{aligned}
\Box\phi
&=
\partial_{\mu}\partial^{\mu}\phi\\
&=
-k_{\mu}k^{\mu}\phi.
\end{aligned}
\]

Thus the massless wave equation implies

\[
\boxed{
k_{\mu}k^{\mu}=0
}.
\]

The wave four-vector is null for a wave with dispersion relation

\[
\omega^{2}=c^{2}|\mathbf{k}|^{2}.
\]

\subsection{Product rule}

The four-derivative obeys the ordinary product rule:

\[
\boxed{
\partial_{\mu}(fg)
=
(\partial_{\mu}f)g
+
f(\partial_{\mu}g)
}.
\]

For a scalar \(f\) and vector \(A^{\mu}\),

\[
\partial_{\mu}(fA^{\mu})
=
(\partial_{\mu}f)A^{\mu}
+
f\partial_{\mu}A^{\mu}.
\]

This identity is the covariant form used in spacetime integration by parts and in derivations of Noether currents.

\subsection{Commutation of partial derivatives}

For a sufficiently smooth field in Cartesian Minkowski coordinates,

\[
\boxed{
\partial_{\mu}\partial_{\nu}\phi
=
\partial_{\nu}\partial_{\mu}\phi
}.
\]

The order of ordinary partial derivatives may therefore be exchanged.

This property will later imply that the antisymmetrized second derivative of a scalar vanishes:

\[
\partial_{\mu}\partial_{\nu}\phi
-
\partial_{\nu}\partial_{\mu}\phi
=0.
\]

In curved spacetime or in gauge theory, generalized covariant derivatives need not commute. Their noncommutativity carries geometric or physical information.

\subsection{Dimensions}

Since

\[
[x^{\mu}]=\mathrm{m},
\]

all components of \(\partial_{\mu}\) have dimensions

\[
[\partial_{\mu}]=\mathrm{m}^{-1}.
\]

The d'Alembertian has dimensions

\[
[\Box]=\mathrm{m}^{-2}.
\]

Thus every term added to \(\Box\phi\) in a field equation must have compatible dimensions.

\subsection{Structural checks}

A covariant derivative calculation should satisfy:

\begin{enumerate}
    \item \(\partial_{\mu}\) has a lower index;
    \item raising the index changes the signs of the spatial components;
    \item the contraction \(\partial_{\mu}A^{\mu}\) is a scalar;
    \item \(\Box\) expands to \(c^{-2}\partial_{t}^{2}-\nabla^{2}\) in the chosen convention;
    \item a plane wave produces the expected dispersion relation;
    \item all terms have compatible dimensions.
\end{enumerate}

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item writing \(\partial_{0}=\partial/\partial t\) instead of \(c^{-1}\partial/\partial t\) when \(x^{0}=ct\);
    \item treating \(\partial^{\mu}\) and \(\partial_{\mu}\) as numerically identical;
    \item using the wrong sign in front of the Laplacian;
    \item calling \(\nabla\phi\) by itself a four-vector;
    \item forgetting that the wave phase must be a Lorentz scalar;
    \item confusing the d'Alembertian with the spatial Laplacian;
    \item applying commutation of derivatives without sufficient smoothness or in a context involving covariant derivatives.
\end{enumerate}

\subsection{What has been established}

The covariant four-gradient is

\[
\partial_{\mu}
=
\left(
\frac{1}{c}\partial_{t},
\nabla
\right),
\]

while

\[
\partial^{\mu}
=
\left(
\frac{1}{c}\partial_{t},
-\nabla
\right).
\]

Their contraction defines the d'Alembert operator

\[
\Box
=
\partial_{\mu}\partial^{\mu}
=
\frac{1}{c^{2}}\partial_{t}^{2}
-
\nabla^{2}.
\]

The ordinary wave equation becomes the covariant scalar equation \(\Box\phi=0\). The next section develops a broader collection of relativistic invariants and shows how scalar contractions encode observer-independent physical statements.