\section{Understanding Quadric Surfaces Through Sections}

A three-dimensional surface should not be recognised only from a memorised
formula. We study it by intersecting it with simple planes. Fixing $z=k$ gives a
horizontal section; fixing $x=k$ or $y=k$ gives vertical sections. The remaining
equation describes a planar curve already studied in this chapter.

The same questions will be asked repeatedly:
\begin{enumerate}
    \item Which curve appears in a chosen cutting plane?
    \item For which values of the parameter does that curve exist?
    \item How does the section change as the plane moves?
    \item What does that family of sections reveal about the whole surface?
\end{enumerate}
This repetition is intentional. It is a method of investigation, not a list of
unrelated recognition tricks.

\subsection{Ellipsoid}

For
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}+\frac{z^2}{c^2}=1,
\]
the section $z=k$ satisfies
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}=1-\frac{k^2}{c^2}.
\]
Thus $|k|<c$ gives an ellipse, $|k|=c$ gives one point, and $|k|>c$ gives no
real points. The surface is understood as ellipses that grow from one endpoint,
reach their largest size at $z=0$, and shrink to the other endpoint.

\subsection{Elliptic paraboloid}

For
\[
z=\frac{x^2}{a^2}+\frac{y^2}{b^2},
\]
the section $z=k$ is
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}=k.
\]
There are no points for $k<0$, one vertex for $k=0$, and growing ellipses for
$k>0$. Setting $y=0$ gives the parabola
\[
z=\frac{x^2}{a^2}.
\]
The horizontal and vertical sections together explain both the opening direction
and the shape near the vertex.

\subsection{Hyperbolic paraboloid}

For
\[
z=\frac{x^2}{a^2}-\frac{y^2}{b^2},
\]
the horizontal section $z=k$ is
\[
\frac{x^2}{a^2}-\frac{y^2}{b^2}=k.
\]
For $k>0$ the hyperbolas open in the $x$ direction; for $k<0$ they open in the
$y$ direction. At $k=0$ the equation factors into two lines:
\[
\left(\frac xa-\frac yb\right)
\left(\frac xa+\frac yb\right)=0.
\]
The vertical sections $y=0$ and $x=0$ are parabolas opening in opposite
directions. These changing sections explain the saddle shape.

\subsection{Hyperboloid of one sheet}

For
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}-\frac{z^2}{c^2}=1,
\]
the section $z=k$ becomes
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}=1+\frac{k^2}{c^2}.
\]
Every height gives an ellipse, smallest at $k=0$ and larger as $|k|$ grows.
Because a nonempty section exists at every height, the surface remains connected
and has a central waist.

\subsection{Hyperboloid of two sheets}

For
\[
-\frac{x^2}{a^2}-\frac{y^2}{b^2}+\frac{z^2}{c^2}=1,
\]
the section $z=k$ satisfies
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}=\frac{k^2}{c^2}-1.
\]
There are no points when $|k|<c$, one point when $|k|=c$, and ellipses when
$|k|>c$. The empty band around $z=0$ separates the surface into two components.
The algebraic absence of sections explains the geometric gap.

\subsection{Elliptic cone}

For
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}-\frac{z^2}{c^2}=0,
\]
the section $z=k$ is
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}=\frac{k^2}{c^2}.
\]
At $k=0$ the section is one point. For positive and negative $k$ we obtain
ellipses of the same size whenever the absolute heights agree. This symmetry
produces two nappes meeting at the origin.

\subsection{Cylinders}

If one coordinate is absent, moving in that coordinate direction never changes
whether the equation is satisfied. For example,
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}=1
\]
has the same elliptical section for every value of $z$. The ellipse is therefore
carried unchanged along the $z$ direction, producing an elliptic cylinder.
Likewise,
\[
y=x^2
\]
produces the same parabola in every plane $z=k$, hence a parabolic cylinder.

\begin{summarybox}
A standard equation names a surface, but its sections explain the surface. Empty
sections reveal gaps, shrinking sections reveal endpoints or vertices, changing
conic types reveal saddle behaviour, and an absent coordinate reveals a direction
of translation. The atlas below records the standard forms; this method explains
why those forms have the geometry shown in their diagrams.
\end{summarybox}
