\section{Problem Set 1: Matrices}

\begin{enumerate}
\item \textbf{Problem 1}\par
For
\[
A=\begin{pmatrix}
2&-1&4\\
0&3&5
\end{pmatrix},
\]
state the dimensions of \(A\) and identify \(a_{12}\), \(a_{21}\), and \(a_{23}\).

\item \textbf{Problem 2}\par
Construct the \(3\times 2\) matrix \(B=(b_{ij})\) defined by
\[
b_{ij}=2i-j.
\]

\item \textbf{Problem 3}\par
Let
\[
A=\begin{pmatrix}1&2\\-3&4\end{pmatrix},
\qquad
B=\begin{pmatrix}5&-1\\2&0\end{pmatrix}.
\]
Compute \(A+B\), \(A-B\), \(3A\), and \(2A-B\).

\item \textbf{Problem 4}\par
Compute the transpose of
\[
C=\begin{pmatrix}
1&0&-2\\
4&3&5
\end{pmatrix}.
\]
Then state the dimensions of \(C^{\mathsf T}\).

\item \textbf{Problem 5}\par
Determine which of the following matrices are symmetric:
\[
A=\begin{pmatrix}2&-1\\-1&3\end{pmatrix},
\qquad
B=\begin{pmatrix}0&2\\-2&0\end{pmatrix},
\qquad
C=\begin{pmatrix}1&0\\0&1\end{pmatrix}.
\]

\item \textbf{Problem 6}\par
Let \(A\) be \(2\times3\), \(B\) be \(3\times4\), and \(C\) be \(4\times2\). Determine which of the products \(AB\), \(BA\), \(BC\), \(CB\), \(AC\), and \(CA\) are defined. State the dimensions of every defined product.

\item \textbf{Problem 7}\par
Compute
\[
AB
\quad\text{for}\quad
A=\begin{pmatrix}1&2&-1\\0&3&4\end{pmatrix},
\qquad
B=\begin{pmatrix}2&1\\-1&0\\3&-2\end{pmatrix}.
\]

\item \textbf{Problem 8}\par
For
\[
A=\begin{pmatrix}1&2\\0&1\end{pmatrix},
\qquad
B=\begin{pmatrix}2&0\\1&3\end{pmatrix},
\]
compute both \(AB\) and \(BA\). Explain what the result shows about matrix multiplication.

\item \textbf{Problem 9}\par
Let
\[
A=\begin{pmatrix}2&-1\\3&4\end{pmatrix},
\qquad
x=\begin{pmatrix}5\\-2\end{pmatrix}.
\]
Compute \(Ax\) and interpret the result as a new column vector.

\item \textbf{Problem 10}\par
Verify directly that
\[
AI_2=I_2A=A
\]
for
\[
A=\begin{pmatrix}3&-2\\1&5\end{pmatrix}.
\]

\item \textbf{Problem 11}\par
Let
\[
D=\operatorname{diag}(2,-1,3).
\]
Compute \(D^2\) and \(D^3\). Describe the pattern.

\item \textbf{Problem 12}\par
Starting with
\[
A=\begin{pmatrix}
1&2&-1\\
3&0&4\\
-2&1&5
\end{pmatrix},
\]
perform the row operation \(R_2\leftarrow R_2-3R_1\).

\item \textbf{Problem 13}\par
Starting with
\[
A=\begin{pmatrix}
1&2\\
3&4
\end{pmatrix},
\]
first interchange \(R_1\) and \(R_2\), and then perform \(R_2\leftarrow R_2-\frac13R_1\). Write the matrix after each step.

\item \textbf{Problem 14}\par
Find \(x,y,z,w\) if
\[
\begin{pmatrix}
x&2y\\
z-1&w+3
\end{pmatrix}
=
\begin{pmatrix}
4&6\\
2&1
\end{pmatrix}.
\]

\item \textbf{Problem 15}\par
Find the matrix \(X\) satisfying
\[
2X+\begin{pmatrix}1&-1\\0&3\end{pmatrix}
=
\begin{pmatrix}5&3\\4&-1\end{pmatrix}.
\]

\item \textbf{Problem 16}\par
Verify the distributive law \(A(B+C)=AB+AC\) for
\[
A=\begin{pmatrix}1&2\\-1&0\end{pmatrix},
\quad
B=\begin{pmatrix}2&1\\0&3\end{pmatrix},
\quad
C=\begin{pmatrix}-1&4\\2&0\end{pmatrix}.
\]

\item \textbf{Problem 17}\par
The columns of
\[
A=\begin{pmatrix}
1&0\\
2&1\\
-1&3
\end{pmatrix}
\]
are vectors \(u\) and \(v\). Write \(u\) and \(v\), and express
\[
A\begin{pmatrix}2\\-1\end{pmatrix}
\]
as a linear combination of these columns.

\item \textbf{Problem 18}\par
The matrix
\[
T=\begin{pmatrix}2&0\\0&-1\end{pmatrix}
\]
acts on points in the plane by matrix multiplication. Find the images of
\[
(1,2)^{\mathsf T},\quad (-3,1)^{\mathsf T},\quad (0,4)^{\mathsf T}.
\]
Describe what \(T\) does geometrically.

\item \textbf{Problem 19}\par
A shop records sales of two products on three days in the matrix
\[
S=\begin{pmatrix}
12&8\\
15&11\\
9&14
\end{pmatrix}.
\]
Explain what the rows and columns could represent. Compute the total sales of each product using matrix or vector notation.

\item \textbf{Problem 20}\par
For each expression below, decide whether it is defined. If it is defined, compute it; if it is not, explain the dimension mismatch:
\[
A=\begin{pmatrix}1&2&3\\0&1&-1\end{pmatrix},
\quad
B=\begin{pmatrix}2&0\\1&4\\-2&3\end{pmatrix},
\quad
u=\begin{pmatrix}1\\-1\\2\end{pmatrix}.
\]
Consider \(AB\), \(BA\), \(Au\), \(Bu\), and \(u^{\mathsf T}B\).
\end{enumerate}

\section{Problem Set 2: Determinants}

\begin{enumerate}
\item \textbf{Problem 1}\par
Compute
\[
\det\begin{pmatrix}4&-3\\2&5\end{pmatrix}.
\]

\item \textbf{Problem 2}\par
Find the determinant of
\[
A=\begin{pmatrix}
1&2&0\\
-1&3&4\\
2&0&5
\end{pmatrix}
\]
by expanding along the first row.

\item \textbf{Problem 3}\par
Compute the determinant of
\[
B=\begin{pmatrix}
2&0&0\\
-1&3&0\\
4&5&-2
\end{pmatrix}
\]
without a Laplace expansion. Explain why your method works.

\item \textbf{Problem 4}\par
Expand
\[
\det\begin{pmatrix}
1&0&2\\
3&-1&4\\
0&5&2
\end{pmatrix}
\]
along the row or column that gives the shortest calculation.

\item \textbf{Problem 5}\par
A matrix \(B\) is obtained from a \(3\times3\) matrix \(A\) by interchanging two rows. If \(\det A=-7\), find \(\det B\).

\item \textbf{Problem 6}\par
A matrix \(C\) is obtained from a \(4\times4\) matrix \(A\) by multiplying one row by \(5\). If \(\det A=3\), find \(\det C\).

\item \textbf{Problem 7}\par
A matrix \(D\) is obtained from \(A\) by adding \(4\) times the first row to the third row. If \(\det A=-2\), determine \(\det D\).

\item \textbf{Problem 8}\par
Suppose \(\det A=3\) and \(\det B=-4\) for two \(3\times3\) matrices. Find \(\det(AB)\), \(\det(A^{\mathsf T})\), and \(\det(2A)\).

\item \textbf{Problem 9}\par
Determine whether
\[
A=\begin{pmatrix}
1&2&3\\
2&4&6\\
0&1&1
\end{pmatrix}
\]
is singular without carrying out a full determinant expansion.

\item \textbf{Problem 10}\par
Find all values of \(k\) for which
\[
A(k)=\begin{pmatrix}
k&2\\
3&k-1
\end{pmatrix}
\]
is singular.

\item \textbf{Problem 11}\par
Find \(a\) if
\[
\det\begin{pmatrix}
a&1\\
2&3
\end{pmatrix}=10.
\]

\item \textbf{Problem 12}\par
Compute
\[
\det\begin{pmatrix}
1&2&3\\
0&4&5\\
0&0&6
\end{pmatrix}
\]
and explain the general rule suggested by the result.

\item \textbf{Problem 13}\par
Let
\[
A=\begin{pmatrix}
1&2&0\\
3&1&4\\
2&-1&5
\end{pmatrix}.
\]
Use row operations to transform \(A\) into an upper triangular matrix, keeping track of every change to the determinant. Then compute \(\det A\).

\item \textbf{Problem 14}\par
A \(3\times3\) matrix has determinant \(6\). What happens to its determinant if:
\begin{enumerate}
\item the first and second rows are interchanged;
\item the third row is multiplied by \(-2\);
\item twice the first row is added to the second row?
\end{enumerate}

\item \textbf{Problem 15}\par
The columns of a \(3\times3\) matrix satisfy \(c_3=2c_1-c_2\). What is the determinant? Explain the structural reason.

\item \textbf{Problem 16}\par
The linear transformation
\[
T(x)=
\begin{pmatrix}
2&1\\
0&3
\end{pmatrix}x
\]
acts on the plane. By what factor does \(T\) scale areas? Does it preserve orientation?

\item \textbf{Problem 17}\par
Find the area of the parallelogram spanned by
\[
u=(3,1),
\qquad
v=(-2,4)
\]
using a determinant.

\item \textbf{Problem 18}\par
Compute the Vandermonde determinant
\[
\det\begin{pmatrix}
1&1&1\\
x_1&x_2&x_3\\
x_1^2&x_2^2&x_3^2
\end{pmatrix}
\]
for \(x_1=1\), \(x_2=2\), and \(x_3=4\), first directly and then using the product formula.

\item \textbf{Problem 19}\par
Consider
\[
A(t)=\begin{pmatrix}
1&t&0\\
0&1&t\\
t&0&1
\end{pmatrix}.
\]
Compute \(\det A(t)\) and determine all real values of \(t\) for which the matrix is singular.

\item \textbf{Problem 20}\par
For
\[
A=\begin{pmatrix}
1&2&3\\
2&5&7\\
0&1&4
\end{pmatrix},
\]
compute the determinant in two different ways: by a Laplace expansion and by row reduction. Compare the amount of work and explain which method is more efficient.
\end{enumerate}

\section{Problem Set 3: Matrix Inversion}

\begin{enumerate}
\item \textbf{Problem 1}\par
Decide whether each matrix is invertible:
\[
A=\begin{pmatrix}2&1\\3&4\end{pmatrix},
\qquad
B=\begin{pmatrix}1&2\\2&4\end{pmatrix}.
\]

\item \textbf{Problem 2}\par
Find the inverse of
\[
A=\begin{pmatrix}3&1\\2&1\end{pmatrix}
\]
using the \(2\times2\) inverse formula.

\item \textbf{Problem 3}\par
Verify your answer to the previous problem by computing both \(AA^{-1}\) and \(A^{-1}A\).

\item \textbf{Problem 4}\par
Find the inverse of
\[
D=\operatorname{diag}(2,-3,5).
\]

\item \textbf{Problem 5}\par
Find all values of \(k\) for which
\[
A(k)=\begin{pmatrix}
k&1\\
2&k
\end{pmatrix}
\]
has an inverse.

\item \textbf{Problem 6}\par
Use Gauss--Jordan elimination to find the inverse of
\[
A=\begin{pmatrix}
1&2\\
3&5
\end{pmatrix}.
\]

\item \textbf{Problem 7}\par
Use Gauss--Jordan elimination to find the inverse of
\[
A=\begin{pmatrix}
1&1&0\\
0&1&1\\
1&0&1
\end{pmatrix}.
\]

\item \textbf{Problem 8}\par
Find the adjugate of
\[
A=\begin{pmatrix}
4&1\\
2&3
\end{pmatrix}
\]
and use it to compute \(A^{-1}\).

\item \textbf{Problem 9}\par
Let
\[
A=\begin{pmatrix}2&1\\1&1\end{pmatrix},
\qquad
b=\begin{pmatrix}5\\3\end{pmatrix}.
\]
Solve \(Ax=b\) using \(A^{-1}\).

\item \textbf{Problem 10}\par
Solve
\[
XA=B
\]
for \(X\), where
\[
A=\begin{pmatrix}1&1\\0&2\end{pmatrix},
\qquad
B=\begin{pmatrix}3&1\\4&2\end{pmatrix}.
\]
Pay attention to the side on which \(A^{-1}\) must be multiplied.

\item \textbf{Problem 11}\par
Solve
\[
2X-A=B
\]
for \(X\), where
\[
A=\begin{pmatrix}1&0\\2&-1\end{pmatrix},
\qquad
B=\begin{pmatrix}3&4\\0&5\end{pmatrix}.
\]

\item \textbf{Problem 12}\par
If
\[
A^{-1}=\begin{pmatrix}2&-1\\0&3\end{pmatrix}
\quad\text{and}\quad
B^{-1}=\begin{pmatrix}1&0\\4&2\end{pmatrix},
\]
find \((AB)^{-1}\).

\item \textbf{Problem 13}\par
Prove by direct multiplication that
\[
(A^{\mathsf T})^{-1}=(A^{-1})^{\mathsf T}
\]
for
\[
A=\begin{pmatrix}1&2\\3&5\end{pmatrix}.
\]

\item \textbf{Problem 14}\par
Find the inverse of the elementary matrix
\[
E=\begin{pmatrix}
1&0&0\\
0&1&0\\
-4&0&1
\end{pmatrix}
\]
and explain which row operation \(E\) represents.

\item \textbf{Problem 15}\par
A proposed inverse of
\[
A=\begin{pmatrix}2&-1\\1&0\end{pmatrix}
\]
is
\[
C=\begin{pmatrix}0&1\\-1&2\end{pmatrix}.
\]
Check whether the proposal is correct.

\item \textbf{Problem 16}\par
Explain why a matrix with two identical rows cannot have an inverse. Use both a determinant argument and an information-loss interpretation.

\item \textbf{Problem 17}\par
The transformation
\[
y=
\begin{pmatrix}2&1\\1&1\end{pmatrix}x
\]
maps an unknown vector \(x\) to
\[
y=\begin{pmatrix}8\\5\end{pmatrix}.
\]
Recover \(x\).

\item \textbf{Problem 18}\par
Choose the most efficient method for finding the inverse of each matrix and justify the choice:
\[
A=\begin{pmatrix}5&0\\0&-2\end{pmatrix},
\quad
B=\begin{pmatrix}1&2\\3&7\end{pmatrix},
\quad
C=\begin{pmatrix}1&1&0\\0&1&1\\0&0&2\end{pmatrix}.
\]
Then compute the inverses.

\item \textbf{Problem 19}\par
Let
\[
A=\begin{pmatrix}
1&a\\
2&2a
\end{pmatrix}.
\]
Explain, without attempting Gauss--Jordan elimination, why \(A^{-1}\) does not exist for any real \(a\).

\item \textbf{Problem 20}\par
For
\[
A=\begin{pmatrix}
1&2\\
3&4
\end{pmatrix},
\]
find \(A^{-1}\), solve \(Ax=b\) for
\[
b=\begin{pmatrix}5\\11\end{pmatrix},
\]
and verify the solution in the original equation. Finally, explain what the inverse accomplishes conceptually.
\end{enumerate}

\section{Problem Set 4: Systems of Linear Equations}

\begin{enumerate}
\item \textbf{Problem 1}\par
Solve by substitution:
\[
\begin{cases}
x+y=7,\\
2x-y=2.
\end{cases}
\]

\item \textbf{Problem 2}\par
Write the augmented matrix for
\[
\begin{cases}
2x-y+3z=5,\\
x+4y-z=2,\\
3x+2z=7.
\end{cases}
\]

\item \textbf{Problem 3}\par
Solve by Gaussian elimination:
\[
\begin{cases}
x+2y=5,\\
3x-y=4.
\end{cases}
\]

\item \textbf{Problem 4}\par
Solve by Gaussian elimination:
\[
\begin{cases}
x+y+z=6,\\
2x-y+z=3,\\
x+3y-2z=1.
\end{cases}
\]

\item \textbf{Problem 5}\par
Reduce the augmented matrix
\[
\left(
\begin{array}{cc|c}
1&2&3\\
2&4&8
\end{array}
\right)
\]
to echelon form and classify the system.

\item \textbf{Problem 6}\par
Reduce
\[
\left(
\begin{array}{ccc|c}
1&2&-1&3\\
2&4&-2&6\\
0&1&1&2
\end{array}
\right)
\]
and express all solutions using a parameter.

\item \textbf{Problem 7}\par
Determine whether
\[
\begin{cases}
x-2y=1,\\
2x-4y=5
\end{cases}
\]
has no solution, one solution, or infinitely many solutions. Interpret the result geometrically.

\item \textbf{Problem 8}\par
Solve
\[
\begin{cases}
2x+y=4,\\
x-3y=-5
\end{cases}
\]
using Cramer's rule.

\item \textbf{Problem 9}\par
Solve the system in the previous problem using the inverse matrix method. Compare the two methods.

\item \textbf{Problem 10}\par
For which values of \(k\) does
\[
\begin{cases}
kx+y=2,\\
2x+ky=3
\end{cases}
\]
have a unique solution?

\item \textbf{Problem 11}\par
Classify the system according to the value of \(a\):
\[
\begin{cases}
x+y=2,\\
2x+2y=a.
\end{cases}
\]

\item \textbf{Problem 12}\par
Find all values of \(k\) for which
\[
\begin{cases}
x+y+z=1,\\
2x+2y+2z=2,\\
x+y+kz=3
\end{cases}
\]
is consistent. Describe the solution set in each consistent case.

\item \textbf{Problem 13}\par
The row-reduced augmented matrix of a system is
\[
\left(
\begin{array}{cccc|c}
1&0&2&0&3\\
0&1&-1&0&4\\
0&0&0&1&2
\end{array}
\right).
\]
Identify the pivot and free variables and write the complete solution parametrically.

\item \textbf{Problem 14}\par
Determine whether the point \((2,-1)\) is the solution of
\[
\begin{cases}
3x+y=5,\\
x-2y=4.
\end{cases}
\]
If it is not, find the correct solution.

\item \textbf{Problem 15}\par
Find the intersection point of the lines
\[
2x-y-1=0,
\qquad
x+3y-9=0.
\]

\item \textbf{Problem 16}\par
Find the intersection of the three planes
\[
\begin{cases}
x+y+z=3,\\
x-y+z=1,\\
2x+z=2.
\end{cases}
\]

\item \textbf{Problem 17}\par
A cinema sells adult tickets for \(12\) units and student tickets for \(8\) units. In one evening it sells \(180\) tickets for a total of \(1840\) units. Form and solve a linear system to find the number of each type of ticket.

\item \textbf{Problem 18}\par
A system reduces to a row
\[
\begin{pmatrix}0&0&0\mid 6\end{pmatrix}.
\]
What does this row mean, and what can be concluded about the original system?

\item \textbf{Problem 19}\par
Construct a system of two linear equations in two variables that has:
\begin{enumerate}
\item exactly one solution;
\item no solution;
\item infinitely many solutions.
\end{enumerate}
Explain how the coefficients produce each case.

\item \textbf{Problem 20}\par
Consider
\[
\begin{cases}
x+ay=1,\\
ax+y=1.
\end{cases}
\]
Classify the system for every real value of \(a\). For each case, determine whether the solution is unique, nonexistent, or non-unique, and find the solution set where possible.
\end{enumerate}
