\section{Problem Set 1: Matrices}

\begin{enumerate}
\item \textbf{Problem 1}\par
For
\[
A=\begin{pmatrix}
2&-1&4\\
0&3&5
\end{pmatrix},
\]
state the dimensions of \(A\) and identify \(a_{12}\), \(a_{21}\), and \(a_{23}\).

\item \textbf{Problem 2}\par
Construct the \(3\times 2\) matrix \(B=(b_{ij})\) defined by
\[
b_{ij}=2i-j.
\]

\item \textbf{Problem 3}\par
Let
\[
A=\begin{pmatrix}1&2\\-3&4\end{pmatrix},
\qquad
B=\begin{pmatrix}5&-1\\2&0\end{pmatrix}.
\]
Compute \(A+B\), \(A-B\), \(3A\), and \(2A-B\).

\item \textbf{Problem 4}\par
Compute the transpose of
\[
C=\begin{pmatrix}
1&0&-2\\
4&3&5
\end{pmatrix}.
\]
Then state the dimensions of \(C^{\mathsf T}\).

\item \textbf{Problem 5}\par
Determine which of the following matrices are symmetric:
\[
A=\begin{pmatrix}2&-1\\-1&3\end{pmatrix},
\qquad
B=\begin{pmatrix}0&2\\-2&0\end{pmatrix},
\qquad
C=\begin{pmatrix}1&0\\0&1\end{pmatrix}.
\]

\item \textbf{Problem 6}\par
Let \(A\) be \(2\times3\), \(B\) be \(3\times4\), and \(C\) be \(4\times2\). Determine which of the products \(AB\), \(BA\), \(BC\), \(CB\), \(AC\), and \(CA\) are defined. State the dimensions of every defined product.

\item \textbf{Problem 7}\par
Compute
\[
AB
\quad\text{for}\quad
A=\begin{pmatrix}1&2&-1\\0&3&4\end{pmatrix},
\qquad
B=\begin{pmatrix}2&1\\-1&0\\3&-2\end{pmatrix}.
\]

\item \textbf{Problem 8}\par
For
\[
A=\begin{pmatrix}1&2\\0&1\end{pmatrix},
\qquad
B=\begin{pmatrix}2&0\\1&3\end{pmatrix},
\]
compute both \(AB\) and \(BA\). Explain what the result shows about matrix multiplication.

\item \textbf{Problem 9}\par
Let
\[
A=\begin{pmatrix}2&-1\\3&4\end{pmatrix},
\qquad
x=\begin{pmatrix}5\\-2\end{pmatrix}.
\]
Compute \(Ax\) and interpret the result as a new column vector.

\item \textbf{Problem 10}\par
Verify directly that
\[
AI_2=I_2A=A
\]
for
\[
A=\begin{pmatrix}3&-2\\1&5\end{pmatrix}.
\]

\item \textbf{Problem 11}\par
Let
\[
D=\operatorname{diag}(2,-1,3).
\]
Compute \(D^2\) and \(D^3\). Describe the pattern.

\item \textbf{Problem 12}\par
Starting with
\[
A=\begin{pmatrix}
1&2&-1\\
3&0&4\\
-2&1&5
\end{pmatrix},
\]
perform the row operation \(R_2\leftarrow R_2-3R_1\).

\item \textbf{Problem 13}\par
Starting with
\[
A=\begin{pmatrix}
1&2\\
3&4
\end{pmatrix},
\]
first interchange \(R_1\) and \(R_2\), and then perform \(R_2\leftarrow R_2-\frac13R_1\). Write the matrix after each step.

\item \textbf{Problem 14}\par
Find \(x,y,z,w\) if
\[
\begin{pmatrix}
x&2y\\
z-1&w+3
\end{pmatrix}
=
\begin{pmatrix}
4&6\\
2&1
\end{pmatrix}.
\]

\item \textbf{Problem 15}\par
Find the matrix \(X\) satisfying
\[
2X+\begin{pmatrix}1&-1\\0&3\end{pmatrix}
=
\begin{pmatrix}5&3\\4&-1\end{pmatrix}.
\]

\item \textbf{Problem 16}\par
Verify the distributive law \(A(B+C)=AB+AC\) for
\[
A=\begin{pmatrix}1&2\\-1&0\end{pmatrix},
\quad
B=\begin{pmatrix}2&1\\0&3\end{pmatrix},
\quad
C=\begin{pmatrix}-1&4\\2&0\end{pmatrix}.
\]

\item \textbf{Problem 17}\par
The columns of
\[
A=\begin{pmatrix}
1&0\\
2&1\\
-1&3
\end{pmatrix}
\]
are vectors \(u\) and \(v\). Write \(u\) and \(v\), and express
\[
A\begin{pmatrix}2\\-1\end{pmatrix}
\]
as a linear combination of these columns.

\item \textbf{Problem 18}\par
The matrix
\[
T=\begin{pmatrix}2&0\\0&-1\end{pmatrix}
\]
acts on points in the plane by matrix multiplication. Find the images of
\[
(1,2)^{\mathsf T},\quad (-3,1)^{\mathsf T},\quad (0,4)^{\mathsf T}.
\]
Describe what \(T\) does geometrically.

\item \textbf{Problem 19}\par
A shop records sales of two products on three days in the matrix
\[
S=\begin{pmatrix}
12&8\\
15&11\\
9&14
\end{pmatrix}.
\]
Explain what the rows and columns could represent. Compute the total sales of each product using matrix or vector notation.

\item \textbf{Problem 20}\par
For each expression below, decide whether it is defined. If it is defined, compute it; if it is not, explain the dimension mismatch:
\[
A=\begin{pmatrix}1&2&3\\0&1&-1\end{pmatrix},
\quad
B=\begin{pmatrix}2&0\\1&4\\-2&3\end{pmatrix},
\quad
u=\begin{pmatrix}1\\-1\\2\end{pmatrix}.
\]
Consider \(AB\), \(BA\), \(Au\), \(Bu\), and \(u^{\mathsf T}B\).
\end{enumerate}
