\section{Problem Set 6: Lines}

\begin{enumerate}
\item \textbf{Problem 1}\par
Write a vector-parametric equation of the line through
\[
A=(2,-1)
\]
with direction vector
\[
v=(3,4).
\]

\item \textbf{Problem 2}\par
Find a parametric equation of the line through
\[
A=(-1,2)
\quad\text{and}\quad
B=(5,-4).
\]

\item \textbf{Problem 3}\par
Determine whether the point \(P=(7,3)\) lies on the line
\[
(x,y)=(1,-1)+t(2,1).
\]

\item \textbf{Problem 4}\par
Write the coordinate parametric equations of
\[
P=(2,3)+t(-1,4).
\]
Find the points corresponding to \(t=-2\), \(t=0\), and \(t=3\).

\item \textbf{Problem 5}\par
Find the point-slope equation of the line through \((3,-2)\) with slope \(5\).

\item \textbf{Problem 6}\par
Convert
\[
y-4=-2(x+1)
\]
to slope-intercept form and general form.

\item \textbf{Problem 7}\par
Write equations for:
\begin{enumerate}
\item the vertical line through \((3,-1)\);
\item the horizontal line through \((-2,5)\).
\end{enumerate}

\item \textbf{Problem 8}\par
Convert the line
\[
(x,y)=(1,2)+t(3,-1)
\]
to general form.

\item \textbf{Problem 9}\par
Convert
\[
2x-5y+10=0
\]
to a parametric equation.

\item \textbf{Problem 10}\par
Determine whether the lines
\[
\ell_1:(x,y)=(1,0)+t(2,3),
\qquad
\ell_2:(x,y)=(-1,4)+s(4,6)
\]
are distinct parallel lines or the same line.

\item \textbf{Problem 11}\par
Determine whether
\[
\ell_1:2x-y+3=0
\quad\text{and}\quad
\ell_2:x+2y-4=0
\]
are perpendicular.

\item \textbf{Problem 12}\par
Find the intersection point of
\[
3x-y-5=0,
\qquad
x+2y-8=0.
\]

\item \textbf{Problem 13}\par
Find the distance from \(P=(4,-1)\) to the line
\[
2x+3y-6=0.
\]

\item \textbf{Problem 14}\par
For the oriented line
\[
x-2y+1=0,
\]
determine on which side of the line the points \(P=(0,0)\) and \(Q=(4,1)\) lie. Are they on the same side?

\item \textbf{Problem 15}\par
Write a parametric equation of the line in space through
\[
A=(1,-2,3)
\]
with direction vector
\[
v=(2,0,-1).
\]

\item \textbf{Problem 16}\par
Convert
\[
(x,y,z)=(2,1,-3)+t(4,-2,5)
\]
to symmetric form.

\item \textbf{Problem 17}\par
Write an appropriate symmetric description of
\[
(x,y,z)=(1,-2,3)+t(4,0,-1),
\]
taking proper account of the zero component.

\item \textbf{Problem 18}\par
Classify the following pair of lines in space as parallel, intersecting, or skew:
\[
\ell_1:(x,y,z)=(0,0,0)+t(1,1,0),
\]
\[
\ell_2:(x,y,z)=(1,0,1)+s(0,1,1).
\]

\item \textbf{Problem 19}\par
A line in the plane is given by a point \(A=(2,1)\) and a normal vector \(n=(3,-4)\). Write its normal equation, general equation, and one direction vector.

\item \textbf{Problem 20}\par
The line passes through \(A=(-1,3)\) and \(B=(5,-1)\). Find:
\begin{enumerate}
\item a direction vector;
\item a parametric equation;
\item a normal vector;
\item a general equation;
\item its slope, if defined;
\item the distance from the origin to the line.
\end{enumerate}
\end{enumerate}
