\section{Problem Set 7: Planes}

\begin{enumerate}
\item \textbf{Problem 1}\par
Write the normal equation and general equation of the plane through
\[
P_0=(1,-2,3)
\]
with normal vector
\[
n=(2,1,-1).
\]

\item \textbf{Problem 2}\par
For the plane
\[
3x-2y+4z-7=0,
\]
state a normal vector and find one convenient point on the plane.

\item \textbf{Problem 3}\par
Determine whether \(P=(2,1,-1)\) lies on
\[
x+2y-z-5=0.
\]

\item \textbf{Problem 4}\par
Write a parametric equation of the plane through
\[
P_0=(1,0,2)
\]
with direction vectors
\[
u=(1,2,0),
\qquad
v=(0,1,3).
\]

\item \textbf{Problem 5}\par
Find a parametric equation of the plane through
\[
A=(1,0,0),\quad B=(0,2,0),\quad C=(0,0,3).
\]

\item \textbf{Problem 6}\par
Find a normal vector to the plane through
\[
A=(1,0,1),\quad B=(3,1,0),\quad C=(0,2,4).
\]

\item \textbf{Problem 7}\par
Find the general equation of the plane through the three points in the previous problem.

\item \textbf{Problem 8}\par
Convert the parametric plane
\[
P=(1,2,0)+s(1,0,1)+t(0,2,-1)
\]
to general form.

\item \textbf{Problem 9}\par
Convert
\[
2x-y+z-4=0
\]
to one possible parametric form.

\item \textbf{Problem 10}\par
Determine whether the planes
\[
\pi_1:x-2y+3z-1=0,
\qquad
\pi_2:2x-4y+6z+5=0
\]
are distinct parallel planes or the same plane.

\item \textbf{Problem 11}\par
Determine whether
\[
\pi_1:x+y-z=0,
\qquad
\pi_2:2x-y+z-3=0
\]
are perpendicular.

\item \textbf{Problem 12}\par
Find the acute angle between
\[
\pi_1:x+2y+2z=0
\quad\text{and}\quad
\pi_2:2x-y+2z-4=0.
\]

\item \textbf{Problem 13}\par
Find the distance from
\[
P=(3,-1,2)
\]
to the plane
\[
2x-y+2z-7=0.
\]

\item \textbf{Problem 14}\par
Find the intersection point of the line
\[
(x,y,z)=(1,0,2)+t(2,-1,1)
\]
with the plane
\[
x+y+z-6=0.
\]

\item \textbf{Problem 15}\par
Determine whether the line
\[
(x,y,z)=(1,2,0)+t(1,-1,2)
\]
lies in, is parallel to, or intersects the plane
\[
2x+2y-z-6=0.
\]

\item \textbf{Problem 16}\par
Find the equation of the plane through \(P=(2,-1,4)\) parallel to
\[
3x-y+2z+5=0.
\]

\item \textbf{Problem 17}\par
Find a plane through \(P=(1,2,3)\) that is perpendicular to the plane
\[
x-y+z=0.
\]
Give one valid answer and explain why it works.

\item \textbf{Problem 18}\par
Find a parametric equation of the line of intersection of
\[
x+y+z=3,
\qquad
x-y+z=1.
\]

\item \textbf{Problem 19}\par
A plane contains the line
\[
\ell:(x,y,z)=(1,0,2)+t(2,1,-1)
\]
and the point \(Q=(0,3,1)\). Find a general equation of the plane.

\item \textbf{Problem 20}\par
For
\[
A=(1,0,0),\quad B=(0,1,0),\quad C=(0,0,1),
\]
find a parametric and a general equation of the plane through the points, verify that all three points satisfy the equation, find the distance from the origin to the plane, and determine where the line
\[
(x,y,z)=t(1,1,1)
\]
intersects it.
\end{enumerate}
