\section{Problem Set 12: Applications of Calculus}

\begin{enumerate}
\item \textbf{Problem 1}\par
For each question, state whether the main tool is a derivative, a definite integral, optimisation, or a differential model:
\begin{enumerate}
\item find instantaneous velocity;
\item find total water discharged from a known flow rate;
\item choose dimensions giving the largest area;
\item model a population whose growth rate is proportional to its size.
\end{enumerate}

\item \textbf{Problem 2}\par
A rectangle has perimeter \(40\). Express its area as a function of one side and find the dimensions that maximise the area.

\item \textbf{Problem 3}\par
A rectangular enclosure is built beside a straight wall, so only three sides require fencing. If \(60\) metres of fencing are available, find the dimensions that maximise the area.

\item \textbf{Problem 4}\par
Squares of side \(x\) are cut from the corners of a \(20\text{ cm}\times30\text{ cm}\) sheet to form an open box. Build the volume function, determine the meaningful domain, and find the value of \(x\) that maximises the volume.

\item \textbf{Problem 5}\par
Find the area enclosed by
\[
y=4-x^2
\]
and the \(x\)-axis.

\item \textbf{Problem 6}\par
The curve
\[
y=\sqrt{x},\qquad 0\le x\le9,
\]
is rotated about the \(x\)-axis. Find the resulting volume.

\item \textbf{Problem 7}\par
Set up, but do not necessarily evaluate, the arc-length integral for
\[
y=x^2,\qquad 0\le x\le2.
\]

\item \textbf{Problem 8}\par
A particle has position
\[
s(t)=t^3-6t^2+9t,\qquad t\ge0.
\]
Find its velocity and acceleration.

\item \textbf{Problem 9}\par
For the particle in the previous problem, determine when it is at rest and on which intervals it is speeding up or slowing down.

\item \textbf{Problem 10}\par
A particle has acceleration
\[
a(t)=6t-4,
\]
initial velocity \(v(0)=3\), and initial position \(s(0)=2\). Find \(v(t)\) and \(s(t)\).

\item \textbf{Problem 11}\par
A spring obeys Hooke's law \(F(x)=40x\) newtons. Find the work required to stretch it from \(x=0.1\) m to \(x=0.25\) m.

\item \textbf{Problem 12}\par
Water flows into a tank at the non-negative rate
\[
r(t)=12-2t
\]
litres per minute for \(0\le t\le6\). How much water enters during the first four minutes?

\item \textbf{Problem 13}\par
The temperature over a six-hour interval is modelled by
\[
T(t)=18+2t-\frac{t^2}{3},\qquad 0\le t\le6.
\]
Find its average value on the interval.

\item \textbf{Problem 14}\par
A population satisfies
\[
N'(t)=0.04N(t),
\qquad
N(0)=500.
\]
Find \(N(t)\) and the doubling time.

\item \textbf{Problem 15}\par
A radioactive substance satisfies
\[
M'(t)=-0.08M(t),
\qquad
M(0)=120.
\]
Find \(M(t)\) and its half-life.

\item \textbf{Problem 16}\par
A quantity \(Q(t)\) has rate of change
\[
Q'(t)=3t^2-4t+1
\]
and \(Q(0)=7\). Find \(Q(t)\) and explain the role of the initial condition.

\item \textbf{Problem 17}\par
A closed cylindrical can must have fixed volume \(V\). Derive the relation between radius and height that minimises its surface area.

\item \textbf{Problem 18}\par
A factory is \(400\) m downstream on the opposite side of a river \(80\) m wide. Underwater cable costs three times as much per metre as land cable. Let \(x\) be the downstream distance from the point directly opposite the station to the landing point. Build the cost function and determine the optimal \(x\).

\item \textbf{Problem 19}\par
A car's velocity is
\[
v(t)=t^2-6t+8,\qquad 0\le t\le6.
\]
Find the signed displacement and the total distance travelled on the interval.

\item \textbf{Problem 20}\par
A company models daily profit by
\[
P(x)=-2x^3+30x^2-96x-50,
\qquad 0\le x\le10,
\]
where \(x\) is production in hundreds of units. Determine the production levels at which profit is locally maximal or minimal, compare all relevant endpoint and critical values, and interpret the economically best choice within the model.
\end{enumerate}
