\section{Introduction}

The Cartesian coordinate system is something students have already encountered many times, both in primary school and in secondary school. They have learned how to draw the axes and how to use coordinates, but usually without a constructive explanation of where this system comes from.

The main purpose of this chapter is to introduce the Cartesian coordinate system constructively. We will begin with the basic objects and constructions provided by Euclidean geometry and show how they can be used to build a coordinate system.

\begin{center}
\begin{tikzpicture}[x=1cm,y=1cm]
    \draw[->,line width=0.8pt] (0,0) -- (11,0);

    \fill (1,0) circle (2pt);
    \node[above=5pt,align=center] at (1,0) {c. 300 BC\\Euclid's \emph{Elements}};

    \fill (9.5,0) circle (2pt);
    \node[above=5pt,align=center] at (9.5,0) {1637\\Descartes' \emph{La G\'eom\'etrie}};

    \draw[<->] (1,-0.55) -- (9.5,-0.55);
    \node[below=4pt] at (5.25,-0.55) {about 1900 years};
\end{tikzpicture}
\end{center}

We will focus mainly on the plane and on low-dimensional spaces, because they are easy to draw, study, and grasp as a whole. This makes them the natural setting in which to build geometric intuition and understand the central ideas. At the same time, many of the constructions and descriptions considered here extend easily to higher-dimensional structures once they are expressed in Cartesian coordinates and, later, in vector notation. We will nevertheless remain close to drawable and directly accessible situations so that the underlying ideas stay visible.

We will then look at elementary ways of using Cartesian coordinates to describe geometric objects and their relations. This will only be the beginning of a language and a collection of methods that will be developed further in later material.