\section{Review Exercises}

\subsection{Elementary computational and graphical exercises}

\begin{enumerate}
\item For $f(x)=x^2-4x+3$, compute $f(-1)$, $f(0)$, $f(2)$, and $f(5)$. Find the zeros and determine where the function is positive or negative.

\item Determine the natural domain of
\[
f(x)=\frac1{x-3},
\qquad
g(x)=\sqrt{x+2},
\qquad
h(x)=\frac{\sqrt{5-x}}{x+1}.
\]

\item Prepare a value table for $f(x)=|x-1|$ at $x=-2,-1,0,1,2,3,4$ and sketch the graph.

\item Starting from $y=x^2$, sketch
\[
y=(x-2)^2+1,
\qquad
y=-x^2,
\qquad
y=2x^2,
\]
and state each transformation.

\item Let $f(x)=2x-1$ and $g(x)=x^2+3$. Compute $f\circ g$, $g\circ f$, and both values at $x=2$.

\item Find the inverse of $f(x)=3x-5$ and verify both compositions.

\item An arithmetic sequence has $a_1=4$ and $d=-3$. Find $a_2,a_3,a_{10}$ and an explicit formula.

\item A geometric sequence has $b_1=2$ and $q=\frac12$. Find $b_2,b_3,b_6$ and an explicit formula.

\item For $a_1=1$ and $a_{n+1}=2a_n+1$, compute the first six terms and plot $(n,a_n)$.

\item Decide whether each sequence is increasing, decreasing, bounded above, or bounded below:
\[
\frac1n,
\qquad n-3,
\qquad(-1)^n,
\qquad2-\frac1n.
\]
Justify each answer briefly.
\end{enumerate}

\subsection{Development and reasoning exercises}

\begin{enumerate}
\item For
\[
f(x)=\frac{x+1}{x-2},
\]
determine the domain, make a small value table on both sides of $x=2$, find the zero and axis intercepts, and sketch the main features. Explain which information came from algebra and which from the table.

\item Let
\[
f(x)=x^2-4,
\qquad
g(x)=2x+1.
\]
Find the domains and formulas for $f\circ g$ and $g\circ f$, solve $(f\circ g)(x)=0$, and explain why the order of composition matters.

\item The function $f(x)=x^2$ is not one-to-one on $\mathbb R$. Restrict its domain in two different natural ways, find the corresponding inverse functions, and compare their graphs.

\item A function is described by
\[
f(x)=|x-2|-1.
\]
Produce a table, graph, zeros, sign intervals, minimum value, and a description of the transformations from $|x|$.

\item The arithmetic sequence satisfies $a_4=11$ and $a_{10}=29$. Find $a_1$, the common difference, and the explicit formula. Explain how two known terms determine the whole sequence.

\item The geometric sequence satisfies $b_2=6$ and $b_5=162$. Find all real possibilities for the ratio and determine the corresponding explicit formulas.

\item Starting from the recursion
\[
a_{n+1}=a_n+3,
\qquad a_1=2,
\]
derive the explicit formula. Then prove it by induction and explain the roles of the recursive and explicit descriptions.

\item Prove directly that $a_n=2-1/n$ is increasing and bounded above. Identify the inequalities that express direction and confinement.

\item Explain why a sequence is a function on a discrete domain. Compare a graph of isolated sequence points with the graph of a real function and state which operations or questions remain meaningful in both settings.

\item Give a complete analysis of one elementary function and one elementary sequence: specify their domains, representations, transformations or recurrence, monotonicity, boundedness, and the long-term question that leads naturally to limits.
\end{enumerate}
