\section{Variation of the Action Under a Symmetry}
\label{sec:variation_action_symmetry}

Noether's theorem follows from a variational identity. The identity separates the change of the action into a term proportional to the Euler--Lagrange equations and a spacetime divergence. Once this separation is visible, a symmetry immediately produces a conserved current.

\subsection{General field-theory action}

Consider \(N\) fields \(\phi_a\), with

\[
a=1,\ldots,N,
\]

and action

\[
\boxed{
S[\phi]
=
\int_{\Omega}
\mathcal{L}
\bigl(
\phi_a,
\partial_{\mu}\phi_a,
x
\bigr)
\,\mathrm{d}^{4}x
}.
\]

Define the canonical derivative momenta

\[
\boxed{
\pi_a^{\mu}
=
\frac{\partial\mathcal{L}}
{\partial(\partial_{\mu}\phi_a)}
}.
\]

The Euler--Lagrange expressions are

\[
\boxed{
\mathcal{E}_a
=
\frac{\partial\mathcal{L}}
{\partial\phi_a}
-
\partial_{\mu}\pi_a^{\mu}
}.
\]

Physical fields satisfy

\[
\mathcal{E}_a=0.
\]

\subsection{Infinitesimal transformation}

Let

\[
x'^{\mu}
=
x^{\mu}
+
\varepsilon\xi^{\mu}(x),
\]

and

\[
\phi'_a(x')
=
\phi_a(x)
+
\varepsilon\Psi_a(x,\phi).
\]

The characteristic is

\[
Q_a
=
\Psi_a
-
\xi^{\nu}\partial_{\nu}\phi_a,
\]

so the fixed-point field variation is

\[
\delta_0\phi_a
=
\varepsilon Q_a.
\]

Because differentiation is linear,

\[
\delta_0
\left(
\partial_{\mu}\phi_a
\right)
=
\partial_{\mu}
\left(
\delta_0\phi_a
\right)
=
\varepsilon\partial_{\mu}Q_a.
\]

\subsection{Variation of the volume element}

The Jacobian matrix is

\[
\frac{\partial x'^{\mu}}
{\partial x^{\nu}}
=
\delta^{\mu}{}_{\nu}
+
\varepsilon
\partial_{\nu}\xi^{\mu}.
\]

For small \(\varepsilon\),

\[
\det
\left(
I+\varepsilon A
\right)
=
1+\varepsilon\operatorname{tr}A
+
O(\varepsilon^{2}).
\]

Therefore

\[
\boxed{
\mathrm{d}^{4}x'
=
\left(
1+
\varepsilon\partial_{\mu}\xi^{\mu}
\right)
\mathrm{d}^{4}x
+
O(\varepsilon^{2})
}.
\]

The change of the measure is essential for spacetime transformations.

\subsection{Fixed-point variation of the Lagrangian density}

At fixed coordinates,

\[
\delta_0\mathcal{L}
=
\frac{\partial\mathcal{L}}
{\partial\phi_a}
\delta_0\phi_a
+
\pi_a^{\mu}
\delta_0
\left(
\partial_{\mu}\phi_a
\right).
\]

Substituting

\[
\delta_0\phi_a
=
\varepsilon Q_a,
\]

gives

\[
\delta_0\mathcal{L}
=
\varepsilon
\left[
\frac{\partial\mathcal{L}}
{\partial\phi_a}
Q_a
+
\pi_a^{\mu}\partial_{\mu}Q_a
\right].
\]

Use the product rule:

\[
\pi_a^{\mu}\partial_{\mu}Q_a
=
\partial_{\mu}
\left(
\pi_a^{\mu}Q_a
\right)
-
\left(
\partial_{\mu}\pi_a^{\mu}
\right)Q_a.
\]

Hence

\[
\delta_0\mathcal{L}
=
\varepsilon
\left[
\mathcal{E}_aQ_a
+
\partial_{\mu}
\left(
\pi_a^{\mu}Q_a
\right)
\right].
\]

This is the field-theory analogue of integration by parts in the derivation of the Euler--Lagrange equations.

\subsection{Combining field and coordinate variations}

The transformed action is integrated over the transformed region. To first order, the change of the action can be written over the original region as

\[
\delta S
=
\int_{\Omega}
\left[
\delta_0\mathcal{L}
+
\varepsilon
\partial_{\mu}
\left(
\mathcal{L}\xi^{\mu}
\right)
\right]
\mathrm{d}^{4}x.
\]

The divergence term contains both the Jacobian change and the displacement of the integration domain.

Substituting the fixed-point variation gives

\[
\delta S
=
\varepsilon
\int_{\Omega}
\left[
\mathcal{E}_aQ_a
+
\partial_{\mu}
\left(
\pi_a^{\mu}Q_a
+
\mathcal{L}\xi^{\mu}
\right)
\right]
\mathrm{d}^{4}x.
\]

Therefore the fundamental variational identity is

\[
\boxed{
\delta S
=
\varepsilon
\int_{\Omega}
\left[
\mathcal{E}_aQ_a
+
\partial_{\mu}
\Theta^{\mu}
\right]
\mathrm{d}^{4}x
},
\]

where

\[
\boxed{
\Theta^{\mu}
=
\pi_a^{\mu}Q_a
+
\mathcal{L}\xi^{\mu}
}.
\]

\subsection{Transformations at fixed coordinates}

For an internal transformation,

\[
\xi^{\mu}=0.
\]

Then

\[
Q_a=\Psi_a,
\]

and the identity reduces to

\[
\boxed{
\delta S
=
\varepsilon
\int_{\Omega}
\left[
\mathcal{E}_a\Psi_a
+
\partial_{\mu}
\left(
\pi_a^{\mu}\Psi_a
\right)
\right]
\mathrm{d}^{4}x
}.
\]

This is the form used for shift symmetries and global phase transformations.

\subsection{Boundary-term symmetries}

A transformation need not satisfy

\[
\delta\mathcal{L}=0
\]

pointwise. It is sufficient that the total action variation be a divergence:

\[
\boxed{
\delta S
=
\varepsilon
\int_{\Omega}
\partial_{\mu}K^{\mu}
\,\mathrm{d}^{4}x
}.
\]

Comparing this with the variational identity gives

\[
\mathcal{E}_aQ_a
+
\partial_{\mu}
\left(
\Theta^{\mu}-K^{\mu}
\right)
=
0.
\]

The current suggested by the symmetry is therefore

\[
\boxed{
j^{\mu}
=
\pi_a^{\mu}Q_a
+
\mathcal{L}\xi^{\mu}
-
K^{\mu}
}.
\]

The next section will state the theorem precisely and analyze this identity on solutions of the field equations.

\subsection{Worked example: the shift variation}

For the massless scalar density

\[
\mathcal{L}
=
\frac{1}{2}
\partial_{\mu}\phi
\partial^{\mu}\phi,
\]

consider

\[
\delta_0\phi
=
\varepsilon.
\]

Then

\[
Q=1,
\qquad
\xi^{\mu}=0,
\qquad
K^{\mu}=0.
\]

The canonical momentum is

\[
\pi^{\mu}
=
\partial^{\mu}\phi.
\]

The variational identity becomes

\[
\delta S
=
\varepsilon
\int_{\Omega}
\left[
-\Box\phi
+
\partial_{\mu}\partial^{\mu}\phi
\right]
\mathrm{d}^{4}x
=
0.
\]

The two terms cancel identically. On shell, the divergence term alone vanishes:

\[
\partial_{\mu}
\left(
\partial^{\mu}\phi
\right)
=
\Box\phi
=
0.
\]

\subsection{Structural check: explicit coordinate dependence}

If

\[
\frac{\partial\mathcal{L}}
{\partial x^{\nu}}
\neq 0
\]

explicitly, a translation generally changes the action. This signals that the external background distinguishes spacetime points. Translation currents then need not be conserved.

For example, an externally prescribed source term

\[
J(x)\phi(x)
\]

breaks translation symmetry unless the source is transformed as part of the system or is constant in the relevant direction.

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item varying the field but forgetting the variation of its derivative;
    \item using \(\Psi_a\) where the characteristic \(Q_a\) is required;
    \item omitting the Jacobian contribution for coordinate transformations;
    \item discarding a total divergence before identifying the associated current;
    \item assuming that \(K^{\mu}=0\) for every symmetry;
    \item applying the field equations too early and thereby hiding the off-shell identity.
\end{enumerate}

\subsection{What has been established}

The variation of a field-theory action under a general infinitesimal transformation separates as

\[
\delta S
=
\varepsilon
\int_{\Omega}
\left[
\mathcal{E}_aQ_a
+
\partial_{\mu}
\left(
\pi_a^{\mu}Q_a+\mathcal{L}\xi^{\mu}
\right)
\right]
\mathrm{d}^{4}x.
\]

This identity is valid before the equations of motion are imposed. When the transformation changes the action only by a divergence, the difference between the two boundary terms is the Noether current.
