\section{Rotational and Lorentz Symmetry}
\label{sec:rotational_lorentz_symmetry}

Translations produce energy and momentum. Rotations and Lorentz transformations produce angular-momentum and boost currents. For scalar fields, the current is orbital. For fields with intrinsic spin, an additional spin current is required.

\subsection{Infinitesimal Lorentz transformations}

A Lorentz transformation preserves the Minkowski metric:

\[
\eta_{\rho\sigma}
\Lambda^{\rho}{}_{\mu}
\Lambda^{\sigma}{}_{\nu}
=
\eta_{\mu\nu}.
\]

Near the identity,

\[
\Lambda^{\mu}{}_{\nu}
=
\delta^{\mu}{}_{\nu}
+
\frac{1}{2}
\varepsilon
\omega^{\mu}{}_{\nu}.
\]

Metric preservation implies

\[
\boxed{
\omega_{\mu\nu}
=
-\omega_{\nu\mu}
}.
\]

The coordinate change is

\[
\boxed{
\delta x^{\mu}
=
\frac{1}{2}
\varepsilon
\omega^{\mu}{}_{\nu}x^{\nu}
}.
\]

The six independent components of \(\omega_{\mu\nu}\) describe three spatial rotations and three boosts.

\subsection{Scalar-field transformation}

A Lorentz scalar has no intrinsic component transformation:

\[
\Psi=0.
\]

Its fixed-point variation is

\[
\delta_0\phi
=
-\delta x^{\rho}\partial_{\rho}\phi.
\]

Therefore

\[
\boxed{
\delta_0\phi
=
-\frac{1}{2}
\varepsilon
\omega_{\rho\sigma}
\left(
x^{\sigma}\partial^{\rho}
-
x^{\rho}\partial^{\sigma}
\right)\phi
}.
\]

The operator in parentheses is the orbital Lorentz generator.

\subsection{Lorentz invariance of the scalar Lagrangian}

For

\[
\mathcal{L}
=
\frac{1}{2}
\partial_{\mu}\phi
\partial^{\mu}\phi
-
V(\phi),
\]

the potential is invariant because \(\phi\) is a scalar. The kinetic term is a complete Lorentz contraction:

\[
\partial_{\mu}\phi
\partial^{\mu}\phi.
\]

Hence the action is Lorentz invariant when the spacetime measure and integration region are transformed consistently.

\subsection{Angular-momentum current}

For a symmetric conserved energy-momentum tensor, define

\[
\boxed{
M^{\lambda\mu\nu}
=
x^{\mu}T^{\lambda\nu}
-
x^{\nu}T^{\lambda\mu}
}.
\]

The current is antisymmetric in its final two indices:

\[
M^{\lambda\mu\nu}
=
-
M^{\lambda\nu\mu}.
\]

Its divergence is

\[
\begin{aligned}
\partial_{\lambda}M^{\lambda\mu\nu}
&=
\partial_{\lambda}
\left(
x^{\mu}T^{\lambda\nu}
-
x^{\nu}T^{\lambda\mu}
\right)
\\
&=
T^{\mu\nu}
+
x^{\mu}\partial_{\lambda}T^{\lambda\nu}
-
T^{\nu\mu}
-
x^{\nu}\partial_{\lambda}T^{\lambda\mu}.
\end{aligned}
\]

If

\[
\partial_{\lambda}T^{\lambda\nu}=0
\]

and

\[
T^{\mu\nu}=T^{\nu\mu},
\]

then

\[
\boxed{
\partial_{\lambda}M^{\lambda\mu\nu}=0
}.
\]

\subsection{Conserved Lorentz charges}

Define

\[
\boxed{
J^{\mu\nu}
=
\frac{1}{c}
\int
M^{0\mu\nu}
\,\mathrm{d}^{3}x
}.
\]

The antisymmetry gives

\[
J^{\mu\nu}
=
-J^{\nu\mu}.
\]

The spatial components

\[
J^{ij}
\]

represent ordinary angular momentum. The mixed components

\[
J^{0i}
\]

represent boost charges.

Under suitable boundary conditions,

\[
\frac{\mathrm{d}J^{\mu\nu}}{\mathrm{d}t}=0.
\]

\subsection{Three-dimensional angular momentum}

The ordinary angular-momentum vector is

\[
\boxed{
L^{i}
=
\frac{1}{2}
\epsilon^{ijk}
J^{jk}
}.
\]

For a scalar field with momentum density \(\mathbf{g}\),

\[
\boxed{
\mathbf{L}
=
\int
\mathbf{x}\times\mathbf{g}
\,\mathrm{d}^{3}x
}.
\]

This is purely orbital angular momentum.

\subsection{Rotation about the z-axis}

For a rotation in the \(x\)-\(y\) plane,

\[
\delta x
=
-\varepsilon y,
\qquad
\delta y
=
\varepsilon x.
\]

The fixed-point variation of a scalar field is

\[
\boxed{
\delta_0\phi
=
-\varepsilon
\left(
-y\partial_x+x\partial_y
\right)\phi
}.
\]

The associated conserved charge is the \(z\)-component of angular momentum:

\[
L_z
=
\int
\left(
xg_y-yg_x
\right)
\,\mathrm{d}^{3}x.
\]

A spherically symmetric configuration has zero orbital angular momentum density because \(\nabla\phi\) is radial and cannot produce a tangential momentum flow by itself.

\subsection{Boost charges}

A boost mixes time with one spatial coordinate. For a boost in the \(x^i\) direction, the corresponding charge is

\[
J^{0i}
=
\frac{1}{c}
\int
\left(
x^{0}T^{0i}
-
x^{i}T^{00}
\right)
\,\mathrm{d}^{3}x.
\]

Since

\[
x^{0}=ct,
\]

this becomes

\[
J^{0i}
=
t
\int
T^{0i}
\,\mathrm{d}^{3}x
-
\frac{1}{c}
\int
x^{i}T^{00}
\,\mathrm{d}^{3}x.
\]

The conservation of \(J^{0i}\) relates the motion of the center of energy to the total momentum.

\subsection{Fields with intrinsic spin}

A vector or spinor field transforms not only because its argument changes but also because its components transform. Write

\[
\delta_0\phi_a
=
-\delta x^{\rho}\partial_{\rho}\phi_a
+
\frac{1}{2}
\varepsilon
\omega_{\mu\nu}
\left(
\Sigma^{\mu\nu}
\right)_a{}^b
\phi_b.
\]

The matrices \(\Sigma^{\mu\nu}\) represent the Lorentz generators in the field's internal component space.

The canonical angular-momentum current then has the form

\[
\boxed{
M_{\mathrm{can}}^{\lambda\mu\nu}
=
x^{\mu}T_{\mathrm{can}}^{\lambda\nu}
-
x^{\nu}T_{\mathrm{can}}^{\lambda\mu}
+
S^{\lambda\mu\nu}
},
\]

where \(S^{\lambda\mu\nu}\) is the spin current.

For a scalar field,

\[
S^{\lambda\mu\nu}=0.
\]

\subsection{Why a Symmetric Energy--Momentum Tensor Matters}

From

\[
\partial_{\lambda}
\left(
x^{\mu}T^{\lambda\nu}
-
x^{\nu}T^{\lambda\mu}
\right)
=
T^{\mu\nu}-T^{\nu\mu},
\]

we see that the purely orbital current is conserved only when the energy-momentum tensor is symmetric.

For spinning fields, the antisymmetric part of the canonical tensor is balanced by the divergence of the spin current. An improved symmetric tensor combines orbital and spin contributions into a form particularly suitable for relativistic mechanics and gravity.

\subsection{Structural check: rotationally invariant potentials}

A scalar potential

\[
V(\phi)
\]

does not depend on spatial direction and therefore does not break rotational symmetry.

An external term such as

\[
J(\mathbf{x})\phi
\]

preserves rotations only if

\[
J(R\mathbf{x})
=
J(\mathbf{x})
\]

for the relevant rotations. A source localized away from the origin generally breaks rotational symmetry about the origin.

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item forgetting the antisymmetry \(\omega_{\mu\nu}=-\omega_{\nu\mu}\);
    \item treating boosts as ordinary spatial rotations;
    \item assuming that the canonical energy-momentum tensor is always symmetric;
    \item omitting the spin current for vector or spinor fields;
    \item identifying \(J^{0i}\) with ordinary angular momentum;
    \item claiming rotational conservation when an external source selects a preferred direction.
\end{enumerate}

\subsection{What has been established}

Lorentz symmetry produces the conserved current

\[
M^{\lambda\mu\nu}
=
x^{\mu}T^{\lambda\nu}
-
x^{\nu}T^{\lambda\mu}
\]

for a scalar field with symmetric energy-momentum tensor. Spatial components give angular momentum, while mixed time-space components give boost charges. The final section applies Noether's theorem to an internal global phase symmetry of a complex scalar field.
