\subsection{Computational and graphical exercises}

\begin{enumerate}

\item For the circle
\[
x^2+y^2=16,
\]
write a trigonometric parameterisation and a polar equation centred at the
circle centre. Mark the points corresponding to parameters
$0,\pi/2,\pi,3\pi/2$.

\item For the ellipse
\[
\frac{x^2}{25}+\frac{y^2}{9}=1,
\]
find $a$, $b$, $c$, both foci, and the eccentricity. Draw a complete sketch
with the semiaxes and foci marked.

\item A parabola has focus $F=(2,0)$ and directrix $x=-2$. Derive its Cartesian
equation, find its vertex, and draw the equal-distance construction for one
point of the curve.

\item For the hyperbola
\[
\frac{x^2}{16}-\frac{y^2}{9}=1,
\]
find the vertices, foci, eccentricity, and asymptotes. Draw both branches and
mark the centre, vertices, and foci.

\item Convert the points
\[
(\sqrt3,1),\qquad (-1,\sqrt3)
\]
from Cartesian to polar coordinates, and convert
\[
(r,\theta)=(4,\pi/6),\qquad (3,\pi)
\]
from polar to Cartesian coordinates. Plot all four points.

\item Convert
\[
r=6\cos\theta
\]
to Cartesian form. Identify the centre and radius, sketch the circle, and
explain from the equation why it passes through the pole.

\item For
\[
r=\frac{6}{1+\frac12\cos\theta},
\]
identify $p$ and $e$, classify the conic, and compute its nearest and farthest
focal distances. Mark these points on a sketch.

\item Classify and sketch the three focal conics
\[
r=\frac{3}{1+0.4\cos\theta},
\qquad
r=\frac{3}{1+\cos\theta},
\qquad
r=\frac{3}{1+1.5\cos\theta}.
\]
For each curve, mark the focus and the direction $\theta=0$.

\item Sketch
\[
r=\frac{4}{1+0.6\cos(\theta-\pi/4)}.
\]
Mark the focus, the periapsis direction, the nearest point, and the farthest
point. State how the angle shift changes the picture.

\item Consider the hyperbolic polar equation
\[
r=\frac{4}{1+2\cos\theta}.
\]
Find the angles at which the denominator vanishes, determine the interval of
angles for which $r\geq0$, and sketch the traced branch together with its two
asymptotic directions.

\end{enumerate}
