\subsection{Development and reasoning exercises}

\begin{enumerate}

\item Draw one point in both Cartesian and polar coordinates. Derive
\[
x=r\cos\theta,
\qquad
y=r\sin\theta,
\qquad
r=\sqrt{x^2+y^2}.
\]
Explain why the polar description is better adapted to a problem with one
distinguished focus.

\item The same circle can be described by
\[
x^2+y^2=R^2
\]
and by
\[
r=R.
\]
Explain why the polar equation is simpler. Then move the centre away from the
pole and discuss what simplicity is lost.

\item Starting from the focus--directrix condition
\[
d(P,F)=e\,d(P,D),
\]
place the pole at $F$ and the directrix at $x=d$. Derive
\[
r=\frac{p}{1+e\cos\theta},
\qquad p=ed,
\]
using a fully labelled geometric diagram.

\item Fix $p=2$ and compare the ellipses obtained for
\[
e=0.2,\qquad e=0.6,\qquad e=0.9.
\]
Explain the changes using the denominator and the values of $r_{\min}$ and
$r_{\max}$ rather than only visual appearance.

\item Explain directly from
\[
r=\frac{p}{1+e\cos\theta}
\]
why $e=1$ is the transition between a closed ellipse and an open conic. Your
argument should discuss whether the denominator can vanish.

\item For $e>1$ and $r\geq0$, the focal polar equation naturally traces one
branch of a hyperbola. Explain why this does not mean that the other branch has
disappeared. Describe how the complete two-branch curve is represented more
naturally in Cartesian coordinates.

\item Compare the centre-based ellipse
\[
\frac{x^2}{a^2}+\frac{y^2}{b^2}=1
\]
with the focus-based equation
\[
r=\frac{a(1-e^2)}{1+e\cos\theta}.
\]
State which geometric information is immediate in each description and why
neither description is universally superior.

\item The parameterisation
\[
x=a\cos t,
\qquad y=b\sin t
\]
and the polar equation
\[
r=\frac{a(1-e^2)}{1+e\cos\theta}
\]
describe the same ellipse. Explain why $t$ is not generally the polar angle
$\theta$ measured from a focus. Support the explanation with a labelled
sketch.

\item An orbit has nearest focal distance $4$ and farthest focal distance $12$.
Determine $a$, $e$, and $p$, and write its polar equation with periapsis at
$\theta=0$. Explain how the two measured distances determine both the size and
the shape of the ellipse.

\item Explain separately the roles of $p$, $e$, and $\theta_0$ in
\[
r=\frac{p}{1+e\cos(\theta-\theta_0)}.
\]
Then summarise the chapter as a change of viewpoint: classical Cartesian
curves first, direction and distance in polar coordinates second, and one
focus--directrix law last. Explain why this sequence prepares a later study of
orbits in mechanics.

\end{enumerate}
