\section{Massless and Massive Field Equations}
\label{sec:massless_and_massive_field_equations}

The free real scalar field has two structurally important cases. When the mass parameter vanishes, the equation is the ordinary relativistic wave equation. When the mass parameter is nonzero, a local restoring term is added and the result is the classical Klein--Gordon equation.

The purpose of this section is to compare the two equations directly, identify their limiting cases, and show how the mass term changes both temporal and spatial behaviour.

\subsection{The massless equation}

For

\[
\mu=0,
\]

the free Lagrangian density is

\[
\boxed{
\mathcal{L}_{\mathrm{massless}}
=
\frac{1}{2}
\partial_{\mu}\phi\partial^{\mu}\phi
}.
\]

The field equation is

\[
\boxed{
\Box\phi=0
}.
\]

Using the chosen metric convention,

\[
\boxed{
\frac{1}{c^{2}}\phi_{tt}
-
\nabla^{2}\phi
=0
}.
\]

Equivalently,

\[
\boxed{
\phi_{tt}
=
c^{2}\nabla^{2}\phi
}.
\]

The field propagates disturbances with characteristic speed \(c\).

\subsection{The massive equation}

For

\[
\mu>0,
\]

the density is

\[
\boxed{
\mathcal{L}_{\mathrm{massive}}
=
\frac{1}{2}
\partial_{\mu}\phi\partial^{\mu}\phi
-
\frac{1}{2}\mu^{2}\phi^{2}
}.
\]

The equation is

\[
\boxed{
(\Box+\mu^{2})\phi=0
}.
\]

In expanded form,

\[
\boxed{
\frac{1}{c^{2}}\phi_{tt}
-
\nabla^{2}\phi
+
\mu^{2}\phi
=0
}.
\]

Multiplying by \(c^{2}\),

\[
\boxed{
\phi_{tt}
-
c^{2}\nabla^{2}\phi
+
c^{2}\mu^{2}\phi
=0
}.
\]

The local term

\[
c^{2}\mu^{2}\phi
\]

acts as a restoring force toward \(\phi=0\).

\subsection{Why the equation is called Klein--Gordon}

The equation

\[
(\Box+\mu^{2})\phi=0
\]

is called the Klein--Gordon equation. In quantum theory, the parameter is commonly written as

\[
\mu=rac{mc}{\hbar},
\]

so the equation becomes

\[
\left(
\Box+
\frac{m^{2}c^{2}}{\hbar^{2}}
\right)
\phi=0.
\]

Here it is treated entirely as a classical relativistic partial differential equation. The field amplitude is a classical real function, and no probabilistic interpretation is assigned to it.

\subsection{Spatially uniform fields}

Suppose the field is independent of position:

\[
\phi(\mathbf{x},t)=q(t).
\]

Then

\[
\nabla^{2}\phi=0.
\]

For the massless theory,

\[
q''(t)=0,
\]

so

\[
\boxed{
q(t)=A+Bt
}.
\]

A spatially uniform massless field may remain constant or change linearly in time.

For the massive theory,

\[
q''(t)+c^{2}\mu^{2}q(t)=0.
\]

Thus

\[
\boxed{
q(t)
=
A\cos(c\mu t)
+
B\sin(c\mu t)
}.
\]

The natural angular frequency of the uniform mode is

\[
\boxed{
\omega_{0}=c\mu
}.
\]

The mass term therefore permits oscillation even when the field has no spatial variation.

\subsection{Static fields}

Suppose the field is time independent:

\[
\phi(\mathbf{x},t)=\Phi(\mathbf{x}).
\]

Then

\[
\phi_{tt}=0.
\]

The massless equation becomes

\[
\boxed{
\nabla^{2}\Phi=0
}.
\]

Thus static massless configurations satisfy the Laplace equation.

The massive equation becomes

\[
\boxed{
\nabla^{2}\Phi
-
\mu^{2}\Phi
=0
}.
\]

This is the homogeneous modified Helmholtz equation. Its solutions typically decay or grow exponentially rather than varying linearly or harmonically in space.

\subsection{One-dimensional static solutions}

In one spatial dimension, the massive static equation is

\[
\frac{\mathrm{d}^{2}\Phi}{\mathrm{d}x^{2}}
-
\mu^{2}\Phi
=0.
\]

The characteristic equation is

\[
r^{2}-\mu^{2}=0,
\]

with roots

\[
r=\pm\mu.
\]

Therefore

\[
\boxed{
\Phi(x)
=
A e^{\mu x}
+
B e^{-\mu x}
}.
\]

On a half-line \(x\geq0\), a bounded solution at infinity requires

\[
A=0,
\]

so

\[
\boxed{
\Phi(x)=B e^{-\mu x}
}.
\]

The characteristic decay length is

\[
\boxed{
\ell_{\mu}=rac{1}{\mu}
}.
\]

The mass parameter therefore introduces a natural spatial length scale.

\subsection{Sourced equations}

An external scalar source \(J(x)\) may be included through

\[
\mathcal{L}
=
\frac{1}{2}
\partial_{\mu}\phi\partial^{\mu}\phi
-
\frac{1}{2}\mu^{2}\phi^{2}
+
J\phi.
\]

The source is prescribed, so

\[
\delta J=0.
\]

The field equation becomes

\[
\boxed{
(\Box+\mu^{2})\phi=J
}.
\]

In the massless case,

\[
\Box\phi=J.
\]

The source-free equation describes free propagation. The sourced equation describes a field generated or driven by external matter.

\subsection{Static point-source behaviour}

The static sourced massive equation has the form

\[
\nabla^{2}\Phi
-
\mu^{2}\Phi
=
-\rho(\mathbf{x}).
\]

Outside the source region,

\[
\rho=0,
\]

and a spherically symmetric solution in three dimensions behaves as

\[
\boxed{
\Phi(r)
\propto
\frac{e^{-\mu r}}{r}
}.
\]

The exponential factor suppresses the field at distances much larger than \(1/\mu\). In the limit

\[
\mu\to0,
\]

this reduces to the massless behaviour

\[
\Phi(r)
\propto
\frac{1}{r}.
\]

This comparison shows how a mass parameter can produce a finite interaction range in a sourced linear theory.

\subsection{Cauchy data}

Both equations are second order in time. To specify a solution, one generally prescribes the field and its first time derivative on an initial spatial surface:

\[
\boxed{
\phi(\mathbf{x},t_{0})
=
f(\mathbf{x})
}
\]

and

\[
\boxed{
\phi_{t}(\mathbf{x},t_{0})
=
g(\mathbf{x})
}.
\]

Boundary conditions may also be required when space is restricted to a finite region.

The mass term does not change the number of initial data. It changes the evolution generated from those data.

\subsection{Finite propagation speed}

The highest-derivative part of both equations is

\[
\Box\phi.
\]

Therefore the characteristic light cone is the same in the massive and massless theories. The lower-order mass term changes the dispersion of modes but does not permit signals to propagate outside the light cone.

This distinction is important:

\begin{quote}
The mass term changes mode propagation without changing the causal characteristic speed of the differential equation.
\end{quote}

\subsection{Symmetries of the free equations}

Both equations are invariant under spacetime translations and Lorentz transformations. They are also invariant under the discrete field reflection

\[
\boxed{
\phi\longrightarrow-\phi
}.
\]

For the massless free theory, the equation is additionally invariant under the constant shift

\[
\boxed{
\phi\longrightarrow\phi+C
}.
\]

Indeed, derivatives remove the constant. The mass term breaks this shift symmetry because

\[
\mu^{2}(\phi+C)
\neq
\mu^{2}\phi
\]

when \(C\neq0\).

\subsection{Superposition}

Both free equations are linear. If

\[
(\Box+\mu^{2})\phi_{1}=0
\]

and

\[
(\Box+\mu^{2})\phi_{2}=0,
\]

then

\[
\boxed{
\phi=a\phi_{1}+b\phi_{2}
}
\]

is also a solution.

This property allows the field to be expanded into normal modes. It will fail once nonlinear powers such as \(\phi^{3}\) appear in the equation.

\subsection{Limiting cases}

Several checks clarify the equations.

\begin{enumerate}
    \item Massless limit:
    \[
    \mu\to0
    \quad\Longrightarrow\quad
    (\Box+\mu^{2})\phi=0
    \longrightarrow
    \Box\phi=0.
    \]

    \item Spatially uniform limit:
    \[
    \nabla\phi=0
    \quad\Longrightarrow\quad
    \phi_{tt}+c^{2}\mu^{2}\phi=0.
    \]

    \item Static limit:
    \[
    \phi_{t}=0
    \quad\Longrightarrow\quad
    \nabla^{2}\phi-\mu^{2}\phi=0.
    \]

    \item Long-wavelength regime:
    spatial derivatives are small, so the local mass oscillation becomes dominant.

    \item Short-wavelength regime:
    gradient terms dominate, and massive modes approach massless wave behaviour.
\end{enumerate}

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item calling \(\Box\phi=0\) the massive Klein--Gordon equation;
    \item using \(-\mu^{2}\phi\) in the equation after choosing the mostly-minus signature and the stated Lagrangian;
    \item assuming that a nonzero mass term changes the characteristic light cone;
    \item confusing the static modified Helmholtz equation with the ordinary Helmholtz equation;
    \item forgetting that two initial functions are required for a second-order evolution equation;
    \item claiming that the massless shift symmetry survives when \(\mu\neq0\);
    \item interpreting the exponential static length scale as a propagation speed.
\end{enumerate}

\subsection{What has been established}

The massless real scalar field satisfies

\[
\Box\phi=0,
\]

while the massive field satisfies

\[
(\Box+\mu^{2})\phi=0.
\]

The mass parameter introduces a natural frequency \(c\mu\) and a natural length scale \(1/\mu\), while leaving the causal light cone determined by the principal wave operator unchanged. The next section constructs explicit plane-wave solutions and shows how their real combinations generate general free-field configurations.