\section{Interacting Scalar Fields and Potentials}
\label{sec:interacting_scalar_fields_and_potentials}

The free scalar theory is quadratic, so its equation is linear and its modes evolve independently. Interactions are introduced by replacing the quadratic mass potential with a more general function \(V(\phi)\), or by coupling several scalar fields in one potential.

The shape of the potential determines equilibrium values, nonlinear forces, stability, and the possible existence of spatially varying static configurations.

\subsection{General interacting action}

Consider the Lagrangian density

\[
\boxed{
\mathcal{L}
=
\frac{1}{2}
\partial_{\mu}\phi\partial^{\mu}\phi
-
V(\phi)
+
J(x)\phi
}.
\]

The field \(\phi\) is dynamical, while \(J(x)\) is a prescribed external source.

The action is

\[
S[\phi]
=
\int_{\Omega}
\mathcal{L}
\,\mathrm{d}^{4}x.
\]

The derivative term is Lorentz invariant because it is a complete contraction. The potential is also a Lorentz scalar because it depends only on the scalar field value.

\subsection{Variation of the potential}

The derivative contribution varies as

\[
\delta
\left(
\frac{1}{2}
\partial_{\mu}\phi\partial^{\mu}\phi
\right)
=
\partial_{\mu}\phi
\partial^{\mu}(\delta\phi).
\]

The potential contribution is

\[
\delta[-V(\phi)]
=
-V'(\phi)\delta\phi.
\]

The source contribution is

\[
\delta(J\phi)
=
J\delta\phi.
\]

After integration by parts and removal of the boundary term under fixed boundary values, stationarity gives

\[
\boxed{
\Box\phi
+
V'(\phi)
=
J
}.
\]

In components,

\[
\boxed{
\frac{1}{c^{2}}\phi_{tt}
-
\nabla^{2}\phi
+
V'(\phi)
=
J
}.
\]

The local force is determined by the derivative of the potential, not by the potential itself.

\subsection{Energy density}

The canonical energy density is

\[
\boxed{
\mathcal{E}
=
\frac{1}{2c^{2}}
\phi_{t}^{2}
+
\frac{1}{2}
|\nabla\phi|^{2}
+
V(\phi)
-
J\phi
}.
\]

When the source is time independent, this expression defines a conserved total energy under suitable boundary conditions. If \(J\) depends explicitly on time, the external system can supply or remove energy.

Adding a constant to the potential,

\[
V(\phi)
\longrightarrow
V(\phi)+V_{0},
\]

does not change the field equation because

\[
\frac{\mathrm{d}V_{0}}{\mathrm{d}\phi}=0.
\]

It does shift the absolute energy density by \(V_{0}\).

\subsection{Uniform equilibria}

For a constant, time-independent field

\[
\phi(x)=\phi_{0},
\]

all derivatives vanish. The field equation reduces to

\[
\boxed{
V'(\phi_{0})=J
}.
\]

In the source-free case,

\[
\boxed{
V'(\phi_{0})=0
}.
\]

Thus uniform equilibria are stationary points of the potential.

A stationary point is not automatically stable. The sign of

\[
V''(\phi_{0})
\]

determines the leading restoring behaviour and will be analyzed systematically in the next section.

\subsection{Quadratic potential as the free theory}

For

\[
V(\phi)
=
\frac{1}{2}\mu^{2}\phi^{2},
\]

we have

\[
V'(\phi)=\mu^{2}\phi.
\]

The equation becomes

\[
\Box\phi+\mu^{2}\phi=J.
\]

When \(J=0\), this is the free massive Klein--Gordon equation.

The free theory is therefore the special case in which the potential is purely quadratic.

\subsection{Quartic self-interaction}

Consider

\[
\boxed{
V(\phi)
=
\frac{1}{2}\mu^{2}\phi^{2}
+
\frac{\lambda}{4}\phi^{4}
}.
\]

Then

\[
V'(\phi)
=
\mu^{2}\phi
+
\lambda\phi^{3}.
\]

The source-free equation is

\[
\boxed{
\Box\phi
+
\mu^{2}\phi
+
\lambda\phi^{3}
=0
}.
\]

The cubic force term makes the equation nonlinear. If \(\phi_{1}\) and \(\phi_{2}\) are solutions, their sum is generally not a solution because

\[
(\phi_{1}+\phi_{2})^{3}
\neq
\phi_{1}^{3}+\phi_{2}^{3}.
\]

Different modes can exchange energy through the interaction.

\subsection{Boundedness of the quartic potential}

For large \(|\phi|\), the quartic term dominates:

\[
V(\phi)
\sim
\frac{\lambda}{4}\phi^{4}.
\]

If

\[
\lambda>0,
\]

the potential grows toward positive infinity and is bounded below.

If

\[
\lambda<0,
\]

the potential tends to negative infinity as \(|\phi|\to\infty\). The classical energy is then unbounded below, so the model has no stable ground configuration without additional higher-order terms.

\subsection{Double-well potential}

A central nonlinear example is

\[
\boxed{
V(\phi)
=
\frac{\lambda}{4}
(\phi^{2}-v^{2})^{2}
},
\]

where

\[
\lambda>0,
\qquad
v>0.
\]

The derivative is

\[
V'(\phi)
=
\lambda\phi
(\phi^{2}-v^{2}).
\]

The field equation is

\[
\boxed{
\Box\phi
+
\lambda\phi
(\phi^{2}-v^{2})
=0
}.
\]

The potential is invariant under

\[
\boxed{
\phi\longrightarrow-\phi
}.
\]

This is a discrete reflection symmetry.

\subsection{Equilibria of the double well}

Uniform equilibria satisfy

\[
\lambda\phi_{0}
(\phi_{0}^{2}-v^{2})=0.
\]

Therefore

\[
\boxed{
\phi_{0}=0,
\qquad
\phi_{0}=v,
\qquad
\phi_{0}=-v
}.
\]

The second derivative is

\[
V''(\phi)
=
\lambda(3\phi^{2}-v^{2}).
\]

At the origin,

\[
V''(0)
=
-\lambda v^{2}<0,
\]

so \(\phi=0\) is a local maximum.

At the outer equilibria,

\[
V''(\pm v)
=
2\lambda v^{2}>0,
\]

so \(\phi=\pm v\) are local minima.

The theory has two degenerate lowest-energy uniform configurations.

\subsection{Static interacting fields}

For a time-independent configuration,

\[
\phi(\mathbf{x},t)=\Phi(\mathbf{x}),
\]

the field equation becomes

\[
\boxed{
-\nabla^{2}\Phi
+
V'(\Phi)
=
J(\mathbf{x})
}.
\]

Equivalently,

\[
\nabla^{2}\Phi
=
V'(\Phi)-J.
\]

This is a nonlinear boundary-value problem when \(V'\) is nonlinear.

The static energy functional is

\[
\boxed{
E[\Phi]
=
\int
\left[
\frac{1}{2}|\nabla\Phi|^{2}
+
V(\Phi)
-
J\Phi
\right]
\,\mathrm{d}^{3}x
}.
\]

Its stationary points satisfy the static field equation.

\subsection{A one-dimensional transition solution}

For the double-well potential in one spatial dimension, a static field obeys

\[
\frac{\mathrm{d}^{2}\Phi}{\mathrm{d}x^{2}}
=
\lambda\Phi
(\Phi^{2}-v^{2}).
\]

A solution connecting the two minima is

\[
\boxed{
\Phi(x)
=
v\tanh
\left[
v\sqrt{
\frac{\lambda}{2}
}
(x-x_{0})
\right]
}.
\]

As

\[
x\to-\infty,
\]

we have

\[
\Phi(x)\to-v,
\]

while as

\[
x\to+\infty,
\]

we have

\[
\Phi(x)\to+v.
\]

The parameter \(x_{0}\) locates the center of the transition. This configuration is often called a kink or domain wall in one spatial dimension.

\subsection{Verification of the transition solution}

Let

\[
a
=
v\sqrt{
\frac{\lambda}{2}
}.
\]

Then

\[
\Phi(x)=v\tanh[a(x-x_{0})].
\]

The first derivative is

\[
\Phi'
=
va\operatorname{sech}^{2}[a(x-x_{0})].
\]

The second derivative is

\[
\Phi''
=
-2va^{2}
\tanh[a(x-x_{0})]
\operatorname{sech}^{2}[a(x-x_{0})].
\]

Using

\[
\operatorname{sech}^{2}z
=
1-	anh^{2}z
\]

and \(a^{2}=\lambda v^{2}/2\), one finds

\[
\Phi''
=
\lambda\Phi(\Phi^{2}-v^{2}).
\]

Thus the proposed field satisfies the static nonlinear equation.

\subsection{Coupled scalar fields}

Let two real fields \(\phi\) and \(\chi\) have density

\[
\mathcal{L}
=
\frac{1}{2}
\partial_{\mu}\phi\partial^{\mu}\phi
+
\frac{1}{2}
\partial_{\mu}\chi\partial^{\mu}\chi
-
V(\phi,\chi).
\]

Variation with respect to each field gives

\[
\boxed{
\Box\phi
+
\frac{\partial V}{\partial\phi}
=0
}
\]

and

\[
\boxed{
\Box\chi
+
\frac{\partial V}{\partial\chi}
=0
}.
\]

For the interaction

\[
V_{\mathrm{int}}
=
\frac{g}{2}
\phi^{2}\chi^{2},
\]

the equations contain

\[
\frac{\partial V_{\mathrm{int}}}{\partial\phi}
=
g\phi\chi^{2}
\]

and

\[
\frac{\partial V_{\mathrm{int}}}{\partial\chi}
=
g\phi^{2}\chi.
\]

Each field changes the effective restoring force acting on the other.

\subsection{External sources and effective potential}

For a constant source \(J\), define

\[
\boxed{
V_{\mathrm{eff}}(\phi)
=
V(\phi)-J\phi
}.
\]

Uniform equilibria satisfy

\[
V_{\mathrm{eff}}'(\phi_{0})=0.
\]

Thus a source tilts the potential and shifts the equilibrium. In a symmetric double well, a nonzero source generally makes one minimum energetically preferred over the other.

\subsection{Finite-energy conditions}

For a static configuration on an unbounded region, finite energy requires the field to approach a minimum of the potential at spatial infinity:

\[
\Phi(\mathbf{x})
\longrightarrow
\phi_{\mathrm{vac}}
\qquad
\text{as }|\mathbf{x}|\to\infty,
\]

where

\[
V'(\phi_{\mathrm{vac}})=0.
\]

It is convenient to shift the potential so that

\[
V(\phi_{\mathrm{vac}})=0.
\]

The gradient must also decay sufficiently rapidly for the total energy integral to converge.

\subsection{Common mistakes}

Typical errors include:

\begin{enumerate}
    \item placing \(V(\phi)\) rather than \(V'(\phi)\) in the field equation;
    \item assuming that superposition survives a nonlinear interaction;
    \item varying an external source that is meant to remain fixed;
    \item calling every stationary point of \(V\) stable;
    \item choosing a quartic coupling with negative leading coefficient and overlooking the unbounded energy;
    \item forgetting that adding a constant to \(V\) changes the energy but not the classical equation;
    \item solving a nonlinear static equation without imposing boundary behaviour.
\end{enumerate}

\subsection{What has been established}

A general interacting scalar field satisfies

\[
\Box\phi+V'(\phi)=J.
\]

The potential determines nonlinear forces, uniform equilibria, energy boundedness, and possible static transition configurations. Quartic and double-well potentials provide the basic examples, while mixed potentials couple several scalar fields. The next section develops the systematic approximation of a nonlinear theory by linearizing around a chosen background solution.