\section*{The Intellectual Roadmap of the Course}
\addcontentsline{toc}{section}{The Intellectual Roadmap of the Course}

The chapters developed so far form one argument about how mathematics grows from
questions. They are not intended as separate catalogues of techniques. At each
stage an earlier language proves insufficient: geometry needs a numerical
representation, positions need a language of change, vectors need to be turned
into geometric sets, and familiar linear objects need to be replaced by objects
created through more complicated conditions. New notation and new operations are
introduced only when they answer a need that has already become visible.

The course therefore asks the student to follow two stories at once. The first is
the mathematical story of geometry, coordinates, vectors, lines, planes, and
quadratic objects. The second is a story about mathematical practice: how one
separates an object from its representation, how definitions are motivated, how
assumptions enter a theory, how a formula is derived from a demand, and how an old
idea is reused in a more complicated setting.

\subsection*{Chapter 1: from geometry to representation}

The first chapter is meant to show that coordinates are not the beginning of
geometry. Points, lines, circles, angles, congruence, parallelism, and
perpendicularity already belong to an underlying Euclidean world. The chapter
first makes the rules of that world explicit and distinguishes assumptions,
permitted constructions, and consequences that must be proved. Its purpose is to
teach that mathematical objects do not appear merely because we have drawn them,
and that a construction is complete only when the required property has been
justified.

The Cartesian system is then built as an additional structure placed on that
geometry. Origin, directions, perpendicular axes, unit length, and continuity are
not invisible facts; they are choices and assumptions needed to assign numerical
labels to points. The intellectual aim is to make the student conscious of what a
representation contains and what it leaves unchanged. By the end of the chapter,
the student should understand coordinates and angle measures as deliberately
constructed languages, and should be able to trace familiar formulas back to the
geometric facts from which they arise.

\subsection*{Chapter 2: from position to change and from object to coordinates}

Once points can be located, the next need is to describe change of position. The
vector is introduced to answer that need. The chapter is designed to prevent
vector algebra from appearing as arbitrary component manipulation: addition means
combining successive changes, scalar multiplication changes the scale and
orientation of a change, linear combinations describe what can be built from
available directions, and the dot product and projection answer questions about
angle, perpendicularity, and directional content. The transition to three
dimensions introduces the cross product only when a genuinely new geometric demand
appears: finding a direction perpendicular to two given directions.

The later parts of the chapter deepen the distinction between an object and its
numerical description. A vector remains fixed while its coordinates change with
the chosen basis. Polar coordinates then show that not every useful coordinate
system arises from a linear basis at all. The intended lesson is broader than the
ability to calculate new coordinates. The student should learn to choose a
representation according to the structure of the problem, to recognise which
properties belong to the geometric object and which belong only to the chosen
language, and to understand algebraic operations as compressed expressions of
geometric requirements.

\subsection*{Chapter 3: constructing geometric sets from permitted motion}

The third chapter gives the vector language a geometric task. A line is created by
allowing motion from one point in one independent direction; a plane is created by
allowing two independent directions. Their parametric descriptions are therefore
not formulas selected from a list, but records of what motion is permitted. Normal
equations arise from the complementary demand that motion in a perpendicular
direction must vanish. This is where the dot product, cross product, projection,
and linear combinations are deliberately reused rather than replaced by unrelated
rules.

The chapter also shows why dimension matters. In the plane, nonparallel lines must
meet; in space they may be skew. A single algebraic test may describe directions
without settling the full relation between geometric objects. The student should
leave this chapter able to move between several descriptions of the same line or
plane, select the description suited to the question, and distinguish directional
information from incidence information. More importantly, the student should see
that increasingly complicated geometry can be constructed from a small number of
previously understood ideas.

\subsection*{Chapter 4: from geometric demands to quadratic objects}

The fourth chapter changes the kind of requirement imposed on a moving point.
Instead of restricting motion to fixed directions, it imposes relations between
distances. Circles, parabolas, ellipses, and hyperbolas are therefore introduced as
answers to geometric demands before they are treated as equations. Their
second-degree equations arise because Euclidean distance contains squares and
square roots. The philosophical point is that the equation records an object that
has already been conceptually constructed; it does not create the object by
itself.

Only after this constructive viewpoint is established does the chapter reverse the
problem and ask how to recognise a geometric object when an equation is given
first. Completing squares, changing coordinates, discriminants, and standard forms
become tools of recognition. In three dimensions, planar sections provide the
central method for understanding quadric surfaces: a difficult object is examined
through families of simpler objects already known. The student should learn both
to derive equations from geometric requirements and to recover geometry from
equations, while remaining aware of which conclusions are fully explained by the
current theory and which will receive a deeper structural account later.

\subsection*{The common intention}

Across all four chapters, the desired habit is the same. The student should ask
what object is being studied, which features are fixed, which are allowed to vary,
what requirement defines the situation, why a particular representation is useful,
and what the result means after the calculation is finished. Progress is not
measured only by the number of formulas remembered. It is measured by the ability
to reconstruct why an idea was needed, to justify how it was obtained, and to
recognise the same method when the surface details of the problem change.

The central ambition of the course is therefore not merely to teach a sequence of
techniques. It is to make visible a way in which mathematics is built: by turning
needs into definitions, definitions into consequences, geometric structure into
algebraic language, and calculations back into understanding.