\subsection{Basic families and elementary questions}

The first examples should remain computationally simple: constants, linear functions, low-degree polynomials, reciprocal functions, absolute value, the exponential function, and the basic sine and cosine functions.

\begin{examplebox}[A quadratic function]
For
\[
f(x)=x^2-4,
\]
the domain is $\mathbb R$, the zeros are $x=\pm2$, and the function is negative between the zeros and positive outside them.
\end{examplebox}

\begin{examplebox}[A simple exponential]
For
\[
g(x)=e^x,
\]
the domain is $\mathbb R$, every value is positive, and the graph increases as $x$ increases.
\end{examplebox}

\begin{examplebox}[A periodic function]
For
\[
h(x)=\sin x,
\]
the values repeat every $2\pi$. The function oscillates between $-1$ and $1$.
\end{examplebox}

\begin{center}
\begin{tikzpicture}[x=0.72cm,y=0.55cm,>=Latex]
  % Quadratic function
  \begin{scope}
    \draw[->] (-2.8,0) -- (2.8,0) node[right] {$x$};
    \draw[->] (0,-4.6) -- (0,3.2) node[above] {$y$};
    \draw[thick,blue,domain=-2.65:2.65,samples=80]
      plot (\x,{\x*\x-4});
    \fill (-2,0) circle (2pt) node[above left] {$-2$};
    \fill (2,0) circle (2pt) node[above right] {$2$};
    \node[below] at (0,-4.6) {$f(x)=x^2-4$};
  \end{scope}

  % Exponential function
  \begin{scope}[xshift=5.3cm]
    \draw[->] (-2.4,0) -- (2.1,0) node[right] {$x$};
    \draw[->] (0,-0.5) -- (0,4.8) node[above] {$y$};
    \draw[thick,teal!80!black,domain=-2.2:1.5,samples=80]
      plot (\x,{exp(\x)});
    \fill (0,1) circle (2pt) node[left] {$1$};
    \node[below] at (0,-0.5) {$g(x)=e^x$};
  \end{scope}

  % Sine function
  \begin{scope}[xshift=10.4cm]
    \draw[->] (-3.5,0) -- (3.5,0) node[right] {$x$};
    \draw[->] (0,-1.6) -- (0,1.7) node[above] {$y$};
    \draw[thick,red!75!black,domain=-3.3:3.3,samples=100]
      plot (\x,{sin(\x r)});
    \draw[dashed] (-3.3,1) -- (3.3,1);
    \draw[dashed] (-3.3,-1) -- (3.3,-1);
    \node[left] at (0,1) {$1$};
    \node[left] at (0,-1) {$-1$};
    \node[below] at (0,-1.6) {$h(x)=\sin x$};
  \end{scope}
\end{tikzpicture}
\end{center}

\begin{interpretationbox}[Three graphs, three kinds of behavior]
The parabola is organized by its zeros and sign, the exponential by positivity and persistent growth, and the sine function by bounded oscillation and repetition. These visible differences motivate the later questions of limits, derivatives, and integrals.
\end{interpretationbox}

\begin{corebox}[Questions before calculus]
For an elementary function ask about its domain, values, zeros, sign, symmetry, and visible long-term behavior. Calculus will later make statements about change and limits precise.
\end{corebox}

The purpose of these examples is not to catalogue every possible function. They provide a small vocabulary rich enough to illustrate the central ideas of analysis without hiding them behind difficult algebra.
