Modern AI coding assistants can generate small interactive applications quickly. For a probability course, this creates an opportunity to turn definitions and formulas into controlled computational experiments: sample spaces can be explored, probability laws can be changed with sliders, repeated samples can be generated, and long-run behaviour can be compared with exact theory.

The purpose of this chapter is not to replace mathematical work with generated software. Its purpose is to help students design visualisations and simulations that make the mathematical structure visible. Every project begins with a model, states its assumptions, identifies the quantities that should remain invariant, and requires the generated application to compare simulation with theory whenever possible.

The prompts below are deliberately more detailed than ordinary chat requests. They specify the learning objective, model, controls, diagrams, numerical checks, and expected behaviour of the application. A student may copy a prompt into an AI coding assistant, but should then inspect the code, test the mathematics, and be able to explain every displayed quantity. The projects are suggestions rather than rigid specifications: they may be simplified or extended as long as the mathematical objective and validation requirements are preserved.

\input{chapters/ai_assisted_visualisation_projects/working_with_ai_coding_assistants}
\input{chapters/ai_assisted_visualisation_projects/introduction_projects}
\input{chapters/ai_assisted_visualisation_projects/combinatorics_projects}
\input{chapters/ai_assisted_visualisation_projects/discovering_formalism_projects}
\input{chapters/ai_assisted_visualisation_projects/sample_spaces_projects}
\input{chapters/ai_assisted_visualisation_projects/conditional_probability_projects}
\input{chapters/ai_assisted_visualisation_projects/discrete_random_variables_projects}
\input{chapters/ai_assisted_visualisation_projects/continuous_random_variables_projects}
\input{chapters/ai_assisted_visualisation_projects/empirical_distributions_projects}
\input{chapters/ai_assisted_visualisation_projects/sampling_distributions_projects}
\input{chapters/ai_assisted_visualisation_projects/limit_theorems_and_inference_projects}
\input{chapters/ai_assisted_visualisation_projects/final_checklist}
