\subsection{Reading change and solving elementary optimisation problems}

\begin{theorembox}[Derivative sign and monotonicity]
Let $f$ be continuous on an interval and differentiable in its interior. If $f'(x)>0$ throughout the interior, then $f$ is increasing. If $f'(x)<0$ throughout the interior, then $f$ is decreasing.
\end{theorembox}

\begin{interpretationbox}[Why local slopes control an interval]
For two points $a<b$, the mean value theorem gives a point $c\in(a,b)$ such that
\[
\frac{f(b)-f(a)}{b-a}=f'(c).
\]
If every derivative value is positive, then the secant slope is positive and therefore $f(b)>f(a)$. This is the bridge from local derivative information to global monotonicity. The full proof of the mean value theorem is not developed here.
\end{interpretationbox}

A critical point occurs where $f'(x)=0$ or where the derivative does not exist. A critical point is only a candidate. A change from positive to negative derivative gives a local maximum; a change from negative to positive gives a local minimum.

\begin{examplebox}[Reading a cubic]
For
\[
f(x)=x^3-3x,
\qquad
f'(x)=3(x-1)(x+1).
\]
The derivative is positive on $(-\infty,-1)$, negative on $(-1,1)$, and positive on $(1,\infty)$. Hence $f$ has a local maximum at $x=-1$ and a local minimum at $x=1$.
\end{examplebox}

\begin{corebox}[Optimisation workflow]
Choose the quantity to optimise, express it as a function of one variable, determine the allowed domain, find critical points, check endpoints when relevant, and interpret the result in the original problem.
\end{corebox}

\begin{examplebox}[Maximum area with fixed perimeter]
A rectangle has perimeter $20$. If one side is $x$, the other is $10-x$, so
\[
A(x)=x(10-x)=10x-x^2,
\qquad 0\le x\le10.
\]
Then
\[
A'(x)=10-2x,
\]
so the only interior critical point is $x=5$. The endpoint areas are zero, while $A(5)=25$. Therefore the maximum occurs for a $5\times5$ square.
\end{examplebox}

The derivative calculation is short. The substantive work is identifying the correct function, its restrictions, the relevant candidates, and the meaning of the result.
