\chapter{Determinants}

This chapter studies the determinant as a number that detects structural loss in a square matrix. After completing it, the student should be able to:
\begin{itemize}
\item compute determinants of small matrices and use simple structural shortcuts;
\item explain why row dependence and a zero determinant signal lost information;
\item connect determinants with invertibility, products, and efficient calculations.
\end{itemize}

\noindent\textit{Lecture emphasis.} The live core is the $2\times2$ determinant, one transparent $3\times3$ example, the meaning of zero, and the effect of row operations. General Laplace expansion and broader identities complete the written reference but need not dominate the presentation.

\input{chapters/chapter_06/determinant_as_function/section}
\input{chapters/chapter_06/small_determinants/section}
\input{chapters/chapter_06/properties_from_formulas/section}
\input{chapters/chapter_06/zero_determinant/section}
\input{chapters/chapter_06/laplace_expansion/section}
\input{chapters/chapter_06/efficient_calculation/section}

\noindent\textbf{Determinants.} The determinant begins with formulas for small square matrices and extends through minors and cofactors. Its row and column properties explain efficient calculations and show why a zero determinant records a relation among rows, columns, or possible matrix outputs. General identities used later are distinguished from properties that were fully derived in this chapter.

\input{chapters/chapter_06/review_questions}
