\section{General properties and efficient calculation}

The recursive definition extends the determinant to every square size. The properties observed for \(2\times2\) and \(3\times3\) matrices continue to hold in arbitrary dimension.

\subsection{Multilinearity and alternating behaviour}

\begin{theorembox}
The determinant is linear in each row separately when all other rows are fixed. It is also linear in each column separately.
\end{theorembox}

If the \(i\)-th row is written as a sum \(r_i=u+v\), then
\[
\det(r_1,\ldots,u+v,\ldots,r_n)
=
\det(r_1,\ldots,u,\ldots,r_n)
+
\det(r_1,\ldots,v,\ldots,r_n).
\]
If the row is multiplied by \(\lambda\), then the determinant is multiplied by \(\lambda\).

\begin{theorembox}
Exchanging two rows changes the sign of the determinant. Consequently, a matrix with two equal rows has determinant zero.
\end{theorembox}

Combining multilinearity with the equal-row property gives a stronger conclusion. If one row is a linear combination of the other rows, then the determinant is zero.

For example, if
\[
r_3=2r_1-r_2,
\]
then
\[
\begin{aligned}
\det(r_1,r_2,r_3)
&=\det(r_1,r_2,2r_1-r_2)\\
&=2\det(r_1,r_2,r_1)-\det(r_1,r_2,r_2)\\
&=0.
\end{aligned}
\]
Both terms contain repeated rows.

Because \(\det(A^{\mathsf T})=\det A\), all corresponding column statements also hold.

\subsection{Elementary row operations}

The effect of the three elementary row operations can now be summarized.

\begin{center}
\begin{tabular}{ll}
\textbf{Row operation} & \textbf{Effect on the determinant}\\
\hline
Exchange two rows & multiply by \(-1\)\\
Multiply one row by \(\lambda\) & multiply by \(\lambda\)\\
Add a multiple of one row to another & no change
\end{tabular}
\end{center}

These rules allow us to simplify a matrix while keeping track of the determinant.

\begin{examplebox}
Compute
\[
\det\begin{pmatrix}
1&2&1\\
2&5&3\\
1&0&4
\end{pmatrix}.
\]
Apply
\[
R_2\leftarrow R_2-2R_1,
\qquad
R_3\leftarrow R_3-R_1.
\]
These operations do not change the determinant:
\[
\begin{pmatrix}
1&2&1\\
2&5&3\\
1&0&4
\end{pmatrix}
\longrightarrow
\begin{pmatrix}
1&2&1\\
0&1&1\\
0&-2&3
\end{pmatrix}.
\]
Next use
\[
R_3\leftarrow R_3+2R_2,
\]
which again leaves the determinant unchanged:
\[
\begin{pmatrix}
1&2&1\\
0&1&1\\
0&-2&3
\end{pmatrix}
\longrightarrow
\begin{pmatrix}
1&2&1\\
0&1&1\\
0&0&5
\end{pmatrix}.
\]
The final matrix is upper triangular, so
\[
\det A=1\cdot1\cdot5=5.
\]
\end{examplebox}

\subsection{Triangular matrices}

\begin{theorembox}
The determinant of an upper or lower triangular matrix is the product of its diagonal entries:
\[
\det A=a_{11}a_{22}\cdots a_{nn}.
\]
\end{theorembox}

This follows from Laplace expansion: the zeros force the calculation to continue through smaller triangular matrices until only the diagonal entries remain.

A useful consequence is that a triangular matrix has determinant zero exactly when at least one diagonal entry is zero.

\subsection{Scaling the whole matrix}

If every row of an \(n\times n\) matrix is multiplied by \(\lambda\), multilinearity contributes one factor of \(\lambda\) from each row. Therefore
\[
\det(\lambda A)=\lambda^n\det A.
\]
This is why the determinant is not a linear function of the whole matrix.

\subsection{The determinant of a product}

A further fundamental property is
\[
\det(AB)=\det A\det B
\]
for square matrices \(A\) and \(B\) of the same size.

The full proof is not needed for the present calculations, but the statement is consistent with the rules already established. For diagonal matrices it is immediate: the diagonal entries multiply position by position, and so do their products.

\begin{remarkbox}
The formula \(\det(AB)=\det A\det B\) concerns matrix multiplication. It is different from entry-by-entry multiplication and should not be confused with linearity in rows or columns.
\end{remarkbox}

\subsection{A structured example}

Consider the Vandermonde-type determinant
\[
V=\begin{pmatrix}
1&1&1\\
x&y&z\\
x^2&y^2&z^2
\end{pmatrix}.
\]
Apply the determinant-preserving column operations
\[
C_2\leftarrow C_2-C_1,
\qquad
C_3\leftarrow C_3-C_1.
\]
Then
\[
\det V
=
\det\begin{pmatrix}
1&0&0\\
x&y-x&z-x\\
x^2&y^2-x^2&z^2-x^2
\end{pmatrix}.
\]
Expanding along the first row gives
\[
\det V
=
\det\begin{pmatrix}
y-x&z-x\\
y^2-x^2&z^2-x^2
\end{pmatrix}.
\]
Factor the two columns:
\[
y^2-x^2=(y-x)(y+x),
\qquad
z^2-x^2=(z-x)(z+x).
\]
Therefore
\[
\begin{aligned}
\det V
&=(y-x)(z-x)
\det\begin{pmatrix}1&1\\y+x&z+x\end{pmatrix}\\
&=(y-x)(z-x)(z-y).
\end{aligned}
\]
The factorization immediately shows when the determinant is zero: whenever two of \(x,y,z\) are equal, two columns of the original matrix are equal.

\subsection*{What to remember}

\begin{summarybox}
Before computing a determinant:
\begin{itemize}
\item check whether rows or columns are equal, proportional, or related by a simple linear combination;
\item look for zeros and triangular structure;
\item decide whether row or column operations will simplify the matrix;
\item track sign changes and extracted factors;
\item use Laplace expansion only after choosing a useful row or column.
\end{itemize}

The determinant begins as a definition. Its mathematical content is revealed by the algebraic properties that the definition satisfies.
\end{summarybox}
