\section{Determinants of larger matrices}

The formulas for \(2\times2\) and \(3\times3\) matrices are explicit, but they do not provide a practical definition for every size. We now extend the determinant recursively: a determinant of size \(n\times n\) will be expressed in terms of determinants of size \((n-1)\times(n-1)\).

\begin{remarkbox}
In this section colour is used only as a visual guide. Usually
\begin{itemize}
\item \textcolor{purple}{purple} marks the entry on which we are focusing,
\item \textcolor{red}{red} marks the row or column that will be deleted or used for expansion,
\item \textcolor{green!50!black}{green} marks the smaller submatrix that remains,
\item \textcolor{blue}{blue} marks the row or column along which we expand.
\end{itemize}
The colours are not part of the definition. They only help us see the substructures inside the matrix.
\end{remarkbox}

\subsection{Minors}

Let \(A=(a_{ij})\) be an \(n\times n\) matrix. To form a minor, we choose one entry, delete its row and its column, and then take the determinant of the smaller matrix that remains.

\begin{definitionbox}
The determinant of the matrix obtained by deleting row \(i\) and column \(j\) is called the minor of the entry \(a_{ij}\). It is denoted by
\[
M_{ij}.
\]
\end{definitionbox}

\begin{examplebox}
Let
\[
A=
\begin{pmatrix}
\textcolor{red}{2}&\textcolor{purple}{1}&\textcolor{red}{3}\\
\textcolor{green!50!black}{0}&\textcolor{red}{-1}&\textcolor{green!50!black}{4}\\
\textcolor{green!50!black}{5}&\textcolor{red}{2}&\textcolor{green!50!black}{1}
\end{pmatrix}.
\]
We want to find \(M_{12}\), so we focus on the \textcolor{purple}{entry in row 1, column 2}. To form its minor, we delete the whole \textcolor{red}{first row} and the whole \textcolor{red}{second column}. What remains is the \textcolor{green!50!black}{smaller \(2\times2\) matrix}
\[
M_{12}
=
\det
\begin{pmatrix}
\textcolor{green!50!black}{0}&\textcolor{green!50!black}{4}\\
\textcolor{green!50!black}{5}&\textcolor{green!50!black}{1}
\end{pmatrix}.
\]
Now compute its determinant:
\[
M_{12}=0\cdot1-4\cdot5=-20.
\]
\end{examplebox}

The important point is not only the answer \(-20\), but also the structure: one chosen entry points to one smaller determinant.

\subsection{Cofactors}

When we expand a determinant, the minors appear with alternating signs. It is useful to separate the sign from the minor itself.

The signs follow a checkerboard pattern:
\[
\begin{pmatrix}
\textcolor{blue}{+}&\textcolor{blue}{-}&\textcolor{blue}{+}&-&\cdots\\
-&+&-&+&\cdots\\
+&-&+&-&\cdots\\
-&+&-&+&\cdots\\
\vdots&\vdots&\vdots&\vdots&\ddots
\end{pmatrix}.
\]
The sign in position \((i,j)\) is \((-1)^{i+j}\).

\begin{definitionbox}
The cofactor of the entry \(a_{ij}\) is
\[
C_{ij}=(-1)^{i+j}M_{ij}.
\]
\end{definitionbox}

For the previous example,
\[
M_{12}=-20,
\qquad
(-1)^{1+2}=-1,
\]
so
\[
C_{12}=(-1)M_{12}=(-1)(-20)=20.
\]
Thus the minor and the cofactor use the same smaller determinant, but the cofactor also remembers the correct sign for expansion.

\subsection{The recursive definition}

We already defined the determinant of a \(1\times1\) matrix by
\[
\det\begin{pmatrix}a\end{pmatrix}=a.
\]
For \(n\ge2\), we define the determinant by expanding along the first row.

\begin{definitionbox}
Let \(A=(a_{ij})\) be an \(n\times n\) matrix. Its determinant is
\[
\det A
=a_{11}C_{11}+a_{12}C_{12}+\cdots+a_{1n}C_{1n}.
\]
Equivalently,
\[
\det A=\sum_{j=1}^{n}a_{1j}C_{1j}.
\]
\end{definitionbox}

This definition reduces the size by one at every step. Eventually every computation reaches determinants of size \(1\times1\).

For a \(2\times2\) matrix,
\[
A=
\begin{pmatrix}
\textcolor{blue}{a}&\textcolor{blue}{b}\\
c&d
\end{pmatrix},
\]
expansion along the \textcolor{blue}{first row} gives
\[
\det A
=\textcolor{blue}{a}\det\begin{pmatrix}d\end{pmatrix}
-\textcolor{blue}{b}\det\begin{pmatrix}c\end{pmatrix}
=ad-bc.
\]
Thus the recursive definition agrees with the earlier definition.

For a \(3\times3\) matrix,
\[
A=
\begin{pmatrix}
\textcolor{blue}{a}&\textcolor{blue}{b}&\textcolor{blue}{c}\\
d&e&f\\
g&h&i
\end{pmatrix},
\]
expansion along the \textcolor{blue}{first row} gives
\[
\begin{aligned}
\det A
&=\textcolor{blue}{a}
\begin{vmatrix}
\textcolor{green!50!black}{e}&\textcolor{green!50!black}{f}\\
\textcolor{green!50!black}{h}&\textcolor{green!50!black}{i}
\end{vmatrix}
-\textcolor{blue}{b}
\begin{vmatrix}
\textcolor{green!50!black}{d}&\textcolor{green!50!black}{f}\\
\textcolor{green!50!black}{g}&\textcolor{green!50!black}{i}
\end{vmatrix}
+\textcolor{blue}{c}
\begin{vmatrix}
\textcolor{green!50!black}{d}&\textcolor{green!50!black}{e}\\
\textcolor{green!50!black}{g}&\textcolor{green!50!black}{h}
\end{vmatrix}\\[1ex]
&=a(ei-fh)-b(di-fg)+c(dh-eg)\\
&=aei+bfg+cdh-ceg-bdi-afh.
\end{aligned}
\]
The colour helps us see the structure:
\begin{itemize}
\item each \textcolor{blue}{entry of the first row} produces one term,
\item each term is multiplied by the determinant of the \textcolor{green!50!black}{smaller matrix left after deleting its row and column},
\item the signs follow the pattern \(+,-,+\).
\end{itemize}
Thus the recursive definition also agrees with the earlier \(3\times3\) formula.

\subsection{Expansion along any row or column}

Although the determinant was defined using the first row, it can be proved that the same value is obtained by expanding along any row or any column.

\begin{theorembox}
Expansion along row \(i\) gives
\[
\det A=\sum_{j=1}^{n}a_{ij}C_{ij}.
\]
Expansion along column \(j\) gives
\[
\det A=\sum_{i=1}^{n}a_{ij}C_{ij}.
\]
\end{theorembox}

This freedom is computationally important. A row or column containing many zeros usually gives the shortest expansion.

\begin{examplebox}
Consider
\[
B=
\begin{pmatrix}
4&\textcolor{red}{0}&1&2\\
0&\textcolor{purple}{3}&0&-1\\
2&\textcolor{red}{0}&5&0\\
1&\textcolor{red}{0}&0&6
\end{pmatrix}.
\]
The \textcolor{red}{second column} contains only one non-zero entry, namely the \textcolor{purple}{3}. Therefore expansion along that column is especially convenient.

Expanding along the \textcolor{red}{second column},
\[
\det B
=0\cdot C_{12}+3\cdot C_{22}+0\cdot C_{32}+0\cdot C_{42}=3C_{22}.
\]
The sign in position \((2,2)\) is positive, so
\[
\det B
=3
\det
\begin{pmatrix}
\textcolor{green!50!black}{4}&\textcolor{green!50!black}{1}&\textcolor{green!50!black}{2}\\
\textcolor{green!50!black}{2}&\textcolor{green!50!black}{5}&\textcolor{green!50!black}{0}\\
\textcolor{green!50!black}{1}&\textcolor{green!50!black}{0}&\textcolor{green!50!black}{6}
\end{pmatrix}.
\]
Again the colour shows the substructure: the \textcolor{red}{chosen column} disappears, and the \textcolor{green!50!black}{remaining \(3\times3\) matrix} is exactly the minor.

Now compute the smaller determinant. Using the \(3\times3\) formula,
\[
\begin{aligned}
\det
\begin{pmatrix}
4&1&2\\
2&5&0\\
1&0&6
\end{pmatrix}
&=4\cdot5\cdot6+1\cdot0\cdot1+2\cdot2\cdot0\\
&\quad-2\cdot5\cdot1-1\cdot2\cdot6-4\cdot0\cdot0\\
&=120+0+0-10-12-0\\
&=98.
\end{aligned}
\]
Therefore
\[
\det B=3\cdot98=294.
\]
\end{examplebox}

\subsection{Using Laplace expansion intelligently}

Laplace expansion is a definition and a general method, but it should not be applied blindly. Before expanding:
\begin{itemize}
\item inspect the matrix for zeros;
\item choose a row or column that produces few non-zero terms;
\item write the cofactor signs carefully;
\item keep the smaller determinants visible;
\item use row or column properties when they shorten the calculation.
\end{itemize}

A good expansion is not only a formal step. It is a way of noticing the right substructure inside the matrix.