\section{Properties discovered from the formulas}

The formulas for \(2\times2\) and \(3\times3\) determinants are now known. We will not treat the next properties as a new list of rules to memorize. We will obtain them from the formulas.

For each property we follow the same pattern:
\begin{enumerate}
\item perform a small numerical experiment;
\item describe what seems to happen;
\item repeat the calculation with general entries;
\item only then write the result as a theorem.
\end{enumerate}

The numerical examples are not proofs. They show us what question to ask. The calculation with general entries explains why the observation is always true.

\subsection{Changing one row at a time}

We begin with two matrices that have the same second row:
\[
A=
\begin{pmatrix}
1&2\\
3&5
\end{pmatrix},
\qquad
B=
\begin{pmatrix}
4&-1\\
3&5
\end{pmatrix}.
\]
Their determinants are
\[
\det A=1\cdot5-2\cdot3=5-6=-1
\]
and
\[
\det B=4\cdot5-(-1)\cdot3=20+3=23.
\]

Now add only the first rows:
\[
(1,2)+(4,-1)=(5,1).
\]
Keeping the second row unchanged gives
\[
C=
\begin{pmatrix}
5&1\\
3&5
\end{pmatrix}.
\]
Its determinant is
\[
\det C=5\cdot5-1\cdot3=25-3=22.
\]
We notice that
\[
22=(-1)+23,
\]
so in this example
\[
\det C=\det A+\det B.
\]

The example suggests that adding two possible first rows may add the corresponding determinants. We now check this from the formula.

Take the two possible first rows
\[
(a,b)
\qquad\text{and}\qquad
(p,q),
\]
and keep the second row
\[
(c,d)
\]
fixed. After adding the first rows, the matrix is
\[
\begin{pmatrix}
a+p&b+q\\
c&d
\end{pmatrix}.
\]
Using the definition of a \(2\times2\) determinant,
\[
\begin{aligned}
\det
\begin{pmatrix}
a+p&b+q\\
c&d
\end{pmatrix}
&=(a+p)d-(b+q)c\\
&=ad+pd-bc-qc\\
&=(ad-bc)+(pd-qc)\\
&=
\det
\begin{pmatrix}
a&b\\
c&d
\end{pmatrix}
+
\det
\begin{pmatrix}
p&q\\
c&d
\end{pmatrix}.
\end{aligned}
\]

Every line has a specific purpose. The first line applies the determinant formula. The second expands the brackets. The third groups together the terms coming from the first row \((a,b)\) and the terms coming from the second possible row \((p,q)\). The last line recognizes the two determinants.

\subsection{Multiplying one row by a number}

Consider
\[
A=
\begin{pmatrix}
2&1\\
3&4
\end{pmatrix}.
\]
We first calculate
\[
\det A=2\cdot4-1\cdot3=8-3=5.
\]

Multiply only the first row by \(3\):
\[
(2,1)\longrightarrow(6,3).
\]
The new matrix is
\[
B=
\begin{pmatrix}
6&3\\
3&4
\end{pmatrix}.
\]
Its determinant is
\[
\det B=6\cdot4-3\cdot3=24-9=15.
\]
Since
\[
15=3\cdot5,
\]
the determinant was also multiplied by \(3\).

Now replace the numerical factor \(3\) by an arbitrary number \(k\). For
\[
\begin{pmatrix}
ka&kb\\
c&d
\end{pmatrix}
\]
we obtain
\[
\begin{aligned}
\det
\begin{pmatrix}
ka&kb\\
c&d
\end{pmatrix}
&=(ka)d-(kb)c\\
&=kad-kbc\\
&=k(ad-bc)\\
&=
k\det
\begin{pmatrix}
a&b\\
c&d
\end{pmatrix}.
\end{aligned}
\]

The factor \(k\) appears in both products, so it can be taken in front of the determinant.

\begin{theorembox}
For a \(2\times2\) determinant, addition in one row becomes addition of determinants, and multiplication of one row by a number multiplies the determinant by the same number. The other row must remain fixed.
\end{theorembox}

This property is called \emph{linearity in one row}. The phrase ``in one row'' matters. It does not mean that
\[
\det(A+B)=\det A+\det B
\]
for arbitrary matrices.

For example, let
\[
A=B=
\begin{pmatrix}
1&0\\
0&1
\end{pmatrix}.
\]
Then
\[
\det A+\det B=1+1=2,
\]
but
\[
A+B=
\begin{pmatrix}
2&0\\
0&2
\end{pmatrix}
\]
and
\[
\det(A+B)=2\cdot2-0\cdot0=4.
\]
Thus linearity applies to one selected row while the other rows are fixed. It is not linearity of the whole matrix.

\subsection{A check for a \(3\times3\) determinant}

The same mechanism can already be seen in the \(3\times3\) formula. We check multiplication of the first row by \(k\).

Start with
\[
\begin{pmatrix}
ka&kb&kc\\
d&e&f\\
g&h&i
\end{pmatrix}.
\]
Applying the six-term formula gives
\[
\begin{aligned}
&\det
\begin{pmatrix}
ka&kb&kc\\
d&e&f\\
g&h&i
\end{pmatrix}\\
&=(ka)ei+(kb)fg+(kc)dh-(kc)eg-(kb)di-(ka)fh\\
&=k(aei)+k(bfg)+k(cdh)-k(ceg)-k(bdi)-k(afh)\\
&=k\bigl(aei+bfg+cdh-ceg-bdi-afh\bigr)\\
&=
k\det
\begin{pmatrix}
a&b&c\\
d&e&f\\
g&h&i
\end{pmatrix}.
\end{aligned}
\]

The calculation is longer than in the \(2\times2\) case, but the idea is unchanged: every term contains exactly one entry from the first row, so multiplying that row by \(k\) multiplies every term by \(k\).

We can also test addition in the first row without introducing another large collection of letters. Keep the last two rows fixed and consider
\[
A=
\begin{pmatrix}
1&2&0\\
1&0&2\\
0&1&1
\end{pmatrix},
\qquad
B=
\begin{pmatrix}
2&-1&1\\
1&0&2\\
0&1&1
\end{pmatrix}.
\]
Using the \(3\times3\) formula,
\[
\det A=-4,
\qquad
\det B=-2.
\]
Adding the first rows gives
\[
(1,2,0)+(2,-1,1)=(3,1,1),
\]
so
\[
C=
\begin{pmatrix}
3&1&1\\
1&0&2\\
0&1&1
\end{pmatrix}.
\]
A direct calculation gives
\[
\det C=-6=-4+(-2).
\]
This numerical calculation is evidence of the same pattern. Later, after the general definition by Laplace expansion, we will prove linearity for every size without writing an increasingly long explicit formula.

\subsection{Exchanging two rows}

Take
\[
A=
\begin{pmatrix}
1&2\\
3&4
\end{pmatrix}.
\]
Then
\[
\det A=1\cdot4-2\cdot3=4-6=-2.
\]

Exchange the rows:
\[
B=
\begin{pmatrix}
3&4\\
1&2
\end{pmatrix}.
\]
Now
\[
\det B=3\cdot2-4\cdot1=6-4=2.
\]
The value changed from \(-2\) to \(2\). The sign changed.

To see why this always happens, start with
\[
A=
\begin{pmatrix}
a&b\\
c&d
\end{pmatrix}.
\]
Its determinant is
\[
\det A=ad-bc.
\]
After exchanging the rows,
\[
B=
\begin{pmatrix}
c&d\\
a&b
\end{pmatrix}.
\]
Therefore
\[
\begin{aligned}
\det B
&=cb-da\\
&=bc-ad\\
&=-(ad-bc)\\
&=-\det A.
\end{aligned}
\]
The equality \(cb=bc\) uses ordinary multiplication of numbers. The third line takes a minus sign in front of the whole expression.

\begin{theorembox}
Exchanging the two rows of a \(2\times2\) determinant changes its sign.
\end{theorembox}

\subsection{Equal rows}

Suppose both rows are equal:
\[
A=
\begin{pmatrix}
a&b\\
a&b
\end{pmatrix}.
\]
The determinant can be calculated directly:
\[
\begin{aligned}
\det A
&=a\cdot b-b\cdot a\\
&=ab-ab\\
&=0.
\end{aligned}
\]

There is also a useful reasoning argument. If we exchange two equal rows, the matrix does not change. Therefore its determinant should still be the same number. But we have just proved that exchanging two rows changes the sign. Hence the same determinant must satisfy
\[
\det A=-\det A.
\]
Adding \(\det A\) to both sides gives
\[
2\det A=0,
\]
and therefore
\[
\det A=0.
\]

The direct calculation checks the result from the formula. The second argument shows how one proved property can be used to obtain another.

\subsection{Proportional rows}

Now suppose the second row is \(k\) times the first row:
\[
A=
\begin{pmatrix}
a&b\\
ka&kb
\end{pmatrix}.
\]
Applying the determinant formula,
\[
\begin{aligned}
\det A
&=a(kb)-b(ka)\\
&=kab-kab\\
&=0.
\end{aligned}
\]
The two products are equal, so they cancel.

We can see the same cancellation directly in the \(3\times3\) formula. Let the second row be \(k\) times the first:
\[
A=
\begin{pmatrix}
a&b&c\\
ka&kb&kc\\
g&h&i
\end{pmatrix}.
\]
Then
\[
\begin{aligned}
\det A
&=a(kb)i+b(kc)g+c(ka)h\\
&\quad-c(kb)g-b(ka)i-a(kc)h\\
&=kabi+kbcg+kach-kbcg-kabi-kach\\
&=0.
\end{aligned}
\]
We have written all six terms so that the cancellation is visible.

\begin{theorembox}
For the \(2\times2\) and \(3\times3\) determinants already defined, equal or proportional rows give determinant zero.
\end{theorembox}

\subsection{Adding a multiple of one row to another}

Consider
\[
A=
\begin{pmatrix}
2&1\\
5&3
\end{pmatrix}.
\]
Its determinant is
\[
\det A=2\cdot3-1\cdot5=6-5=1.
\]

Replace the second row by the second row minus twice the first:
\[
(5,3)-2(2,1)=(5,3)-(4,2)=(1,1).
\]
The new matrix is
\[
B=
\begin{pmatrix}
2&1\\
1&1
\end{pmatrix}.
\]
Its determinant is
\[
\det B=2\cdot1-1\cdot1=2-1=1.
\]
The determinant did not change.

Now check the general calculation. Start with
\[
\begin{pmatrix}
a&b\\
c&d
\end{pmatrix}
\]
and replace the second row by
\[
(c,d)+k(a,b)=(c+ka,d+kb).
\]
The new determinant is
\[
\begin{aligned}
\det
\begin{pmatrix}
a&b\\
c+ka&d+kb
\end{pmatrix}
&=a(d+kb)-b(c+ka)\\
&=ad+kab-bc-kab\\
&=ad-bc\\
&=
\det
\begin{pmatrix}
a&b\\
c&d
\end{pmatrix}.
\end{aligned}
\]
The new terms \(kab\) and \(-kab\) cancel. This is why the determinant remains unchanged.

\begin{theorembox}
Adding a multiple of one row to another row does not change a \(2\times2\) determinant.
\end{theorembox}

This property can also be reconstructed from the earlier discoveries:
\begin{itemize}
\item addition in one row separates into a sum of determinants;
\item the added part is a multiple of the other row;
\item a determinant with proportional rows is zero.
\end{itemize}
The direct calculation above is still important because it shows the cancellation explicitly.

\subsection{Rows and columns}

So far we have worked mainly with rows. Columns satisfy the same properties. For \(2\times2\) matrices this can be checked by transposition.

Let
\[
A=
\begin{pmatrix}
a&b\\
c&d
\end{pmatrix},
\qquad
A^{\mathsf T}=
\begin{pmatrix}
a&c\\
b&d
\end{pmatrix}.
\]
Then
\[
\det A=ad-bc,
\]
while
\[
\begin{aligned}
\det(A^{\mathsf T})
&=ad-cb\\
&=ad-bc\\
&=\det A.
\end{aligned}
\]
The middle equality again uses \(cb=bc\).

Transposition changes rows into columns without changing the determinant. Therefore a row calculation can be transferred to columns. For example, proportional columns become proportional rows after transposition, so their determinant is also zero.

\begin{theorembox}
For the small determinants considered here,
\[
\det(A^{\mathsf T})=\det A.
\]
Consequently, the corresponding row properties also hold for columns.
\end{theorembox}

\subsection*{What has been learned}

\begin{summarybox}
The formulas were the starting point, not the final lesson.

We first performed concrete calculations. They suggested patterns. We then returned to the definitions and showed the patterns by expanding, regrouping, and cancelling terms. Only after this work did we state the properties:
\begin{itemize}
\item one row may be added or multiplied by a number while the other rows remain fixed;
\item exchanging two rows changes the sign;
\item equal or proportional rows give determinant zero;
\item adding a multiple of one row to another does not change the determinant;
\item the corresponding statements hold for columns.
\end{itemize}

A formula summarizes a piece of reasoning. In a complete solution, the intermediate steps must remain visible so that the reader can see why the conclusion follows.
\end{summarybox}
