\section{Combinations without repetition}

In combinations, order does not matter. We are interested only in which objects are selected, not in the order in which they are written.

\begin{definitionbox}
Let \(A\) be a finite set with \(n\) distinct elements, and let \(k\leq n\). A combination without repetition of size \(k\) is a \(k\)-element subset of \(A\).
\end{definitionbox}

For example, if
\[
A=\{a,b,c\},
\]
then the combinations without repetition of size \(2\) are
\[
\{a,b\},\qquad \{a,c\},\qquad \{b,c\}.
\]
The expressions \(\{a,b\}\) and \(\{b,a\}\) describe the same set, so they are not counted separately.

\begin{center}
\begin{tikzpicture}[scale=1.05, transform shape]
\foreach \first/\second/\x in {a/b/0, a/c/3, b/c/6} {
    \node[draw, circle, fill=blue!5, minimum size=0.65cm] at (\x,0) {$\first$};
    \node[draw, circle, fill=blue!5, minimum size=0.65cm] at (\x+0.8,0) {$\second$};
    \node at (\x+0.4,-0.75) {$\{\first,\second\}$};
}
\end{tikzpicture}
\end{center}

\begin{theorembox}
The number of combinations without repetition of size \(k\) chosen from \(n\) distinct elements is
\[
\binom{n}{k}=\frac{n!}{k!(n-k)!}.
\]
\end{theorembox}

To see why, first count ordered selections without repetition. There are
\[
\frac{n!}{(n-k)!}
\]
of them. But each unordered group of \(k\) elements can be ordered in \(k!\) ways. Therefore each combination has been counted \(k!\) times. Dividing by \(k!\) gives
\[
\frac{1}{k!}\cdot \frac{n!}{(n-k)!}=\frac{n!}{k!(n-k)!}.
\]

\begin{examplebox}
From six distinct books, choose three books to take on a trip. The order of choosing does not matter. The number of possible selections is
\[
\binom{6}{3}=\frac{6!}{3!3!}=20.
\]
\end{examplebox}

\begin{remarkbox}
Use combinations without repetition when order does not matter and no object may be selected more than once.
\end{remarkbox}
