\section{Counting begins with elementary outcomes}

Combinatorics is often presented as a list of formulas. In probability this is dangerous. Before a formula can be used, we must know what one elementary outcome is. A physical experiment does not determine a sample space by itself. The sample space depends on what is recorded.

Suppose that three balls are drawn from a box. If the balls are labelled and the order of drawing is recorded, then an outcome is a sequence of labelled objects. If only the selected balls are recorded, then an outcome is a set. If only the colors are recorded, then labelled balls of the same color may become indistinguishable. If only the number of red balls is recorded, then the outcome is no longer the full draw, but only the value of a random variable.

Thus the first question in a counting problem is not
\[
\text{Which formula should be used?}
\]
The first question is
\[
\text{What is one outcome?}
\]

\begin{definitionbox}
An \textbf{elementary outcome} is a complete description of one possible result of an experiment, according to the recording rule chosen for the model.
\end{definitionbox}

The phrase ``according to the recording rule'' is essential. The same physical situation may produce different mathematical sample spaces if different information is recorded.

\begin{examplebox}
A box contains three balls labelled \(R_1,R_2,B_1\). Two balls are drawn without replacement.

If order and labels are recorded, then examples of outcomes are
\[
(R_1,B_1),\qquad (B_1,R_1),\qquad (R_1,R_2).
\]
If only the set of labelled balls is recorded, then examples of outcomes are
\[
\{R_1,B_1\},\qquad \{R_1,R_2\}.
\]
If only the sequence of colors is recorded, then examples of outcomes are
\[
(R,B),\qquad (B,R),\qquad (R,R).
\]
If only the number of red balls is recorded, then the possible recorded values are
\[
0,1,2.
\]
These are not four different answers to the same counting question. They are four different mathematical models of the same physical experiment.
\end{examplebox}

\begin{remarkbox}
A counting error often begins before any arithmetic is done. It begins when the outcome type is chosen incorrectly.
\end{remarkbox}

\section{Recording rules and sample spaces}

A recording rule tells us which differences between physical results are mathematically visible. Some differences are recorded; others are ignored. Once the recording rule is fixed, the sample space is the set of all possible recorded outcomes.

\begin{definitionbox}
A \textbf{recording rule} is a decision about what information is included in the mathematical outcome and what information is ignored.
\end{definitionbox}

For finite experiments, the construction usually follows this pattern:
\[
\text{physical situation}
\longrightarrow
\text{recording rule}
\longrightarrow
\text{elementary outcome}
\longrightarrow
\text{sample space}.
\]

The following table shows common outcome types.

\begin{center}
\begin{tabular}{lll}
\toprule
Recording rule & Outcome type & Typical notation \\
\midrule
Order is recorded & sequence & \((a_1,a_2,\ldots,a_k)\) \\
Only membership is recorded & set & \(\{a_1,a_2,\ldots,a_k\}\) \\
All objects are arranged in a row & linear arrangement & \((a_{\sigma(1)},\ldots,a_{\sigma(n)})\) \\
Objects are arranged on a circle & circular arrangement & relative order \\
Only numbers of types are recorded & count vector & \((n_1,n_2,\ldots,n_r)\) \\
Only a numerical summary is recorded & random variable value & \(X(\omega)\) \\
\bottomrule
\end{tabular}
\end{center}

\begin{examplebox}
A committee of three people is chosen from ten people. If the committee is recorded only as a group, then
\[
\{A,B,C\}=\{C,A,B\}.
\]
The order of listing the names is irrelevant. The outcome is a set.

If the same three people are assigned the roles of chair, secretary, and treasurer, then
\[
(A,B,C)\neq (C,A,B).
\]
The outcome is now an ordered assignment of roles.
\end{examplebox}

The two situations may look similar in everyday language: in both cases three people are selected. But mathematically they are different because the recording rules are different.

\section{Diagnostic questions before counting}

Before using a formula, answer the following questions.

\begin{enumerate}
    \item What is one elementary outcome?
    \item Is the outcome a sequence, a set, an arrangement, or a count vector?
    \item Does order matter?
    \item Are all objects used, or only some of them?
    \item Can the same object be used more than once?
    \item Are the objects distinguishable or indistinguishable?
    \item Are we first counting labelled descriptions and then identifying descriptions that represent the same outcome?
\end{enumerate}

These questions are more important than the names of formulas. The names are only labels attached to structures that have already been recognized.

\begin{remarkbox}
Do not begin a problem by asking which formula you remember. Begin by asking what kind of outcome is being constructed.
\end{remarkbox}

\section{The multiplication principle as a construction rule}

Many counting problems can be solved by describing how an outcome is constructed step by step.

\begin{theorembox}
\textbf{Multiplication principle.} Suppose a construction is performed in \(k\) successive stages. If there are \(m_1\) choices for the first stage, \(m_2\) choices for the second stage, and so on up to \(m_k\) choices for the \(k\)-th stage, then the number of possible complete constructions is
\[
m_1m_2\cdots m_k.
\]
\end{theorembox}

The multiplication principle is not only a formula. It is a way of reading an outcome as a construction process.

\begin{examplebox}
A four-digit PIN code is formed from the digits \(0,1,\ldots,9\), and digits may repeat. One outcome is a sequence such as
\[
(0,7,7,3).
\]
There are \(10\) choices for each position, hence
\[
10\cdot 10\cdot 10\cdot 10=10^4
\]
possible PIN codes.

The reason is not that the formula \(n^k\) was memorized. The reason is that a PIN code is a sequence of four independent choices from a set of ten symbols.
\end{examplebox}

\begin{examplebox}
A three-letter code is formed from the letters \(A,B,C,D\), and letters may not repeat. One outcome is an ordered triple, for example
\[
(A,D,B).
\]
There are \(4\) choices for the first position, \(3\) for the second, and \(2\) for the third. Therefore the number of codes is
\[
4\cdot 3\cdot 2=24.
\]
\end{examplebox}

When the same number of choices is available at every stage, powers appear. When the number of choices decreases, falling products appear. When order is later ignored, division appears. Most elementary combinatorial formulas can be understood from these three ideas.

\section{Sequences with repetition}

A sequence with repetition is an ordered list in which each position is filled independently from the same set of symbols.

\begin{definitionbox}
Let \(S\) be a set with \(n\) elements. A \textbf{sequence of length \(k\) with repetition} over \(S\) is an ordered \(k\)-tuple
\[
(a_1,a_2,\ldots,a_k),
\]
where each \(a_i\in S\). The same element may occur many times.
\end{definitionbox}

The number of such sequences is
\[
n^k.
\]

This model is used when:
\begin{itemize}
    \item order is recorded,
    \item each position is chosen separately,
    \item repetition is allowed,
    \item the outcome is a full sequence.
\end{itemize}

Typical examples include PIN codes, repeated coin tosses, repeated die rolls, strings over an alphabet, and repeated selections with replacement.

\begin{examplebox}
The sample space for three tosses of a coin is
\[
\Omega=\{H,T\}^3.
\]
It has
\[
2^3=8
\]
elementary outcomes. Each outcome is a sequence, for example
\[
(H,T,H).
\]
The outcomes \((H,T,H)\) and \((T,H,H)\) are different because the order of tosses is recorded.
\end{examplebox}

\section{Ordered selections without repetition}

Sometimes only some objects are chosen, but the order or position of the chosen objects matters.

\begin{definitionbox}
An \textbf{ordered selection without repetition} of length \(k\) from \(n\) distinct objects is an ordered list of \(k\) different objects chosen from the \(n\) available objects.
\end{definitionbox}

The number of ordered selections without repetition is
\[
P(n,k)=n(n-1)\cdots(n-k+1)=\frac{n!}{(n-k)!}.
\]

This model is used when:
\begin{itemize}
    \item only some objects are selected,
    \item order or position matters,
    \item no object may be used twice.
\end{itemize}

\begin{examplebox}
Gold, silver, and bronze medals are assigned among \(12\) runners. An outcome is not a set of three runners. It is an ordered triple
\[
(\text{gold},\text{silver},\text{bronze}).
\]
There are
\[
12\cdot 11\cdot 10=P(12,3)
\]
possible assignments.
\end{examplebox}

In some European terminology, ordered selections are called variations. In these notes we use the name ordered selections without repetition, or \(k\)-permutations, because it describes directly what the outcome is.

\section{Arrangements of all objects}

A permutation is an arrangement of all objects in a row.

\begin{definitionbox}
A \textbf{permutation} of \(n\) distinct objects is a linear arrangement of all \(n\) objects.
\end{definitionbox}

There are
\[
n!
\]
permutations of \(n\) distinct objects.

This is a special case of ordered selection without repetition with \(k=n\):
\[
P(n,n)=n!.
\]

\begin{examplebox}
If \(7\) students are arranged in a line, then one outcome is a complete ordered list of all students. The number of arrangements is
\[
7!.
\]
\end{examplebox}

Permutations are used when:
\begin{itemize}
    \item all objects are used,
    \item order matters,
    \item the objects are distinguishable,
    \item the arrangement is linear.
\end{itemize}

\section{Circular arrangements}

A circular arrangement is different from a linear arrangement because rotating the whole arrangement does not create a new relative seating order.

\begin{examplebox}
Suppose four people \(A,B,C,D\) sit around a round table. The linear lists
\[
(A,B,C,D),\qquad (B,C,D,A),\qquad (C,D,A,B),\qquad (D,A,B,C)
\]
represent the same circular arrangement if only relative positions matter.
\end{examplebox}

For \(n\) distinct objects placed around a circle, with rotations identified, the number of circular arrangements is
\[
(n-1)!.
\]

One way to see this is to fix one object as a reference point. The remaining \(n-1\) objects can then be arranged around it in
\[
(n-1)!
\]
ways.

\begin{remarkbox}
A circular arrangement requires a modelling decision. If seats are numbered or if there is a distinguished position, then rotations may become different outcomes and the linear count may be appropriate.
\end{remarkbox}

\section{Overcounting and equivalent descriptions}

Many combinatorial formulas arise from a two-step idea:
\begin{enumerate}
    \item Count descriptions that are easy to construct.
    \item Divide by the number of descriptions representing the same outcome.
\end{enumerate}

This is the idea of overcounting.

\begin{definitionbox}
\textbf{Overcounting} occurs when the same intended outcome is counted more than once because it has several different descriptions.
\end{definitionbox}

Overcounting is not a mistake if it is controlled. It becomes a method when we know exactly how many times each outcome has been counted.

\begin{examplebox}
Suppose we want to choose a committee of three people from \(n\) people. The order of names in the committee does not matter.

It is easier first to count ordered selections:
\[
n(n-1)(n-2).
\]
But each committee of three people has been counted \(3!\) times, because the same three people can be listed in \(3!\) different orders. Therefore the number of committees is
\[
\frac{n(n-1)(n-2)}{3!}.
\]
\end{examplebox}

The essential idea is not division itself. The essential idea is identifying which descriptions should be treated as the same outcome.

\section{Unordered selections: combinations}

A combination is an unordered selection.

\begin{definitionbox}
A \textbf{combination} of size \(k\) from an \(n\)-element set is a subset with \(k\) elements.
\end{definitionbox}

The number of combinations is
\[
\binom{n}{k}=\frac{n!}{k!(n-k)!}.
\]

This formula can be derived from ordered selections. There are
\[
P(n,k)=\frac{n!}{(n-k)!}
\]
ordered selections of length \(k\). But each unordered selection of \(k\) objects can be ordered in \(k!\) ways. Therefore
\[
\binom{n}{k}=\frac{P(n,k)}{k!}=\frac{n!}{k!(n-k)!}.
\]

Combinations are used when:
\begin{itemize}
    \item only membership is recorded,
    \item order is ignored,
    \item each object may be chosen at most once,
    \item the outcome is a set.
\end{itemize}

\begin{examplebox}
A five-card hand from a standard deck is a set of five cards, not a sequence of five draws, if the final hand is all that is recorded. The number of possible hands is
\[
\binom{52}{5}.
\]
If the order of drawing is recorded, the correct count is instead
\[
P(52,5)=52\cdot 51\cdot 50\cdot 49\cdot 48.
\]
These are different sample spaces.
\end{examplebox}

\section{Distinguishable and indistinguishable objects}

Objects are distinguishable if the model treats them as separate individuals. Objects are indistinguishable if only their type matters.

\begin{examplebox}
The objects
\[
R_1,R_2,R_3
\]
are three distinguishable red balls. The symbols
\[
R,R,R
\]
represent three indistinguishable red balls of the same type.
\end{examplebox}

A physical object may be distinguishable in one model and indistinguishable in another. This depends on the recording rule.

\begin{examplebox}
An urn contains red and blue balls. If each ball has a label, and labels are recorded, then two red balls are different outcomes. If only colors are recorded, then the red balls are treated as indistinguishable.
\end{examplebox}

This distinction is crucial in probability. If the physical random mechanism selects labelled balls uniformly, but the sample space records only colors, then the color outcomes may not be equally likely.

\section{Permutations with repeated elements}

When some objects are indistinguishable, a linear arrangement may have several labelled descriptions that represent the same visible arrangement.

Suppose we arrange \(n\) objects, where there are \(n_1\) objects of type 1, \(n_2\) objects of type 2, and so on, with
\[
n_1+n_2+\cdots+n_r=n.
\]
The number of distinct visible arrangements is
\[
\frac{n!}{n_1!n_2!\cdots n_r!}.
\]

\begin{examplebox}
The word \(\text{BANANA}\) contains six letters: three \(A\)'s, two \(N\)'s, and one \(B\). If all six positions are arranged, then the number of distinct words is
\[
\frac{6!}{3!2!1!}.
\]
The division appears because permuting the three \(A\)'s among themselves does not change the visible word, and neither does permuting the two \(N\)'s.
\end{examplebox}

This formula should be read as an overcounting correction:
\[
\text{labelled arrangements}
\quad \longrightarrow \quad
\text{visible arrangements}.
\]

\section{Combinations with repetition}

Sometimes we choose \(k\) objects from \(n\) types, and the same type may be chosen several times. The order of the chosen types is ignored. The outcome is then not a sequence but a vector of counts.

\begin{definitionbox}
A \textbf{combination with repetition} of size \(k\) from \(n\) types is a choice of non-negative integers
\[
(x_1,x_2,\ldots,x_n)
\]
such that
\[
x_1+x_2+\cdots+x_n=k.
\]
Here \(x_i\) records how many objects of type \(i\) are chosen.
\end{definitionbox}

The number of such choices is
\[
\binom{n+k-1}{k}=\binom{n+k-1}{n-1}.
\]

This is often called the stars and bars formula. The idea is to represent the \(k\) chosen objects as stars and use \(n-1\) bars to separate the counts of the \(n\) types.

\begin{examplebox}
How many ways are there to choose \(5\) donuts from \(3\) flavors if only the numbers of each flavor matter?

An outcome is a count vector
\[
(x_1,x_2,x_3),\qquad x_1+x_2+x_3=5.
\]
The number of outcomes is
\[
\binom{3+5-1}{5}=\binom{7}{5}=\binom{7}{2}.
\]
\end{examplebox}

\begin{remarkbox}
Combinations with repetition do not describe ordered repeated choices. If order matters, the model is a sequence with repetition and the count is \(n^k\). If order is ignored and only counts of types matter, the model is combinations with repetition.
\end{remarkbox}

\section{Counting favorable events in probability}

Combinatorics becomes part of probability when it is used to count sample spaces and events.

If \(\Omega\) is finite and all elementary outcomes are equally likely, then
\[
P(A)=\frac{|A|}{|\Omega|}.
\]
This formula is valid only after the sample space has been correctly chosen and the elementary outcomes are equally likely.

\begin{examplebox}
A fair die is rolled twice. If order is recorded, then
\[
\Omega=\{1,2,3,4,5,6\}^2,\qquad |\Omega|=36.
\]
The event that the sum is \(7\) is
\[
A=\{(1,6),(2,5),(3,4),(4,3),(5,2),(6,1)\}.
\]
Thus
\[
P(A)=\frac{6}{36}=\frac{1}{6}.
\]
\end{examplebox}

\begin{examplebox}
Five cards are drawn from a standard deck, and the final hand is recorded. If every five-card hand is equally likely, then
\[
|\Omega|=\binom{52}{5}.
\]
The number of hands with exactly two hearts is
\[
\binom{13}{2}\binom{39}{3}.
\]
Therefore
\[
P(\text{exactly two hearts})=
\frac{\binom{13}{2}\binom{39}{3}}{\binom{52}{5}}.
\]
\end{examplebox}

The important point is that the denominator and numerator must be counted inside the same model. It is wrong to count ordered outcomes in the denominator and unordered outcomes in the numerator, or conversely.

\section{Changing the recording rule changes the model}

The same experiment may be counted in different ways because different questions require different sample spaces.

\begin{examplebox}
Three balls are drawn without replacement from an urn containing labelled balls. Compare the following recording rules.

\begin{center}
\begin{tabular}{lll}
\toprule
Recorded information & Outcome type & Typical count \\
\midrule
labels in order & sequence without repetition & \(P(n,3)\) \\
labels only, order ignored & subset & \(\binom{n}{3}\) \\
colors in order & sequence of colors & depends on color counts \\
color counts only & count vector & depends on available colors \\
number of red balls only & value of a random variable & \(0,1,2,3\) possible \\
\bottomrule
\end{tabular}
\end{center}
\end{examplebox}

The physical action is the same: three balls are drawn. The mathematical model is not the same, because the recorded outcome is different.

\section{Recognizing the correct counting model}

The following diagnostic table summarizes the main finite counting structures.

\begin{center}
\begin{tabular}{llll}
\toprule
Outcome type & Order? & Repetition? & Count \\
\midrule
sequence with repetition & yes & yes & \(n^k\) \\
ordered selection & yes & no & \(P(n,k)=\dfrac{n!}{(n-k)!}\) \\
permutation & yes & no, all used & \(n!\) \\
circular arrangement & relative order & no, all used & \((n-1)!\) \\
combination & no & no & \(\binom{n}{k}\) \\
combination with repetition & no & yes & \(\binom{n+k-1}{k}\) \\
multiset permutation & positions yes & identical objects & \(\dfrac{n!}{n_1!\cdots n_r!}\) \\
\bottomrule
\end{tabular}
\end{center}

The table should be used only after the outcome type has been identified. It is a summary, not a substitute for modelling.

\begin{remarkbox}
A formula is a compressed argument. If the argument does not match the situation, the formula gives the wrong count even when the arithmetic is correct.
\end{remarkbox}

\section{Common traps}

\subsection*{Trap 1: confusing a sequence with a set}

The ordered outcomes
\[
(A,B,C)\quad\text{and}\quad(C,B,A)
\]
are different sequences but the same set:
\[
\{A,B,C\}=\{C,B,A\}.
\]
Use sequences when order is recorded. Use sets when only membership is recorded.

\subsection*{Trap 2: treating indistinguishable objects as distinguishable}

The letters in \(\text{BANANA}\) are not six distinct letters. The three \(A\)'s are indistinguishable, and the two \(N\)'s are indistinguishable. Therefore \(6!\) overcounts the number of visible arrangements.

\subsection*{Trap 3: using combinations when roles matter}

A committee is a set. A medal assignment is an ordered assignment. A group of three athletes and a gold-silver-bronze result are not the same kind of outcome.

\subsection*{Trap 4: assuming all recorded outcomes are equally likely}

If labelled balls are selected uniformly but only colors are recorded, then color outcomes may have different probabilities. A smaller recorded sample space does not automatically become equally likely.

\subsection*{Trap 5: mixing models in one probability computation}

In the formula
\[
P(A)=\frac{|A|}{|\Omega|},
\]
both \(|A|\) and \(|\Omega|\) must be counted using the same outcome type. Do not count ordered draws in one part and unordered hands in the other.

\section{From combinatorics to probability models}

The role of combinatorics in this course is not to memorize more formulas. Its role is to build sample spaces and events precisely.

The general workflow is
\[
\text{situation}
\longrightarrow
\text{recording rule}
\longrightarrow
\text{elementary outcome}
\longrightarrow
\text{sample space}
\longrightarrow
\text{event}
\longrightarrow
\text{probability}.
\]

In later chapters, many probability distributions will arise by recording only a numerical summary of a richer experiment. For example, in \(n\) Bernoulli trials the full elementary outcome may be a sequence such as
\[
(S,F,S,S,F),
\]
while the random variable records only the number of successes:
\[
X=3.
\]
The binomial coefficient appears because many different sequences have the same number of successes.

Thus combinatorics is the bridge between concrete experiments and probability laws. It explains how many elementary outcomes produce the same event, the same count, or the same value of a random variable.

\section{Final checklist}

When solving a counting problem, write down the following before doing arithmetic:

\begin{enumerate}
    \item What is the physical situation?
    \item What information is recorded?
    \item What is one elementary outcome?
    \item What is the sample space?
    \item Are outcomes sequences, sets, arrangements, or count vectors?
    \item Are the elementary outcomes equally likely if probability is being computed?
    \item Which construction or overcounting argument gives the count?
    \item What does the final number mean in the original situation?
\end{enumerate}

This checklist is the main lesson of the chapter. The formulas are useful, but they are useful only when the underlying structure has been discovered first.
