\section{Recognizing the correct counting model}

The following diagnostic table summarizes the main finite counting structures.

\begin{center}
\begin{tabular}{llll}
\toprule
Outcome type & Order? & Repetition? & Count \\
\midrule
sequence with repetition & yes & yes & \(n^k\) \\
ordered selection & yes & no & \(P(n,k)=\dfrac{n!}{(n-k)!}\) \\
permutation & yes & no, all used & \(n!\) \\
circular arrangement & relative order & no, all used & \((n-1)!\) \\
combination & no & no & \(\binom{n}{k}\) \\
combination with repetition & no & yes & \(\binom{n+k-1}{k}\) \\
multiset permutation & positions yes & identical objects & \(\dfrac{n!}{n_1!\cdots n_r!}\) \\
\bottomrule
\end{tabular}
\end{center}

The table should be used only after the outcome type has been identified. It is a summary, not a substitute for modelling.

\begin{remarkbox}
A formula is a compressed argument. If the argument does not match the situation, the formula gives the wrong count even when the arithmetic is correct.
\end{remarkbox}

\section{Common traps}

\subsection*{Trap 1: confusing a sequence with a set}

The ordered outcomes
\[
(A,B,C)\quad\text{and}\quad(C,B,A)
\]
are different sequences but the same set:
\[
\{A,B,C\}=\{C,B,A\}.
\]
Use sequences when order is recorded. Use sets when only membership is recorded.

\subsection*{Trap 2: treating indistinguishable objects as distinguishable}

The letters in \(\text{BANANA}\) are not six distinct letters. The three \(A\)'s are indistinguishable, and the two \(N\)'s are indistinguishable. Therefore \(6!\) overcounts the number of visible arrangements.

\subsection*{Trap 3: using combinations when roles matter}

A committee is a set. A medal assignment is an ordered assignment. A group of three athletes and a gold-silver-bronze result are not the same kind of outcome.

\subsection*{Trap 4: assuming all recorded outcomes are equally likely}

If labelled balls are selected uniformly but only colors are recorded, then color outcomes may have different probabilities. A smaller recorded sample space does not automatically become equally likely.

\subsection*{Trap 5: mixing models in one probability computation}

In the formula
\[
P(A)=\frac{|A|}{|\Omega|},
\]
both \(|A|\) and \(|\Omega|\) must be counted using the same outcome type. Do not count ordered draws in one part and unordered hands in the other.
