\section{Variations without repetition}

A permutation uses all available objects. A variation uses only some of them. The order still matters.

\begin{definitionbox}
Let \(A\) be a finite set with \(n\) distinct elements, and let \(k\leq n\). A variation without repetition of length \(k\) is an ordered selection of \(k\) distinct elements from \(A\).
\end{definitionbox}

For example, from
\[
A=\{a,b,c\}
\]
the variations without repetition of length \(2\) are
\[
ab,\quad ac,\quad ba,\quad bc,\quad ca,\quad cb.
\]
The selections \(ab\) and \(ba\) are different, because order matters.

\begin{center}
\begin{tikzpicture}[scale=1.05, transform shape]
\foreach \word/\x/\y in {ab/0/2, ac/2.5/2, ba/5/2, bc/0/0, ca/2.5/0, cb/5/0} {
    \node[draw, rounded corners, fill=green!5, minimum width=1.6cm, minimum height=0.7cm] at (\x,\y) {$\word$};
}
\end{tikzpicture}
\end{center}

\begin{theorembox}
The number of variations without repetition of length \(k\) chosen from \(n\) distinct elements is
\[
V(n,k)=n(n-1)(n-2)\cdots(n-k+1)=\frac{n!}{(n-k)!}.
\]
\end{theorembox}

There are \(n\) choices for the first position, \(n-1\) choices for the second position, and so on. Since no element may be repeated, the number of choices decreases after each position is filled.

\begin{examplebox}
A code consists of three different letters chosen from the six letters
\[
A,B,C,D,E,F.
\]
Order matters. The number of possible codes is
\[
V(6,3)=6\cdot5\cdot4=120.
\]
\end{examplebox}

\begin{remarkbox}
Use variations without repetition when order matters, only \(k\) positions are filled, and no object may be used more than once.
\end{remarkbox}
