\section{Changing the Basis: New Coordinates for the Same Vector}

The preceding section developed the basic geometry and algebra of vectors:
addition, scalar multiplication, linear combinations, length, angle, the dot
product, parallelism, perpendicularity, and orthogonal projection. We now use
that toolkit to study how the coordinate description of one fixed geometric
vector changes when the basis is changed.

\noindent\textbf{What changes and what does not.} Changing the basis does not change the geometric vector. It changes only the
pair of coefficients used to reconstruct that vector from the chosen basis
directions. Coordinates belong to a vector relative to a basis; the vector
itself is independent of that description.

\input{chapters/chapter_02/change_of_basis/basis_definition_and_geometry}
\input{chapters/chapter_02/change_of_basis/same_vector_two_grids}
\input{chapters/chapter_02/change_of_basis/coordinates_from_linear_system}
\input{chapters/chapter_02/change_of_basis/general_coordinate_formula}
\input{chapters/chapter_02/change_of_basis/change_between_bases_and_projection}
\input{chapters/chapter_02/change_of_basis/negative_coordinate_example}

\noindent\textbf{Basis coordinates.} A basis is an ordered pair of nonparallel vectors. Every vector in the plane has
one coordinate pair relative to that basis. Positive and negative coefficients
are signed instructions telling us how much of each basis vector is required.
Different bases give different coordinate pairs for the same geometric vector.
To change coordinates, reconstruct the vector from the old basis and then solve
for its coefficients in the new basis.
