\section{Review Exercises}

\subsection{Elementary computational and graphical exercises}

\begin{enumerate}

\item Let $A=(1,2)$ and $B=(5,-1)$. Compute the free vector
$[\overrightarrow{AB}]$. Draw equal anchored representatives beginning at
$O=(0,0)$ and at $P=(-2,3)$, and write the endpoint of each representative.

\item For
\[
\mathbf u=[3,-2],\qquad \mathbf v=[-1,4],
\]
compute
\[
\mathbf u+\mathbf v,\qquad
\mathbf u-\mathbf v,\qquad
2\mathbf u,\qquad
-3\mathbf v.
\]
Draw $\mathbf u$, $\mathbf v$, and $\mathbf u+\mathbf v$ using the
parallelogram rule.

\item For
\[
\mathbf u=[1,2],\qquad \mathbf v=[2,-1],
\]
compute $2\mathbf u-3\mathbf v$. Reproduce the same displacement by placing
anchored representatives head to tail.

\item Compute the lengths and corresponding unit vectors for
\[
\mathbf u=[3,4],\qquad
\mathbf v=[-5,12],\qquad
\mathbf w=[2,-2].
\]

\item Determine which of the following pairs are parallel, perpendicular, or
neither:
\[
[2,1]\text{ and }[6,3],\qquad
[2,1]\text{ and }[1,-2],\qquad
[3,-1]\text{ and }[2,5].
\]
Use scalar multiples for parallelism and dot products for perpendicularity.

\item Find the angles between
\[
\mathbf u=[1,0],\qquad \mathbf v=[1,1]
\]
and between
\[
\mathbf v=[1,1],\qquad \mathbf w=[-1,1].
\]
Draw each pair and compare the computed angle with the picture.

\item The sides from one vertex of a triangle are
\[
\mathbf u=[3,1],\qquad \mathbf v=[-1,2].
\]
Compute the third side length $\|\mathbf u-\mathbf v\|$ directly and again
from the cosine theorem using the angle between $\mathbf u$ and $\mathbf v$.

\item Compute the projection of $\mathbf u=[4,3]$ onto
\[
\mathbf v=[1,0]
\]
and onto
\[
\mathbf w=[1,1].
\]
In each case write $\mathbf u$ as a parallel component plus a perpendicular
remainder and verify the perpendicularity by a dot product.

\item Let
\[
\mathbf b_1=[1,1],\qquad
\mathbf b_2=[1,-1],\qquad
\mathbf u=[4,2].
\]
Find $\alpha,\beta$ such that
\[
\mathbf u=\alpha\mathbf b_1+\beta\mathbf b_2.
\]
State $[\mathbf u]_B$ and reconstruct $\mathbf u$ from the obtained
coordinates.

\item A point starts at $P=(2,-1)$ and undergoes the successive displacements
\[
\mathbf u=[3,2],\qquad
\mathbf v=[-1,4],\qquad
-2\mathbf u.
\]
Find the final point. Then combine the three displacements first and verify
that the same endpoint is obtained.

\end{enumerate}

\subsection{Development and reasoning exercises}

\begin{enumerate}

\item Two anchored vectors are declared equal when they represent the same
translation. Explain, using a parallelogram construction, why moving an arrow
without changing its length and direction does not change the free vector.
Give one numerical example with two different initial points.

\item Use three concrete vectors to draw both
\[
(\mathbf u+\mathbf v)+\mathbf w
\qquad\text{and}\qquad
\mathbf u+(\mathbf v+\mathbf w).
\]
Explain geometrically, without relying only on coordinate arithmetic, why the
endpoints coincide.

\item Show that two nonzero vectors in the plane are parallel exactly when one
is a scalar multiple of the other. Explain separately what positive and
negative scalar multiples do to orientation.

\item Choose two nonzero vectors of equal length such that
\[
\|\mathbf u+\mathbf v\|<\|\mathbf u\|.
\]
Compute the lengths and explain how the angle between the vectors produces
partial cancellation.

\item Starting from the cosine theorem for the triangle formed by
$\mathbf u$, $\mathbf v$, and $\mathbf u-\mathbf v$, derive
\[
\mathbf u\cdot\mathbf v
=\|\mathbf u\|\,\|\mathbf v\|\cos\theta.
\]
Explain why the sign of the dot product distinguishes acute, right, and obtuse
angles.

\item Let
\[
\mathbf r
=\mathbf u-\operatorname{proj}_{\mathbf v}\mathbf u.
\]
Prove directly from the projection formula that $\mathbf r\cdot\mathbf v=0$.
Then interpret the result geometrically on a drawing.

\item Suppose $\mathbf b_1$ and $\mathbf b_2$ are nonparallel. Explain why a
vector cannot have two different representations
\[
\mathbf u=\alpha\mathbf b_1+\beta\mathbf b_2
=\alpha'\mathbf b_1+\beta'\mathbf b_2.
\]
Reduce the question to a linear combination equal to the zero vector.

\item Explain the difference between the point $P=(3,2)$ and the vector
$\mathbf v=[3,2]$. Construct one situation in which the same ordered pair of
numbers refers to the location of a point and another in which it refers to a
displacement.

\item The vector $\mathbf u=[6,2]$ is awkward in the standard basis but simple
in the basis
\[
\mathbf b_1=[1,1],\qquad \mathbf b_2=[1,-1].
\]
Find its coordinates in both bases and explain what it means for one basis to
be better adapted to a problem.

\item A student claims that every pair of nonzero vectors forms a basis of the
plane. Give one counterexample and one valid example. For each pair, explain
geometrically whether its linear combinations fill a line or the whole plane.

\end{enumerate}
