\section{Vectors from Coordinate Displacements}

In the previous chapter a point of the Euclidean plane was identified with an
ordered pair of real numbers.  We now use that identification immediately.
The aim is not to collect formulas about vectors.  We shall begin with concrete
geometric questions and allow the formulas to emerge from coordinate
calculations.

Choose two points
\[
A=(1,2),
\qquad
B=(4,6).
\]
To pass from $A$ to $B$, the first coordinate changes by $3$ and the second by
$4$.  The pair
\[
(3,4)
\]
records the complete displacement.

\subsection{A Vector Records a Change of Coordinates}

For arbitrary points
\[
A=(x_1,y_1),
\qquad
B=(x_2,y_2),
\]
the displacement from $A$ to $B$ is
\[
\overrightarrow{AB}
=
(x_2-x_1,\,y_2-y_1).
\]

\begin{definitionbox}
A vector in the Cartesian plane is an ordered pair
\[
\mathbf v=(v_1,v_2)
\]
interpreted as a coordinate displacement: $v_1$ units horizontally and $v_2$
units vertically.
\end{definitionbox}

This makes the phrase ``the same vector starting elsewhere'' precise.  If
$P=(p_1,p_2)$ and
\[
Q=(p_1+v_1,\,p_2+v_2),
\]
then
\[
\overrightarrow{PQ}=(v_1,v_2)=\mathbf v.
\]
No arrow is moved by an undefined geometric operation: the same two coordinate
changes are simply applied at another point.

\begin{center}
\begin{tikzpicture}[scale=0.85]
    \draw[axis] (-0.5,0) -- (7.0,0) node[right] {$x$};
    \draw[axis] (0,-0.5) -- (0,5.0) node[above] {$y$};
    \coordinate (A) at (1,1);
    \coordinate (B) at (4,3);
    \coordinate (P) at (2.5,2.2);
    \coordinate (Q) at (5.5,4.2);
    \draw[subset,->] (A) -- (B) node[midway,below right] {$(3,2)$};
    \draw[subset,->] (P) -- (Q) node[midway,above left] {$(3,2)$};
    \fill[geometricSubsetColor] (A) circle (2pt) node[below left] {$A$};
    \fill[geometricSubsetColor] (B) circle (2pt) node[below right] {$B$};
    \fill[geometricSubsetColor] (P) circle (2pt) node[above left] {$P$};
    \fill[geometricSubsetColor] (Q) circle (2pt) node[above right] {$Q$};
\end{tikzpicture}
\end{center}

The zero displacement is $\mathbf0=(0,0)$.  Reversing
$\mathbf v=(v_1,v_2)$ gives
\[
-\mathbf v=(-v_1,-v_2).
\]

\subsection{Two Successive Displacements Force Vector Addition}

Start at an arbitrary point $(x,y)$.  First apply
\[
\mathbf u=(u_1,u_2),
\]
and then
\[
\mathbf v=(v_1,v_2).
\]
The successive endpoints are
\[
(x,y)
\longmapsto
(x+u_1,y+u_2)
\longmapsto
(x+u_1+v_1,y+u_2+v_2).
\]
Therefore the single displacement with the same effect must be
\[
\mathbf u+\mathbf v
=
(u_1+v_1,\,u_2+v_2).
\]
The rule is forced by the experiment; it is not introduced as a convention to
memorise.

Because addition of real numbers is commutative, applying $\mathbf u$ and then
$\mathbf v$ reaches the same endpoint as applying $\mathbf v$ and then
$\mathbf u$.  This produces the parallelogram rule.

\begin{center}
\begin{tikzpicture}[scale=0.95]
    \coordinate (O) at (0,0);
    \coordinate (U) at (3.2,0.9);
    \coordinate (V) at (1.0,2.5);
    \coordinate (S) at (4.2,3.4);
    \draw[subset,->] (O) -- (U) node[midway,below] {$\mathbf u$};
    \draw[subset,->] (O) -- (V) node[midway,left] {$\mathbf v$};
    \draw[helper] (U) -- (S);
    \draw[helper] (V) -- (S);
    \draw[very thick,->] (O) -- (S) node[midway,above left] {$\mathbf u+\mathbf v$};
\end{tikzpicture}
\end{center}

Subtraction is the displacement that undoes another displacement:
\[
\mathbf u-\mathbf v
=
(u_1-v_1,\,u_2-v_2).
\]
In particular,
\[
\overrightarrow{AB}=B-A.
\]

\subsection{Repeating and Reversing a Displacement}

Repeating $\mathbf v=(v_1,v_2)$ twice gives $(2v_1,2v_2)$, while reversing it
gives $(-v_1,-v_2)$.  The natural extension is
\[
\lambda\mathbf v=(\lambda v_1,\lambda v_2),
\qquad \lambda\in\mathbb R.
\]
For positive $\lambda$ the orientation is unchanged; for negative $\lambda$ it
is reversed.

\subsection{When Do Two Vectors Determine Parallel Directions?}

Take two nonzero vectors
\[
\mathbf u=(u_1,u_2),
\qquad
\mathbf v=(v_1,v_2)
\]
and choose an arbitrary point $P$.  Construct
\[
Q=P+\mathbf u,
\qquad
R=P+\mathbf v.
\]
The line through $P,Q$ has direction $\mathbf u$, and the line through $P,R$
has direction $\mathbf v$.

Suppose first that both directions are nonvertical.  Their slopes are
\[
m_{\mathbf u}=\frac{u_2}{u_1},
\qquad
m_{\mathbf v}=\frac{v_2}{v_1}.
\]
The two direction lines coincide precisely when these slopes are equal:
\[
\frac{u_2}{u_1}=\frac{v_2}{v_1}.
\]
After cross multiplication,
\[
u_2v_1=u_1v_2.
\]
This equality means that the two coordinate pairs have the same ratio.  Hence
there is a nonzero number $\lambda$ such that
\[
\mathbf u=\lambda\mathbf v.
\]
The same conclusion is immediate for vertical directions: both first
coordinates are zero, and one nonzero vertical displacement is a scalar
multiple of the other.

Thus the coordinate condition has been obtained from the geometry of their
direction lines:
\[
\mathbf u\parallel\mathbf v
\iff
\mathbf u=\lambda\mathbf v
\quad\text{for some }\lambda\neq0.
\]
Here parallel vectors mean vectors whose direction lines are parallel or
coincident.  The formula is the conclusion of the calculation, not its starting
point.

\begin{examplebox}
The vectors
\[
\mathbf u=(6,-4),
\qquad
\mathbf v=(-3,2)
\]
determine the same line direction because
\[
\mathbf u=-2\mathbf v.
\]
Their orientations are opposite, but their carrier lines are parallel.
\end{examplebox}

\subsection{Length from the Pythagorean Theorem}

Draw $\mathbf v=(v_1,v_2)$ from any point.  The horizontal and vertical changes
form the legs of a right triangle, so the Pythagorean theorem gives
\[
\|\mathbf v\|^2=v_1^2+v_2^2,
\qquad
\|\mathbf v\|=\sqrt{v_1^2+v_2^2}.
\]
Consequently,
\[
d(A,B)=\|B-A\|.
\]

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \coordinate (O) at (0,0);
    \coordinate (P) at (4,2.7);
    \coordinate (R) at (4,0);
    \draw[subset,->] (O) -- (P) node[midway,above left] {$\mathbf v$};
    \draw[helper] (O) -- (R) node[midway,below] {$|v_1|$};
    \draw[helper] (R) -- (P) node[midway,right] {$|v_2|$};
    \draw (R) ++(-0.28,0) -- ++(0,0.28) -- ++(0.28,0);
\end{tikzpicture}
\end{center}

A nonzero vector can be rescaled to length one:
\[
\widehat{\mathbf v}=\frac{\mathbf v}{\|\mathbf v\|}.
\]

\subsection{What Coordinate Condition Expresses a Right Angle?}

We now ask a geometric question before introducing any new algebraic operation.
Given two nonzero vectors
\[
\mathbf u=(u_1,u_2),
\qquad
\mathbf v=(v_1,v_2),
\]
when are their direction lines perpendicular?

Choose any common initial point
\[
P=(x_0,y_0)
\]
and form
\[
Q=P+\mathbf u,
\qquad
R=P+\mathbf v.
\]
The point $P$ will disappear from the calculation because slopes depend only on
coordinate differences.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \coordinate (P) at (0.7,0.6);
    \coordinate (Q) at (4.5,2.1);
    \coordinate (R) at (-0.5,3.6);
    \draw[helper] (-0.25,0.225) -- (5.45,2.475);
    \draw[helper] (1.0,-0.15) -- (-0.8,4.35);
    \draw[subset,->] (P) -- (Q) node[midway,below right] {$\mathbf u$};
    \draw[subset,->] (P) -- (R) node[midway,left] {$\mathbf v$};
    \fill[geometricSubsetColor] (P) circle (2pt) node[below left] {$P$};
    \fill[geometricSubsetColor] (Q) circle (2pt) node[above right] {$Q$};
    \fill[geometricSubsetColor] (R) circle (2pt) node[above left] {$R$};
\end{tikzpicture}
\end{center}

Assume first that neither direction is vertical.  Then
\[
m_{\mathbf u}=\frac{u_2}{u_1},
\qquad
m_{\mathbf v}=\frac{v_2}{v_1}.
\]
The previously established coordinate criterion for perpendicular nonvertical
lines is
\[
m_{\mathbf u}m_{\mathbf v}=-1.
\]
Substitution gives
\[
\frac{u_2}{u_1}\frac{v_2}{v_1}=-1.
\]
Multiplying by $u_1v_1$ yields
\[
u_1v_1+u_2v_2=0.
\]

The remaining perpendicular case consists of a horizontal and a vertical
direction.  Their vectors have the forms
\[
(u_1,0)
\qquad\text{and}\qquad
(0,v_2),
\]
and the same expression again vanishes:
\[
u_1\cdot0+0\cdot v_2=0.
\]

We have therefore discovered the coordinate expression that detects a right
angle.  Since it will occur repeatedly, we give it a name only now.

\begin{definitionbox}
For vectors in the plane, their \emph{dot product} is
\[
\mathbf u\cdot\mathbf v
=
u_1v_1+u_2v_2.
\]
In three-dimensional space,
\[
(u_1,u_2,u_3)\cdot(v_1,v_2,v_3)
=
u_1v_1+u_2v_2+u_3v_3.
\]
\end{definitionbox}

\begin{theorembox}
For nonzero vectors in the plane,
\[
\mathbf u\perp\mathbf v
\iff
\mathbf u\cdot\mathbf v=0,
\]
where vector perpendicularity means perpendicularity of the direction lines
constructed from the vectors at a common point.
\end{theorembox}

\begin{examplebox}
Take
\[
\mathbf u=(3,2),
\qquad
\mathbf v=(-2,3).
\]
Their direction-line slopes are $2/3$ and $-3/2$, whose product is $-1$.  The
same geometric fact appears in the expression just discovered:
\[
\mathbf u\cdot\mathbf v
=3(-2)+2(3)=0.
\]
\end{examplebox}

A vector perpendicular to a given direction is called a normal vector.  This
idea will be used to construct equations of lines and planes.

\subsection{What Does the Dot Product Measure for an Arbitrary Angle?}

The preceding calculation discovered the dot product by studying one special
geometric question: when is the angle a right angle?  We now ask what the same
coordinate expression says about an arbitrary angle.

Take two arbitrary nonzero vectors beginning at a common point:
\[
\mathbf u=(u_1,u_2),
\qquad
\mathbf v=(v_1,v_2).
\]
Let $\alpha$ and $\beta$ be their respective angles of inclination measured
from the positive horizontal direction.  The vector $\mathbf u$ has length
$\|\mathbf u\|$.  Its horizontal and vertical coordinate changes are therefore
\[
u_1=\|\mathbf u\|\cos\alpha,
\qquad
u_2=\|\mathbf u\|\sin\alpha.
\]
Likewise,
\[
v_1=\|\mathbf v\|\cos\beta,
\qquad
v_2=\|\mathbf v\|\sin\beta.
\]
Thus the two vectors can be written without making either of them a coordinate
axis:
\[
\mathbf u
=
\|\mathbf u\|(\cos\alpha,\sin\alpha),
\]
\[
\mathbf v
=
\|\mathbf v\|(\cos\beta,\sin\beta).
\]

Now substitute these coordinates into the dot product discovered above:
\[
\begin{aligned}
\mathbf u\cdot\mathbf v
&=u_1v_1+u_2v_2\\
&=\|\mathbf u\|\cos\alpha\,
  \|\mathbf v\|\cos\beta
 +\|\mathbf u\|\sin\alpha\,
  \|\mathbf v\|\sin\beta\\
&=\|\mathbf u\|\,\|\mathbf v\|
  \bigl(\cos\alpha\cos\beta+
        \sin\alpha\sin\beta\bigr).
\end{aligned}
\]
The angle between the two directions is
\[
\theta=|\alpha-\beta|.
\]
The cosine-difference identity gives
\[
\cos(\alpha-\beta)
=
\cos\alpha\cos\beta+\sin\alpha\sin\beta.
\]
Because cosine is even,
\[
\cos(\alpha-\beta)=\cos|\alpha-\beta|=\cos\theta.
\]
Therefore
\[
\mathbf u\cdot\mathbf v
=
\|\mathbf u\|\,\|\mathbf v\|\cos\theta.
\]

\begin{theorembox}
For arbitrary nonzero vectors $\mathbf u$ and $\mathbf v$ in the plane, if
$\theta$ is the angle between their direction lines, then
\[
\mathbf u\cdot\mathbf v
=
\|\mathbf u\|\,\|\mathbf v\|\cos\theta.
\]
\end{theorembox}

This formula was not used to define the dot product.  The dot product first
appeared as the coordinate expression forced by perpendicular direction lines.
The calculation above shows that the same expression records every angle, not
only a right angle.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \coordinate (O) at (0,0);
    \coordinate (U) at (4.2,1.6);
    \coordinate (V) at (1.6,3.6);
    \draw[axis] (-0.4,0) -- (5.2,0) node[right] {$x$};
    \draw[subset,->] (O) -- (U) node[above right] {$\mathbf u$};
    \draw[subset,->] (O) -- (V) node[above] {$\mathbf v$};
    \draw (1.0,0) arc[start angle=0,end angle=20.9,radius=1.0];
    \node at (1.22,0.25) {$\alpha$};
    \draw (0.78,0.30) arc[start angle=20.9,end angle=66.0,radius=0.84];
    \node at (0.72,0.72) {$\theta$};
    \draw (0.55,1.24) arc[start angle=66.0,end angle=90,radius=1.36];
    \node at (0.28,1.42) {$\beta$};
\end{tikzpicture}
\end{center}

\subsection{The Cosine Theorem Emerges from Vector Subtraction}

The two vectors $\mathbf u$ and $\mathbf v$, drawn from a common point, form two
sides of a triangle.  The displacement from the endpoint of $\mathbf v$ to the
endpoint of $\mathbf u$ is
\[
\mathbf u-\mathbf v.
\]
Hence the three side lengths are
\[
\|\mathbf u\|,
\qquad
\|\mathbf v\|,
\qquad
\|\mathbf u-\mathbf v\|.
\]
Let $\theta$ be the angle between $\mathbf u$ and $\mathbf v$.

We calculate the square of the third side directly from coordinates and the dot
product:
\[
\begin{aligned}
\|\mathbf u-\mathbf v\|^2
&=(\mathbf u-\mathbf v)\cdot(\mathbf u-\mathbf v)\\
&=\mathbf u\cdot\mathbf u
  -\mathbf u\cdot\mathbf v
  -\mathbf v\cdot\mathbf u
  +\mathbf v\cdot\mathbf v.
\end{aligned}
\]
The dot product is symmetric, so
\[
\mathbf u\cdot\mathbf v=\mathbf v\cdot\mathbf u.
\]
Also,
\[
\mathbf u\cdot\mathbf u=\|\mathbf u\|^2,
\qquad
\mathbf v\cdot\mathbf v=\|\mathbf v\|^2.
\]
Therefore
\[
\|\mathbf u-\mathbf v\|^2
=
\|\mathbf u\|^2+
\|\mathbf v\|^2-
2\mathbf u\cdot\mathbf v.
\]
Substituting the angle formula just derived gives
\[
\|\mathbf u-\mathbf v\|^2
=
\|\mathbf u\|^2+
\|\mathbf v\|^2-
2\|\mathbf u\|\,\|\mathbf v\|\cos\theta.
\]

If we write
\[
a=\|\mathbf u\|,
\qquad
b=\|\mathbf v\|,
\qquad
c=\|\mathbf u-\mathbf v\|,
\]
then the calculation becomes
\[
c^2=a^2+b^2-2ab\cos\theta.
\]

\begin{theorembox}
For a triangle with side lengths $a$, $b$, and $c$, where $\theta$ is the angle
between the sides of lengths $a$ and $b$,
\[
c^2=a^2+b^2-2ab\cos\theta.
\]
\end{theorembox}

The cosine theorem is therefore not inserted as an external rule.  It is the
coordinate expansion of the displacement between the endpoints of two arbitrary
vectors.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \coordinate (O) at (0,0);
    \coordinate (U) at (4.4,1.5);
    \coordinate (V) at (1.3,3.5);
    \draw[subset,->] (O) -- (U) node[midway,below] {$\mathbf u$};
    \draw[subset,->] (O) -- (V) node[midway,left] {$\mathbf v$};
    \draw[very thick,->] (V) -- (U) node[midway,above right] {$\mathbf u-\mathbf v$};
    \draw (0.72,0.25) arc[start angle=19,end angle=69.6,radius=0.76];
    \node at (0.70,0.65) {$\theta$};
\end{tikzpicture}
\end{center}

\subsection{Projection Is Forced by the Right-Angle Condition}

Let $\mathbf v\neq\mathbf0$.  We seek a multiple $\lambda\mathbf v$ such that
the remainder
\[
\mathbf u-\lambda\mathbf v
\]
is perpendicular to $\mathbf v$.  The criterion just derived requires
\[
(\mathbf u-\lambda\mathbf v)\cdot\mathbf v=0.
\]
Expanding and solving for $\lambda$ gives
\[
\lambda
=
\frac{\mathbf u\cdot\mathbf v}{\mathbf v\cdot\mathbf v}.
\]
Thus the parallel part of $\mathbf u$ is not introduced by a memorised formula;
it is the unique multiple that leaves a perpendicular remainder:
\[
\operatorname{proj}_{\mathbf v}\mathbf u
=
\frac{\mathbf u\cdot\mathbf v}{\mathbf v\cdot\mathbf v}\,\mathbf v.
\]

\subsection{The Same Coordinate Ideas in Three Dimensions}

In $\mathbb R^3$, a vector is a displacement
\[
\mathbf v=(v_1,v_2,v_3),
\]
and all addition and scaling rules remain componentwise.  Its length is obtained
by applying the Pythagorean theorem twice:
\[
\|\mathbf v\|
=
\sqrt{v_1^2+v_2^2+v_3^2}.
\]

Two nonparallel directions in space determine a plane.  To construct a normal
direction to that plane, seek a vector perpendicular to both.  The coordinate
expression
\[
\mathbf u\times\mathbf v
=
\bigl(
 u_2v_3-u_3v_2,
 u_3v_1-u_1v_3,
 u_1v_2-u_2v_1
\bigr)
\]
has exactly this property, because direct substitution gives
\[
(\mathbf u\times\mathbf v)\cdot\mathbf u=0,
\qquad
(\mathbf u\times\mathbf v)\cdot\mathbf v=0.
\]
Thus the cross product is introduced as a constructed solution to two
perpendicularity conditions.

\begin{summarybox}
Every operation in this chapter arose from a geometric problem expressed in
coordinates.  Successive displacements produced vector addition.  Repeated
displacements produced scalar multiplication.  Direction lines produced the
criterion for parallel vectors.  Perpendicular direction lines produced the
coordinate expression later named the dot product.  Writing two arbitrary
vectors by their lengths and inclination angles showed that this expression is
$\|\mathbf u\|\,\|\mathbf v\|\cos\theta$.  Expanding the displacement
$\mathbf u-\mathbf v$ then produced the cosine theorem.  Projection and the
cross product were obtained by solving explicit perpendicularity conditions.
\end{summarybox}
