\section{Vectors from Coordinates}

In the previous chapter we learned to replace a point of the Euclidean plane by
an ordered pair of real numbers.  We now use that correspondence immediately.
Instead of beginning with an abstract object, let us begin with an experiment.

Choose two points
\[
A=(1,2),
\qquad
B=(4,6).
\]
To move from $A$ to $B$, the first coordinate must increase by $3$ and the
second coordinate by $4$:
\[
4-1=3,
\qquad
6-2=4.
\]
The ordered pair
\[
(3,4)
\]
records the complete displacement from $A$ to $B$.

\subsection{A Vector as a Coordinate Displacement}

Let
\[
A=(x_1,y_1),
\qquad
B=(x_2,y_2).
\]
The displacement from $A$ to $B$ is
\[
\overrightarrow{AB}=(x_2-x_1,\,y_2-y_1).
\]

\begin{definitionbox}
A vector in the Cartesian plane is an ordered pair
\[
\mathbf v=(v_1,v_2)
\]
interpreted as a displacement: $v_1$ units in the horizontal direction and
$v_2$ units in the vertical direction.
\end{definitionbox}

This definition makes the phrase ``the same vector at another point'' precise.
If $C=(c_1,c_2)$ and
\[
D=(c_1+v_1,\,c_2+v_2),
\]
then
\[
\overrightarrow{CD}=(v_1,v_2)=\mathbf v.
\]
Nothing is moved vaguely.  The endpoint is obtained by adding the same two
coordinate changes.

\begin{center}
\begin{tikzpicture}[scale=0.85]
    \draw[axis] (-0.5,0) -- (7.0,0) node[right] {$x$};
    \draw[axis] (0,-0.5) -- (0,5.0) node[above] {$y$};
    \coordinate (A) at (1,1);
    \coordinate (B) at (4,3);
    \coordinate (C) at (2.5,2.2);
    \coordinate (D) at (5.5,4.2);
    \draw[subset,->] (A) -- (B) node[midway,below right] {$(3,2)$};
    \draw[subset,->] (C) -- (D) node[midway,above left] {$(3,2)$};
    \fill[geometricSubsetColor] (A) circle (2pt) node[below left] {$A$};
    \fill[geometricSubsetColor] (B) circle (2pt) node[below right] {$B$};
    \fill[geometricSubsetColor] (C) circle (2pt) node[above left] {$C$};
    \fill[geometricSubsetColor] (D) circle (2pt) node[above right] {$D$};
\end{tikzpicture}
\end{center}

The two arrows in the picture have different initial points, but both change
coordinates by exactly $(3,2)$.  This is the coordinate meaning of equality of
vectors.

\begin{examplebox}
Let
\[
P=(-2,1),
\qquad
Q=(3,-4).
\]
Then
\[
\overrightarrow{PQ}
=
(3-(-2),\,-4-1)
=
(5,-5).
\]
Starting from any point $(a,b)$, the same displacement ends at
\[
(a+5,\,b-5).
\]
\end{examplebox}

The zero vector is
\[
\mathbf 0=(0,0).
\]
It changes neither coordinate.  The opposite of
\[
\mathbf v=(v_1,v_2)
\]
is
\[
-\mathbf v=(-v_1,-v_2),
\]
which reverses the displacement.

\subsection{Adding Displacements}

Suppose we first move by
\[
\mathbf u=(u_1,u_2)
\]
and then by
\[
\mathbf v=(v_1,v_2).
\]
Starting at $(x,y)$, the first move gives
\[
(x+u_1,\,y+u_2),
\]
and the second gives
\[
(x+u_1+v_1,\,y+u_2+v_2).
\]
Therefore the combined displacement is
\[
\mathbf u+\mathbf v
=
(u_1+v_1,\,u_2+v_2).
\]

This formula is not an arbitrary rule.  It is forced by performing two
coordinate changes one after another.

\begin{examplebox}
From $(1,1)$ move first by
\[
\mathbf u=(3,1)
\]
and then by
\[
\mathbf v=(-1,2).
\]
The intermediate point is
\[
(1,1)+(3,1)=(4,2),
\]
and the final point is
\[
(4,2)+(-1,2)=(3,4).
\]
The single displacement from start to finish is
\[
(3,4)-(1,1)=(2,3),
\]
which is exactly
\[
\mathbf u+\mathbf v=(3,1)+(-1,2)=(2,3).
\]
\end{examplebox}

\begin{center}
\begin{tikzpicture}[scale=0.95]
    \coordinate (O) at (0,0);
    \coordinate (U) at (3.2,0.9);
    \coordinate (V) at (1.0,2.5);
    \coordinate (S) at (4.2,3.4);
    \draw[subset,->] (O) -- (U) node[midway,below] {$\mathbf u$};
    \draw[subset,->] (O) -- (V) node[midway,left] {$\mathbf v$};
    \draw[helper] (U) -- (S);
    \draw[helper] (V) -- (S);
    \draw[very thick,->] (O) -- (S) node[midway,above left] {$\mathbf u+\mathbf v$};
\end{tikzpicture}
\end{center}

The coordinate formula also explains the parallelogram rule.  Beginning at the
same point, the two paths
\[
\mathbf u\text{ then }\mathbf v
\qquad\text{and}\qquad
\mathbf v\text{ then }\mathbf u
\]
end at the same coordinates because real-number addition is commutative.
Hence the four points form a parallelogram, and its diagonal represents
$\mathbf u+\mathbf v$.

Subtraction is simply the displacement needed to undo another displacement:
\[
\mathbf u-\mathbf v
=
(u_1-v_1,\,u_2-v_2).
\]
In particular,
\[
\overrightarrow{AB}=B-A
\]
when points are identified with their coordinate pairs.

\subsection{Scaling a Vector}

Let
\[
\mathbf v=(v_1,v_2)
\]
and let $\lambda\in\mathbb R$.  Repeating the displacement twice gives
\[
2\mathbf v=(2v_1,2v_2),
\]
while reversing it gives
\[
-\mathbf v=(-v_1,-v_2).
\]
This suggests the general rule
\[
\lambda\mathbf v=(\lambda v_1,\lambda v_2).
\]

For $\lambda>0$, the direction is preserved.  For $\lambda<0$, it is reversed.
The factor $|\lambda|$ tells us how the length changes.

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \coordinate (O) at (0,0);
    \draw[subset,->] (O) -- (2,1) node[above] {$\mathbf v$};
    \draw[very thick,->] (O) -- (4,2) node[above right] {$2\mathbf v$};
    \draw[subset,->] (O) -- (-2,-1) node[below left] {$-\mathbf v$};
    \fill[geometricSubsetColor] (O) circle (2pt);
\end{tikzpicture}
\end{center}

Two nonzero vectors point along the same line precisely when one is a scalar
multiple of the other:
\[
\mathbf u\parallel\mathbf v
\iff
\mathbf u=\lambda\mathbf v
\quad\text{for some }\lambda\neq0.
\]
This criterion will later become the coordinate test for parallel lines.

\subsection{The Standard Coordinate Vectors}

The displacements
\[
\mathbf e_1=(1,0),
\qquad
\mathbf e_2=(0,1)
\]
move one unit along the coordinate axes.  Every vector can therefore be written
as
\[
(v_1,v_2)=v_1\mathbf e_1+v_2\mathbf e_2.
\]

This is exactly why coordinates are useful: they tell us how much of each basic
direction is present.

\begin{examplebox}
For
\[
\mathbf v=(3,-2),
\]
we have
\[
\mathbf v=3\mathbf e_1-2\mathbf e_2.
\]
Geometrically, this means three units to the right and two units downward.
\end{examplebox}

The same idea works in space.  There we use
\[
\mathbf e_1=(1,0,0),
\qquad
\mathbf e_2=(0,1,0),
\qquad
\mathbf e_3=(0,0,1),
\]
and
\[
(v_1,v_2,v_3)
=
v_1\mathbf e_1+v_2\mathbf e_2+v_3\mathbf e_3.
\]
For points
\[
A=(x_1,y_1,z_1),
\qquad
B=(x_2,y_2,z_2),
\]
the displacement is
\[
\overrightarrow{AB}
=
(x_2-x_1,\,y_2-y_1,\,z_2-z_1).
\]

\subsection{Length from the Pythagorean Theorem}

Take
\[
\mathbf v=(v_1,v_2).
\]
Draw it from the origin.  Its endpoint is $(v_1,v_2)$, so the horizontal and
vertical sides of the associated right triangle have lengths $|v_1|$ and
$|v_2|$.  The Pythagorean theorem gives
\[
\|\mathbf v\|
=
\sqrt{v_1^2+v_2^2}.
\]

\begin{center}
\begin{tikzpicture}[scale=0.9]
    \coordinate (O) at (0,0);
    \coordinate (P) at (4,2.7);
    \coordinate (R) at (4,0);
    \draw[subset,->] (O) -- (P) node[midway,above left] {$\mathbf v$};
    \draw[helper] (O) -- (R) node[midway,below] {$|v_1|$};
    \draw[helper] (R) -- (P) node[midway,right] {$|v_2|$};
    \draw (R) ++(-0.28,0) -- ++(0,0.28) -- ++(0.28,0);
\end{tikzpicture}
\end{center}

In three dimensions, applying the same theorem twice gives
\[
\|(v_1,v_2,v_3)\|
=
\sqrt{v_1^2+v_2^2+v_3^2}.
\]
The distance between two points is now simply the length of their displacement:
\[
d(A,B)=\|B-A\|.
\]

\begin{examplebox}
For
\[
A=(1,-1),
\qquad
B=(5,2),
\]
we have
\[
B-A=(4,3),
\]
so
\[
d(A,B)=\sqrt{4^2+3^2}=5.
\]
The familiar $3$--$4$--$5$ triangle appears directly in the coordinate
calculation.
\end{examplebox}

A vector of length one is called a unit vector.  For $\mathbf v\neq\mathbf0$,
\[
\widehat{\mathbf v}
=
\frac{\mathbf v}{\|\mathbf v\|}
\]
has the same direction and length one.

\subsection{Recovering Angles from Coordinates}

Length came from the Pythagorean theorem.  Angles can also be recovered from
coordinates.

Let $\mathbf u$ and $\mathbf v$ begin at the same point, and let $\theta$ be the
angle between them.  The third side of the triangle is $\mathbf u-\mathbf v$.
The cosine rule gives
\[
\|\mathbf u-\mathbf v\|^2
=
\|\mathbf u\|^2+\|\mathbf v\|^2
-2\|\mathbf u\|\,\|\mathbf v\|\cos\theta.
\]
On the other hand, if
\[
\mathbf u=(u_1,u_2),
\qquad
\mathbf v=(v_1,v_2),
\]
then direct expansion gives
\[
\|\mathbf u-\mathbf v\|^2
=
\|\mathbf u\|^2+\|\mathbf v\|^2
-2(u_1v_1+u_2v_2).
\]
Comparing the two expressions shows that the coordinate combination
\[
u_1v_1+u_2v_2
\]
measures the angle.

\begin{definitionbox}
The dot product of
\[
\mathbf u=(u_1,u_2)
\qquad\text{and}\qquad
\mathbf v=(v_1,v_2)
\]
is
\[
\mathbf u\cdot\mathbf v=u_1v_1+u_2v_2.
\]
In $\mathbb R^3$ one adds the third product:
\[
(u_1,u_2,u_3)\cdot(v_1,v_2,v_3)
=
u_1v_1+u_2v_2+u_3v_3.
\]
\end{definitionbox}

The preceding calculation proves
\[
\mathbf u\cdot\mathbf v
=
\|\mathbf u\|\,\|\mathbf v\|\cos\theta.
\]
Thus
\[
\cos\theta
=
\frac{\mathbf u\cdot\mathbf v}
{\|\mathbf u\|\,\|\mathbf v\|}.
\]

\begin{examplebox}
Let
\[
\mathbf u=(1,2),
\qquad
\mathbf v=(3,1).
\]
Then
\[
\mathbf u\cdot\mathbf v=1\cdot3+2\cdot1=5,
\]
while
\[
\|\mathbf u\|=\sqrt5,
\qquad
\|\mathbf v\|=\sqrt{10}.
\]
Therefore
\[
\cos\theta=
\frac{5}{\sqrt5\sqrt{10}}
=
\frac{1}{\sqrt2},
\]
so the angle is $45^\circ$.
\end{examplebox}

\subsection{Perpendicularity as a Coordinate Test}

A right angle has cosine zero.  Therefore, for nonzero vectors,
\[
\mathbf u\perp\mathbf v
\iff
\mathbf u\cdot\mathbf v=0.
\]

This is the first major payoff of introducing vectors through coordinates.  A
geometric statement about a right angle becomes one arithmetic calculation.

\begin{examplebox}
Take
\[
\mathbf u=(3,2),
\qquad
\mathbf v=(-2,3).
\]
Then
\[
\mathbf u\cdot\mathbf v
=3(-2)+2(3)=0.
\]
Hence the vectors are perpendicular.  Notice that $(-2,3)$ is obtained from
$(3,2)$ by a quarter-turn: the coordinates are exchanged and one sign changes.
\end{examplebox}

A vector perpendicular to a given direction is called a normal vector.  In the
next chapter, normal vectors will let us write equations of lines and planes.

\subsection{Projection: Splitting a Vector into Parallel and Perpendicular Parts}

Let $\mathbf v\neq\mathbf0$.  We want to split an arbitrary vector $\mathbf u$
into one part parallel to $\mathbf v$ and one part perpendicular to it.

A vector parallel to $\mathbf v$ has the form $\lambda\mathbf v$.  We require
\[
\mathbf u-\lambda\mathbf v
\]
to be perpendicular to $\mathbf v$.  Therefore
\[
(\mathbf u-\lambda\mathbf v)\cdot\mathbf v=0.
\]
Expanding gives
\[
\mathbf u\cdot\mathbf v
-
\lambda(\mathbf v\cdot\mathbf v)=0,
\]
so
\[
\lambda
=
\frac{\mathbf u\cdot\mathbf v}{\mathbf v\cdot\mathbf v}.
\]
Hence the parallel part is
\[
\operatorname{proj}_{\mathbf v}\mathbf u
=
\frac{\mathbf u\cdot\mathbf v}{\mathbf v\cdot\mathbf v}\,\mathbf v.
\]

\begin{center}
\begin{tikzpicture}[scale=0.95]
    \coordinate (O) at (0,0);
    \coordinate (V) at (5.2,0);
    \coordinate (U) at (3.2,2.7);
    \coordinate (P) at (3.2,0);
    \draw[subset,->] (O) -- (V) node[right] {$\mathbf v$};
    \draw[subset,->] (O) -- (U) node[above] {$\mathbf u$};
    \draw[very thick,->] (O) -- (P) node[midway,below] {$\operatorname{proj}_{\mathbf v}\mathbf u$};
    \draw[helper] (U) -- (P);
    \draw (P) ++(-0.28,0) -- ++(0,0.28) -- ++(0.28,0);
\end{tikzpicture}
\end{center}

The remainder
\[
\mathbf u-
\operatorname{proj}_{\mathbf v}\mathbf u
\]
is perpendicular to $\mathbf v$.  This decomposition will later describe
shortest distances to lines and planes.

\subsection{Two Directions and the Parallelogram They Generate}

Given two vectors $\mathbf u$ and $\mathbf v$, every expression
\[
s\mathbf u+t\mathbf v
\]
is called a linear combination of them.  In the plane, two nonparallel vectors
allow us to reach every point from the origin by choosing suitable $s$ and $t$.
They form a basis of the plane.

If the vectors are parallel, then every combination remains on one line.  If
they are not parallel, their combinations fill the whole plane.  This is the
geometric distinction between one available direction and two genuinely
independent directions.

The parallelogram generated by $\mathbf u$ and $\mathbf v$ has vertices
\[
\mathbf0,
\qquad
\mathbf u,
\qquad
\mathbf v,
\qquad
\mathbf u+\mathbf v.
\]
Its area in the plane is
\[
|u_1v_2-u_2v_1|.
\]
The expression vanishes exactly when the two vectors are parallel.

\begin{examplebox}
For
\[
\mathbf u=(3,1),
\qquad
\mathbf v=(1,2),
\]
the area is
\[
|3\cdot2-1\cdot1|=5.
\]
Because the area is nonzero, the vectors are not parallel and form a basis of
the plane.
\end{examplebox}

\subsection{The Cross Product in Three Dimensions}

In space, two nonparallel vectors determine a plane.  It is often useful to
construct a vector perpendicular to both of them.

For
\[
\mathbf u=(u_1,u_2,u_3),
\qquad
\mathbf v=(v_1,v_2,v_3),
\]
define
\[
\mathbf u\times\mathbf v
=
\bigl(
 u_2v_3-u_3v_2,
 u_3v_1-u_1v_3,
 u_1v_2-u_2v_1
\bigr).
\]
A direct dot-product calculation gives
\[
(\mathbf u\times\mathbf v)\cdot\mathbf u=0,
\qquad
(\mathbf u\times\mathbf v)\cdot\mathbf v=0.
\]
Thus $\mathbf u\times\mathbf v$ is normal to the plane generated by
$\mathbf u$ and $\mathbf v$.

Its length equals the area of their parallelogram:
\[
\|\mathbf u\times\mathbf v\|
=
\|\mathbf u\|\,\|\mathbf v\|\sin\theta.
\]
Consequently,
\[
\mathbf u\times\mathbf v=\mathbf0
\]
exactly when the vectors are parallel.

\begin{examplebox}
Let
\[
\mathbf u=(1,0,1),
\qquad
\mathbf v=(0,2,1).
\]
Then
\[
\mathbf u\times\mathbf v
=
(-2,-1,2).
\]
Indeed,
\[
(-2,-1,2)\cdot(1,0,1)=0
\]
and
\[
(-2,-1,2)\cdot(0,2,1)=0.
\]
So $(-2,-1,2)$ is perpendicular to both original directions.
\end{examplebox}

\begin{summarybox}
A vector is introduced here directly as a coordinate displacement.
For points $A$ and $B$,
\[
\overrightarrow{AB}=B-A.
\]
Successive displacements force componentwise addition, scaling forces
componentwise multiplication by a scalar, and the Pythagorean theorem gives the
norm.  Comparing the cosine rule with coordinate expansion produces the dot
product and the test
\[
\mathbf u\perp\mathbf v
\iff
\mathbf u\cdot\mathbf v=0.
\]
In three dimensions, the cross product constructs a normal direction to two
nonparallel vectors.  These tools are exactly what we need to describe lines
and planes in the next chapter.
\end{summarybox}
